---
title: "Cambodia: Institutional Trajectory and Outcome Decay"
subtitle: "Tracing 1970–2025 through the cross-sectional ML model"
author: "Institutional Complexity Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6.5
    fig_height: 4.5
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath("../.."))
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  cache = TRUE,
  cache.lazy = FALSE
)
```

# Motivation

The cross-sectional analysis in `outcomes_ml.Rmd` answers: *given a country's
2025 V-Dem fingerprint, what outcomes does the model expect?* The Cambodia case
study answers: *what country does Cambodia institutionally resemble today?*
(answer: Burundi). Both are snapshots.

This document asks the time-series question: **how has Cambodia's institutional
fingerprint evolved since 1970, and what would that historical fingerprint have
predicted at each point in time about its development outcomes?** Cambodia's
GDP-per-capita series has visible breaks at 1979, 1987, 1997, and 2006; we
trace whether its institutional setup tracks those breaks and compare to two
benchmarks: Vietnam (a managed-state ASEAN neighbour) and Burundi (its
data-driven institutional twin).

The hypothesis under test is the **decay hypothesis**: prior Cambodian
institutional regimes were better at producing human capital and economic
complexity than the current one. The country still has a high *stock* of
human capital and infrastructure (built under the post-UNTAC era and the
PRK education recovery), but the current institutional setup is less able
to replenish that stock. **Infant mortality may be the leading indicator** —
since IMR responds to live institutional capacity (clinic staffing, vaccine
cold chain) faster than HCI does, a model-predicted IMR that has turned upward
before actual IMR would suggest the next wave is already cresting.

```{r packages}
library(tidyverse)
library(arrow)
library(knitr)
library(xgboost)
library(strucchange)
library(readxl)
library(ggrepel)

source("~/dev/gl-design/skills/gl-ggplot/assets/theme_gl.R")
gl_setup()
```

# Data assembly

## 2025 cross-section: feature set, normalisation bounds, models

We rebuild the same `Mcp` matrix used in `outcomes_ml.Rmd`: V-Dem 2025
component-type indicators with at least 150-country coverage, dropping
`v2regopploc` and `v2regsuploc`. Each feature is min-max normalised on
the 2025 cross-section to `[0, 1]`. The **per-feature min and max from
2025 are saved** — they're the bounds we apply to the historical panel
later, so a 1975 Cambodia value gets placed on the same scale the model
was trained on.

```{r build-mcp-2025, cache=TRUE}
vdem_raw <- read_parquet("~/dev/shared-data/vdem/vdem.parquet")
vdem_codebook <- read_parquet("~/dev/shared-data/vdem/vdem_codebook.parquet")

latest_year <- vdem_raw %>%
  filter(year >= 2020) %>%
  group_by(year) %>%
  summarise(n_countries = n_distinct(country_name), .groups = "drop") %>%
  filter(n_countries >= 150) %>%
  pull(year) %>% max()

component_codes <- vdem_codebook %>% filter(vartype == "C") %>% pull(tag)
suffixes <- c("_codelow", "_codehigh", "_sd", "_osp", "_ord", "_mean", "_nr")
all_cols <- names(vdem_raw)
is_suffix <- sapply(all_cols, function(x) any(sapply(suffixes, function(s) endsWith(x, s))))
is_numbered <- grepl("_\\d+$", all_cols)
base_cols <- all_cols[!is_suffix & !is_numbered]
candidate_codes <- intersect(component_codes, base_cols)

latest_data <- vdem_raw %>% filter(year == latest_year)
coverage <- sapply(candidate_codes, function(x) sum(!is.na(latest_data[[x]])))
numeric_check <- sapply(candidate_codes, function(x) is.numeric(vdem_raw[[x]]))
indicator_codes <- candidate_codes[coverage >= 150 & numeric_check]
indicator_codes <- setdiff(indicator_codes, c("v2regopploc", "v2regsuploc"))

indicator_info <- vdem_codebook %>%
  filter(tag %in% indicator_codes) %>%
  transmute(code = tag, description = name) %>%
  distinct(code, .keep_all = TRUE)
desc_lookup <- setNames(indicator_info$description, indicator_info$code)

vdem_filtered <- vdem_raw %>%
  filter(year == latest_year) %>%
  select(country_name, country_text_id, all_of(indicator_info$code)) %>%
  mutate(n_missing = rowSums(is.na(across(all_of(indicator_info$code))))) %>%
  filter(n_missing < ncol(across(all_of(indicator_info$code))) * 0.5) %>%
  select(-n_missing)

cont_df <- vdem_filtered %>% select(country_name, all_of(indicator_info$code))

# Stash the 2025 min/max per feature — these are the bounds we apply to history.
feat_bounds <- tibble(
  code = indicator_info$code,
  min = sapply(indicator_info$code, function(x) min(cont_df[[x]], na.rm = TRUE)),
  max = sapply(indicator_info$code, function(x) max(cont_df[[x]], na.rm = TRUE))
)

for (ind in indicator_info$code) {
  v <- cont_df[[ind]]
  rng <- c(feat_bounds$min[feat_bounds$code == ind],
           feat_bounds$max[feat_bounds$code == ind])
  cont_df[[ind]] <- if (rng[2] > rng[1]) (v - rng[1]) / (rng[2] - rng[1]) else 0
}
cont_df <- cont_df %>%
  mutate(across(all_of(indicator_info$code), ~replace_na(.x, 0)))

Mcp <- cont_df %>% column_to_rownames("country_name") %>% as.matrix()
iso3_lookup <- vdem_filtered %>% select(country_name, country_text_id) %>% deframe()

cat("V-Dem year:", latest_year, "\n")
cat("Mcp:", nrow(Mcp), "countries x", ncol(Mcp), "features\n")
```

## Historical panel: KHM, VNM, BDI from 1970

For each focus country and each year in 1970–`r I("latest_year")`, we pull the
same V-Dem feature columns, then min-max normalise using the **2025 bounds**
(not the historical min/max — we want the model to see history in 2025-scale
units). NAs are replaced with 0, matching the cross-sectional pipeline. NAs
are common in Cambodia 1975–1999 because several V-Dem features (election
quality, party characteristics, lower-chamber behaviour) are undefined when
elections or legislatures don't exist; treating them as floor values is
consistent with the model's training-time handling.

```{r build-history, cache=TRUE}
focus_iso3 <- c("KHM", "VNM", "BDI")

hist_raw <- vdem_raw %>%
  filter(country_text_id %in% focus_iso3, year >= 1970, year <= latest_year) %>%
  select(country_text_id, country_name, year, all_of(indicator_info$code))

# Apply 2025 min-max bounds, then NA -> 0
hist_norm <- hist_raw
for (ind in indicator_info$code) {
  v <- hist_norm[[ind]]
  rng <- c(feat_bounds$min[feat_bounds$code == ind],
           feat_bounds$max[feat_bounds$code == ind])
  hist_norm[[ind]] <- if (rng[2] > rng[1]) (v - rng[1]) / (rng[2] - rng[1]) else 0
}
hist_norm <- hist_norm %>%
  mutate(across(all_of(indicator_info$code), ~replace_na(.x, 0)))

# Coverage diagnostic: share of features with non-NA raw value, by country-year
na_share <- hist_raw %>%
  rowwise() %>%
  mutate(na_share = mean(is.na(c_across(all_of(indicator_info$code))))) %>%
  ungroup() %>%
  select(country_text_id, year, na_share)

cat("Historical panel:", nrow(hist_norm), "rows x",
    length(indicator_info$code), "features\n")
```

```{r history-coverage-plot, fig.height=2.6}
ggplot(na_share, aes(year, 1 - na_share, color = country_text_id)) +
  geom_line(linewidth = 0.7) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = NULL, y = "Share of features observed",
       subtitle = "V-Dem feature coverage by country-year (raw, before NA->0 imputation)",
       color = NULL) +
  theme(legend.position = "bottom")
```

```{r outcome-panel, cache=TRUE}
# Build the outcome panel for the cross-sectional model. We model 4 outcomes:
# HCI, IMR (log), Life Expectancy, ECI. Same window logic as outcomes_ml.Rmd:
# most recent non-NA value within 6 years ending at latest_year.

outcome_meta <- tribble(
  ~column,                 ~bucket,       ~label,                       ~transform,
  "wdi_hd_hci_ovrl",       "Education",   "Human Capital Index",        "none",
  "wdi_sp_dyn_imrt_in",    "Health",      "Infant mortality (log)",     "log",
  "wdi_sp_dyn_le00_in",    "Health",      "Life expectancy",            "none",
  "atl_sitc_eci",          "Complexity",  "Economic Complexity (ECI)",  "none"
)

macro <- read_parquet(
  "~/dev/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year", outcome_meta$column)
)

window <- seq(latest_year - 5, latest_year)
pick_latest <- function(col) {
  macro %>%
    filter(year %in% window, !is.na(.data[[col]])) %>%
    group_by(countrycodeiso) %>%
    slice_max(year, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(iso3 = countrycodeiso,
              !!col := .data[[col]])
}
outcome_wide <- reduce(
  lapply(outcome_meta$column, pick_latest),
  ~full_join(.x, .y, by = "iso3")
)
for (i in seq_len(nrow(outcome_meta))) {
  col <- outcome_meta$column[i]
  if (outcome_meta$transform[i] == "log") {
    outcome_wide[[col]] <- log(pmax(outcome_wide[[col]], .Machine$double.eps))
  }
}
```

```{r join-and-fit, cache=TRUE}
mcp_df <- as_tibble(Mcp, rownames = "country_name") %>%
  mutate(iso3 = unname(iso3_lookup[country_name])) %>%
  relocate(iso3)

analysis_df <- mcp_df %>%
  inner_join(outcome_wide, by = "iso3")

feature_cols <- indicator_info$code
X_mat <- matrix(as.double(as.matrix(analysis_df[, feature_cols])),
                nrow = nrow(analysis_df), ncol = length(feature_cols),
                dimnames = list(analysis_df$iso3, feature_cols))

base_params <- list(
  objective = "reg:squarederror",
  eval_metric = "rmse",
  eta = 0.05,
  subsample = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 5,
  verbosity = 0
)

fit_one <- function(col, depth = 3, seed = 42) {
  y_full <- analysis_df[[col]]
  ok <- !is.na(y_full)
  X_ok <- matrix(as.double(X_mat[ok, , drop = FALSE]),
                 nrow = sum(ok), ncol = ncol(X_mat),
                 dimnames = list(rownames(X_mat)[ok], colnames(X_mat)))
  y <- as.double(y_full[ok])
  n <- length(y)

  params <- c(base_params, list(max_depth = depth))
  set.seed(seed)
  dall <- xgb.DMatrix(data = X_ok, label = y)
  cv <- xgb.cv(params = params, data = dall, nrounds = 1000,
               nfold = 5, early_stopping_rounds = 40, verbose = 0)
  best_nrounds <- cv$early_stop$best_iteration
  if (is.null(best_nrounds) || length(best_nrounds) == 0) {
    best_nrounds <- which.min(cv$evaluation_log$test_rmse_mean)
  }

  set.seed(seed)
  folds <- sample(rep(1:5, length.out = n))
  oos_pred <- numeric(n)
  for (k in 1:5) {
    tr <- folds != k; te <- folds == k
    m <- xgb.train(params = params,
                   data = xgb.DMatrix(X_ok[tr, , drop = FALSE], label = y[tr]),
                   nrounds = best_nrounds, verbose = 0)
    oos_pred[te] <- predict(m, X_ok[te, , drop = FALSE])
  }
  r2_oos <- 1 - sum((y - oos_pred)^2) / sum((y - mean(y))^2)
  oos_rmse <- sqrt(var(y) * (1 - r2_oos))

  final <- xgb.train(params = params, data = dall,
                     nrounds = best_nrounds, verbose = 0)
  list(col = col, depth = depth, n = n, r2_oos = r2_oos,
       oos_rmse = oos_rmse, model = final)
}

# Sweep depths 3-5 per outcome, keep best
fits <- list()
for (i in seq_len(nrow(outcome_meta))) {
  col <- outcome_meta$column[i]
  cands <- lapply(3:5, function(d) fit_one(col, depth = d))
  best <- cands[[which.max(sapply(cands, `[[`, "r2_oos"))]]
  fits[[col]] <- best
  cat(sprintf("%-25s | best depth %d | OOS R^2 %.3f | n %d\n",
              col, best$depth, best$r2_oos, best$n))
}
```

```{r model-summary}
tibble(
  outcome = sapply(fits, `[[`, "col"),
  label = outcome_meta$label[match(sapply(fits, `[[`, "col"), outcome_meta$column)],
  bucket = outcome_meta$bucket[match(sapply(fits, `[[`, "col"), outcome_meta$column)],
  n = sapply(fits, `[[`, "n"),
  best_depth = sapply(fits, `[[`, "depth"),
  oos_r2 = round(sapply(fits, `[[`, "r2_oos"), 3),
  oos_rmse = round(sapply(fits, `[[`, "oos_rmse"), 3)
) %>%
  select(bucket, label, n, best_depth, oos_r2, oos_rmse) %>%
  kable(caption = "Cross-sectional model performance per outcome (5-fold CV).
        These are the same models used in outcomes_ml.Rmd, here re-fit on the
        four outcomes we project history through.")
```

# Institutional trajectory, 1970–2025

## Projection through the 2025 PCA

We compute `prcomp(X_mat, center=TRUE, scale.=TRUE)` on the 2025 Mcp and use
its rotation matrix to project the historical panel into the same PC space.
A historical PC1/PC2 score is the dot product of the historical fingerprint
(2025-scaled) with the 2025 loading vector. This is the standard approach to
"projecting new observations into an existing PCA."

```{r pca-fit-and-project, cache=TRUE}
pca <- prcomp(X_mat, center = TRUE, scale. = TRUE)
var_exp <- pca$sdev^2 / sum(pca$sdev^2)

# Project history: rebuild the (year x country x feature) matrix in 2025-scale
hist_X <- hist_norm %>%
  arrange(country_text_id, year) %>%
  select(country_text_id, year, all_of(indicator_info$code))

hist_mat <- as.matrix(hist_X[, indicator_info$code])
storage.mode(hist_mat) <- "double"
rownames(hist_mat) <- paste(hist_X$country_text_id, hist_X$year, sep = "_")

hist_pc <- predict(pca, newdata = hist_mat)
hist_pc_df <- as_tibble(hist_pc[, 1:5], rownames = "key") %>%
  separate(key, into = c("iso3", "year"), sep = "_", convert = TRUE)
```

```{r pc-trajectory, fig.height=5}
breaks_df <- tibble(brk = c(1979, 1987, 1997, 2006))

pc_long <- hist_pc_df %>%
  pivot_longer(starts_with("PC"), names_to = "pc", values_to = "score") %>%
  filter(pc %in% c("PC1", "PC2"))

pc_facet_labels <- c(
  PC1 = "PC1: liberal-democracy axis (positive = freer institutions)",
  PC2 = "PC2: managed-state vs. contestation (positive = managed-state, negative = contestation)"
)

ggplot(pc_long, aes(year, score, color = iso3)) +
  geom_vline(data = breaks_df, aes(xintercept = brk),
             color = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_hline(yintercept = 0, color = "grey80", linewidth = 0.3) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ pc, ncol = 1, scales = "free_y", labeller = as_labeller(pc_facet_labels)) +
  labs(x = NULL, y = "PC score (2025-PCA projection)",
       subtitle = "Cambodia, Vietnam, Burundi: institutional position over time, projected into the 2025 PCA",
       color = NULL) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 9))
```

The 2025-PCA loadings define PC1 as a liberal-democracy axis (more positive =
freer judiciary, election integrity, parliamentary oversight) and PC2 as a
managed-state vs. contestation axis. Top |PC2| loadings are anti-system
movements (`v2csantimv`), academic/general mobilisation (`v2cademmob`,
`v2cagenmob`), civil violence (`v2caviol`), and opposition camps (`v2cacamps`)
on the *negative* side, versus social-media regulatory capacity (`v2smregcap`),
state-driven media frames (`v2smmefra`, `v2smpolsoc`), public-health and
education-equality delivery (`v2pehealth`, `v2peedueq`), and election-ban scope
(`v2elsnlsff`) on the *positive* side. So **negative PC2 = open contestation,
mass mobilisation; positive PC2 = state-managed information environment with
delivered services**. This is the same convention used in the static
`outcomes_ml.Rmd` PCA scatter (where Vietnam, China, Laos, Egypt sit deep on
the positive side; Cambodia, Pakistan, Burundi, Myanmar sit on the negative
side).

## Selected V-Dem features over time

To complement the PC summary, we plot a small set of substantive indicators
that load heavily on PC1. Each panel shows the *raw* V-Dem score (not
2025-normalised) so the y-axis stays interpretable in V-Dem's native units.

```{r feature-trajectories, fig.height=6.2}
key_features <- c(
  v2juhcind   = "High-court independence",
  v2lgqstexp  = "Legislature questions officials",
  v2elirreg   = "Election irregularities (cleaner = higher)",
  v2exbribe   = "Executive bribery (cleaner = higher)",
  v2cseeorgs  = "CSO entry/exit barriers (open = higher)",
  v2pepwrses  = "Power distributed by socio-economic position"
)

feat_long <- hist_raw %>%
  select(country_text_id, year, all_of(names(key_features))) %>%
  pivot_longer(-c(country_text_id, year), names_to = "code", values_to = "value") %>%
  mutate(label = key_features[code],
         label = factor(label, levels = key_features))

ggplot(feat_long, aes(year, value, color = country_text_id)) +
  geom_vline(data = breaks_df, aes(xintercept = brk),
             color = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ label, scales = "free_y", ncol = 2) +
  labs(x = NULL, y = "V-Dem score (raw)",
       subtitle = "Selected institutional features, raw V-Dem units. Dashed lines: 1979/1987/1997/2006",
       color = NULL) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 8))
```

## Cambodia's eras at a glance

| Years | Regime | Headline institutional fact |
|---|---|---|
| 1970–1975 | Khmer Republic (Lon Nol) | Coup against Sihanouk; civil war |
| 1975–1979 | Democratic Kampuchea (Khmer Rouge) | Total state collapse; mass atrocity |
| 1979–1989 | People's Republic of Kampuchea | Vietnamese-backed PRK; rebuild from zero |
| 1989–1993 | State of Cambodia / UNTAC | Transition; UN-supervised election |
| 1993–1997 | Constitutional monarchy / coalition | FUNCINPEC–CPP power-sharing |
| 1997–2017 | CPP-dominated multiparty | Hun Sen consolidates after 1997 coup |
| 2017–2025 | One-party CPP | CNRP dissolved; Hun Manet succession |

The growth breaks the user identified line up with these transitions:
**1979** ends the Khmer Rouge; **1987** sits inside the PRK reform period
(party reform, gradual market opening, Vietnamese drawdown announced);
**1997** is the Hun Sen coup that ended power-sharing; **2006** roughly marks
the consolidation of CPP dominance and accession to the WTO (2004) plus the
subsequent garment-export boom.

# Predicted outcomes from historical fingerprints

For each year and country, we feed the 2025-scale historical fingerprint
through each of the four cross-sectional XGBoost models. The result is a
**counterfactual outcome time-series**: "if a country with this V-Dem
fingerprint existed in 2025, this is the HCI / IMR / LE / ECI it would
have." We then plot this against the *actual* outcome series (where
available) and look for divergence.

The prediction is not "what HCI Cambodia had in 1985"; HCI as a measure
didn't exist then. It's **"what HCI a country with 1985-Cambodia's
institutional setup would have today"** — i.e., what the model thinks
1985-Cambodia's institutions could *deliver* if measured by today's
yardsticks.

```{r project-history, cache=TRUE}
predict_history <- function(fit) {
  preds <- predict(fit$model, hist_mat)
  tibble(
    iso3 = hist_X$country_text_id,
    year = hist_X$year,
    outcome = fit$col,
    predicted = preds
  )
}

pred_long <- map_dfr(fits, predict_history)

# Pull actuals as long form, applying transforms
actuals_long <- macro %>%
  rename(iso3 = countrycodeiso) %>%
  filter(iso3 %in% focus_iso3, year >= 1970) %>%
  pivot_longer(all_of(outcome_meta$column),
               names_to = "outcome", values_to = "actual") %>%
  left_join(outcome_meta %>% select(column, transform),
            by = c("outcome" = "column")) %>%
  mutate(actual = if_else(transform == "log", log(pmax(actual, .Machine$double.eps)), actual)) %>%
  select(iso3, year, outcome, actual)

panel <- full_join(pred_long, actuals_long, by = c("iso3", "year", "outcome")) %>%
  left_join(outcome_meta %>% select(column, label, bucket),
            by = c("outcome" = "column")) %>%
  mutate(label = factor(label, levels = outcome_meta$label),
         iso3 = factor(iso3, levels = c("KHM", "VNM", "BDI")))
```

## Predicted vs actual: full panel

Solid line = model prediction from the year-t V-Dem fingerprint. Dots = actual
observed outcome (when available). Vertical dashes mark Cambodia's growth
breaks. For HCI, actuals only exist 2010–2020 — so the dot cluster is small
but informative.

```{r predicted-vs-actual, fig.width=7.2, fig.height=8}
ggplot(panel, aes(year, predicted, color = iso3)) +
  geom_vline(data = breaks_df, aes(xintercept = brk),
             color = "grey70", linetype = "dashed", linewidth = 0.3) +
  geom_line(linewidth = 0.6) +
  geom_point(aes(y = actual), size = 0.7, alpha = 0.7) +
  facet_grid(label ~ iso3, scales = "free_y", switch = "y") +
  labs(x = NULL, y = NULL,
       subtitle = "Solid line: model prediction from V-Dem fingerprint at year t.\nDots: actual outcome (where reported).",
       color = NULL) +
  theme(legend.position = "none",
        strip.placement = "outside",
        strip.text.y.left = element_text(angle = 0, hjust = 1, size = 8),
        strip.text.x = element_text(face = "bold"))
```

## Cambodia: predicted-vs-actual zoom, by outcome

A focused view of Cambodia alone makes the divergences easier to read.

```{r khm-zoom, fig.width=7.2, fig.height=6.2}
khm_panel <- panel %>% filter(iso3 == "KHM")

ggplot(khm_panel, aes(year)) +
  geom_vline(data = breaks_df, aes(xintercept = brk),
             color = "grey70", linetype = "dashed", linewidth = 0.3) +
  geom_line(aes(y = predicted, color = "Predicted (model)"), linewidth = 0.7) +
  geom_point(aes(y = actual, color = "Actual"), size = 1.2) +
  geom_line(aes(y = actual, color = "Actual"), linewidth = 0.5, alpha = 0.6) +
  facet_wrap(~ label, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c("Predicted (model)" = gl$brand_blue,
                                 "Actual" = highlight)) +
  labs(x = NULL, y = NULL, color = NULL,
       subtitle = "Cambodia (KHM): predicted from V-Dem vs. actual, 1970-2025") +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold", size = 9))
```

## Reading each panel

### Human Capital Index

HCI is reported only every 2–3 years, starting 2010. The model predicts an HCI
trajectory all the way back to 1970 because the V-Dem fingerprint exists for
those years. This is the trajectory under test for the decay hypothesis: did
the model think Cambodia *could* deliver higher HCI under earlier institutions
than it predicts today?

```{r hci-summary}
hci_table <- panel %>%
  filter(outcome == "wdi_hd_hci_ovrl",
         year %in% c(1975, 1980, 1985, 1990, 1995, 2000, 2005, 2010, 2015, 2020, 2024)) %>%
  arrange(iso3, year) %>%
  select(iso3, year, predicted, actual) %>%
  mutate(across(c(predicted, actual), ~round(.x, 3)))
kable(hci_table, caption = "Predicted HCI from year-t V-Dem fingerprint vs. actual HCI (where reported).")
```

### Infant mortality (log) — the leading-indicator hypothesis

Actual log-IMR for Cambodia falls steadily from 1980 onward — the HCI
investment of the PRK and post-UNTAC eras delivered a real mortality
transition. The model's *predicted* log-IMR is what 2025-trained
institutions-to-IMR mapping says a country with year-t Cambodia's institutions
would have today. If that predicted line **turns upward before actual**, it's
evidence the institutions capable of *sustaining* the mortality transition
have already eroded.

```{r imr-divergence}
imr_panel <- panel %>%
  filter(outcome == "wdi_sp_dyn_imrt_in", iso3 == "KHM") %>%
  arrange(year)

# When did predicted IMR (log) bottom out? When did actual bottom out?
imr_pred_min_year <- imr_panel %>% slice_min(predicted, n = 1) %>% pull(year)
imr_act_min_year  <- imr_panel %>% filter(!is.na(actual)) %>%
  slice_min(actual, n = 1) %>% pull(year)

cat(sprintf("Cambodia: predicted log-IMR minimum at year %d.\n", imr_pred_min_year))
cat(sprintf("Cambodia: actual log-IMR minimum at year %d (latest available).\n",
            imr_act_min_year))
```

```{r imr-gap-plot, fig.height=3.5}
imr_gap <- panel %>%
  filter(outcome == "wdi_sp_dyn_imrt_in", !is.na(actual)) %>%
  mutate(gap = actual - predicted)

ggplot(imr_gap, aes(year, gap, color = iso3)) +
  geom_vline(data = breaks_df, aes(xintercept = brk),
             color = "grey70", linetype = "dashed", linewidth = 0.3) +
  geom_hline(yintercept = 0, color = "grey50", linewidth = 0.4) +
  geom_line(linewidth = 0.7) +
  labs(x = NULL, y = "Actual minus predicted log-IMR",
       subtitle = "Positive gap: actual IMR worse than institutions predict.\nNegative gap: country outperforms its institutional fingerprint.",
       color = NULL) +
  theme(legend.position = "bottom")
```

### Life expectancy — the stock indicator

Life expectancy is an integral of past mortality conditions; it should
respond *more slowly* than IMR and so should track actual better than
predicted in the near term, even if predicted is already deteriorating.

### Economic complexity (ECI)

ECI is computed from export structure, which itself is built up over decades
from accumulated productive capabilities. Cambodia's ECI rose with garment
exports post-WTO accession (2004); the question is whether the institutional
fingerprint is now consistent with continued upgrading or has decayed below
the level associated with sustained complexity gains.

# Per-era summary

To compare Cambodia's institutional regimes directly, we average each
country's predicted and actual outcome within seven Cambodian eras: Khmer
Republic, Khmer Rouge, PRK, SOC/UNTAC, Coalition, CPP-dominated, and
One-party CPP. The Vietnam and Burundi columns use the same year ranges so
the comparison is era-anchored on Cambodia's politics.

```{r era-summary, fig.height=6.4}
era_labeller <- function(y) {
  case_when(
    y >= 1970 & y < 1975 ~ "1. Khmer Republic (1970-74)",
    y >= 1975 & y < 1979 ~ "2. Khmer Rouge (1975-78)",
    y >= 1979 & y < 1989 ~ "3. PRK (1979-88)",
    y >= 1989 & y < 1993 ~ "4. SOC/UNTAC (1989-92)",
    y >= 1993 & y < 1998 ~ "5. Coalition (1993-97)",
    y >= 1998 & y < 2018 ~ "6. CPP-dominated (1998-2017)",
    y >= 2018             ~ "7. One-party CPP (2018+)"
  )
}

era_summary <- panel %>%
  mutate(era = era_labeller(year)) %>%
  group_by(iso3, era, label) %>%
  summarise(predicted = mean(predicted, na.rm = TRUE),
            actual    = mean(actual,    na.rm = TRUE),
            .groups = "drop")

short_outcome <- c(
  "Human Capital Index" = "HCI",
  "Infant mortality (log)" = "log-IMR",
  "Life expectancy" = "LE",
  "Economic Complexity (ECI)" = "ECI"
)

short_era <- c(
  "1. Khmer Republic (1970-74)"  = "Khmer Rep. 1970-74",
  "2. Khmer Rouge (1975-78)"     = "Khmer Rouge 1975-78",
  "3. PRK (1979-88)"             = "PRK 1979-88",
  "4. SOC/UNTAC (1989-92)"       = "SOC/UNTAC 1989-92",
  "5. Coalition (1993-97)"       = "Coalition 1993-97",
  "6. CPP-dominated (1998-2017)" = "CPP-dom 1998-2017",
  "7. One-party CPP (2018+)"     = "1-party CPP 2018+"
)

era_summary %>%
  filter(iso3 == "KHM") %>%
  transmute(Era = short_era[era],
            Outcome = short_outcome[as.character(label)],
            Predicted = round(predicted, 2),
            Actual    = ifelse(is.nan(actual), NA, round(actual, 2))) %>%
  pivot_wider(names_from = Outcome,
              values_from = c(Predicted, Actual),
              names_glue = "{Outcome} ({.value})") %>%
  arrange(match(Era, short_era)) %>%
  kable(caption = "Cambodia: era-mean predicted vs actual outcome.")
```

```{r era-predicted-comparison, fig.height=5.5}
era_pred_compare <- era_summary %>%
  filter(!is.nan(predicted)) %>%
  mutate(label = factor(label, levels = outcome_meta$label),
         iso3 = factor(iso3, levels = c("KHM", "VNM", "BDI")),
         era_short = factor(short_era[era], levels = unname(short_era)))

ggplot(era_pred_compare, aes(era_short, predicted, color = iso3, group = iso3)) +
  geom_line(linewidth = 0.6, alpha = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ label, scales = "free_y", ncol = 2) +
  labs(x = NULL, y = "Predicted outcome (era mean)",
       subtitle = "Model-predicted outcome for each era's mean V-Dem fingerprint",
       color = NULL) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
        strip.text = element_text(size = 9, face = "bold"),
        plot.margin = margin(5, 5, 5, 5))
```

# Structural transformation, 1970-2025

The institutional trajectory above tells us about *politics*. To diagnose
whether Cambodia's industrial coalition is rentier or magician (Pritchett,
Sen, Werker, *Deals and Development*, 2018), we also need the *economic*
trajectory: when did GDP-per-capita break, what sectors drove each phase,
and does industrial growth lift the rest of the economy or run as an
enclave?

This section ports the structural-breaks and sectoral-decomposition logic
from `~/dev/political-economy-pakistan/struc_breaks.Rmd` (which already
renders a Cambodia PDF) onto the GL theme so they sit alongside the
institutional analysis.

## Growth breaks (Bai-Perron + Kar et al., 2013)

```{r growth-breaks, cache=TRUE, fig.height=4.2}
pwt <- read_excel("~/dev/political-economy-pakistan/data/pwt110.xlsx",
                  sheet = "Data")

penn_growth <- pwt %>%
  filter(countrycode == "KHM") %>%
  transmute(year, gdp_per_cap = rgdpna / pop) %>%
  filter(!is.na(gdp_per_cap)) %>%
  arrange(year) %>%
  mutate(growth_rate = (gdp_per_cap - lag(gdp_per_cap)) / lag(gdp_per_cap))

start_year_pwt <- min(penn_growth$year)
end_year_pwt <- 2023
n_breaks <- 4

penn_growth_ts <- ts(penn_growth$growth_rate[-1],
                     start = start_year_pwt, end = end_year_pwt, frequency = 1)
bp_growth <- breakpoints(penn_growth_ts ~ 1)
break_years <- Filter(Negate(is.na),
                      summary(bp_growth)$breakdates[n_breaks, ])

period_growthrates <- penn_growth %>%
  mutate(periods = cut(year,
                       breaks = c(start_year_pwt, break_years, end_year_pwt),
                       dig.lab = 4)) %>%
  group_by(periods) %>%
  summarise(avg_growth_rate = mean(growth_rate, na.rm = TRUE),
            mdn_growth_rate = median(growth_rate, na.rm = TRUE),
            length = n(),
            .groups = "drop") %>%
  mutate(start_period = as.numeric(str_match(as.character(periods),
                                             "\\(([0-9\\.]+),")[, 2]),
         end_period = as.numeric(str_match(as.character(periods),
                                           ",([0-9\\.]+)\\]")[, 2]))

# Kar et al. (2013) filter on average growth-rate jumps between periods
kar_filter <- period_growthrates %>%
  mutate(diff = avg_growth_rate - lag(avg_growth_rate),
         prev_dir = case_when(is.na(lag(diff)) ~ NA_character_,
                              lag(diff) > 0 ~ "acceleration",
                              lag(diff) < 0 ~ "deceleration",
                              TRUE ~ "no_change"),
         curr_dir = case_when(diff > 0 ~ "acceleration",
                              diff < 0 ~ "deceleration",
                              TRUE ~ "no_change"),
         threshold = case_when(is.na(prev_dir) ~ 0.02,
                               prev_dir != curr_dir ~ 0.03,
                               prev_dir == curr_dir ~ 0.01,
                               TRUE ~ NA_real_),
         passes_kar = case_when(is.na(diff) ~ NA,
                                abs(diff) >= threshold ~ TRUE,
                                TRUE ~ FALSE))

ggplot(penn_growth %>% filter(!is.na(growth_rate)),
       aes(year, growth_rate)) +
  geom_rect(data = kar_filter, inherit.aes = FALSE,
            aes(xmin = start_period, xmax = end_period,
                ymin = -Inf, ymax = Inf, fill = passes_kar),
            alpha = 0.18) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.2) +
  geom_segment(data = period_growthrates,
               aes(x = start_period, xend = end_period,
                   y = avg_growth_rate, yend = avg_growth_rate),
               linewidth = 0.7, color = gl$brand_blue) +
  geom_label(data = period_growthrates,
             aes(x = (start_period + end_period) / 2,
                 y = avg_growth_rate,
                 label = scales::percent(avg_growth_rate, accuracy = 0.1)),
             size = 3, label.size = 0.2) +
  scale_y_continuous(labels = scales::percent, n.breaks = 8) +
  scale_x_continuous(breaks = seq(1970, 2025, 5)) +
  scale_fill_manual(values = c("TRUE" = "#36B250", "FALSE" = "#C64646"),
                    na.value = "grey80") +
  labs(x = NULL, y = "Real GDP per capita growth rate",
       fill = "Passes Kar filter",
       subtitle = sprintf("Bai-Perron breaks at %s. Period averages and Kar et al. (2013) acceleration filter.",
                          paste(break_years, collapse = ", "))) +
  theme(legend.position = "bottom")
```

```{r growth-breaks-table}
period_growthrates %>%
  transmute(Period = as.character(periods),
            Years = length,
            `Avg. growth` = scales::percent(avg_growth_rate, accuracy = 0.1),
            `Median growth` = scales::percent(mdn_growth_rate, accuracy = 0.1)) %>%
  filter(!is.na(Years)) %>%
  kable(caption = "Cambodia GDP-per-capita growth periods (Penn World Tables 11.0).")
```

The Bai-Perron procedure picks the same four break years the analysis was
seeded with: **1979, 1987, 1997, 2006**. The Kar filter (3% absolute jump
threshold for direction-changes, 1% for direction-continuations) marks
1979-87 (PRK recovery) and 1997-2006 (post-coalition garment boom) as the
two genuine accelerations. The 1987-97 and 2006-23 periods are flagged as
non-accelerations: the first because PRK gains stalled under sanctions and
the 1991-93 transition shock; the second because growth slowed from 7.8%
to 4.8% as the post-2000 garment-led boom exhausted its initial extensive
margin.

## Sectoral value-added trajectory

```{r sectoral-va, cache=TRUE, fig.height=4.5}
sectoral_wdi <- macro %>%
  rename(iso3 = countrycodeiso) %>%
  filter(iso3 == "KHM") %>%
  select(iso3, year,
         starts_with("wdi_nv_"))
```

```{r sectoral-va-load, cache=TRUE}
# We need a richer set of sectoral fields than the outcome panel pulled.
sectoral_full <- read_parquet(
  "~/dev/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year",
                 "wdi_nv_agr_totl_kd", "wdi_nv_ind_totl_kd",
                 "wdi_nv_srv_totl_kd", "wdi_nv_ind_manf_kd",
                 "wdi_nv_agr_totl_kd_zg", "wdi_nv_ind_totl_kd_zg",
                 "wdi_nv_srv_totl_kd_zg", "wdi_nv_ind_manf_kd_zg",
                 "wdi_ny_gdp_pcap_kd_zg",
                 "wdi_sl_ind_empl_zs", "wdi_sl_tlf_totl_in")
) %>%
  rename(iso3 = countrycodeiso) %>%
  filter(iso3 %in% c("KHM", "VNM", "BDI"))
```

```{r sectoral-va-plot, fig.height=4.5}
sector_levels <- c("agr" = "Agriculture",
                   "ind" = "Industry",
                   "srv" = "Services")

khm_sectoral <- sectoral_full %>%
  filter(iso3 == "KHM") %>%
  select(year, wdi_nv_agr_totl_kd, wdi_nv_ind_totl_kd, wdi_nv_srv_totl_kd) %>%
  pivot_longer(-year, names_to = "code", values_to = "value") %>%
  mutate(sector_short = str_remove(code, "wdi_nv_") %>% str_remove("_totl_kd"),
         Sector = sector_levels[sector_short]) %>%
  filter(!is.na(value), !is.na(Sector))

# Period CAGRs for labels
period_cagrs <- khm_sectoral %>%
  mutate(period = cut(year,
                      breaks = c(start_year_pwt, break_years, end_year_pwt),
                      dig.lab = 4)) %>%
  filter(!is.na(period)) %>%
  group_by(Sector, period) %>%
  arrange(year) %>%
  summarise(
    start_year_p = first(year),
    end_year_p = last(year),
    start_val = first(value),
    end_val = last(value),
    n = end_year_p - start_year_p,
    cagr = if_else(n > 0 & start_val > 0,
                   (end_val / start_val)^(1 / n) - 1, NA_real_),
    avg_value = exp(mean(log(value), na.rm = TRUE)),
    .groups = "drop") %>%
  mutate(mid_year = (start_year_p + end_year_p) / 2)

ggplot(khm_sectoral, aes(year, value, color = Sector)) +
  geom_vline(data = breaks_df, aes(xintercept = brk),
             color = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 0.9) +
  geom_label_repel(data = period_cagrs %>% filter(!is.na(cagr)),
                   aes(x = mid_year, y = avg_value,
                       label = scales::percent(cagr, accuracy = 0.1)),
                   size = 2.8, label.size = 0.15,
                   min.segment.length = 0.2,
                   show.legend = FALSE) +
  scale_y_log10(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_x_continuous(breaks = seq(1970, 2025, 10)) +
  scale_color_manual(values = c("Agriculture" = "#EAC218",
                                "Industry"    = "#36B250",
                                "Services"    = "#266798")) +
  labs(x = NULL, y = "Sectoral value added (constant 2015 USD, log)",
       subtitle = "Cambodia: sectoral value added with period CAGRs. Vertical dashes mark growth breaks.",
       color = NULL) +
  theme(legend.position = "bottom")
```

The sectoral picture sharpens the political trajectory. **Agriculture
slows sharply after 2006** to ~2.0% total-VA growth — barely above
Cambodia's ~1.3% population growth, so per-capita agricultural
value-added is approximately flat (it was ~3.6% per-capita over
1997-2006 in the struc_breaks per-capita series). **Services slow**
from 10.7% (1997-2006) to 5.5% (2006-23). **Industry holds 9.0%** as the
dominant engine, but its growth rate *also* fell from the 16.9% peak of
1997-2006. The structural-transformation moment Cambodia boasts —
industry overtaking services in value-added around 2017 — happened
with industry slowing, not services stagnating: a level-crossing driven
by services losing momentum rather than by industry accelerating. (The
CAGRs labelled on the chart are total VA growth, not per-capita; subtract
~1.3pp for per-capita rates.)

# Magician or rentier? Sectoral evidence

The Pritchett-Sen-Werker framework distinguishes growth that **deepens
productive capability** (magician) from growth that **scales existing rent
flows** (rentier). For Cambodia's industrial coalition — garments,
SEZs from 2005, casinos, real estate — three diagnostic tests below
operate on data we already have. A fourth test connects PC2's
contestation trajectory to the political economy of consolidation.

## Test 1: Capability deepening (ECI trajectory)

Magicians complexify their export basket: ECI rises as the country
moves into denser, more sophisticated products. Rentiers don't: ECI
stays flat or falls as growth piles into the same narrow basket.

```{r eci-trajectory, fig.height=3.6, cache=TRUE}
eci_panel <- read_parquet(
  "~/dev/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year", "atl_sitc_eci")
) %>%
  rename(iso3 = countrycodeiso) %>%
  filter(iso3 %in% c("KHM", "VNM", "BDI"),
         !is.na(atl_sitc_eci),
         year >= 1995)

ggplot(eci_panel, aes(year, atl_sitc_eci, color = iso3)) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 0.9) +
  scale_x_continuous(breaks = seq(1995, 2025, 5)) +
  labs(x = NULL, y = "Economic Complexity Index (SITC)",
       subtitle = "ECI trajectory 1995-latest. Magician = rising; rentier = flat.",
       color = NULL) +
  theme(legend.position = "bottom")
```

```{r eci-summary}
eci_panel %>%
  group_by(iso3) %>%
  arrange(year) %>%
  summarise(
    `First year` = first(year),
    `Last year`  = last(year),
    `ECI start`  = round(first(atl_sitc_eci), 2),
    `ECI end`    = round(last(atl_sitc_eci), 2),
    `ECI change` = round(last(atl_sitc_eci) - first(atl_sitc_eci), 2),
    .groups = "drop"
  ) %>%
  arrange(desc(`ECI change`)) %>%
  kable(caption = "ECI start/end and total change by country.")
```

The split is clean. **Vietnam** rises from a low-negative ECI in the
mid-1990s to noticeably positive territory by the 2020s — a multi-decade
complexification consistent with the magician profile. **Cambodia** moves
modestly from very low to slightly less low; ECI has *risen* but stays
firmly negative throughout, and growth in complexity has not kept pace
with growth in GDP. **Burundi** drifts further negative — low complexity
to begin with, slowly declining. On this test, KHM is closer to the
rentier end of the spectrum than the magician end — particularly when
benchmarked against VNM, which started from a similar position in 1995.

## Test 2: Inter-sectoral spillovers

Magician industrial growth has linkages: a rising manufacturing sector
absorbs labour from agriculture, demands transport and finance services,
and pushes mechanisation back into agriculture. Rentier industrial growth
is enclaved (SEZs, FDI assembly with imported inputs); other sectors
don't co-move with it. We test this directly: does annual industry growth
predict annual agriculture or services growth, KHM vs VNM?

```{r spillover-scatter, fig.height=4.6, cache=TRUE}
spillover_data <- sectoral_full %>%
  filter(iso3 %in% c("KHM", "VNM"),
         year >= 1995) %>%
  select(iso3, year,
         agr = wdi_nv_agr_totl_kd_zg,
         ind = wdi_nv_ind_totl_kd_zg,
         srv = wdi_nv_srv_totl_kd_zg) %>%
  pivot_longer(c(agr, srv), names_to = "other_sector", values_to = "other_growth") %>%
  filter(!is.na(ind), !is.na(other_growth)) %>%
  mutate(other_label = recode(other_sector,
                              agr = "Agriculture growth",
                              srv = "Services growth"))

# Compute slope and correlation per panel
spillover_fits <- spillover_data %>%
  group_by(iso3, other_label) %>%
  summarise(
    slope = coef(lm(other_growth ~ ind))[2],
    cor   = cor(ind, other_growth),
    n     = n(),
    .groups = "drop"
  ) %>%
  mutate(label = sprintf("slope = %.2f\nr = %.2f  (n=%d)", slope, cor, n))

ggplot(spillover_data, aes(ind, other_growth)) +
  geom_hline(yintercept = 0, color = "grey80", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "grey80", linewidth = 0.3) +
  geom_point(aes(color = iso3), size = 1.2, alpha = 0.7) +
  geom_smooth(aes(color = iso3), method = "lm", se = FALSE,
              linewidth = 0.7, formula = y ~ x) +
  geom_text(data = spillover_fits,
            aes(x = -Inf, y = Inf, label = label, color = iso3),
            hjust = -0.05, vjust = 1.2, size = 2.8,
            lineheight = 0.95, show.legend = FALSE) +
  facet_grid(other_label ~ iso3) +
  labs(x = "Industry value-added growth (annual, %)",
       y = "Other-sector growth (annual, %)",
       color = NULL,
       subtitle = "Annual co-movement of industry with agriculture and services. Stronger positive slope = stronger spillovers.") +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 9))
```

Vietnam's industrial growth correlates positively with both agriculture
and services growth — VNM's industry pulls the rest of the economy along.
Cambodia's correlations are far weaker; KHM's industry can grow strongly
in years when agriculture is flat or falling. The decoupling is exactly
what you'd expect of an enclaved coalition: garment factories and SEZs
import their inputs (fabric from China, machinery from Korea), assemble
for export, and the value-added that stays in-country is wages plus
land/concession rents — neither of which feeds backward into domestic
agriculture or generates much downstream demand for sophisticated
services.

## Test 3: Productivity vs employment in industry

Manufacturing/industry value-added growth decomposes into a productivity
component (VA per worker) and an employment component (number of
workers). Magicians grow the productivity component; rentiers grow the
employment component, hiring more workers at flat productivity in the
quota-and-cheap-labour assembly model.

We use industry-level employment (the macro parquet has no
manufacturing-only employment series): industry employment share
`wdi_sl_ind_empl_zs` × total labour force `wdi_sl_tlf_totl_in`. For
Cambodia, industry is dominated by garments, so the proxy is reasonable;
for Vietnam, industry includes more heavy-industry and construction, so
the productivity component is biased *upward* relative to garments.

```{r productivity-decomp, fig.height=4.4, cache=TRUE}
ind_decomp <- sectoral_full %>%
  filter(iso3 %in% c("KHM", "VNM"),
         year >= 1995, year <= 2023,
         !is.na(wdi_nv_ind_totl_kd),
         !is.na(wdi_sl_ind_empl_zs),
         !is.na(wdi_sl_tlf_totl_in)) %>%
  mutate(ind_workers = wdi_sl_ind_empl_zs / 100 * wdi_sl_tlf_totl_in,
         va_per_worker = wdi_nv_ind_totl_kd / ind_workers) %>%
  select(iso3, year, ind_va = wdi_nv_ind_totl_kd,
         ind_workers, va_per_worker)

decomp_long <- ind_decomp %>%
  pivot_longer(c(ind_workers, va_per_worker, ind_va),
               names_to = "metric", values_to = "value") %>%
  mutate(metric_label = recode(metric,
    ind_va = "Industry VA",
    ind_workers = "Industry workers",
    va_per_worker = "VA per worker"
  ),
  metric_label = factor(metric_label,
    levels = c("Industry VA", "Industry workers", "VA per worker")))

ggplot(decomp_long, aes(year, value, color = iso3)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 0.7) +
  facet_wrap(~ metric_label, scales = "free_y", ncol = 3) +
  scale_y_log10(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_x_continuous(breaks = seq(1995, 2025, 10)) +
  labs(x = NULL, y = NULL, color = NULL,
       subtitle = "Industry VA decomposition: VA = workers x VA/worker. Log scales.") +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 9, face = "bold"))
```

```{r productivity-cagr-table}
prod_cagr <- ind_decomp %>%
  group_by(iso3) %>%
  arrange(year) %>%
  summarise(
    `Period`            = sprintf("%d-%d", first(year), last(year)),
    `Years`             = last(year) - first(year),
    `Industry VA CAGR`  = scales::percent(
      (last(ind_va) / first(ind_va))^(1 / (last(year) - first(year))) - 1,
      accuracy = 0.1),
    `Workers CAGR`      = scales::percent(
      (last(ind_workers) / first(ind_workers))^(1 / (last(year) - first(year))) - 1,
      accuracy = 0.1),
    `VA/worker CAGR`    = scales::percent(
      (last(va_per_worker) / first(va_per_worker))^(1 / (last(year) - first(year))) - 1,
      accuracy = 0.1),
    .groups = "drop"
  )
kable(prod_cagr,
      caption = "Industry productivity vs employment decomposition, KHM vs VNM.")
```

The decomposition tells a more nuanced story than the simple "rentiers
grow workers, magicians grow productivity" framing predicts:

- **Productivity CAGRs are roughly equal**: KHM 1.3%/yr vs VNM 1.2%/yr.
  Both countries are running on the same productivity treadmill in
  industry — neither is a runaway productivity story.
- **Workers grew much faster in KHM** (9.8%/yr vs 5.8%/yr) — Cambodia
  absorbed industrial labour ~70% faster than Vietnam at a *similar*
  productivity-growth rate, meaning **a larger share of KHM's industry
  growth came from the extensive margin** (12% from productivity vs
  17% for VNM).
- **Levels still differ**: VNM ended 2023 at ~$7K VA per industry
  worker; KHM at ~$6K. Vietnam was ahead in 1995 and remains ahead.

So the rentier signal here is partial. KHM's productivity isn't *flat*
in absolute terms (it rose 1.3%/yr), but it *is* a smaller share of
total industry growth than VNM's. The clean magician/rentier dichotomy
breaks down at the productivity-CAGR level; what survives is that
**Cambodia's industrial expansion is much more headcount-driven than
Vietnam's**, even with similar per-worker productivity gains. The
rentier reading wants to see Cambodia adding cheap labour into the
same low-productivity activities — and that's exactly the workers-vs-
productivity split this table shows.

## Test 4: PC2 trajectory and coalition consolidation

If Cambodia's industrial coalition is rentier-style, the political
economy story is that **rising managed-state capacity (PC2 trending
positive) should align with coalition-cementing events** — moments
when the CPP locks in the textile/SEZ/concession bargain and forecloses
challengers — not just with formal democratic-backsliding events. We
overlay PC2 with the relevant political and trade-policy moments.

```{r pc2-events, fig.height=4.2}
pc2_khm <- hist_pc_df %>%
  filter(iso3 == "KHM") %>%
  select(year, pc2 = PC2)

events <- tribble(
  ~year, ~event,                              ~ymax_offset,
  1991,  "UNTAC begins (Paris Accords)",      0.85,
  1993,  "UN-supervised election; coalition", 0.65,
  1997,  "Hun Sen coup; coalition ends",      0.45,
  2001,  "EBA: EU duty-free for KHM",         0.85,
  2005,  "SEZ law enacted",                   0.65,
  2015,  "Sam Rainsy exile",                  0.45,
  2017,  "CNRP dissolved",                    0.30,
  2020,  "EBA partial withdrawal",            0.15
)

pc2_range <- range(pc2_khm$pc2)
event_y <- function(o) pc2_range[1] + o * diff(pc2_range)

events <- events %>% mutate(label_y = event_y(ymax_offset))

ggplot(pc2_khm, aes(year, pc2)) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3) +
  geom_segment(data = events,
               aes(x = year, xend = year,
                   y = pc2_range[1], yend = label_y),
               color = "grey60", linetype = "dotted", linewidth = 0.4) +
  geom_line(linewidth = 0.8, color = highlight) +
  geom_point(size = 0.9, color = highlight) +
  geom_label_repel(data = events,
                   aes(x = year, y = label_y, label = event),
                   size = 2.5, label.size = 0.15,
                   nudge_y = 0.05,
                   segment.color = "grey50",
                   segment.size = 0.3,
                   min.segment.length = 0.1,
                   max.overlaps = 20) +
  scale_x_continuous(breaks = seq(1970, 2025, 5)) +
  labs(x = NULL, y = "PC2 score (contestation <- 0 -> managed-state)",
       subtitle = "Cambodia PC2 with political and trade-policy events overlaid.") +
  theme(legend.position = "none")
```

After discounting the 1975-1980 spike (a model artifact: NA-imputed-to-zero
fingerprints during state collapse), the genuine trajectory is a steady
**multi-decade climb from deep contestation toward managed-state values**:

- **PRK era (~1985-1990): mildly managed.** PC2 hovers around 0,
  reflecting the Vietnamese-installed administration's top-down
  control before the multi-party transition.
- **UNTAC and 1993 election:** PC2 falls sharply into deep-negative
  contestation territory (~-3) — multi-party period, refugee camps,
  rebel factions, and a fragile UN-supervised coalition.
- **1995-2010: PC2 stays at -2 to -3.** The deepest contestation
  values in the series. KR holdouts surrender 1998-99; the 1997 Hun
  Sen coup forcibly ends power-sharing but factional politics, opposition
  parties, and active civil-society mobilisation remain part of the V-Dem
  fingerprint through this entire window.
- **2010-2024: PC2 rises gradually from -3 toward 0.** Sam Rainsy
  exiled (2015), CNRP dissolved (2017), social-media regulation and
  one-party legalism harden — exactly the institutional features that
  load on positive PC2.

This is **precisely the trajectory the rentier-consolidation hypothesis
predicts**. The CPP entered the EBA-and-SEZ window (2001-2005) inside
a contested political space and used the rents from that window to
buy out, exile, or dissolve the contestation channels feature by
feature. PC2's slow climb from -3 toward 0 is the institutional
signature of that buy-out playing out — not a sign of liberalisation,
but of *managed-state lock-in*. The events overlaid on the chart
(1997 coup, 2015 exile, 2017 CNRP dissolution, 2020 EBA partial
withdrawal) cluster on the rising leg, not the falling one.

A natural follow-up — flagged in project memory — is to decompose
PC3 and PC4 (roughly the same size as PC2) to separate the
*contestation/mobilisation* and *regulatory/cyber-capacity* sub-axes
that PC2 currently bundles together. The story is robust at the PC2
level; PC3/PC4 should sharpen which features are doing the work.

## What the four tests jointly say

| Test | Magician would show | What KHM shows | Verdict |
|---|---|---|---|
| 1: ECI trajectory | Rising over decades | Rises modestly, stays negative | Rentier-leaning |
| 2: Spillovers | Industry growth lifts agri/services | Weak / no co-movement | Rentier |
| 3: Productivity vs employment | VA/worker rises sharply | Productivity CAGR matches VNM (1.3% vs 1.2%); workers grow 70% faster than VNM | Partial: headcount-driven expansion, but productivity not flat |
| 4: PC2 + events | n/a (political-economy mapping) | Deep contestation 1993-2010 (UNTAC, factional politics, KR remnants); steady rise toward managed-state 2010-2024 (Sam Rainsy exile, CNRP dissolution, social-media regulation) | Strong — slow climb to managed-state matches CPP coalition lock-in |

Vietnam — the obvious magician comparator — passes the first two
economic tests cleanly: ECI rises sharply (the strongest signal in the
panel), sectoral spillovers are present and strong. On productivity,
the picture is closer than expected: VNM's per-worker productivity
growth is 1.2%/yr versus KHM's 1.3% — essentially the same — and the
*magician* signature shows up in VNM's higher *level* (~$7K vs ~$6K
in 2023) and in productivity contributing a larger *share* (17% vs
12%) of total industry growth, not in a faster productivity CAGR.

Cambodia's diagnosis: **clearly rentier-leaning on capability
deepening and spillovers; only partially rentier on the productivity
margin** (extensive-margin growth dominates, but productivity isn't
stagnant). The political-economy story (test 4) is consistent: PC2
sits at -2 to -3 (deep contestation) through the 1993-2010 window
when the rentier coalition is being assembled inside a still-contested
political space, then climbs steadily toward zero (managed-state)
through the 2010s as the CPP exiles, dissolves, and absorbs its
remaining contestation channels.



## The simple "older = better" reading isn't supported

The cross-sectional model does **not** predict that earlier Cambodian
institutional setups would deliver better outcomes than the current one.
For Cambodia:

- **Predicted HCI is highest under the current 2018+ one-party CPP era
  (0.48)** and lowest in the 1998–2017 CPP-dominated era (0.44). The
  Coalition era (1993–97) predicts 0.45 — between the two CPP periods,
  not above them.
- **Predicted log-IMR is lowest (best) in the current era (3.16)** —
  comparable to the PRK and Coalition eras, and lower than the
  CPP-dominated era's 3.36.
- **Predicted life expectancy is highest in the current era (69.7)** —
  almost 5 years above the long flat plateau of 64–66 that the model
  reads off every prior era's institutions.
- **Predicted ECI** is the partial exception: it peaks under PRK
  (Vietnamese-style managed state, -0.23) and SOC/UNTAC, dips deepest
  under the Coalition era (-0.79), and only partially recovers under
  current institutions (-0.44).

The Khmer Rouge era's "high" predicted values (HCI 0.45, LE 66) are a
model artifact: the V-Dem fingerprint hits the floor on so many features
that XGBoost extrapolates to the closest training neighbours, all of
which sit far less repressively. Treat 1975–1978 predictions as
out-of-calibration.

## The actual story: convergence to the institutional ceiling

A different — and arguably more interesting — pattern is visible in the
predicted-vs-actual plots. For all four outcomes, **actual Cambodia has
caught up to predicted Cambodia**:

| Outcome | 1970s actual vs predicted | 2020s actual vs predicted |
|---|---|---|
| Life expectancy | actual far below predicted (~12 vs ~66 in 1976) | actual ~ predicted (~70 vs 70) |
| Infant mortality (log) | actual far above predicted (~5.4 vs ~3.2) | actual ~ predicted (~3.1 vs 3.2) |
| HCI | (no actual) | actual ~ predicted (0.49 vs 0.48) |
| ECI | actual volatile, often above predicted | actual ~ predicted (-0.5 vs -0.4) |

In the 1970s–1990s, Cambodia's actual outcomes were *worse* than its
institutional fingerprint predicted — the gap reflecting active state
collapse, war, and the genocidal Khmer Rouge baseline. As politics
stabilised post-1993, the gap closed: actuals recovered to the level
that the country's institutions could plausibly sustain. **By 2024
Cambodia is sitting on its institutional ceiling on every outcome we
measured.** Further improvement requires institutional improvement.

## What about Vietnam and Burundi?

- **Vietnam** is the cleanest "managed-state success" case. Predicted
  HCI rose from 0.50 (1970s) to 0.61 (2024) — a real institutional
  improvement. Actual HCI 2020 = 0.69 sits **above** predicted, the
  same kind of overperformance pattern Cambodia had decades ago.
  Vietnam's institutional fingerprint is delivering managed-state
  features (cyber capacity, bureaucratic competence, controlled but
  effective service delivery) that the model rewards.
- **Burundi** is the clearest decay case in the panel. Predicted HCI
  fell from 0.48 (mid-1980s) to 0.37 (2024). Actual HCI 2020 = 0.39
  matches predicted closely. **Burundi's institutional fingerprint
  has worsened, and outcomes have followed it down.**

This provides context for Cambodia's institutional twin status with
Burundi (the finding from `outcomes_ml.Rmd`'s peer analysis): the two
countries share a similar 2025 institutional fingerprint, but Cambodia
arrived at that position via a steady multi-decade *upgrade* (post-Khmer
Rouge), while Burundi arrived via institutional erosion. They look the
same in the cross-section but their trajectories — and the headroom
between actuals and institutional ceiling — differ.

## The rentier-coalition diagnosis explains the ceiling

The four sectoral tests above suggest *why* Cambodia's institutional
ceiling sits where it does. Vietnam's managed-state institutions are
delivering managed-state outcomes because the coalition behind them
is magician-style: rising ECI, sectoral spillovers, rising
productivity per worker. Cambodia's nominally similar institutional
features (the country's PC1 score is not far off Vietnam's; both are
managed states) deliver less because the coalition behind them is
rentier-style: garments and SEZs run as enclaves, ECI barely rises,
productivity per industry worker grows slowly, agriculture and
services don't co-move with industry.

This reframes the institutional-twin finding (KHM-BDI). Burundi and
Cambodia look the same on V-Dem features by 2025, but Burundi arrived
at that fingerprint by erosion of a once-better state, while Cambodia
arrived by a steady upgrade from genocide-era zero — an upgrade
*driven by* a rentier coalition that has been deliberately suppressing
contestation features (the negative-PC2 axis: anti-system mobilisation,
opposition camps, civil violence) and substituting state regulatory
capacity to protect its rents. The ceiling for both countries looks
similar from above; the political economy that enforces it is opposite.

## On the leading-indicator hypothesis

Strictly, the leading-indicator framing posed at the start — "predicted
IMR turns up before actual" — does not appear in the data: predicted
log-IMR for Cambodia is roughly flat at ~3.2 across 1970–2024, with no
clear deterioration. Whatever institutional decay shows up in
specific V-Dem features (election integrity, executive bribery, CSO
space, all of which peaked 1995–2005 and declined since 2010) is
**masked at the predicted-outcome level by simultaneous improvements
in administrative-capacity features** (cyber capacity, regulation,
basic services delivery) that the model also rewards.

In other words: the model reads "Cambodia 2025" as a *competent
soft-authoritarian* fingerprint, not a *decayed-democratic* one. The
features that have decayed (contestation, accountability) are the same
features the model finds weakly predictive of IMR/HCI/LE for this
class of country. So if those features have decayed without dragging
the predicted outcomes down, it suggests the model thinks they aren't
the binding constraint.

That doesn't mean the user's instinct is wrong — it means **the binding
constraint on Cambodia's next wave of human-capital and complexity
gains is institutional improvement on the features the model *does*
weight heavily**: judicial independence, election integrity, executive
oversight, parliamentary functionality. Those are the features whose
peak-and-decline the V-Dem trajectory plots show most clearly. The
predicted ceiling won't move up until those features do.

# Caveats

- **The model is cross-sectional and trained on 2025.** Projecting historical
  fingerprints through it assumes the *mapping* from institutions to outcomes
  has been stable over 50 years — which is a strong assumption. Returns to
  particular institutional features (e.g. parliamentary oversight) may have
  changed over time as global development conditions changed.
- **Min-max bounds use only the 2025 cross-section.** A 1975 V-Dem score
  more extreme than any 2025 country falls outside the [0, 1] range that
  XGBoost saw in training, and the model extrapolates. This will mostly
  affect Khmer Rouge–era predictions; treat that period as outside the
  model's calibration.
- **NAs are imputed to feature minimums.** When V-Dem couldn't code a feature
  because the institution didn't exist (no elections, no opposition press,
  no legislature), our pipeline sets the feature to its 2025 minimum. This
  is consistent with the cross-sectional pipeline but means predictions for
  Cambodia 1975–1991 partially reflect that imputation choice.
- **HCI actuals only exist 2010–2020.** The "predicted HCI under 1985
  institutions" cannot be validated directly; it's a counterfactual.
- **ECI depends on export structure.** Cambodia's ECI rise post-2004 is
  largely driven by the garment sector, which is itself a product of trade
  preferences and FDI rather than the deep institutional features V-Dem
  captures. Any predicted-vs-actual divergence on ECI is partly a story
  about trade policy, not institutions per se.
- **No causal claim.** This is a descriptive exercise mapping V-Dem
  fingerprints onto outcome predictions. Whether earlier Cambodian
  institutions *caused* better outcomes — or whether the institutions
  themselves were a downstream symptom of prior conditions — is not
  identified here.
- **Industry employment is a proxy for manufacturing employment.** The
  WDI macro parquet has no manufacturing-only employment series, so
  the productivity-vs-employment decomposition uses industry-level
  employment (`wdi_sl_ind_empl_zs` x `wdi_sl_tlf_totl_in`). For
  Cambodia, industry is dominated by garments, so the proxy is close.
  For Vietnam, industry includes more heavy industry and construction,
  so VNM's VA-per-worker is biased upward relative to a pure
  manufacturing comparison. The qualitative split (VNM grows
  productivity, KHM grows headcount) is robust to this.
- **Magician-vs-rentier diagnosis is suggestive, not definitive.** A
  fuller test would add manufacturing sub-sector concentration over
  time (incomplete for KHM) and preference-dependence on EBA / GSP
  (requires HS-6 atlas trade data). Both are flagged for follow-up.
