---
title: "Cambodia: Growth Breaks and Institutional Position"
subtitle: "Periodisation, sectoral trajectory, and country-space neighbours"
author: "Institutional Complexity Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6.5
    fig_height: 4.5
    latex_engine: xelatex
    includes:
      in_header: ~/dev/gl-design/skills/gl-ggplot/assets/gl_pdf.tex
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath("../.."))
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  cache = TRUE,
  cache.lazy = FALSE
)
```

```{r packages, cache=FALSE}
library(tidyverse)
library(arrow)
library(strucchange)
library(readxl)
library(ggrepel)
library(uwot)
library(knitr)
library(patchwork)
```

```{r gl-setup, cache=FALSE}
source("~/dev/gl-design/skills/gl-ggplot/assets/theme_gl.R")
gl_setup()
```

```{r constants}
ccode <- "KHM"
cname <- "Cambodia"
```

# Periodisation: Cambodia's growth breaks

We identify structural breaks in Cambodia's real GDP-per-capita growth using the
Bai-Perron procedure on Penn World Table 11.0 and overlay the Kar et al. (2013)
acceleration filter. The four-break solution gives the periodisation we use
throughout this document.

```{r growth-breaks-build, cache=TRUE}
pwt <- read_excel("~/dev/political-economy-pakistan/data/pwt110.xlsx",
                  sheet = "Data")

penn_growth <- pwt %>%
  filter(countrycode == ccode) %>%
  transmute(year, gdp_per_cap = rgdpna / pop) %>%
  filter(!is.na(gdp_per_cap)) %>%
  arrange(year) %>%
  mutate(growth_rate = (gdp_per_cap - lag(gdp_per_cap)) / lag(gdp_per_cap))

start_year <- min(penn_growth$year)
end_year   <- 2023
n_breaks   <- 4

penn_growth_ts <- ts(penn_growth$growth_rate[-1],
                     start = start_year, end = end_year, frequency = 1)
bp_growth    <- breakpoints(penn_growth_ts ~ 1)
break_years  <- Filter(Negate(is.na),
                       summary(bp_growth)$breakdates[n_breaks, ])

period_growthrates <- penn_growth %>%
  mutate(periods = cut(year,
                       breaks = c(start_year, break_years, end_year),
                       dig.lab = 4)) %>%
  group_by(periods) %>%
  summarise(avg_growth_rate = mean(growth_rate, na.rm = TRUE),
            mdn_growth_rate = median(growth_rate, na.rm = TRUE),
            length = n(),
            .groups = "drop") %>%
  mutate(start_period = as.numeric(str_match(as.character(periods),
                                             "\\(([0-9\\.]+),")[, 2]),
         end_period   = as.numeric(str_match(as.character(periods),
                                             ",([0-9\\.]+)\\]")[, 2]))

# Kar et al. (2013) acceleration filter on average growth-rate jumps
kar_filter <- period_growthrates %>%
  mutate(diff = avg_growth_rate - lag(avg_growth_rate),
         prev_dir = case_when(is.na(lag(diff)) ~ NA_character_,
                              lag(diff) > 0 ~ "acceleration",
                              lag(diff) < 0 ~ "deceleration",
                              TRUE ~ "no_change"),
         curr_dir = case_when(diff > 0 ~ "acceleration",
                              diff < 0 ~ "deceleration",
                              TRUE ~ "no_change"),
         threshold = case_when(is.na(prev_dir) ~ 0.02,
                               prev_dir != curr_dir ~ 0.03,
                               prev_dir == curr_dir ~ 0.01,
                               TRUE ~ NA_real_),
         passes_kar = case_when(is.na(diff) ~ NA,
                                abs(diff) >= threshold ~ TRUE,
                                TRUE ~ FALSE))

# Helper for breakpoint vector used downstream
all_break_years <- c(start_year, break_years, end_year)
```

```{r growth-breaks-plot, fig.height=4.5}
ggplot(penn_growth %>% filter(!is.na(growth_rate)),
       aes(year, growth_rate)) +
  geom_rect(data = kar_filter, inherit.aes = FALSE,
            aes(xmin = start_period, xmax = end_period,
                ymin = -Inf, ymax = Inf, fill = passes_kar),
            alpha = 0.18) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.2) +
  geom_segment(data = period_growthrates,
               aes(x = start_period, xend = end_period,
                   y = avg_growth_rate, yend = avg_growth_rate),
               linewidth = 0.7, color = gl$brand_blue) +
  geom_label(data = period_growthrates,
             aes(x = (start_period + end_period) / 2,
                 y = avg_growth_rate,
                 label = scales::percent(avg_growth_rate, accuracy = 0.1)),
             size = 3, label.size = 0.2) +
  scale_y_continuous(labels = scales::percent, n.breaks = 8) +
  scale_x_continuous(breaks = seq(1970, 2025, 5)) +
  scale_fill_manual(values = c("TRUE" = "#36B250", "FALSE" = "#C64646"),
                    na.value = "grey80") +
  labs(x = NULL, y = "Real GDP per capita growth rate",
       fill = "Passes Kar filter",
       title = paste(cname, "growth breaks"),
       subtitle = sprintf("Bai-Perron breaks at %s. Source: Penn World Tables 11.0.",
                          paste(break_years, collapse = ", "))) +
  theme(legend.position = "bottom")

ggsave("imgs/cambodia/growth_breaks.png", width = 9, height = 5, dpi = 300)
```

```{r growth-breaks-table}
period_growthrates %>%
  transmute(Period = as.character(periods),
            Years = length,
            `Avg. growth`    = scales::percent(avg_growth_rate, accuracy = 0.1),
            `Median growth`  = scales::percent(mdn_growth_rate, accuracy = 0.1)) %>%
  filter(!is.na(Years)) %>%
  kable(caption = sprintf("%s GDP-per-capita growth periods (PWT 11.0).", cname))
```

The Bai-Perron procedure gives break years **`r paste(break_years, collapse = ", ")`**.
We carry these forward as the periodisation for every chart below.

# Sectoral value-added across the growth breaks

Using WDI value-added series (constant 2015 USD), we plot each sector's
trajectory on a log scale, with vertical dashes at the growth-break years and
per-period CAGR labels.

```{r sectoral-load, cache=TRUE}
macro <- read_parquet(
  "data/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year",
                 "wdi_nv_agr_totl_kd", "wdi_nv_ind_totl_kd",
                 "wdi_nv_srv_totl_kd", "wdi_nv_ind_manf_kd",
                 "wdi_ne_exp_gnfs_zs",
                 "wdi_sl_agr_empl_zs", "wdi_sl_ind_empl_zs",
                 "wdi_sl_srv_empl_zs", "wdi_sl_tlf_totl_in")
) %>%
  rename(iso3 = countrycodeiso) %>%
  filter(iso3 == ccode)
```

```{r sectoral-va-plot, fig.height=4.7}
sector_levels <- c("agr" = "Agriculture",
                   "ind" = "Industry",
                   "srv" = "Services")

khm_sectoral <- macro %>%
  select(year, wdi_nv_agr_totl_kd, wdi_nv_ind_totl_kd, wdi_nv_srv_totl_kd) %>%
  pivot_longer(-year, names_to = "code", values_to = "value") %>%
  mutate(sector_short = str_remove(code, "wdi_nv_") %>% str_remove("_totl_kd"),
         Sector = sector_levels[sector_short]) %>%
  filter(!is.na(value), !is.na(Sector))

# Period CAGRs from observed values within each period
sector_period_cagrs <- function(df, value_col = "value") {
  df %>%
    mutate(period = cut(year,
                        breaks = all_break_years,
                        dig.lab = 4)) %>%
    filter(!is.na(period)) %>%
    group_by(Sector, period) %>%
    arrange(year) %>%
    summarise(start_year_p = first(year),
              end_year_p   = last(year),
              start_val    = first(.data[[value_col]]),
              end_val      = last(.data[[value_col]]),
              n            = end_year_p - start_year_p,
              cagr = if_else(n > 0 & start_val > 0,
                             (end_val / start_val)^(1 / n) - 1, NA_real_),
              avg_value = exp(mean(log(.data[[value_col]]), na.rm = TRUE)),
              .groups = "drop") %>%
    mutate(mid_year = (start_year_p + end_year_p) / 2)
}

va_period_cagrs <- sector_period_cagrs(khm_sectoral, "value")

sector_palette <- c("Agriculture" = "#EAC218",
                    "Industry"    = "#36B250",
                    "Services"    = "#266798")

ggplot(khm_sectoral, aes(year, value, color = Sector)) +
  geom_vline(xintercept = break_years,
             color = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 0.9) +
  geom_label_repel(data = va_period_cagrs %>% filter(!is.na(cagr)),
                   aes(x = mid_year, y = avg_value,
                       label = scales::percent(cagr, accuracy = 0.1)),
                   size = 2.8, label.size = 0.15,
                   min.segment.length = 0.2,
                   show.legend = FALSE) +
  scale_y_log10(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_x_continuous(breaks = seq(1970, 2025, 10)) +
  scale_color_manual(values = sector_palette) +
  labs(x = NULL, y = "Sectoral value added (constant 2015 USD, log)",
       title = sprintf("%s sectoral value added", cname),
       subtitle = sprintf("Vertical dashes mark growth breaks (%s). Labels are period CAGRs.",
                          paste(break_years, collapse = ", ")),
       color = NULL) +
  theme(legend.position = "bottom")

ggsave("imgs/cambodia/sectoral_value_added.png", width = 9, height = 5, dpi = 300)
```

# Exports as % of GDP across the growth breaks

WDI's exports-of-goods-and-services share starts in 1993 for Cambodia, so the
1979 and 1987 break lines fall left of the data. Period growth rates label the
average annual change in the export share within each visible period.

```{r exports-plot, fig.height=4.0}
khm_exports <- macro %>%
  select(year, share = wdi_ne_exp_gnfs_zs) %>%
  filter(!is.na(share))

# Use simple period-mean and period CAGR of the share itself
exports_cagrs <- khm_exports %>%
  mutate(period = cut(year,
                      breaks = all_break_years,
                      dig.lab = 4)) %>%
  filter(!is.na(period)) %>%
  group_by(period) %>%
  arrange(year) %>%
  summarise(start_year_p = first(year),
            end_year_p   = last(year),
            start_val    = first(share),
            end_val      = last(share),
            n            = end_year_p - start_year_p,
            cagr = if_else(n > 0 & start_val > 0,
                           (end_val / start_val)^(1 / n) - 1, NA_real_),
            mid_share = mean(share, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(mid_year = (start_year_p + end_year_p) / 2)

ggplot(khm_exports, aes(year, share)) +
  geom_vline(xintercept = break_years,
             color = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_line(linewidth = 0.7, color = gl$brand_blue) +
  geom_point(size = 1.0, color = gl$brand_blue) +
  geom_label_repel(data = exports_cagrs %>% filter(!is.na(cagr)),
                   aes(x = mid_year, y = mid_share,
                       label = scales::percent(cagr, accuracy = 0.1)),
                   size = 3, label.size = 0.15,
                   min.segment.length = 0.2) +
  scale_x_continuous(breaks = seq(1970, 2025, 5),
                     limits = c(start_year, end_year)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(x = NULL, y = "Exports of goods & services (% of GDP)",
       title = sprintf("%s: export orientation, 1993-present", cname),
       subtitle = "Vertical dashes mark growth breaks. Labels: period CAGR of the share.") +
  theme(legend.position = "none")

ggsave("imgs/cambodia/exports_share.png", width = 9, height = 5, dpi = 300)
```

# Sectoral labour productivity across the growth breaks

Value added per worker by sector. Sectoral employment shares
(`wdi_sl_*_empl_zs`) and total labour force (`wdi_sl_tlf_totl_in`) are available
from 1991 for Cambodia, so the 1979 and 1987 break lines fall left of the data.

```{r productivity-plot, fig.height=4.7}
khm_prod <- macro %>%
  filter(!is.na(wdi_sl_tlf_totl_in)) %>%
  transmute(year,
            agr_workers = wdi_sl_agr_empl_zs / 100 * wdi_sl_tlf_totl_in,
            ind_workers = wdi_sl_ind_empl_zs / 100 * wdi_sl_tlf_totl_in,
            srv_workers = wdi_sl_srv_empl_zs / 100 * wdi_sl_tlf_totl_in,
            agr_va = wdi_nv_agr_totl_kd,
            ind_va = wdi_nv_ind_totl_kd,
            srv_va = wdi_nv_srv_totl_kd) %>%
  mutate(Agriculture = agr_va / agr_workers,
         Industry    = ind_va / ind_workers,
         Services    = srv_va / srv_workers) %>%
  select(year, Agriculture, Industry, Services) %>%
  pivot_longer(-year, names_to = "Sector", values_to = "value") %>%
  filter(!is.na(value), is.finite(value))

prod_period_cagrs <- sector_period_cagrs(khm_prod, "value")

ggplot(khm_prod, aes(year, value, color = Sector)) +
  geom_vline(xintercept = break_years,
             color = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 0.9) +
  geom_label_repel(data = prod_period_cagrs %>% filter(!is.na(cagr)),
                   aes(x = mid_year, y = avg_value,
                       label = scales::percent(cagr, accuracy = 0.1)),
                   size = 2.8, label.size = 0.15,
                   min.segment.length = 0.2,
                   show.legend = FALSE) +
  scale_y_log10(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_x_continuous(breaks = seq(1990, 2025, 5)) +
  scale_color_manual(values = sector_palette) +
  labs(x = NULL, y = "Value added per worker (constant 2015 USD, log)",
       title = sprintf("%s: sectoral labour productivity", cname),
       subtitle = "Sectoral VA divided by sector employment. Vertical dashes: growth breaks. Labels: period CAGR.",
       color = NULL) +
  theme(legend.position = "bottom")

ggsave("imgs/cambodia/sectoral_productivity.png", width = 9, height = 5, dpi = 300)
```

# Country space at each break year

Using the V-Dem-based country-space pipeline from
`analyses/country-space-writeup.Rmd`, we re-build the space at the original
year (1970), at each Cambodia growth-break year (1979, 1987, 1997, 2006), and
at the present (2025). Cambodia is highlighted in each panel.

```{r vdem-load, cache=TRUE}
vdem_raw      <- read_parquet("data/shared-data/vdem/vdem.parquet")
vdem_codebook <- read_parquet("data/shared-data/vdem/vdem_codebook.parquet")
```

```{r country-space-pipeline, cache=TRUE}
# Re-implementation of the country-space pipeline. Kept self-contained here
# rather than sourced from the writeup Rmd; if a third caller appears, lift
# this to an R/ helper.

discrete_vars <- c("v2regopploc", "v2regsuploc")

redundant_drops <- c(
  "v2cldmovew", "v2cldiscw", "v2clslavef", "v2clprptyw", "v2clacjstw",
  "v2juncind", "v2juhccomp", "v2csrlgrep", "v2csrlgcon", "v2exdfdmhs",
  "v2smgovsm", "v2lgqstexp", "v2lglegplo", "v2merange", "v2dlreason",
  "v2psoppaut", "v2cafexch", "v2caconmob"
)

region_labels <- c(
  "1" = "E. Europe & Central Asia",  "2" = "Latin America & Caribbean",
  "3" = "Middle East & N. Africa",   "4" = "Sub-Saharan Africa",
  "5" = "W. Europe & N. America",    "6" = "East Asia & Pacific"
)

region_lookup <- vdem_raw %>%
  group_by(country_name) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(country_name, e_regionpol_6C) %>%
  mutate(region = region_labels[as.character(e_regionpol_6C)]) %>%
  select(country_name, region) %>%
  deframe()

suffixes        <- c("_codelow", "_codehigh", "_sd", "_osp",
                     "_ord", "_mean", "_nr", "_rec")
all_cols        <- names(vdem_raw)
is_suffix       <- vapply(all_cols, function(x)
                          any(vapply(suffixes, endsWith, logical(1), x = x)),
                          logical(1))
is_numbered     <- grepl("_\\d+$", all_cols)
base_cols       <- all_cols[!is_suffix & !is_numbered]
component_codes <- vdem_codebook %>%
                     filter(vartype == "C") %>% pull(tag) %>% unique()
candidate_codes <- intersect(component_codes, base_cols) %>%
                     setdiff(discrete_vars) %>%
                     setdiff(redundant_drops)
candidate_codes <- candidate_codes[
  vapply(candidate_codes, function(x) is.numeric(vdem_raw[[x]]), logical(1))
]

build_country_space <- function(target_year,
                                min_indicator_coverage = 0.5,
                                max_country_missing    = 0.5,
                                seed = 42,
                                n_neighbors = 15,
                                min_dist = 0.2,
                                n_epochs = 500) {
  year_data <- vdem_raw %>% filter(year == target_year)
  n_total   <- nrow(year_data)

  coverage  <- vapply(candidate_codes,
                      function(x) sum(!is.na(year_data[[x]])),
                      integer(1))
  threshold <- min_indicator_coverage * n_total
  indicators <- candidate_codes[coverage >= threshold]

  raw <- year_data %>%
    select(country_name, all_of(indicators)) %>%
    mutate(p_missing = rowSums(is.na(across(all_of(indicators)))) /
                       length(indicators)) %>%
    filter(p_missing < max_country_missing) %>%
    select(-p_missing) %>%
    column_to_rownames("country_name") %>%
    as.matrix()

  for (j in seq_len(ncol(raw))) {
    na_idx <- is.na(raw[, j])
    if (any(na_idx)) raw[na_idx, j] <- mean(raw[, j], na.rm = TRUE)
  }

  Mz <- scale(raw, center = TRUE, scale = TRUE)
  Mz <- Mz[, !is.nan(colSums(Mz)), drop = FALSE]

  set.seed(seed)
  emb <- umap(Mz, n_neighbors = n_neighbors,
              min_dist = min_dist, n_epochs = n_epochs)

  coords <- tibble(
    country = rownames(Mz),
    UMAP1   = emb[, 1],
    UMAP2   = emb[, 2],
    region  = unname(region_lookup[country])
  ) %>%
    mutate(region = replace_na(region, "Other"))

  list(year = target_year,
       n_countries = nrow(Mz),
       n_indicators = ncol(Mz),
       indicators = colnames(Mz),
       Mz = Mz,
       coords = coords)
}
```

```{r build-snapshots, cache=TRUE}
snapshot_years <- c(1970, break_years, 2025)
spaces <- map(snapshot_years, build_country_space)
names(spaces) <- as.character(snapshot_years)

space_summary <- map_dfr(spaces, function(s)
  tibble(year = s$year,
         n_countries  = s$n_countries,
         n_indicators = s$n_indicators))
kable(space_summary,
      caption = "Country-space pipeline output by snapshot year.")
```

```{r country-space-plot-fn}
plot_country_space_khm <- function(sp,
                                   target = "Cambodia",
                                   label_size = 2.4,
                                   point_size = 2.0) {
  d <- sp$coords %>%
    mutate(is_target = country == target)

  ggplot(d, aes(UMAP1, UMAP2)) +
    geom_point(aes(color = region), size = point_size, alpha = 0.65) +
    geom_point(data = ~ filter(.x, is_target),
               color = highlight, size = 4.2) +
    geom_text_repel(
      data = ~ filter(.x, !is_target),
      aes(label = country),
      size = label_size, max.overlaps = 30,
      segment.alpha = 0.25, color = "grey25",
      show.legend = FALSE
    ) +
    geom_label_repel(
      data = ~ filter(.x, is_target),
      aes(label = country),
      size = 3.6, color = highlight, fontface = "bold",
      box.padding = 0.6, point.padding = 0.4,
      label.size = 0.4, label.padding = unit(0.25, "lines"),
      segment.color = highlight, segment.alpha = 0.7,
      show.legend = FALSE
    ) +
    labs(x = "UMAP 1", y = "UMAP 2", color = NULL,
         title = sprintf("Country space, %d", sp$year),
         subtitle = sprintf("%d countries x %d V-Dem indicators",
                            sp$n_countries, sp$n_indicators)) +
    theme(legend.position = "bottom")
}
```

```{r save-individual-spaces, fig.show='hide', cache=FALSE}
for (yr in snapshot_years) {
  p <- plot_country_space_khm(spaces[[as.character(yr)]])
  ggsave(sprintf("imgs/cambodia/country_space_%d.png", yr),
         p, width = 9, height = 9, dpi = 300)
}
```

\newpage

## 1970

```{r plot-1970, fig.width=7, fig.height=8.0, out.width="100%"}
plot_country_space_khm(spaces[["1970"]])
```

\newpage

## 1979

```{r plot-1979, fig.width=7, fig.height=8.0, out.width="100%"}
plot_country_space_khm(spaces[["1979"]])
```

\newpage

## 1987

```{r plot-1987, fig.width=7, fig.height=8.0, out.width="100%"}
plot_country_space_khm(spaces[["1987"]])
```

\newpage

## 1997

```{r plot-1997, fig.width=7, fig.height=8.0, out.width="100%"}
plot_country_space_khm(spaces[["1997"]])
```

\newpage

## 2006

```{r plot-2006, fig.width=7, fig.height=8.0, out.width="100%"}
plot_country_space_khm(spaces[["2006"]])
```

\newpage

## 2025

```{r plot-2025, fig.width=7, fig.height=8.0, out.width="100%"}
plot_country_space_khm(spaces[["2025"]])
```

\newpage

# Cambodia's k nearest peers at each break year

For each snapshot we compute Euclidean distance in the **z-scored V-Dem feature
space** (`Mz`, not the UMAP coordinates) between Cambodia and every other country
present in that year. The UMAP layout is for visualisation; the high-dimensional
space is what defines "institutional similarity."

```{r nearest-peers, cache=TRUE, dependson="build-snapshots"}
k_peers <- 10

khm_neighbors <- function(sp, target = "Cambodia", k = k_peers) {
  if (!target %in% rownames(sp$Mz)) {
    return(tibble(rank = integer(), country = character(),
                  distance = double(), region = character()))
  }
  v   <- sp$Mz[target, , drop = TRUE]
  d   <- sqrt(rowSums((sp$Mz - matrix(v, nrow = nrow(sp$Mz),
                                      ncol = ncol(sp$Mz),
                                      byrow = TRUE))^2))
  d   <- d[names(d) != target]
  ord <- order(d)[seq_len(min(k, length(d)))]
  tibble(rank = seq_along(ord),
         country = names(d)[ord],
         distance = unname(d[ord]),
         region = unname(region_lookup[names(d)[ord]]))
}

peers_long <- map_dfr(snapshot_years, function(yr) {
  khm_neighbors(spaces[[as.character(yr)]]) %>%
    mutate(year = yr) %>%
    relocate(year)
})

# Wide table: one column per snapshot year, k_peers rows.
peers_wide <- peers_long %>%
  mutate(label = sprintf("%s (%.2f)", country, distance)) %>%
  select(year, rank, label) %>%
  pivot_wider(names_from = year, values_from = label,
              names_prefix = "y_") %>%
  arrange(rank)

names(peers_wide) <- str_replace(names(peers_wide), "^y_", "")

kable(peers_wide,
      caption = sprintf("%s's %d nearest peers in V-Dem feature space at each snapshot year. Number in parentheses is Euclidean distance on z-scored features.",
                        cname, k_peers))
```

```{r nearest-peers-2025}
peers_long %>%
  filter(year == 2025) %>%
  transmute(Rank     = rank,
            Country  = country,
            Region   = region,
            Distance = round(distance, 2)) %>%
  kable(caption = sprintf("%s's %d nearest peers today (2025 V-Dem cross-section). Same Euclidean distance on z-scored features as the multi-year table above; the wide table is too narrow to fit the 2025 column on the page so we list it separately here.",
                          cname, k_peers))
```

```{r persist-peers}
write_csv(peers_long, "outputs/cambodia_peers_by_break_year.csv")
```

\newpage

# Institutional position over time (pooled PCA)

The lens here is the pooled-panel PCA fit in
`analyses/pooled-panel-pca.Rmd`. Fitting on every (country, year) cell from
1970-2025 (rather than on 2025 alone) gives axes whose loadings reflect
variance over the full window we care about, which is the right choice for
trajectory analysis. Two axes survive the pool with clear interpretation:

- **PC1 — openness of the regime.** Positive = liberal democracy
  (free expression, free press, autonomous electoral management,
  independent judiciary, peaceful assembly). Negative = personalist /
  ideological autocracy with a state-controlled information environment.
- **PC2 — stateness / encompassment.** Positive = the state encompasses
  society (broad public-goods delivery, uniform subnational
  administration, clean everyday administration; *and* state ownership of
  economy, state-owned media, controlled CSO and party space). Negative =
  society encompasses the state (private economy and media, autonomous
  CSOs and universities, judicial review *or* political violence,
  insurgent movements, weak central authority).

The two axes are orthogonal by construction. Welfare democracies and
developmental autocracies share encompassment (high PC2) despite opposing
each other on PC1; liberal-market democracies and weak / contested
polities share non-state weight (low PC2) despite opposing each other on
PC1. See `pooled-panel-pca.Rmd` for the full loading detail and sign
verification.

```{r load-pooled-pca, cache=TRUE}
pooled_scores <- read_csv("outputs/pooled_pca/country_year_pc_scores.csv",
                          show_col_types = FALSE)

# Sign convention (verified against pooled-panel-pca.Rmd top-loadings table):
# - PC1 positive pole loads on freedom of discussion, peaceful assembly,
#   freedom from torture, EMB autonomy → high PC1 = liberal democracy.
# - PC2 positive pole loads on educational/health equality, state-owned
#   broadcast media, state ownership of economy (after V-Dem reverse
#   coding) → high PC2 = state encompasses society.
# No flips needed.

pooled_scores <- pooled_scores %>%
  mutate(region = unname(region_lookup[country_name]),
         region = replace_na(region, "Other"))
```

```{r pooled-quadrants-plot, fig.width=7.2, fig.height=9.5, out.width="100%"}
pooled_snapshots <- pooled_scores %>%
  filter(year %in% snapshot_years) %>%
  mutate(is_target = country_name == "Cambodia")

axis_lim <- max(abs(c(pooled_snapshots$PC1, pooled_snapshots$PC2)),
                na.rm = TRUE) * 1.05

quadrant_labels_pool <- tibble(
  label = c("State-encompassing\ndemocracy",
            "State-encompassing\nautocracy",
            "Liberal-market\ndemocracy",
            "Weak / contested\nautocracy"),
  x     = c( 1, -1,  1, -1) * axis_lim * 0.95,
  y     = c( 1,  1, -1, -1) * axis_lim * 0.95,
  hjust = c( 1,  0,  1,  0),
  vjust = c( 1,  1,  0,  0)
)

ggplot(pooled_snapshots, aes(PC1, PC2)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey50", alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey50", alpha = 0.6) +
  geom_text(data = quadrant_labels_pool,
            aes(x = x, y = y, label = label,
                hjust = hjust, vjust = vjust),
            color = "grey45", size = 2.2, lineheight = 0.9,
            inherit.aes = FALSE) +
  geom_point(aes(color = is_target, size = is_target, alpha = is_target)) +
  geom_label_repel(data = ~ filter(.x, is_target),
                   aes(label = country_name),
                   color = highlight, fontface = "bold",
                   size = 3, label.size = 0.3,
                   label.padding = unit(0.2, "lines"),
                   box.padding = 0.5, point.padding = 0.3,
                   segment.color = highlight,
                   show.legend = FALSE) +
  facet_wrap(~ year, ncol = 2) +
  scale_color_manual(values = c("FALSE" = "grey55", "TRUE" = highlight)) +
  scale_size_manual(values = c("FALSE" = 0.9, "TRUE" = 2.6)) +
  scale_alpha_manual(values = c("FALSE" = 0.45, "TRUE" = 1)) +
  scale_x_continuous(limits = c(-axis_lim, axis_lim)) +
  scale_y_continuous(limits = c(-axis_lim, axis_lim)) +
  labs(x = "PC1 — openness of regime (high = liberal democracy)",
       y = "PC2 — stateness (high = state encompasses society)",
       title = sprintf("%s in pooled (PC1, PC2) space, by snapshot year", cname),
       subtitle = "Each panel: every country-year that survived the pooled-PCA filter at that year.") +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 10),
        panel.spacing = unit(0.8, "lines"))

ggsave("imgs/cambodia/pooled_quadrants_by_year.png",
       width = 9, height = 12, dpi = 300)
```

\newpage

## Cambodia's PC1 and PC2 over time

Pooled-PCA scores carry the full 1970-`r max(pooled_scores$year)` trajectory
for every country in the panel. We plot Cambodia's two co-ordinates
separately so each axis can be read on its own scale.

```{r khm-pooled-pc1, fig.height=3.5}
khm_pcs <- pooled_scores %>%
  filter(country_name == "Cambodia",
         year >= 1970) %>%
  arrange(year)

ggplot(khm_pcs, aes(year, PC1)) +
  geom_vline(xintercept = break_years,
             color = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  geom_line(linewidth = 0.8, color = highlight) +
  geom_point(size = 1.0, color = highlight) +
  scale_x_continuous(breaks = seq(1970, 2025, 5)) +
  labs(x = NULL, y = "PC1 (high = liberal democracy)",
       title = sprintf("%s PC1 over time (pooled PCA)", cname),
       subtitle = "Vertical dashes mark growth breaks (1979, 1987, 1997, 2006).")

ggsave("imgs/cambodia/khm_pooled_pc1.png", width = 9, height = 4.5, dpi = 300)
```

```{r khm-pooled-pc2, fig.height=3.5}
ggplot(khm_pcs, aes(year, PC2)) +
  geom_vline(xintercept = break_years,
             color = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  geom_line(linewidth = 0.8, color = highlight) +
  geom_point(size = 1.0, color = highlight) +
  scale_x_continuous(breaks = seq(1970, 2025, 5)) +
  labs(x = NULL, y = "PC2 (high = state encompasses society)",
       title = sprintf("%s PC2 over time (pooled PCA)", cname),
       subtitle = "Vertical dashes mark growth breaks (1979, 1987, 1997, 2006).")

ggsave("imgs/cambodia/khm_pooled_pc2.png", width = 9, height = 4.5, dpi = 300)
```

PC1 sits deep in the personalist-autocracy region across the entire window:
the post-Khmer-Rouge PRK years and the 1990s UNTAC interlude move PC1
modestly up, but from 2000 onward Hun Sen's consolidation reverses that.
PC2 traces the substantive state-building story more cleanly: negative
through the post-conflict 1990s and 2000s (society encompasses state — weak
central authority, autonomous donors and NGOs, civil-war hangover), then
trending steadily upward from 2010 as the regime extends administrative
reach, party-system control, and media-environment control. By the end of
the window Cambodia has crossed into PC2-positive territory — a
state-encompassing autocracy.

\newpage

## 2025 cross-section, original single-year PCA

For comparison, we keep the original single-year PCA built on the 2025
V-Dem cross-section. PC1 there reads as openness in the same direction as
the pooled fit; PC2 carries the *managed-vs-contested* reading used in the
deals-development writeup (high = contested politics — Pakistan, Lebanon —
low = managed). Because the loadings are pulled toward post-2000 modules
that don't exist before 2000, this PCA is the right lens for *where
Cambodia sits today* relative to its contemporaries, but a poor lens for
trajectories. The 2025-only fit is also what feeds the PC2 driver
decomposition in the next section.

```{r settlement-pca-2025, cache=TRUE}
# Mirror the institutional-complexity / deals-development pipeline:
# vartype C indicators with >=150-country coverage in 2025, drop the two
# discrete vars, min-max normalize across the 2025 cross-section, NA -> 0.
latest_year_settle <- vdem_raw %>%
  filter(year >= 2020) %>%
  group_by(year) %>%
  summarise(n_countries = n_distinct(country_name), .groups = "drop") %>%
  filter(n_countries >= 150) %>%
  pull(year) %>% max()

settle_candidates <- vdem_codebook %>%
  filter(vartype == "C") %>% pull(tag) %>% unique() %>%
  intersect(base_cols) %>%
  setdiff(discrete_vars)

vdem_2025 <- vdem_raw %>% filter(year == latest_year_settle)
settle_coverage <- vapply(settle_candidates,
                          function(x) sum(!is.na(vdem_2025[[x]])),
                          integer(1))
settle_numeric  <- vapply(settle_candidates,
                          function(x) is.numeric(vdem_raw[[x]]),
                          logical(1))
settle_indicators <- settle_candidates[settle_coverage >= 150 & settle_numeric]

mcp_df_2025 <- vdem_2025 %>%
  select(country_name, all_of(settle_indicators)) %>%
  mutate(p_missing = rowSums(is.na(across(all_of(settle_indicators)))) /
                     length(settle_indicators)) %>%
  filter(p_missing < 0.5) %>%
  select(-p_missing)

# Cache 2025 min/max bounds for projecting history.
feat_bounds <- tibble(
  code = settle_indicators,
  min  = vapply(settle_indicators,
                function(x) min(mcp_df_2025[[x]], na.rm = TRUE),
                double(1)),
  max  = vapply(settle_indicators,
                function(x) max(mcp_df_2025[[x]], na.rm = TRUE),
                double(1))
)

apply_2025_bounds <- function(df) {
  cols_present <- intersect(feat_bounds$code, names(df))
  for (ind in cols_present) {
    rng <- c(feat_bounds$min[feat_bounds$code == ind],
             feat_bounds$max[feat_bounds$code == ind])
    df[[ind]] <- if (rng[2] > rng[1])
      (df[[ind]] - rng[1]) / (rng[2] - rng[1]) else 0
  }
  df %>% mutate(across(all_of(cols_present), ~replace_na(.x, 0)))
}

mcp_2025_norm <- apply_2025_bounds(mcp_df_2025)
Mcp_2025 <- mcp_2025_norm %>%
  column_to_rownames("country_name") %>%
  as.matrix()

pca_2025 <- prcomp(Mcp_2025, center = TRUE, scale. = TRUE)

# Sign conventions for the 2025 PCA:
# - PC1 positive = liberal democracy (anchor on Sweden / Germany).
# - PC2 positive = contested (anchor on Pakistan / Lebanon).
pc1_2025_scores <- pca_2025$x[, "PC1"]
anchor_pc1 <- intersect(c("Sweden", "Germany"), names(pc1_2025_scores))
pc1_sign_2025 <- if (length(anchor_pc1) > 0 &&
                     mean(pc1_2025_scores[anchor_pc1]) < 0) -1 else 1

pc2_2025_scores <- pca_2025$x[, "PC2"]
anchor_countries <- intersect(c("Pakistan", "Lebanon"), names(pc2_2025_scores))
pc2_sign <- if (length(anchor_countries) > 0 &&
                mean(pc2_2025_scores[anchor_countries]) < 0) -1 else 1
```

```{r settlement-2025-quadrant-plot, fig.width=7.2, fig.height=8.0, out.width="100%"}
scores_2025 <- tibble(
  country_name = rownames(pca_2025$x),
  PC1 = pc1_sign_2025 * pca_2025$x[, "PC1"],
  PC2 = pc2_sign      * pca_2025$x[, "PC2"]
) %>%
  mutate(is_target = country_name == "Cambodia",
         region = unname(region_lookup[country_name]),
         region = replace_na(region, "Other"))

axis_lim_2025 <- max(abs(c(scores_2025$PC1, scores_2025$PC2)),
                     na.rm = TRUE) * 1.05

quadrant_labels_2025 <- tibble(
  label = c("Competitive\nclientelist",
            "Vulnerable\nauthoritarian\ncoalition",
            "Consolidated\ndemocracy",
            "Dominant\nparty"),
  x     = c( 1, -1,  1, -1) * axis_lim_2025 * 0.95,
  y     = c( 1,  1, -1, -1) * axis_lim_2025 * 0.95,
  hjust = c( 1,  0,  1,  0),
  vjust = c( 1,  1,  0,  0)
)

ggplot(scores_2025, aes(PC1, PC2)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey50", alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey50", alpha = 0.6) +
  geom_text(data = quadrant_labels_2025,
            aes(x = x, y = y, label = label,
                hjust = hjust, vjust = vjust),
            color = "grey45", size = 2.4, lineheight = 0.9,
            inherit.aes = FALSE) +
  geom_point(aes(color = is_target, size = is_target, alpha = is_target)) +
  geom_text_repel(data = ~ filter(.x, !is_target),
                  aes(label = country_name),
                  size = 2.2, max.overlaps = 30,
                  segment.alpha = 0.25, color = "grey25",
                  show.legend = FALSE) +
  geom_label_repel(data = ~ filter(.x, is_target),
                   aes(label = country_name),
                   color = highlight, fontface = "bold",
                   size = 3.2, label.size = 0.3,
                   label.padding = unit(0.2, "lines"),
                   box.padding = 0.6, point.padding = 0.4,
                   segment.color = highlight,
                   show.legend = FALSE) +
  scale_color_manual(values = c("FALSE" = "grey55", "TRUE" = highlight)) +
  scale_size_manual(values = c("FALSE" = 0.9, "TRUE" = 2.6)) +
  scale_alpha_manual(values = c("FALSE" = 0.45, "TRUE" = 1)) +
  scale_x_continuous(limits = c(-axis_lim_2025, axis_lim_2025)) +
  scale_y_continuous(limits = c(-axis_lim_2025, axis_lim_2025)) +
  labs(x = "PC1 — openness (high = liberal democracy)",
       y = "PC2 — contested vs managed (high = contested)",
       title = sprintf("%s in 2025 cross-sectional PCA space", cname),
       subtitle = "Original single-year PCA on the 2025 V-Dem cross-section.") +
  theme(legend.position = "none")

ggsave("imgs/cambodia/khm_2025_pc1_pc2_quadrant.png",
       width = 9, height = 9.5, dpi = 300)
```

\newpage

# Drivers of Cambodia's PC2 climb, 2006-2025

Cambodia's pooled PC2 climbs from clearly negative in 2006 (society
encompasses state — autonomous donors and NGOs, civil-war hangover, weak
central authority) to positive in 2025 (state encompasses society —
administrative reach, party-system control, and media-environment control
all extending under Hun Sen). We decompose that move indicator by indicator
using the pooled PCA loadings, to see which V-Dem features did the work.

The change in Cambodia's PC2 from year A to year B is additive across
indicators:

$$\Delta \text{PC2} = \sum_j \text{loading}_j \cdot (z_{B,j} - z_{A,j})$$

where $z_{Y,j} = (x_{Y,j} - \mu_j) / \sigma_j$ is the pool-z-scored value of
indicator $j$ in year $Y$, $(\mu_j, \sigma_j)$ are the pool mean and
standard deviation cached by the pooled PCA fit, and $\text{loading}_j$ is
the indicator's coefficient in the PC2 rotation. NAs are imputed to the
pool mean and so contribute zero to that year's z-score by construction.
We sort by absolute contribution and keep the top.

```{r pc2-drivers-compute, cache=TRUE}
n_top_drivers <- 12

pca_pool_obj      <- readRDS("outputs/pooled_pca/pca_pool.rds")
pool_features     <- pca_pool_obj$indicators
pool_means        <- pca_pool_obj$feature_means[pool_features]
pool_sds          <- pca_pool_obj$feature_sds[pool_features]
loadings_pc2_pool <- pca_pool_obj$rotation[pool_features, "PC2"]
# Pooled PC2 sign convention: positive pole = state encompasses society
# (verified in load-pooled-pca chunk; no sign flip needed).

khm_endpoints_raw <- vdem_raw %>%
  filter(country_name == "Cambodia",
         year %in% c(2006, 2025)) %>%
  select(year, any_of(pool_features))

get_raw <- function(yr) {
  row <- khm_endpoints_raw %>% filter(year == yr)
  vals <- setNames(rep(NA_real_, length(pool_features)), pool_features)
  cols_present <- intersect(names(row), pool_features)
  if (nrow(row) == 1)
    vals[cols_present] <- as.numeric(row[1, cols_present, drop = TRUE])
  vals
}
raw_2006 <- get_raw(2006)
raw_2025 <- get_raw(2025)

impute_pool <- function(v) ifelse(is.na(v), pool_means, v)
z_2006 <- (impute_pool(raw_2006) - pool_means) / pool_sds
z_2025 <- (impute_pool(raw_2025) - pool_means) / pool_sds

pc2_drivers <- tibble(
  code         = pool_features,
  raw_2006     = raw_2006,
  raw_2025     = raw_2025,
  delta_z      = z_2025 - z_2006,
  loading      = loadings_pc2_pool,
  contribution = loadings_pc2_pool * (z_2025 - z_2006)
) %>%
  arrange(desc(abs(contribution))) %>%
  left_join(
    vdem_codebook %>%
      transmute(code = tag, name) %>%
      distinct(code, .keep_all = TRUE),
    by = "code"
  )

top_drivers <- pc2_drivers %>%
  slice_head(n = n_top_drivers)

cat(sprintf("Total PC2 change 2006 -> 2025 (pooled PCA): %+.3f\n",
            sum(pc2_drivers$contribution, na.rm = TRUE)))
cat(sprintf("Top %d drivers explain %.1f%% of the absolute movement.\n",
            n_top_drivers,
            100 * sum(abs(top_drivers$contribution), na.rm = TRUE) /
                  sum(abs(pc2_drivers$contribution),  na.rm = TRUE)))
```

```{r pc2-drivers-table}
top_drivers %>%
  transmute(Tag           = code,
            Indicator     = name,
            `2006 (raw)`  = ifelse(is.na(raw_2006), "—",
                                    sprintf("%.2f", raw_2006)),
            `2025 (raw)`  = ifelse(is.na(raw_2025), "—",
                                    sprintf("%.2f", raw_2025)),
            `Delta z`     = round(delta_z, 2),
            `PC2 loading` = round(loading, 3),
            `Contribution` = round(contribution, 3)) %>%
  kable(caption = sprintf("Top %d V-Dem indicators by contribution to Cambodia's pooled-PC2 change, 2006 to 2025. Raw values are V-Dem native units (— = NA in that year, imputed to the pool mean before z-scoring). Contribution = loading * (z_2025 - z_2006); positive = pushed PC2 up = state encompasses society more.",
                          n_top_drivers))
```

```{r pc2-drivers-plot, fig.width=7.4, fig.height=8.5, out.width="100%"}
top_codes <- top_drivers$code

driver_panel <- vdem_raw %>%
  filter(country_name == "Cambodia",
         year >= 1970, year <= 2025) %>%
  select(year, all_of(top_codes)) %>%
  pivot_longer(-year, names_to = "code", values_to = "value") %>%
  filter(!is.na(value)) %>%
  left_join(
    top_drivers %>%
      transmute(code,
                facet_label = sprintf("%s\n(%s, contrib %+.2f)",
                                       name, code, contribution),
                contribution),
    by = "code"
  ) %>%
  mutate(facet_label = factor(facet_label,
                              levels = sprintf("%s\n(%s, contrib %+.2f)",
                                               top_drivers$name,
                                               top_drivers$code,
                                               top_drivers$contribution)))

ggplot(driver_panel, aes(year, value)) +
  geom_vline(xintercept = break_years,
             color = "grey70", linetype = "dashed", linewidth = 0.3) +
  geom_line(linewidth = 0.6, color = highlight) +
  geom_point(size = 0.8, color = highlight) +
  facet_wrap(~ facet_label, ncol = 3, scales = "free_y") +
  scale_x_continuous(breaks = seq(1970, 2025, 15)) +
  labs(x = NULL, y = "V-Dem score (raw, native units)",
       title = sprintf("Top %d drivers of %s's pooled-PC2 shift, 2006-2025",
                       n_top_drivers, cname),
       subtitle = "Each panel: raw V-Dem score over time. Sign of contribution shown in panel title (+ = pushed PC2 up = state encompasses society more; - = pushed it down = society encompasses state more). Vertical dashes: growth breaks.") +
  theme(strip.text = element_text(size = 7.2, lineheight = 0.9),
        panel.spacing = unit(0.6, "lines"))

ggsave("imgs/cambodia/khm_pc2_top_drivers.png",
       width = 9, height = 11, dpi = 300)
```

```{r pc2-drivers-persist}
write_csv(pc2_drivers, "outputs/cambodia_pc2_drivers_2006_2025.csv")
```

\newpage

# Rents space and political settlement, Cambodia and peers

We bring in two views from `analyses/deals-development.Rmd` — the
4×6 stakeholder × policy-demand satisfaction grid (PSW's rents space)
and the horizontal × vertical political-settlement scatter — and focus
both on Cambodia plus three peer cohorts:

- **Institutional peers** — Cambodia's nearest neighbours in the 2025
  V-Dem country space (top 4 from the peers table above).
- **Geographic peers** — its mainland and maritime ASEAN neighbours.
- **Economic peers** — countries within ±25% of Cambodia's 2023
  GDP-per-capita (PWT 11.0), excluding ones already picked up by the
  other cohorts.

The deals-development pipeline writes its per-country axis scores and
settlement scores to CSV, so we just load those rather than re-fitting.

```{r build-peer-set, cache=TRUE, dependson="nearest-peers"}
deals_axes <- read_csv("outputs/deals_axis_scores.csv",
                       show_col_types = FALSE)
settle_scores <- read_csv("outputs/settlement_scores.csv",
                          show_col_types = FALSE)

name_iso <- deals_axes %>% select(country_name, iso3) %>% distinct()

# --- Institutional peers: top 4 in 2025 V-Dem country space ---
institutional_iso <- peers_long %>%
  filter(year == 2025) %>%
  slice_head(n = 4) %>%
  inner_join(name_iso, by = c("country" = "country_name")) %>%
  pull(iso3)

# --- Geographic peers: ASEAN minus Brunei ---
geographic_iso <- c("VNM", "LAO", "THA", "MMR", "MYS", "IDN", "PHL", "SGP")

# --- Economic peers: GDP-per-capita within +/- 25% of Cambodia in 2023 ---
khm_gdp_pc_2023 <- pwt %>%
  filter(countrycode == "KHM", year == 2023) %>%
  mutate(gdp_pc = rgdpna / pop) %>% pull(gdp_pc)

econ_candidates <- pwt %>%
  filter(year == 2023, !is.na(rgdpna), !is.na(pop), pop > 1) %>%
  mutate(gdp_pc = rgdpna / pop) %>%
  filter(countrycode != "KHM",
         abs(gdp_pc - khm_gdp_pc_2023) / khm_gdp_pc_2023 < 0.25) %>%
  arrange(abs(gdp_pc - khm_gdp_pc_2023))

# Drop ones already in institutional/geographic; keep up to 4
already <- c(institutional_iso, geographic_iso)
economic_iso <- econ_candidates %>%
  filter(!countrycode %in% already) %>%
  slice_head(n = 4) %>%
  pull(countrycode)

peer_set <- bind_rows(
  tibble(iso3 = "KHM",            group_raw = "Cambodia"),
  tibble(iso3 = institutional_iso, group_raw = "Institutional"),
  tibble(iso3 = geographic_iso,    group_raw = "Geographic"),
  tibble(iso3 = economic_iso,      group_raw = "Economic")
) %>%
  mutate(group_order = case_when(
    group_raw == "Cambodia"      ~ 1L,
    group_raw == "Institutional" ~ 2L,
    group_raw == "Geographic"    ~ 3L,
    group_raw == "Economic"      ~ 4L
  )) %>%
  group_by(iso3) %>%
  slice_min(group_order, n = 1, with_ties = FALSE) %>%   # dedupe, keep first cohort
  ungroup() %>%
  inner_join(name_iso, by = "iso3") %>%
  arrange(group_order, country_name) %>%
  mutate(group = factor(group_raw,
                        levels = c("Cambodia", "Institutional",
                                   "Geographic", "Economic")))

peer_iso <- peer_set$iso3
```

```{r peer-set-table}
peer_set %>%
  transmute(Group   = group_raw,
            ISO3    = iso3,
            Country = country_name) %>%
  kable(caption = sprintf("Peer set used in the rents-grid and political-settlement charts (n = %d). Each country appears in its first matching cohort only.",
                          length(peer_iso)))
```

## Rents-space report card

For each country in the peer set we colour the 4 (stakeholder) × 6
(policy demand) grid by where the country sits on each policy-demand
axis: red = bottom third of the world distribution, yellow = middle
third, green = top third. Cells where the stakeholder doesn't demand
that policy are left blank (the demand structure is in PSW's Table 1.1
as encoded in `deals-development.Rmd`).

Reading down a column tells you which stakeholder coalitions would be
served by the country's strongest policies; reading across a row tells
you whether a given stakeholder's whole bundle is satisfied. The
Cambodia panel comes first; institutional peers next, then ASEAN,
then GDP-per-capita peers.

```{r rents-grid-build}
psw_prefs <- tribble(
  ~stakeholder,
  ~transparent_rules, ~common_infra, ~specialized_infra,
  ~contract_enforcement, ~subsidies_clientelism, ~exclusivity,
  "Magicians",    1, 1, 1, 1, 1, 0,
  "Workhorses",   1, 1, 0, 1, 0, 0,
  "Rentiers",     0, 0, 1, 1, 1, 1,
  "Powerbrokers", 0, 0, 0, 0, 0, 1
)

stakeholder_levels <- c("Magicians", "Workhorses", "Rentiers", "Powerbrokers")
axis_levels <- c("transparent_rules", "common_infra", "specialized_infra",
                 "contract_enforcement", "subsidies_clientelism", "exclusivity")
axis_short_labels <- c(
  transparent_rules     = "Transp. rules",
  common_infra          = "Common infra",
  specialized_infra     = "Special. infra",
  contract_enforcement  = "Contract enf.",
  subsidies_clientelism = "Subsidies",
  exclusivity           = "Exclusivity"
)

psw_prefs_long <- psw_prefs %>%
  pivot_longer(-stakeholder, names_to = "axis", values_to = "demanded") %>%
  mutate(stakeholder = factor(stakeholder, levels = stakeholder_levels),
         axis        = factor(axis, levels = axis_levels))

axis_long <- deals_axes %>%
  filter(iso3 %in% peer_iso) %>%
  pivot_longer(-c(country_name, iso3),
               names_to = "axis", values_to = "axis_score") %>%
  mutate(axis = factor(axis, levels = axis_levels))

country_order_levels <- peer_set$country_name

rents_grid <- expand_grid(
  iso3        = peer_iso,
  stakeholder = factor(stakeholder_levels, levels = stakeholder_levels),
  axis        = factor(axis_levels, levels = axis_levels)
) %>%
  inner_join(psw_prefs_long, by = c("stakeholder", "axis")) %>%
  left_join(axis_long %>% select(iso3, axis, axis_score),
            by = c("iso3", "axis")) %>%
  inner_join(peer_set %>% select(iso3, country_name, group),
             by = "iso3") %>%
  mutate(
    band = case_when(
      demanded == 0     ~ NA_character_,
      is.na(axis_score) ~ "Missing",
      axis_score < 1/3  ~ "Bottom third",
      axis_score < 2/3  ~ "Middle third",
      TRUE              ~ "Top third"
    ),
    band = factor(band,
                  levels = c("Bottom third", "Middle third",
                             "Top third", "Missing")),
    facet_label = sprintf("%s (%s)", country_name, group),
    facet_label = factor(facet_label,
                         levels = sprintf("%s (%s)",
                                          peer_set$country_name,
                                          peer_set$group))
  )

rents_palette <- c(
  "Bottom third" = "#d73027",
  "Middle third" = "#fee08b",
  "Top third"    = "#1a9850",
  "Missing"      = "grey60"
)
```

```{r rents-grid-plot, fig.width=7.4, fig.height=8.8, out.width="100%"}
ggplot(rents_grid,
       aes(x = axis, y = fct_rev(stakeholder), fill = band)) +
  geom_tile(color = "white", linewidth = 0.6, na.rm = FALSE) +
  facet_wrap(~ facet_label, ncol = 3) +
  scale_fill_manual(values = rents_palette,
                    na.value = NA, drop = FALSE) +
  scale_x_discrete(labels = axis_short_labels) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = sprintf("Policy-demand satisfaction matrix — %s and peers",
                       cname),
       subtitle = "Coloured = stakeholder demands this policy; blank = no demand. Bands are world-wide percentile thirds on each policy-demand axis.") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 7, angle = 45, hjust = 1, vjust = 1),
        axis.text.y = element_text(size = 8),
        strip.text  = element_text(size = 8, face = "bold"))

ggsave("imgs/cambodia/rents_grid_khm_peers.png",
       width = 9, height = 11, dpi = 300)
```

\newpage

## Political settlement — horizontal × contested (PC2 overlay)

The companion view from `deals-development.Rmd` swaps the elite-mass
vertical axis for *managed-vs-contested*: the within-world percentile
of PC2 from the institutional-complexity PCA on the 2025 V-Dem
cross-section. High PC2 = anti-system mobilization, polarization,
political violence (Pakistan, Lebanon); low PC2 = political conflict
absorbed by institutions (Singapore, China, Saudi Arabia). The
quadrants map onto Khan's typology more directly than the elite-mass
read does.

```{r settlement-scatter-pc2, fig.width=6.5, fig.height=6.5, out.width="100%"}
inst_pca_pct <- read_csv("outputs/institutional_pca_scores.csv",
                         show_col_types = FALSE) %>%
  select(country_name, PC2_pct)

settle_with_pc2 <- settle_scores %>%
  left_join(inst_pca_pct, by = "country_name") %>%
  mutate(contested = PC2_pct / 100) %>%
  filter(!is.na(contested)) %>%
  left_join(peer_set %>% select(iso3, group), by = "iso3")

group_palette <- c(
  "Cambodia"      = highlight,
  "Institutional" = "#266798",
  "Geographic"    = "#36B250",
  "Economic"      = "#EAC218"
)

quadrant_labels <- tibble(
  label = c("Competitive\nclientelist",
            "Vulnerable\nauthoritarian\ncoalition",
            "Consolidated\ndemocracy",
            "Dominant\nparty"),
  x     = c(0.05, 0.95, 0.05, 0.95),
  y     = c(0.97, 0.97, 0.03, 0.03),
  hjust = c(0,    1,    0,    1),
  vjust = c(1,    1,    0,    0)
)

ggplot(settle_with_pc2, aes(horizontal, contested)) +
  geom_vline(xintercept = 0.5, linetype = "dashed",
             color = "grey50", alpha = 0.4) +
  geom_hline(yintercept = 0.5, linetype = "dashed",
             color = "grey50", alpha = 0.4) +
  geom_text(data = quadrant_labels,
            aes(x = x, y = y, label = label,
                hjust = hjust, vjust = vjust),
            color = "grey45", size = 2.4, lineheight = 0.9,
            inherit.aes = FALSE) +
  geom_point(data = ~ filter(.x, is.na(group)),
             color = "grey75", alpha = 0.45, size = 1.0) +
  geom_point(data = ~ filter(.x, !is.na(group)),
             aes(color = group), size = 2.6) +
  geom_text_repel(data = ~ filter(.x, !is.na(group)),
                  aes(label = iso3, color = group),
                  size = 3, fontface = "bold",
                  max.overlaps = 30, segment.alpha = 0.4) +
  scale_color_manual(values = group_palette, drop = FALSE) +
  scale_x_continuous(limits = c(0, 1),
                     labels = scales::percent_format(accuracy = 1),
                     breaks = c(0, 0.5, 1)) +
  scale_y_continuous(limits = c(0, 1),
                     labels = scales::percent_format(accuracy = 1),
                     breaks = c(0, 0.5, 1)) +
  labs(x = "Horizontal: concentration among political groups (high = one dominant)",
       y = "Contested state (high = mobilized / chaotic politics, low = managed)",
       color = NULL,
       title = sprintf("Political settlement — %s and peers", cname),
       subtitle = "Horizontal: within-world percentile of group concentration. Vertical: within-world percentile of PC2 from the 2025 institutional-complexity PCA.") +
  theme(legend.position = "bottom")

ggsave("imgs/cambodia/settlement_horiz_pc2_khm_peers.png",
       width = 9, height = 9, dpi = 300)
```

\newpage

# Regime classification over time

V-Dem's `v2x_regime` is the four-category Regimes of the World typology:
0 = closed autocracy, 1 = electoral autocracy, 2 = electoral democracy,
3 = liberal democracy. We plot Cambodia's series with the growth-break
years overlaid.

```{r khm-regime-classification, fig.height=3.8}
khm_regime <- vdem_raw %>%
  filter(country_name == "Cambodia", year >= 1970, !is.na(v2x_regime)) %>%
  select(year, v2x_regime) %>%
  arrange(year)

regime_levels <- c("Closed\nautocracy", "Electoral\nautocracy",
                   "Electoral\ndemocracy", "Liberal\ndemocracy")

ggplot(khm_regime, aes(year, v2x_regime)) +
  geom_vline(xintercept = break_years,
             color = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_step(linewidth = 0.8, color = highlight) +
  geom_point(size = 1.2, color = highlight) +
  scale_x_continuous(breaks = seq(1970, 2025, 5)) +
  scale_y_continuous(breaks = 0:3, labels = regime_levels,
                     limits = c(-0.2, 3.2)) +
  labs(x = NULL, y = NULL,
       title = sprintf("%s regime classification, 1970-%d",
                       cname, max(khm_regime$year)),
       subtitle = "V-Dem Regimes of the World (v2x_regime). Vertical dashes mark growth breaks.")

ggsave("imgs/cambodia/khm_regime_classification.png",
       width = 9, height = 4.5, dpi = 300)
```

\newpage

# Winning coalition size

The Bueno de Mesquita & Smith (2022) winning-coalition index `W` is the mean
of four z-scored V-Dem components (z-scores computed across all country-year
observations): election-monitoring autonomy (`v2elembaut`), opposition-party
autonomy (`v2psoppaut`), barriers to political-party participation
(`v2psbars`), and the negated maximum of the three closed-succession indices
(`v2x_ex_hereditary`, `v2x_ex_military`, `v2x_ex_party`). Higher `W` =
larger winning coalition. See `analyses/winning_coalition.R` for the
construction.

```{r winning-coalition-build, cache=TRUE}
zscore <- function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)

w_panel <- vdem_raw %>%
  select(country_name, year,
         v2elembaut, v2psoppaut, v2psbars,
         v2x_ex_hereditary, v2x_ex_military, v2x_ex_party) %>%
  mutate(closed_succession = -pmax(v2x_ex_hereditary,
                                   v2x_ex_military,
                                   v2x_ex_party,
                                   na.rm = FALSE),
         z_elembaut = zscore(v2elembaut),
         z_oppaut   = zscore(v2psoppaut),
         z_bars     = zscore(v2psbars),
         z_succ     = zscore(closed_succession)) %>%
  rowwise() %>%
  mutate(W = mean(c(z_elembaut, z_oppaut, z_bars, z_succ),
                  na.rm = FALSE)) %>%
  ungroup() %>%
  select(country_name, year, W)
```

```{r khm-winning-coalition, fig.height=3.8}
khm_w <- w_panel %>%
  filter(country_name == "Cambodia", year >= 1970, !is.na(W))

ggplot(khm_w, aes(year, W)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  geom_vline(xintercept = break_years,
             color = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_line(linewidth = 0.8, color = highlight) +
  scale_x_continuous(breaks = seq(1970, 2025, 5)) +
  labs(x = NULL, y = "W (z-score units)",
       title = sprintf("%s winning coalition size, 1970-%d",
                       cname, max(khm_w$year)),
       subtitle = "BdM & Smith (2022) index. Higher = larger winning coalition. Vertical dashes mark growth breaks.")

ggsave("imgs/cambodia/khm_winning_coalition.png",
       width = 9, height = 4.5, dpi = 300)
```

\newpage

## Peer trajectories

The same index for every country that appeared as one of Cambodia's ten
nearest peers at any of the 1970, 1979, 1987, 1997, or 2006 snapshots.

```{r khm-peers-winning-coalition, fig.width=6.5, fig.height=9, out.width="100%"}
peer_countries <- peers_long %>% pull(country) %>% unique()

peers_w <- w_panel %>%
  filter(country_name %in% peer_countries, year >= 1970, !is.na(W))

ggplot(peers_w, aes(year, W)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  geom_vline(xintercept = break_years,
             color = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_line(linewidth = 0.6, color = highlight) +
  facet_wrap(~ country_name, ncol = 5) +
  scale_x_continuous(breaks = c(1980, 2000, 2020)) +
  labs(x = NULL, y = "W (z-score units)",
       title = "Peer winning-coalition trajectories, 1970-2025",
       subtitle = sprintf("%d distinct nearest-peer countries of %s across the 1970-2006 snapshots. Vertical dashes: %s growth breaks.",
                          length(peer_countries), cname, cname))

ggsave("imgs/cambodia/khm_peers_winning_coalition.png",
       width = 9, height = 12, dpi = 300)
```

\newpage

# Tax revenue composition

IMF World Revenue Longitudinal Database (WoRLD, 2026 release). Cambodia
coverage begins in 1999, just after the post-UNTAC fiscal regime stabilised.
Values are % of GDP. The four tax aggregates are income taxes (`TaxInc`),
goods & services / VAT (`TaxSal`), international trade / customs (`TaxTra`),
and a small residual of property + other taxes (`TaxPro` + `TaxOth`).

```{r khm-tax-load, cache=TRUE}
imf_world <- suppressMessages(
  readxl::read_excel("data/shared-data/world-imf2026.xlsx",
                     sheet = "Data", guess_max = 20000)
)

khm_tax <- imf_world %>%
  filter(ISO3 == "KHM") %>%
  mutate(TaxOthAll = coalesce(TaxPro, 0) + coalesce(TaxOth, 0)) %>%
  select(year, TaxInc, TaxSal, TaxTra, TaxOthAll, TaxRev) %>%
  filter(!is.na(TaxRev))

khm_tax_long <- khm_tax %>%
  select(-TaxRev) %>%
  pivot_longer(-year, names_to = "component", values_to = "pct_gdp") %>%
  mutate(component = recode(component,
    "TaxInc"    = "Income taxes",
    "TaxSal"    = "Goods & services (VAT/sales)",
    "TaxTra"    = "International trade",
    "TaxOthAll" = "Property + other"
  ),
  component = factor(component, levels = c(
    "International trade", "Goods & services (VAT/sales)",
    "Income taxes", "Property + other"
  )))
```

```{r khm-tax-stacked, fig.height=4.5}
ggplot(khm_tax_long, aes(year, pct_gdp, fill = component)) +
  geom_vline(xintercept = break_years,
             color = "grey40", linetype = "dashed", linewidth = 0.4) +
  geom_area(alpha = 0.92) +
  scale_x_continuous(breaks = seq(1999, 2025, 4),
                     limits = c(min(khm_tax_long$year),
                                max(khm_tax_long$year))) +
  scale_y_continuous(labels = scales::label_percent(scale = 1, accuracy = 1)) +
  labs(x = NULL, y = "% of GDP", fill = NULL,
       title = sprintf("%s tax revenue composition, %d-%d",
                       cname,
                       min(khm_tax_long$year),
                       max(khm_tax_long$year)),
       subtitle = "Source: IMF WoRLD 2026. Vertical dashes mark growth breaks.") +
  theme(legend.position = "bottom")

ggsave("imgs/cambodia/khm_tax_composition_pct_gdp.png",
       width = 9, height = 5, dpi = 300)
```

The level rises from ~7% of GDP in 1999 to a 2017-2019 peak around 14-16%,
then stabilises through the COVID period. The composition shift is the
clearer story: international-trade taxes were the single largest source in
1999 (~3.2% of GDP, ~45% of tax revenue) and have since fallen in both
absolute and share terms as Cambodia entered ASEAN trade arrangements and
joined the WTO (2004). VAT / sales taxes took over as the dominant base.

```{r khm-tax-shares, fig.height=4}
khm_tax_shares <- khm_tax %>%
  mutate(across(c(TaxInc, TaxSal, TaxTra, TaxOthAll),
                ~ .x / TaxRev)) %>%
  select(-TaxRev) %>%
  pivot_longer(-year, names_to = "component", values_to = "share") %>%
  mutate(component = recode(component,
    "TaxInc"    = "Income taxes",
    "TaxSal"    = "Goods & services (VAT/sales)",
    "TaxTra"    = "International trade",
    "TaxOthAll" = "Property + other"
  ),
  component = factor(component, levels = c(
    "Goods & services (VAT/sales)", "International trade",
    "Income taxes", "Property + other"
  )))

ggplot(khm_tax_shares, aes(year, share, color = component)) +
  geom_vline(xintercept = break_years,
             color = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_line(linewidth = 0.9) +
  scale_x_continuous(breaks = seq(1999, 2025, 4)) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1),
                     limits = c(0, NA)) +
  labs(x = NULL, y = "Share of tax revenue", color = NULL,
       title = sprintf("%s tax mix, share of total tax revenue", cname),
       subtitle = "Source: IMF WoRLD 2026. Vertical dashes mark growth breaks.") +
  theme(legend.position = "bottom")

ggsave("imgs/cambodia/khm_tax_shares.png",
       width = 9, height = 5, dpi = 300)
```

