---
title: "Institutional Complexity Analysis"
subtitle: "Applying Economic Complexity Methods to V-Dem Governance Data"
author: "Institutional Complexity Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 10
    fig_height: 7
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  cache = TRUE,
  cache.lazy = FALSE
)
```

# Introduction

Economic complexity theory, developed by Hidalgo and Hausmann (2009), provides a powerful
framework for understanding why some countries are wealthier than others. The core insight is
elegant: countries that can produce many products -- especially products that few other countries
can produce -- tend to be more economically sophisticated.

Here we apply this same logic to **institutions and governance**. Using data from the
Varieties of Democracy (V-Dem) project, we ask: which countries have built complex institutional
portfolios? Which institutional features are "rare" -- present only in countries with deep
institutional capacity? And how do institutions cluster together in an "institution space"?

The analogy works as follows:

| Economic Complexity | Institutional Complexity |
|---|---|
| Countries export products | Countries have institutional features |
| Revealed Comparative Advantage | Binarized V-Dem indicators |
| Product Complexity Index (PCI) | Institutional Feature Complexity (IFC) |
| Economic Complexity Index (ECI) | Institutional Complexity Index (ICI) |
| Product Space | Institution Space |

# Setup and Data

## Load packages

```{r packages}
library(tidyverse)
library(vdemdata)
library(uwot)
library(ggrepel)
library(countrycode)
library(WDI)
library(pheatmap)
library(igraph)
library(RColorBrewer)
library(knitr)
library(DT)
```

## Load V-Dem data

We use the `vdemdata` package which provides the full V-Dem dataset. We'll work with the most
recent year that has good coverage.

```{r load-vdem, cache=TRUE}
vdem_raw <- vdem

# Check the most recent years and coverage
year_coverage <- vdem_raw %>%
  filter(year >= 2020) %>%
  group_by(year) %>%
  summarise(
    n_countries = n_distinct(country_name),
    .groups = "drop"
  )

kable(year_coverage, caption = "Country coverage by year (2020+)")
```

```{r select-year}
# Use the most recent year with substantial coverage
latest_year <- year_coverage %>%
  filter(n_countries >= 150) %>%
  pull(year) %>%
  max()

cat("Selected year:", latest_year, "\n")
```

## Select institutional indicators

We curate a set of institutional and governance indicators spanning multiple domains. V-Dem
indicators come in several forms:

- **Ordinal (0-4)**: Expert-coded assessments (e.g., `v2x_*` indices are often interval, but
  many component indicators like `v2cl*`, `v2ju*` are ordinal)
- **Interval (0-1)**: Measurement-model aggregated indices
- **Continuous**: Some structural variables

We focus on indicators that capture the *presence* or *quality* of specific institutional features.

```{r select-indicators}
# Define our indicator set with categories
indicator_info <- tribble(
  ~code, ~category, ~description,
  # Rule of Law
  "v2x_rule", "Rule of Law", "Rule of law index",
  "v2xcl_rol", "Rule of Law", "Equality before the law and individual liberty index",
  "v2clrspct", "Rule of Law", "Rigorous and impartial public administration",
  "v2cltrnslw", "Rule of Law", "Transparent laws with predictable enforcement",
  "v2clacjust", "Rule of Law", "Access to justice for men",
  "v2clacjstw", "Rule of Law", "Access to justice for women",
  "v2clsocgrp", "Rule of Law", "Social class equality in respect for civil liberties",
  "v2clrgunev", "Rule of Law", "Even geographic distribution of civil liberties",

  # Judicial Independence
  "v2x_jucon", "Judicial Independence", "Judicial constraints on the executive index",
  "v2juhcind", "Judicial Independence", "High court independence",
  "v2juncind", "Judicial Independence", "Lower court independence",
  "v2juhccomp", "Judicial Independence", "Compliance with high court decisions",
  "v2jucomp", "Judicial Independence", "Compliance with judiciary",
  "v2jureview", "Judicial Independence", "Judicial review",
  "v2juaccnt", "Judicial Independence", "Judicial accountability",
  "v2jupack", "Judicial Independence", "Court packing",
  "v2jupurge", "Judicial Independence", "Purges of the judiciary",

  # Bureaucratic Quality / State Capacity
  "v2stcritrecadm", "State Capacity", "Criteria for appointment decisions in state administration",
  "v2strenadm", "State Capacity", "Remuneration in state administration",
  "v2x_pubcorr", "State Capacity", "Public sector corruption index",
  "v2clrspct", "State Capacity", "Rigorous and impartial public administration",
  "v2svstterr", "State Capacity", "State authority over territory",
  "v2stfisci", "State Capacity", "Fiscal capacity",
  "v2ststatag", "State Capacity", "State statistical agency capacity",

  # Property Rights
  "v2xcl_prpty", "Property Rights", "Property rights index",
  "v2clprptym", "Property Rights", "Property rights for men",
  "v2clprptyw", "Property Rights", "Property rights for women",

  # Media Freedom
  "v2x_freexp_altinf", "Media Freedom", "Freedom of expression and alternative sources of info index",
  "v2mecenefm", "Media Freedom", "Government censorship effort -- media",
  "v2meharjrn", "Media Freedom", "Harassment of journalists",
  "v2meslfcen", "Media Freedom", "Media self-censorship",
  "v2mebias", "Media Freedom", "Media bias",
  "v2mecrit", "Media Freedom", "Print/broadcast media critical",
  "v2merange", "Media Freedom", "Print/broadcast media perspectives",
  "v2x_freexp", "Media Freedom", "Freedom of expression index",
  "v2meaccess", "Media Freedom", "Media access",

  # Electoral Integrity
  "v2x_polyarchy", "Electoral Integrity", "Electoral democracy index",
  "v2xel_frefair", "Electoral Integrity", "Clean elections index",
  "v2elfrfair", "Electoral Integrity", "Free and fair elections",
  "v2elirreg", "Electoral Integrity", "Election irregularities",
  "v2elintim", "Electoral Integrity", "Election government intimidation",
  "v2elpeace", "Electoral Integrity", "Election other voting irregularities",
  "v2elvotbuy", "Electoral Integrity", "Election vote buying",
  "v2elembaut", "Electoral Integrity", "EMB autonomy",
  "v2elembcap", "Electoral Integrity", "EMB capacity",
  "v2elrgstry", "Electoral Integrity", "Election voter registry",
  "v2x_elecoff", "Electoral Integrity", "Elected officials index",

  # Civil Liberties
  "v2x_civlib", "Civil Liberties", "Civil liberties index",
  "v2xcl_disc", "Civil Liberties", "Freedom of discussion index",
  "v2clacfree", "Civil Liberties", "Freedom of academic and cultural expression",
  "v2cldiscm", "Civil Liberties", "Freedom of discussion for men",
  "v2cldiscw", "Civil Liberties", "Freedom of discussion for women",
  "v2clfmove", "Civil Liberties", "Freedom of foreign movement",
  "v2cldmovem", "Civil Liberties", "Freedom of domestic movement for men",
  "v2cldmovew", "Civil Liberties", "Freedom of domestic movement for women",
  "v2clkill", "Civil Liberties", "Freedom from political killings",
  "v2clslavem", "Civil Liberties", "Freedom from forced labor for men",
  "v2clslavef", "Civil Liberties", "Freedom from forced labor for women",
  "v2cltort", "Civil Liberties", "Freedom from torture",
  "v2clstown", "Civil Liberties", "State ownership of economy",
  "v2clrelig", "Civil Liberties", "Freedom of religion",

  # Legislative Constraints
  "v2x_legcon", "Legislative Constraints", "Legislative constraints on executive index",
  "v2lgqstexp", "Legislative Constraints", "Legislature questions officials in practice",
  "v2lginvstp", "Legislative Constraints", "Legislature investigates in practice",
  "v2lgotovst", "Legislative Constraints", "Legislature opposition parties oversight",
  "v2lgcrrpt", "Legislative Constraints", "Legislature corrupt activities",
  "v2lgoppart", "Legislative Constraints", "Legislature opposition parties",
  "v2lgfunds", "Legislative Constraints", "Legislature controls resources",
  "v2lgdsadlo", "Legislative Constraints", "Legislature disbands/self-limits",
  "v2lgello", "Legislative Constraints", "Legislature elections are free/fair",

  # Anti-Corruption
  "v2x_corr", "Anti-Corruption", "Political corruption index",
  "v2excrptps", "Anti-Corruption", "Executive corruption (public sector)",
  "v2exthftps", "Anti-Corruption", "Executive embezzlement",
  "v2exbribe", "Anti-Corruption", "Executive bribery",
  "v2lgcrrpt", "Anti-Corruption", "Legislature corrupt activities",
  "v2jucorrdc", "Anti-Corruption", "Judicial corruption decision",

  # Association and Civil Society
  "v2x_cspart", "Civil Society", "Civil society participation index",
  "v2csprtcpt", "Civil Society", "Civil society participatory environment",
  "v2cseeorgs", "Civil Society", "CSO entry and exit",
  "v2csreprss", "Civil Society", "CSO repression",
  "v2csantimv", "Civil Society", "CSO anti-system movements",
  "v2x_frassoc_thick", "Civil Society", "Freedom of association (thick) index",

  # Academic Freedom
  "v2xca_academ", "Academic Freedom", "Academic and cultural freedom index",
  "v2caassemb", "Academic Freedom", "Freedom of academic exchange",
  "v2cafexch", "Academic Freedom", "Freedom of academic exchange and dissemination",
  "v2casurv", "Academic Freedom", "Surveillance of academic activities",

  # Deliberation
  "v2x_delibdem", "Deliberation", "Deliberative democracy index",
  "v2dlreason", "Deliberation", "Reasoned justification",
  "v2dlcommon", "Deliberation", "Common good justification",
  "v2dlcountr", "Deliberation", "Respect for counterarguments",
  "v2dlconslt", "Deliberation", "Range of consultation",
  "v2dlengage", "Deliberation", "Engaged society",

  # Equality
  "v2x_gender", "Equality", "Women's political empowerment index",
  "v2x_genpp", "Equality", "Women's civil society participation index",
  "v2x_gencl", "Equality", "Women's civil liberties index",
  "v2pepwrgen", "Equality", "Power distributed by gender",
  "v2pepwrses", "Equality", "Power distributed by socioeconomic position",
  "v2pepwrsoc", "Equality", "Power distributed by social group",

  # Accountability
  "v2x_accountability", "Accountability", "Accountability index",
  "v2x_veracc", "Accountability", "Vertical accountability index",
  "v2x_horacc", "Accountability", "Horizontal accountability index",
  "v2x_diagacc", "Accountability", "Diagonal accountability index"
)

# Remove duplicates (some indicators appear in multiple categories -- keep first)
indicator_info <- indicator_info %>% distinct(code, .keep_all = TRUE)

cat("Total unique indicators selected:", nrow(indicator_info), "\n")
cat("Categories:\n")
indicator_info %>% count(category, sort = TRUE) %>% print(n = 20)
```

```{r filter-data}
# Check which of our selected indicators actually exist in the dataset
available_codes <- intersect(indicator_info$code, names(vdem_raw))
missing_codes <- setdiff(indicator_info$code, names(vdem_raw))

if (length(missing_codes) > 0) {
  cat("Indicators not found in dataset (will be dropped):\n")
  cat(paste(missing_codes, collapse = ", "), "\n\n")
}

indicator_info <- indicator_info %>% filter(code %in% available_codes)
cat("Indicators available:", nrow(indicator_info), "\n")

# Filter to selected year and extract relevant columns
vdem_filtered <- vdem_raw %>%
  filter(year == latest_year) %>%
  select(
    country_name, country_text_id, COWcode,
    e_regionpol, e_regionpol_6C,
    all_of(indicator_info$code)
  ) %>%
  # Drop countries with too much missing data (> 50% missing)
  mutate(n_missing = rowSums(is.na(across(all_of(indicator_info$code))))) %>%
  filter(n_missing < ncol(across(all_of(indicator_info$code))) * 0.5) %>%
  select(-n_missing)

cat("Countries retained:", nrow(vdem_filtered), "\n")
```

## Indicator catalog

Here is the full set of institutional indicators we're working with:

```{r indicator-table}
kable(
  indicator_info %>% select(Code = code, Category = category, Description = description),
  caption = "Institutional indicators used in this analysis"
)
```

# Binarization

The key step in economic complexity is constructing a binary **country × feature** matrix. In
trade data, this is done via Revealed Comparative Advantage (RCA). Here we adapt the approach
to V-Dem indicators.

## Strategy

V-Dem indicators come in two main forms:

1. **Ordinal (0-4)**: Expert-coded assessments where higher = more democratic/better governance.
   We binarize at **>= 3**, meaning the feature is "present" when a country scores at least 3
   out of 4.

2. **Interval (0-1)**: Aggregated indices from measurement models. We use an **RCA-like approach**:
   a country "has" the feature if its value exceeds the global mean for that indicator.

```{r classify-indicators}
# Determine type for each indicator based on its range in the data
classify_indicator <- function(values) {
  values <- na.omit(values)
  rng <- range(values)

  # If max is close to 1 and min is close to 0, it's likely interval

  if (rng[2] <= 1.1 & rng[1] >= -0.1) {
    return("interval")
  }
  # If values look like integers in 0-4 range, ordinal
  if (rng[2] <= 5 & all(abs(values - round(values)) < 0.01)) {
    return("ordinal")
  }
  # Default: treat as interval with RCA approach
  return("interval")
}

indicator_types <- tibble(
  code = indicator_info$code,
  type = map_chr(indicator_info$code, ~classify_indicator(vdem_filtered[[.x]]))
)

cat("Indicator types:\n")
count(indicator_types, type)
```

```{r binarize}
# Build the binary matrix
binary_df <- vdem_filtered %>% select(country_name, all_of(indicator_info$code))

for (ind in indicator_info$code) {
  ind_type <- indicator_types$type[indicator_types$code == ind]
  values <- binary_df[[ind]]

  if (ind_type == "ordinal") {
    binary_df[[ind]] <- as.integer(values >= 3)
  } else {
    # RCA-like: above global mean
    global_mean <- mean(values, na.rm = TRUE)
    binary_df[[ind]] <- as.integer(values >= global_mean)
  }
}

# Replace remaining NAs with 0 (absence)
binary_df <- binary_df %>%
  mutate(across(all_of(indicator_info$code), ~replace_na(.x, 0L)))

# Create the Mcp matrix (countries x institutional features)
Mcp <- binary_df %>%
  column_to_rownames("country_name") %>%
  as.matrix()

cat("Mcp matrix dimensions:", nrow(Mcp), "countries x", ncol(Mcp), "institutional features\n")
cat("Matrix density (fraction of 1s):", round(mean(Mcp), 3), "\n")
```

The density of the matrix tells us something important. If it's too high (> 0.8) or too low
(< 0.1), the complexity calculations won't be very informative. A density around 0.3-0.6 is
typical for complexity analyses.

```{r binarization-summary}
# Summary statistics
diversity <- rowSums(Mcp)
ubiquity <- colSums(Mcp)

cat("\nDiversity (# features per country):\n")
cat("  Mean:", round(mean(diversity), 1), "\n")
cat("  Range:", min(diversity), "-", max(diversity), "\n")

cat("\nUbiquity (# countries per feature):\n")
cat("  Mean:", round(mean(ubiquity), 1), "\n")
cat("  Range:", min(ubiquity), "-", max(ubiquity), "\n")
```

# Complexity Calculation

## Diversity and Ubiquity (0th order)

The most basic measures: **diversity** is how many institutional features a country has;
**ubiquity** is how many countries share a given feature.

```{r diversity-ubiquity}
# Diversity: number of features each country has
kc0 <- rowSums(Mcp)
# Ubiquity: number of countries that have each feature
kp0 <- colSums(Mcp)

diversity_df <- tibble(
  country = names(kc0),
  diversity = as.numeric(kc0)
) %>%
  arrange(desc(diversity))

# Top and bottom by diversity
cat("Top 10 countries by diversity (# institutional features):\n")
head(diversity_df, 10) %>% kable()

cat("\nBottom 10 countries by diversity:\n")
tail(diversity_df, 10) %>% kable()
```

```{r ubiquity-plot}
ubiquity_df <- tibble(
  indicator = names(kp0),
  ubiquity = as.numeric(kp0)
) %>%
  left_join(indicator_info, by = c("indicator" = "code")) %>%
  arrange(ubiquity)

ggplot(ubiquity_df, aes(x = reorder(indicator, ubiquity), y = ubiquity, fill = category)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Ubiquity of Institutional Features",
    subtitle = "Number of countries that 'have' each feature",
    x = NULL, y = "Number of countries",
    fill = "Category"
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_text(size = 6))
```

## Method of Reflections

The Method of Reflections iteratively refines diversity and ubiquity. At each iteration,
a country's complexity becomes the average complexity of the features it has, weighted by
how complex those features are -- and vice versa. This creates a recursive, self-consistent
measure.

```{r method-of-reflections}
# Method of Reflections
n_iterations <- 20

# Initialize
kc <- matrix(0, nrow = nrow(Mcp), ncol = n_iterations + 1)
kp <- matrix(0, nrow = ncol(Mcp), ncol = n_iterations + 1)
rownames(kc) <- rownames(Mcp)
rownames(kp) <- colnames(Mcp)

kc[, 1] <- kc0
kp[, 1] <- kp0

for (n in 1:n_iterations) {
  # Update country complexity: average ubiquity-weighted feature complexity
  for (c in 1:nrow(Mcp)) {
    features <- which(Mcp[c, ] == 1)
    if (length(features) > 0 & kc0[c] > 0) {
      kc[c, n + 1] <- (1 / kc0[c]) * sum(kp[features, n])
    }
  }

  # Update feature complexity: average diversity-weighted country complexity
  for (p in 1:ncol(Mcp)) {
    countries <- which(Mcp[, p] == 1)
    if (length(countries) > 0 & kp0[p] > 0) {
      kp[p, n + 1] <- (1 / kp0[p]) * sum(kc[countries, n])
    }
  }
}

# ECI from Method of Reflections (using even iterations for countries)
# The complexity information is in the variance of the even-iteration values
kc_even <- kc[, n_iterations + 1]  # Last even iteration
eci_mor <- (kc_even - mean(kc_even)) / sd(kc_even)

kp_even <- kp[, n_iterations + 1]
pci_mor <- (kp_even - mean(kp_even)) / sd(kp_even)

cat("Method of Reflections converged after", n_iterations, "iterations.\n")
```

## Eigenvalue Method

The eigenvalue method is more elegant and mathematically equivalent to the Method of Reflections
at convergence. We construct the matrix $\tilde{M}_{cc'}$ and take its second eigenvector.

$$\tilde{M}_{cc'} = \sum_p \frac{M_{cp} M_{c'p}}{k_{c,0} k_{p,0}}$$

```{r eigenvalue-method}
# Build Mcc_tilde
# Mcc_tilde = (1/kc0) * Mcp %*% diag(1/kp0) %*% t(Mcp)

# Handle zeros
kc0_safe <- ifelse(kc0 == 0, 1, kc0)
kp0_safe <- ifelse(kp0 == 0, 1, kp0)

Mcc_tilde <- diag(1 / kc0_safe) %*% Mcp %*% diag(1 / kp0_safe) %*% t(Mcp)
rownames(Mcc_tilde) <- rownames(Mcp)
colnames(Mcc_tilde) <- rownames(Mcp)

# Eigen decomposition
eig <- eigen(Mcc_tilde)

# The first eigenvector is trivial (all same sign), use second
eci_eigen_raw <- Re(eig$vectors[, 2])
names(eci_eigen_raw) <- rownames(Mcp)

# Standardize
eci_eigen <- (eci_eigen_raw - mean(eci_eigen_raw)) / sd(eci_eigen_raw)

# Ensure sign convention: positive ECI = more complex
# Check correlation with diversity -- complex countries should have higher diversity
if (cor(eci_eigen, kc0) < 0) {
  eci_eigen <- -eci_eigen
}

# Similarly for PCI via Mpp_tilde
Mpp_tilde <- diag(1 / kp0_safe) %*% t(Mcp) %*% diag(1 / kc0_safe) %*% Mcp
rownames(Mpp_tilde) <- colnames(Mcp)
colnames(Mpp_tilde) <- colnames(Mcp)

eig_p <- eigen(Mpp_tilde)
pci_eigen_raw <- Re(eig_p$vectors[, 2])
names(pci_eigen_raw) <- colnames(Mcp)
pci_eigen <- (pci_eigen_raw - mean(pci_eigen_raw)) / sd(pci_eigen_raw)

if (cor(pci_eigen, kp0, use = "complete.obs") > 0) {
  # More ubiquitous features should be LESS complex
  pci_eigen <- -pci_eigen
}

cat("Correlation between MoR and eigenvalue ECI:",
    round(cor(eci_mor, eci_eigen), 3), "\n")
```

## Results: Institutional Complexity Index

We use the eigenvalue method as our primary measure (it's more robust), with the Method of
Reflections as a check.

```{r eci-results}
eci_df <- tibble(
  country = names(eci_eigen),
  ICI = eci_eigen,
  ICI_MoR = eci_mor,
  diversity = kc0
) %>%
  arrange(desc(ICI))

cat("Top 20 countries by Institutional Complexity Index:\n")
eci_df %>%
  head(20) %>%
  mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  kable()

cat("\nBottom 20 countries by Institutional Complexity Index:\n")
eci_df %>%
  tail(20) %>%
  mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  kable()
```

```{r eci-distribution}
ggplot(eci_df, aes(x = ICI)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Distribution of Institutional Complexity Index",
    x = "ICI (standardized)", y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## Results: Institutional Feature Complexity

Which institutional features are most "complex" -- i.e., found predominantly in
institutionally sophisticated countries?

```{r pci-results}
pci_df <- tibble(
  indicator = names(pci_eigen),
  IFC = pci_eigen,
  ubiquity = kp0
) %>%
  left_join(indicator_info, by = c("indicator" = "code")) %>%
  arrange(desc(IFC))

cat("Top 20 most complex institutional features:\n")
pci_df %>%
  head(20) %>%
  select(indicator, description, category, IFC, ubiquity) %>%
  mutate(IFC = round(IFC, 3)) %>%
  kable()

cat("\nBottom 20 (most ubiquitous/least complex) features:\n")
pci_df %>%
  tail(20) %>%
  select(indicator, description, category, IFC, ubiquity) %>%
  mutate(IFC = round(IFC, 3)) %>%
  kable()
```

```{r pci-by-category}
ggplot(pci_df, aes(x = reorder(category, IFC, median), y = IFC, fill = category)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Institutional Feature Complexity by Category",
    x = NULL, y = "Institutional Feature Complexity (IFC)"
  ) +
  theme_minimal(base_size = 13)
```

# Proximity and Relatedness

## Institution proximity

We calculate the proximity between institutional features using the conditional probability
approach from Hidalgo et al. (2007):

$$\phi_{pp'} = \min\left(P(p | p'), P(p' | p)\right)$$

where $P(p | p')$ is the probability that a country has feature $p$ given that it has
feature $p'$. This measures how likely two features are to co-occur.

```{r institution-proximity, cache=TRUE}
n_features <- ncol(Mcp)
phi_pp <- matrix(0, n_features, n_features)
rownames(phi_pp) <- colnames(Mcp)
colnames(phi_pp) <- colnames(Mcp)

for (i in 1:n_features) {
  for (j in i:n_features) {
    if (i == j) {
      phi_pp[i, j] <- 1
    } else {
      co_occur <- sum(Mcp[, i] & Mcp[, j])
      p_given_j <- if (kp0[j] > 0) co_occur / kp0[j] else 0
      p_given_i <- if (kp0[i] > 0) co_occur / kp0[i] else 0
      phi_pp[i, j] <- min(p_given_i, p_given_j)
      phi_pp[j, i] <- phi_pp[i, j]
    }
  }
}

cat("Institution proximity matrix:", nrow(phi_pp), "x", ncol(phi_pp), "\n")
cat("Mean proximity:", round(mean(phi_pp[upper.tri(phi_pp)]), 3), "\n")
```

## Country proximity

Similarly, we compute proximity between countries:

$$\phi_{cc'} = \min\left(P(c | c'), P(c' | c)\right)$$

```{r country-proximity, cache=TRUE}
n_countries <- nrow(Mcp)
phi_cc <- matrix(0, n_countries, n_countries)
rownames(phi_cc) <- rownames(Mcp)
colnames(phi_cc) <- rownames(Mcp)

for (i in 1:n_countries) {
  for (j in i:n_countries) {
    if (i == j) {
      phi_cc[i, j] <- 1
    } else {
      co_occur <- sum(Mcp[i, ] & Mcp[j, ])
      p_given_j <- if (kc0[j] > 0) co_occur / kc0[j] else 0
      p_given_i <- if (kc0[i] > 0) co_occur / kc0[i] else 0
      phi_cc[i, j] <- min(p_given_i, p_given_j)
      phi_cc[j, i] <- phi_cc[i, j]
    }
  }
}
```

## Institution Space (Network)

The institution space is a network where nodes are institutional features and edges connect
features that tend to co-occur across countries.

```{r institution-space-network}
# Threshold for edges: use a percentile-based cutoff
threshold_p <- quantile(phi_pp[upper.tri(phi_pp)], 0.75)

# Build adjacency matrix
adj_pp <- phi_pp
adj_pp[adj_pp < threshold_p] <- 0
diag(adj_pp) <- 0

# Create igraph object
g_inst <- graph_from_adjacency_matrix(adj_pp, mode = "undirected", weighted = TRUE)

# Add attributes
V(g_inst)$category <- indicator_info$category[match(V(g_inst)$name, indicator_info$code)]
V(g_inst)$complexity <- pci_eigen[V(g_inst)$name]
V(g_inst)$ubiquity <- kp0[V(g_inst)$name]

cat("Institution space: ", vcount(g_inst), "nodes,", ecount(g_inst), "edges\n")
cat("Mean degree:", round(mean(degree(g_inst)), 1), "\n")
```

## Country Space (Network)

```{r country-space-network}
threshold_c <- quantile(phi_cc[upper.tri(phi_cc)], 0.80)

adj_cc <- phi_cc
adj_cc[adj_cc < threshold_c] <- 0
diag(adj_cc) <- 0

g_country <- graph_from_adjacency_matrix(adj_cc, mode = "undirected", weighted = TRUE)

# Add region info
region_lookup <- vdem_filtered %>%
  select(country_name, e_regionpol_6C) %>%
  deframe()

region_labels <- c(
  "1" = "Eastern Europe & Central Asia",
  "2" = "Latin America & Caribbean",
  "3" = "Middle East & North Africa",
  "4" = "Sub-Saharan Africa",
  "5" = "Western Europe & North America",
  "6" = "East Asia & Pacific"
)

V(g_country)$region <- region_labels[as.character(
  region_lookup[V(g_country)$name]
)]
V(g_country)$ICI <- eci_eigen[V(g_country)$name]

cat("Country space:", vcount(g_country), "nodes,", ecount(g_country), "edges\n")
```

# UMAP Visualizations

## Institution Space (UMAP)

We use UMAP to embed the institution proximity matrix into 2D for visualization. UMAP
preserves local structure well, making it ideal for revealing clusters of related institutions.

```{r umap-institutions, cache=TRUE}
# Convert proximity to distance, use cmdscale (MDS) as fallback since uwot
# doesn't support precomputed distance matrices well on all versions
inst_dist <- as.dist(1 - phi_pp)

# Use MDS first, then UMAP on the MDS coordinates for robust embedding
set.seed(42)
mds_inst <- cmdscale(inst_dist, k = min(10, ncol(Mcp) - 1))
umap_inst <- umap(mds_inst, n_neighbors = 15, min_dist = 0.1, n_epochs = 500)

umap_inst_df <- tibble(
  indicator = colnames(Mcp),
  UMAP1 = umap_inst[, 1],
  UMAP2 = umap_inst[, 2],
  ubiquity = kp0,
  IFC = pci_eigen
) %>%
  left_join(indicator_info, by = c("indicator" = "code"))
```

```{r umap-inst-plot, fig.width=12, fig.height=9}
# Color palette for categories
n_cats <- length(unique(umap_inst_df$category))
cat_colors <- colorRampPalette(brewer.pal(12, "Set3"))(n_cats)

ggplot(umap_inst_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = category, size = IFC), alpha = 0.75) +
  geom_text_repel(
    aes(label = indicator),
    size = 2.2, max.overlaps = 20, segment.alpha = 0.3
  ) +
  scale_size_continuous(range = c(2, 8), name = "Feature\nComplexity") +
  scale_color_manual(values = cat_colors, name = "Category") +
  labs(
    title = "Institution Space (UMAP Embedding)",
    subtitle = "Institutional features positioned by co-occurrence similarity",
    x = "UMAP 1", y = "UMAP 2"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 16)
  )
```

## Country Space (UMAP)

```{r umap-countries, cache=TRUE}
set.seed(42)
country_dist <- as.dist(1 - phi_cc)
mds_country <- cmdscale(country_dist, k = min(10, nrow(Mcp) - 1))
umap_country <- umap(mds_country, n_neighbors = 15, min_dist = 0.2, n_epochs = 500)

umap_country_df <- tibble(
  country = rownames(Mcp),
  UMAP1 = umap_country[, 1],
  UMAP2 = umap_country[, 2],
  ICI = eci_eigen,
  diversity = kc0
)

# Add region
umap_country_df <- umap_country_df %>%
  mutate(
    region = region_labels[as.character(region_lookup[country])],
    region = replace_na(region, "Other")
  )
```

```{r umap-country-plot, fig.width=12, fig.height=9}
region_colors <- c(
  "Western Europe & North America" = "#2166AC",
  "Eastern Europe & Central Asia" = "#67A9CF",
  "Latin America & Caribbean" = "#1B9E77",
  "East Asia & Pacific" = "#D95F02",
  "Middle East & North Africa" = "#E7298A",
  "Sub-Saharan Africa" = "#E6AB02",
  "Other" = "grey50"
)

ggplot(umap_country_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = region, size = ICI), alpha = 0.75) +
  geom_text_repel(
    aes(label = country),
    size = 2.5, max.overlaps = 25, segment.alpha = 0.3
  ) +
  scale_size_continuous(range = c(2, 7), name = "Institutional\nComplexity") +
  scale_color_manual(values = region_colors, name = "Region") +
  labs(
    title = "Country Space (UMAP Embedding)",
    subtitle = "Countries positioned by institutional portfolio similarity",
    x = "UMAP 1", y = "UMAP 2"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 16)
  )
```

# Additional Visualizations

## Heatmap of the Mcp Matrix

```{r heatmap, fig.width=14, fig.height=10}
# Cluster and plot heatmap
# Add category annotations for columns
col_annotation <- indicator_info %>%
  filter(code %in% colnames(Mcp)) %>%
  select(code, category) %>%
  column_to_rownames("code")

# Add region annotations for rows
row_annotation <- tibble(
  country = rownames(Mcp),
  Region = region_labels[as.character(region_lookup[rownames(Mcp)])]
) %>%
  mutate(Region = replace_na(Region, "Other")) %>%
  column_to_rownames("country")

# Annotation colors
ann_colors <- list(
  category = setNames(cat_colors, unique(col_annotation$category)),
  Region = region_colors[unique(row_annotation$Region)]
)

pheatmap(
  Mcp,
  color = c("grey95", "steelblue"),
  legend_breaks = c(0, 1),
  legend_labels = c("Absent", "Present"),
  annotation_col = col_annotation,
  annotation_row = row_annotation,
  annotation_colors = ann_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  clustering_method = "ward.D2",
  main = "Country x Institutional Feature Matrix (Mcp)",
  fontsize = 8
)
```

## ICI vs GDP per Capita

```{r gdp-data, cache=TRUE}
# Get GDP per capita from World Development Indicators
gdp_data <- WDI(
  indicator = "NY.GDP.PCAP.PP.KD",
  start = latest_year - 1,
  end = latest_year,
  extra = TRUE
) %>%
  filter(!is.na(NY.GDP.PCAP.PP.KD)) %>%
  group_by(country) %>%
  slice_max(year, n = 1) %>%
  ungroup() %>%
  select(country, iso3c, gdp_pc = NY.GDP.PCAP.PP.KD)

# Match to our country names using countrycode
eci_gdp <- eci_df %>%
  mutate(
    iso3c = countrycode(country, "country.name", "iso3c",
                        warn = FALSE)
  ) %>%
  left_join(gdp_data, by = "iso3c") %>%
  filter(!is.na(gdp_pc))

cat("Countries matched with GDP data:", nrow(eci_gdp), "\n")
```

```{r eci-vs-gdp, fig.width=10, fig.height=8}
# Add regions
eci_gdp <- eci_gdp %>%
  mutate(
    region = region_labels[as.character(
      region_lookup[country.x]
    )],
    region = replace_na(region, "Other")
  )

cor_gdp <- cor(eci_gdp$ICI, log(eci_gdp$gdp_pc), use = "complete.obs")

ggplot(eci_gdp, aes(x = ICI, y = gdp_pc)) +
  geom_point(aes(color = region), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "grey30", linetype = "dashed", se = TRUE, alpha = 0.15) +
  geom_text_repel(
    aes(label = country.x),
    size = 2.5, max.overlaps = 15
  ) +
  scale_y_log10(labels = scales::dollar_format()) +
  scale_color_manual(values = region_colors, name = "Region") +
  labs(
    title = "Institutional Complexity vs GDP per Capita",
    subtitle = sprintf("Correlation with log GDP: %.2f", cor_gdp),
    x = "Institutional Complexity Index (ICI)",
    y = "GDP per capita (PPP, constant $, log scale)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 16))
```

## ICI vs Liberal Democracy Index

```{r eci-vs-libdem, fig.width=10, fig.height=8}
# Get liberal democracy index
libdem <- vdem_raw %>%
  filter(year == latest_year) %>%
  select(country_name, v2x_libdem) %>%
  filter(!is.na(v2x_libdem))

eci_libdem <- eci_df %>%
  left_join(libdem, by = c("country" = "country_name")) %>%
  filter(!is.na(v2x_libdem)) %>%
  mutate(
    region = region_labels[as.character(region_lookup[country])],
    region = replace_na(region, "Other")
  )

cor_libdem <- cor(eci_libdem$ICI, eci_libdem$v2x_libdem, use = "complete.obs")

ggplot(eci_libdem, aes(x = ICI, y = v2x_libdem)) +
  geom_point(aes(color = region), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "grey30", linetype = "dashed", se = TRUE, alpha = 0.15) +
  geom_text_repel(
    aes(label = country),
    size = 2.5, max.overlaps = 15
  ) +
  scale_color_manual(values = region_colors, name = "Region") +
  labs(
    title = "Institutional Complexity vs Liberal Democracy",
    subtitle = sprintf("Correlation: %.2f", cor_libdem),
    x = "Institutional Complexity Index (ICI)",
    y = "V-Dem Liberal Democracy Index"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 16))
```

## ICI by World Region

```{r eci-by-region, fig.width=10, fig.height=7}
eci_regions <- eci_df %>%
  mutate(
    region = region_labels[as.character(region_lookup[country])],
    region = replace_na(region, "Other")
  ) %>%
  filter(region != "Other")

ggplot(eci_regions, aes(x = reorder(region, ICI, median), y = ICI, fill = region)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE, outlier.alpha = 0.3) +
  geom_jitter(aes(color = region), width = 0.15, alpha = 0.5, size = 1.5, show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = region_colors) +
  scale_color_manual(values = region_colors) +
  labs(
    title = "Institutional Complexity Index by World Region",
    x = NULL, y = "Institutional Complexity Index (ICI)"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 16))
```

# Discussion

## What does institutional complexity tell us?

The analogy between economic and institutional complexity reveals several patterns:

**High-complexity countries** tend to be established democracies with deep, interlocking
institutional frameworks. These countries don't just have individual strong institutions --
they have *rare combinations* of institutions that reinforce each other. Just as economically
complex countries produce products that require many complementary capabilities, institutionally
complex countries maintain governance features that require deep state capacity, civic culture,
and historical development.

**The institution space** reveals how governance features cluster. Features that are close
together in the UMAP embedding tend to co-occur -- they represent institutional "packages"
that countries adopt together. The structure of this space suggests that some institutional
reforms may be easier to adopt once neighboring capabilities are in place.

**Feature complexity** is particularly informative. The most "complex" institutional features --
those found primarily in institutionally sophisticated countries -- tend to involve subtle
aspects of governance: rigorous and impartial public administration, judicial accountability,
freedom from self-censorship in media, and deliberative quality of political discourse.
These are features that require not just formal rules but deeply embedded norms and practices.

## Caveats

Several important limitations:

1. **V-Dem data reflects expert assessments**, not objective measures. The binarization choices
   affect results significantly.

2. **The analogy has limits.** Countries don't "choose" institutions the way firms choose
   products to export. Institutional development is path-dependent, often violent, and
   constrained by geography, history, and international power relations.

3. **Correlation with GDP and democracy is expected** given that many V-Dem indicators
   directly measure democratic quality. The interesting question is whether the *complexity*
   measure adds information beyond simple additive indices.

4. **The method is sensitive to indicator selection.** A different set of indicators would
   produce different complexity rankings. Our selection, while broad, reflects choices about
   what counts as an "institution."

5. **Cross-sectional analysis only.** A natural extension would track institutional complexity
   over time to study how countries build (or lose) institutional sophistication.

## Future directions

- Panel analysis: how does institutional complexity evolve over time?
- Predictive power: does ICI predict future economic growth or democratic stability
  beyond what simpler indices capture?
- Institutional "product space" analysis: which institutional upgrades are most
  feasible for countries at different positions in the institution space?
- Comparison with actual Economic Complexity Index: are economically and institutionally
  complex countries the same?

---

*Analysis generated on `r Sys.Date()`. Data from V-Dem v`r packageVersion("vdemdata")` and
World Development Indicators.*
