---
title: "Institutions → Outcomes"
subtitle: "Which institutional bundles explain fiscal, health, education, and complexity outcomes?"
author: "Institutional Complexity Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6.5
    fig_height: 4.5
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  cache = TRUE,
  cache.lazy = FALSE
)
```

# Motivation

The institutional-complexity analysis describes *how* institutional features cluster.
This companion document asks the next question: **which institutional bundles predict
real-world outcomes?** For each of ten cross-country outcomes (fiscal, health,
education, economic complexity) we fit a flexible gradient-boosted model on the full
V-Dem feature set and use SHAP values to surface (1) which individual institutional
features matter most and (2) which *pairs* of features interact most strongly. Shallow
trees (`max_depth = 3`) restrict the model to at-most-three-way interactions — the
"bundle" hypothesis says pairwise and low-order interactions should dominate over
single-feature main effects.

This is a descriptive exercise, not a causal one. V-Dem scores and fiscal/health/
education outcomes are co-determined over long horizons; no design here identifies a
causal effect of institutions on outcomes.

```{r packages}
library(tidyverse)
library(arrow)
library(knitr)
library(xgboost)
library(shapviz)

source("~/dev/gl-design/skills/gl-ggplot/assets/theme_gl.R")
gl_setup()
```

# Data

## Rebuild the institutional Mcp matrix

We reproduce the V-Dem preprocessing from the main analysis: the most recent year
with at least 150-country coverage, all C-type components with good coverage, min-max
normalised to [0, 1]. See `institutional_complexity.Rmd` for detail.

```{r build-mcp, cache=TRUE}
vdem_raw <- read_parquet("~/dev/shared-data/vdem/vdem.parquet")
vdem_codebook <- read_parquet("~/dev/shared-data/vdem/vdem_codebook.parquet")

latest_year <- vdem_raw %>%
  filter(year >= 2020) %>%
  group_by(year) %>%
  summarise(n_countries = n_distinct(country_name), .groups = "drop") %>%
  filter(n_countries >= 150) %>%
  pull(year) %>% max()

component_codes <- vdem_codebook %>% filter(vartype == "C") %>% pull(tag)

suffixes <- c("_codelow", "_codehigh", "_sd", "_osp", "_ord", "_mean", "_nr")
all_cols <- names(vdem_raw)
is_suffix <- sapply(all_cols, function(x) any(sapply(suffixes, function(s) endsWith(x, s))))
is_numbered <- grepl("_\\d+$", all_cols)
base_cols <- all_cols[!is_suffix & !is_numbered]
candidate_codes <- intersect(component_codes, base_cols)

latest_data <- vdem_raw %>% filter(year == latest_year)
coverage <- sapply(candidate_codes, function(x) sum(!is.na(latest_data[[x]])))
numeric_check <- sapply(candidate_codes, function(x) is.numeric(vdem_raw[[x]]))
indicator_codes <- candidate_codes[coverage >= 150 & numeric_check]
indicator_codes <- setdiff(indicator_codes, c("v2regopploc", "v2regsuploc"))

indicator_info <- vdem_codebook %>%
  filter(tag %in% indicator_codes) %>%
  transmute(code = tag, description = name) %>%
  distinct(code, .keep_all = TRUE)

vdem_filtered <- vdem_raw %>%
  filter(year == latest_year) %>%
  select(country_name, country_text_id, all_of(indicator_info$code)) %>%
  mutate(n_missing = rowSums(is.na(across(all_of(indicator_info$code))))) %>%
  filter(n_missing < ncol(across(all_of(indicator_info$code))) * 0.5) %>%
  select(-n_missing)

cont_df <- vdem_filtered %>% select(country_name, all_of(indicator_info$code))
for (ind in indicator_info$code) {
  v <- cont_df[[ind]]; rng <- range(v, na.rm = TRUE)
  cont_df[[ind]] <- if (rng[2] > rng[1]) (v - rng[1]) / (rng[2] - rng[1]) else 0
}
cont_df <- cont_df %>%
  mutate(across(all_of(indicator_info$code), ~replace_na(.x, 0)))

Mcp <- cont_df %>% column_to_rownames("country_name") %>% as.matrix()

iso3_lookup <- vdem_filtered %>%
  select(country_name, country_text_id) %>% deframe()

cat("V-Dem year:", latest_year, "\n")
cat("Mcp:", nrow(Mcp), "countries ×", ncol(Mcp), "features\n")
```

## Load outcome variables from the macro parquet

Ten outcomes across four buckets. V-Dem's latest-coverage year runs ahead of
several macro sources (HCI in particular is released only every 2–3 years),
so for each country × outcome we take the **most recent non-NA value within the
six years ending at `latest_year`**. This trades a little temporal precision for
usable cross-country coverage. The "year used" is reported per outcome.
`qog_bl_asymf` (Barro-Lee avg years of schooling) was dropped — its latest
non-NA observation in this vintage is 2015, too stale to pair with a V-Dem 2025
snapshot.

```{r load-outcomes, cache=TRUE}
outcome_meta <- tribble(
  ~column,                 ~bucket,       ~label,                       ~transform, ~source,
  "weo_ggr_ngdp",          "Fiscal",      "Revenue % GDP",              "none",     "macro",
  "imf_tax_inc",           "Fiscal",      "Income tax % GDP",           "none",     "imf_world",
  "wdi_sp_dyn_le00_in",    "Health",      "Life expectancy",            "none",     "macro",
  "wdi_sp_dyn_imrt_in",    "Health",      "Infant mortality (log)",     "log",      "macro",
  "wdi_sh_sta_mmrt",       "Health",      "Maternal mortality (log)",   "log",      "macro",
  "wdi_hd_hci_ovrl",       "Education",   "Human Capital Index",        "none",     "macro",
  "wdi_se_sec_enrr",       "Education",   "Secondary enrollment %",     "none",     "macro",
  "atl_sitc_eci",          "Complexity",  "Economic Complexity (ECI)",  "none",     "macro",
  "wdi_ne_exp_gnfs_zs",    "Trade",       "Exports % GDP",              "none",     "macro"
)

macro_cols <- outcome_meta %>% filter(source == "macro") %>% pull(column)
macro <- read_parquet(
  "~/dev/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year", macro_cols)
)

# IMF WoRLD 2026: pull income-tax (TaxInc) keyed by ISO3, rename to imf_tax_inc
# so it matches outcome_meta$column.
imf_world <- suppressMessages(
  readxl::read_excel("~/dev/shared-data/world-imf2026.xlsx",
                     sheet = "Data", guess_max = 20000)
) %>%
  transmute(countrycodeiso = ISO3, year, imf_tax_inc = TaxInc) %>%
  filter(!is.na(countrycodeiso))

macro <- full_join(macro, imf_world, by = c("countrycodeiso", "year"))

# For each outcome × country, pick the most recent non-NA value in the six
# years ending at latest_year. This is still "cross-section-ish" but avoids
# losing outcomes whose most recent vintage trails V-Dem.
window <- seq(latest_year - 5, latest_year)

pick_latest <- function(col) {
  macro %>%
    filter(year %in% window, !is.na(.data[[col]])) %>%
    group_by(countrycodeiso) %>%
    slice_max(year, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(iso3 = countrycodeiso,
              !!col := .data[[col]],
              !!paste0(col, "_year") := year)
}

outcome_wide <- reduce(
  lapply(outcome_meta$column, pick_latest),
  ~full_join(.x, .y, by = "iso3")
)

# Apply transforms in-place, carry the transform tag into column names
for (i in seq_len(nrow(outcome_meta))) {
  col <- outcome_meta$column[i]
  tr  <- outcome_meta$transform[i]
  if (tr == "log") {
    outcome_wide[[col]] <- log(pmax(outcome_wide[[col]], .Machine$double.eps))
  } else if (tr == "log1p") {
    outcome_wide[[col]] <- log1p(pmax(outcome_wide[[col]], 0))
  }
}

kable(outcome_meta,
      caption = "Outcome variables. `log` indicates a natural-log transform was applied before modeling. Debt uses log1p because small-debt observations are near zero.")
```

## Join to Mcp

```{r join}
mcp_df <- as_tibble(Mcp, rownames = "country_name") %>%
  mutate(iso3 = unname(iso3_lookup[country_name])) %>%
  relocate(iso3)

analysis_df <- mcp_df %>%
  inner_join(outcome_wide %>% select(iso3, any_of(outcome_meta$column)),
             by = "iso3")

cat("Rows in analysis frame:", nrow(analysis_df), "\n")
cat("Countries in Mcp but missing macro match:",
    sum(!mcp_df$iso3 %in% outcome_wide$iso3), "\n\n")

feature_cols <- indicator_info$code
# Build as a fresh double matrix — xgboost 2.x errors on misaligned pointers
# when the matrix comes from tibble -> as.matrix() coercion.
X_mat <- matrix(as.double(as.matrix(analysis_df[, feature_cols])),
                nrow = nrow(analysis_df), ncol = length(feature_cols),
                dimnames = list(analysis_df$iso3, feature_cols))

year_cols <- paste0(outcome_meta$column, "_year")
year_info <- outcome_wide %>%
  filter(iso3 %in% analysis_df$iso3) %>%
  select(any_of(year_cols)) %>%
  summarise(across(everything(),
    ~sprintf("%s (median %s)",
             paste(range(.x, na.rm = TRUE), collapse = "–"),
             median(.x, na.rm = TRUE)))) %>%
  pivot_longer(everything(), names_to = "ycol", values_to = "year_range") %>%
  mutate(column = sub("_year$", "", ycol))

outcome_coverage <- outcome_meta %>%
  rowwise() %>%
  mutate(n = sum(!is.na(analysis_df[[column]]))) %>%
  ungroup() %>%
  left_join(year_info %>% select(column, year_range), by = "column")
kable(outcome_coverage %>% select(bucket, label, column, transform, n, year_range),
      caption = "Per-outcome country coverage and the range of years used in the joined analysis frame.")
```

# Modeling

## Pipeline

For each outcome we fit XGBoost at each of four tree depths, `max_depth in {3, 4, 5, 6}`.
Shallower trees cap the interaction order the model can learn — depth 3 allows up to
three-way interactions, depth 6 up to six-way. Out-of-sample R^2 is computed from
5-fold CV predictions at each depth; comparing the R^2 curve across depths per
outcome is the "how much does bundle *size* matter" diagnostic.

Other hyperparameters are held fixed: `eta = 0.05`, `subsample = 0.8`,
`colsample_bytree = 0.8`, `min_child_weight = 5` (the small N makes this last one
important — it prevents leaves from specializing on one or two countries).

```{r pipeline}
base_params <- list(
  objective = "reg:squarederror",
  eval_metric = "rmse",
  eta = 0.05,
  subsample = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 5,
  verbosity = 0
)

# Given a fitted booster, walk every leaf back to the root. For each leaf,
# collect the UNIQUE set of features used on splits along the path and
# weight it by the leaf's Cover (sum of Hessians ~ samples reaching the leaf).
# Return a tibble: features (canonical|-joined), size, weight.
extract_bundles <- function(model, feature_names) {
  dt <- xgboost::xgb.model.dt.tree(feature_names = feature_names, model = model)
  internal <- dt[Feature != "Leaf"]
  children <- bind_rows(
    internal %>% select(parent = ID, child = Yes),
    internal %>% select(parent = ID, child = No),
    internal %>% select(parent = ID, child = Missing)
  ) %>% filter(!is.na(child), child != "")
  parent_of  <- setNames(children$parent, children$child)
  feature_of <- setNames(dt$Feature, dt$ID)

  leaves <- dt[Feature == "Leaf"]
  bundle_rows <- vector("list", nrow(leaves))
  for (i in seq_len(nrow(leaves))) {
    id <- leaves$ID[i]
    cover <- leaves$Cover[i]
    feats <- character()
    cur <- id
    while (!is.na(parent_of[cur]) && cur %in% names(parent_of)) {
      par <- unname(parent_of[cur])
      feats <- c(feats, unname(feature_of[par]))
      cur <- par
    }
    fs <- sort(unique(feats))
    if (length(fs) == 0) next
    bundle_rows[[i]] <- tibble(
      features = paste(fs, collapse = "|"),
      size = length(fs),
      weight = cover
    )
  }

  bind_rows(bundle_rows) %>%
    group_by(features, size) %>%
    summarise(weight = sum(weight), .groups = "drop") %>%
    arrange(desc(weight))
}

fit_outcome <- function(col, label, X, y_full, depth, seed = 42) {
  ok <- !is.na(y_full)
  # Fresh double matrix to sidestep xgboost pointer-alignment errors.
  X_ok <- matrix(as.double(X[ok, , drop = FALSE]),
                 nrow = sum(ok), ncol = ncol(X),
                 dimnames = list(rownames(X)[ok], colnames(X)))
  y <- as.double(y_full[ok])
  n <- length(y)

  params <- c(base_params, list(max_depth = depth))
  set.seed(seed)
  dall <- xgb.DMatrix(data = X_ok, label = y)

  cv <- xgb.cv(params = params, data = dall, nrounds = 1000,
               nfold = 5, early_stopping_rounds = 40, verbose = 0)
  best_nrounds <- cv$early_stop$best_iteration
  if (is.null(best_nrounds) || length(best_nrounds) == 0) {
    best_nrounds <- which.min(cv$evaluation_log$test_rmse_mean)
  }

  set.seed(seed)
  folds <- sample(rep(1:5, length.out = n))
  oos_pred <- numeric(n)
  for (k in 1:5) {
    tr <- folds != k; te <- folds == k
    m <- xgb.train(params = params,
                   data = xgb.DMatrix(X_ok[tr, , drop = FALSE], label = y[tr]),
                   nrounds = best_nrounds, verbose = 0)
    oos_pred[te] <- predict(m, X_ok[te, , drop = FALSE])
  }
  r2_oos <- 1 - sum((y - oos_pred)^2) / sum((y - mean(y))^2)

  final <- xgb.train(params = params, data = dall,
                     nrounds = best_nrounds, verbose = 0)

  contrib <- predict(final, X_ok, predcontrib = TRUE)
  shap_main <- contrib[, -ncol(contrib), drop = FALSE]
  colnames(shap_main) <- colnames(X_ok)
  mean_abs_shap <- colMeans(abs(shap_main))

  bundles <- extract_bundles(final, colnames(X_ok))

  list(outcome = col, label = label, depth = depth, n = n,
       r2_oos = r2_oos, best_nrounds = best_nrounds,
       model = final, X = X_ok, y = y,
       shap_main = shap_main, mean_abs_shap = mean_abs_shap,
       bundles = bundles)
}
```

## Depth sweep

```{r run-sweep, cache=TRUE}
depths <- 3:6
results <- list()
for (i in seq_len(nrow(outcome_meta))) {
  col <- outcome_meta$column[i]
  results[[col]] <- list()
  y_full <- analysis_df[[col]]
  for (d in depths) {
    cat("Fitting:", col, "@ depth", d, "\n")
    results[[col]][[as.character(d)]] <-
      fit_outcome(col, outcome_meta$label[i], X_mat, y_full, depth = d)
  }
}
```

```{r sweep-table}
sweep_df <- map_dfr(results, function(outcome_res) {
  map_dfr(outcome_res, function(r) {
    tibble(
      bucket  = outcome_meta$bucket[outcome_meta$column == r$outcome],
      label   = r$label,
      depth   = r$depth,
      n       = r$n,
      r2_oos  = r$r2_oos,
      nrounds = r$best_nrounds
    )
  })
})

sweep_wide <- sweep_df %>%
  select(bucket, label, depth, r2_oos) %>%
  pivot_wider(names_from = depth, values_from = r2_oos,
              names_prefix = "d") %>%
  mutate(across(starts_with("d"), ~round(.x, 3))) %>%
  arrange(bucket, desc(d3))

kable(sweep_wide,
      caption = "Out-of-sample R^2 by outcome and max_depth. Each column is a fit with that interaction cap.")
```

```{r sweep-plot, fig.height=5}
best_depth_df <- sweep_df %>%
  group_by(label) %>%
  slice_max(r2_oos, n = 1, with_ties = FALSE) %>%
  ungroup()

ggplot(sweep_df, aes(x = depth, y = r2_oos, color = bucket, group = label)) +
  geom_line(alpha = 0.8, linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_point(data = best_depth_df, color = highlight, size = 3, shape = 1,
             stroke = 1.2) +
  ggrepel::geom_text_repel(
    data = best_depth_df, aes(label = label),
    size = 2.6, segment.alpha = 0.3, max.overlaps = Inf,
    show.legend = FALSE, box.padding = 0.4
  ) +
  scale_x_continuous(breaks = depths) +
  labs(x = "max_depth (interaction order cap)",
       y = "Out-of-sample R^2",
       subtitle = "Red ring marks the best depth per outcome") +
  theme(legend.position = "bottom")
```

```{r best-depth-table}
best_depth_df %>%
  transmute(Bucket = bucket, Outcome = label,
            `Best depth` = depth, `Best R^2` = round(r2_oos, 3),
            `R^2 @ d3` = round(sweep_wide$d3[match(label, sweep_wide$label)], 3),
            Gain = round(r2_oos - sweep_wide$d3[match(label, sweep_wide$label)], 3)) %>%
  arrange(Bucket, desc(Gain)) %>%
  kable(caption = "Best depth per outcome. `Gain` = best R^2 minus R^2 at depth 3 — i.e. how much moving beyond three-way bundles actually helps.")
```

# Per-outcome detail (at best depth)

Each outcome's section below uses the model at its best-performing depth. We show
(1) a SHAP beeswarm of the top individual features — i.e. where each feature falls
across the country distribution and how it pushes predictions — and (2) the top
**feature bundles** of each size, extracted by walking the tree leaves. Bundle
weight = sum of leaf `Cover` for all leaves whose path uses exactly that set of
features; high-weight bundles are conjunctions of features the model relies on
repeatedly across many countries.

```{r desc-lookup}
desc_lookup <- setNames(indicator_info$description, indicator_info$code)

short_desc <- function(codes, max_len = 40) {
  d <- desc_lookup[codes]
  ifelse(is.na(d), codes, paste0(codes, ": ", str_trunc(d, max_len)))
}

format_bundle <- function(feature_str, max_len = 35) {
  feats <- strsplit(feature_str, "\\|")[[1]]
  labs  <- ifelse(feats %in% names(desc_lookup),
                  paste0(feats, ": ", str_trunc(desc_lookup[feats], max_len)),
                  feats)
  paste(labs, collapse = "\n+ ")
}
```

```{r per-outcome-plots, fig.width=6.5, fig.height=9, results='asis'}
for (i in seq_len(nrow(outcome_meta))) {
  all_depths <- results[[outcome_meta$column[i]]]
  r <- best_depth_df %>%
    filter(label == outcome_meta$label[i]) %>%
    pull(depth) %>%
    as.character() %>%
    (\(d) all_depths[[d]])()
  if (is.null(r)) next

  cat("\n\n## ", r$label, "\n\n", sep = "")
  cat("*N = ", r$n, " | best depth = ", r$depth,
      " | OOS R^2 = ", round(r$r2_oos, 3),
      " | nrounds = ", r$best_nrounds, "*\n\n", sep = "")

  # SHAP beeswarm, top 15 features by mean |SHAP|
  top_feats <- names(sort(r$mean_abs_shap, decreasing = TRUE))[1:15]
  readable <- short_desc(top_feats)
  shap_top <- r$shap_main[, top_feats, drop = FALSE]
  X_top    <- as.data.frame(r$X[, top_feats, drop = FALSE])
  colnames(shap_top) <- readable
  colnames(X_top)    <- readable
  sv <- shapviz(shap_top, X = X_top)
  print(sv_importance(sv, kind = "beeswarm", max_display = 15) +
        labs(title = NULL))

  # Top bundles by size 2, 3, 4 (if depth allows)
  for (k in intersect(2:min(4, r$depth), unique(r$bundles$size))) {
    top_k <- r$bundles %>%
      filter(size == k) %>%
      head(8) %>%
      mutate(bundle_label = map_chr(features, format_bundle),
             bundle_label = fct_reorder(bundle_label, weight))
    if (nrow(top_k) == 0) next
    p <- ggplot(top_k, aes(x = weight, y = bundle_label)) +
      geom_col(fill = gl$brand_blue, alpha = 0.85) +
      labs(x = "Bundle weight (sum of leaf Cover)",
           y = NULL,
           subtitle = paste0("Top size-", k, " feature bundles"))
    print(p)
  }
  cat("\n\n")
}
```

# Two-PC summary

The ML model treats each of ~60 V-Dem features as its own knob. But these features
are highly collinear: if we run a PCA on the Mcp matrix, two or three principal
components will usually capture most of the variation across countries. That
suggests a simpler exercise: **project each country into the first two principal
components, split the plane into quadrants, and ask what each quadrant means for
outcomes.** If the first two PCs explain most of the outcome signal too, the
quadrant labels tell a compact story about what bundles of institutional
arrangements tend to produce what outcomes.

## PCA on Mcp

We center and scale each V-Dem feature before PCA (`prcomp` with `scale. = TRUE`)
so that the min-max normalisation earlier in the pipeline doesn't let
wider-variance features dominate the rotation.

```{r pca-fit, cache=TRUE}
pca <- prcomp(X_mat, center = TRUE, scale. = TRUE)
var_exp <- pca$sdev^2 / sum(pca$sdev^2)
cum_var <- cumsum(var_exp)

tibble(PC = seq_along(var_exp),
       `Variance explained` = round(var_exp, 3),
       `Cumulative` = round(cum_var, 3)) %>%
  head(10) %>%
  kable(caption = "Share of Mcp variance explained by the first 10 principal components.")
```

```{r pca-loadings}
loadings <- as_tibble(pca$rotation[, 1:2], rownames = "code") %>%
  left_join(indicator_info, by = "code")

bind_rows(
  loadings %>% mutate(axis = "PC1", loading = PC1) %>%
    slice_max(abs(loading), n = 10) %>%
    select(axis, code, description, loading),
  loadings %>% mutate(axis = "PC2", loading = PC2) %>%
    slice_max(abs(loading), n = 10) %>%
    select(axis, code, description, loading)
) %>%
  mutate(loading = round(loading, 3),
         description = str_trunc(description, 45)) %>%
  kable(caption = "Top 10 V-Dem features by absolute loading on PC1 and PC2. Loadings are signed; sign indicates direction along the axis.")
```

## Country scatter

```{r pca-scatter, fig.height=6}
scores <- as_tibble(pca$x[, 1:2]) %>%
  mutate(iso3 = analysis_df$iso3) %>%
  mutate(quadrant = case_when(
    PC1 >= 0 & PC2 >= 0 ~ "Q1 (+,+)",
    PC1 <  0 & PC2 >= 0 ~ "Q2 (-,+)",
    PC1 <  0 & PC2 <  0 ~ "Q3 (-,-)",
    PC1 >= 0 & PC2 <  0 ~ "Q4 (+,-)"
  ))

ggplot(scores, aes(PC1, PC2)) +
  geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "grey75", linetype = "dashed") +
  geom_point(aes(color = quadrant), alpha = 0.6, size = 1.4) +
  ggrepel::geom_text_repel(aes(label = iso3, color = quadrant),
                           size = 2.4, max.overlaps = Inf,
                           segment.alpha = 0.25, show.legend = FALSE) +
  labs(x = sprintf("PC1 (%.1f%% var)", 100 * var_exp[1]),
       y = sprintf("PC2 (%.1f%% var)", 100 * var_exp[2])) +
  theme(legend.position = "bottom")
```

```{r pca-scatter-cambodia, fig.height=6}
sea_peers <- c("KHM", "VNM", "LAO", "THA", "MMR")
scores_focus <- scores %>% mutate(highlighted = iso3 %in% sea_peers)

ggplot(scores_focus, aes(PC1, PC2)) +
  geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "grey75", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 1.4) +
  geom_point(data = \(d) filter(d, highlighted),
             color = highlight, size = 3) +
  ggrepel::geom_text_repel(
    data = \(d) filter(d, highlighted),
    aes(label = iso3), color = highlight, size = 3.4,
    max.overlaps = Inf, segment.alpha = 0.5, fontface = "bold"
  ) +
  labs(x = sprintf("PC1 (%.1f%% var)", 100 * var_exp[1]),
       y = sprintf("PC2 (%.1f%% var)", 100 * var_exp[2]),
       subtitle = "Cambodia and mainland Southeast Asian peers (KHM, VNM, LAO, THA, MMR)")
```

## Cambodia's institutional peers

Geography is a guess at peerhood. With the PCA in hand we can ask the data who
Cambodia's actual institutional neighbours are — countries with similar V-Dem
fingerprints rather than similar coordinates on a map. We compute Euclidean
distance from Cambodia to every other country in the first ten PCs (about 69%
of total variance) and rank.

```{r cambodia-institutional-peers}
K_peer <- 10
khm_scores <- pca$x["KHM", 1:K_peer]
peer_dist <- pca$x[, 1:K_peer, drop = FALSE]
peer_dist <- sqrt(rowSums(sweep(peer_dist, 2, khm_scores, "-")^2))

country_lookup <- setNames(analysis_df$country_name, analysis_df$iso3)

peer_rank <- tibble(iso3 = names(peer_dist),
                    distance = peer_dist) %>%
  mutate(country = country_lookup[iso3],
         PC1 = pca$x[iso3, "PC1"],
         PC2 = pca$x[iso3, "PC2"]) %>%
  filter(iso3 != "KHM") %>%
  arrange(distance)

peer_rank %>%
  head(15) %>%
  mutate(across(c(distance, PC1, PC2), ~round(.x, 2))) %>%
  transmute(rank = row_number(), iso3, country, distance, PC1, PC2) %>%
  kable(caption = sprintf("Cambodia's 15 nearest institutional peers in the top-%d PC subspace (cumulative variance %.1f%%). Distance is Euclidean in PC space, which weights each PC by its variance.",
                          K_peer, 100 * cum_var[K_peer]))
```

```{r pca-scatter-institutional-peers, fig.height=6}
inst_peers <- peer_rank %>% head(8) %>% pull(iso3)
scores_inst <- scores %>%
  mutate(highlighted = iso3 %in% c("KHM", inst_peers))

ggplot(scores_inst, aes(PC1, PC2)) +
  geom_hline(yintercept = 0, color = "grey75", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "grey75", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 1.4) +
  geom_point(data = \(d) filter(d, highlighted),
             color = highlight, size = 3) +
  ggrepel::geom_text_repel(
    data = \(d) filter(d, highlighted),
    aes(label = iso3), color = highlight, size = 3.4,
    max.overlaps = Inf, segment.alpha = 0.5, fontface = "bold"
  ) +
  labs(x = sprintf("PC1 (%.1f%% var)", 100 * var_exp[1]),
       y = sprintf("PC2 (%.1f%% var)", 100 * var_exp[2]),
       subtitle = "Cambodia's 8 nearest institutional peers (top-10 PC subspace)")
```

```{r pca-geographic-peer-ranks}
geo <- c("VNM", "LAO", "THA", "MMR")
peer_rank %>%
  filter(iso3 %in% geo) %>%
  mutate(rank = match(iso3, peer_rank$iso3),
         across(c(distance, PC1, PC2), ~round(.x, 2))) %>%
  transmute(iso3, country, rank, distance, PC1, PC2) %>%
  arrange(rank) %>%
  kable(caption = "Where the geographic neighbours rank in Cambodia's institutional-similarity ordering. The closest is well outside the top 25 — geography and institutions disagree.")
```

## Cambodia and peers: actual vs predicted outcomes

For each outcome we ask: **given this country's V-Dem fingerprint, what would
a model that hasn't seen it expect its outcome to be?** The honest version is
**leave-one-out**: for each focus country and each outcome, refit the
best-depth model on every *other* country and predict the held-out one.
Comparing this prediction to the country's actual outcome surfaces
*institutional surprises* — countries that outperform or underperform what
their institutional setup would predict.

The residual is rescaled by the model's out-of-sample RMSE (derived from the
5-fold CV R² in the depth sweep above), so a z-score of +1 means the country's
actual outcome is one OOS-RMSE above what the model predicts.

We look at six focus countries: Cambodia (KHM), its institutional twin
Burundi (BDI), and its four mainland Southeast Asian geographic neighbours
(VNM, LAO, THA, MMR). Outcomes with `log` transforms (infant and maternal
mortality) are reported on the log scale; predictions and residuals are in
log-units of mortality.

```{r predict-focus-countries, cache=TRUE}
focus_iso3 <- c("KHM", "BDI", "VNM", "LAO", "THA", "MMR")

loo_predict_one <- function(r, focus_iso3, depth) {
  X_ok <- r$X
  y <- r$y
  nrounds <- r$best_nrounds
  prm <- list(objective = "reg:squarederror", eval_metric = "rmse",
              eta = 0.05, subsample = 0.8, colsample_bytree = 0.8,
              min_child_weight = 5, max_depth = depth, verbosity = 0)
  preds <- setNames(rep(NA_real_, length(focus_iso3)), focus_iso3)
  for (cc in focus_iso3) {
    if (!cc %in% rownames(X_ok)) next
    keep <- rownames(X_ok) != cc
    set.seed(42)
    m <- xgb.train(params = prm,
                   data = xgb.DMatrix(X_ok[keep, , drop = FALSE], label = y[keep]),
                   nrounds = nrounds, verbose = 0)
    preds[cc] <- predict(m, X_ok[cc, , drop = FALSE])
  }
  preds
}

scorecard <- map_dfr(seq_len(nrow(outcome_meta)), function(i) {
  col <- outcome_meta$column[i]
  bd <- best_depth_df %>% filter(label == outcome_meta$label[i]) %>%
    pull(depth) %>% as.character()
  r <- results[[col]][[bd]]
  if (is.null(r)) return(NULL)
  pred_focus <- loo_predict_one(r, focus_iso3, depth = as.numeric(bd))
  actual_focus <- analysis_df[[col]][match(focus_iso3, analysis_df$iso3)]
  oos_rmse <- sqrt(var(r$y) * (1 - r$r2_oos))
  tibble(iso3 = focus_iso3,
         outcome = col,
         label = outcome_meta$label[i],
         bucket = outcome_meta$bucket[i],
         actual = actual_focus,
         predicted = unname(pred_focus),
         residual = actual_focus - unname(pred_focus),
         oos_rmse = oos_rmse,
         z = residual / oos_rmse)
})
```

```{r residual-heatmap, fig.height=4.2}
heat_df <- scorecard %>%
  filter(!is.na(z)) %>%
  mutate(label = factor(label, levels = outcome_meta$label),
         iso3  = factor(iso3,  levels = focus_iso3))

ggplot(heat_df, aes(label, iso3, fill = z)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%+.1f", z)), size = 3) +
  scale_fill_gradient2(low = highlight, mid = "grey95",
                       high = gl$brand_blue, midpoint = 0,
                       limits = c(-3, 3), oob = scales::squish,
                       name = "Residual (SD units)") +
  labs(x = NULL, y = NULL,
       subtitle = "Actual minus model prediction, in residual SDs.\nBlue = country outperforms what its institutions predict; red = underperforms") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1),
        legend.position = "bottom",
        panel.grid = element_blank())
```

```{r scorecard-table}
scorecard %>%
  filter(!is.na(actual)) %>%
  arrange(factor(iso3, levels = focus_iso3),
          factor(label, levels = outcome_meta$label)) %>%
  transmute(Country = iso3, Bucket = bucket, Outcome = label,
            Actual = round(actual, 2),
            Predicted = round(predicted, 2),
            `Residual (raw)` = round(residual, 2),
            `Residual (SD)`  = round(z, 2)) %>%
  kable(caption = "Per-country actual vs predicted on each outcome. Residual SD > 0 means the country does better than its V-Dem fingerprint predicts; SD < 0 means worse.")
```

## Do two PCs explain the outcomes?

For each outcome we fit a simple OLS regression `y ~ PC1 + PC2 + PC1:PC2`. The
interaction term is what makes this a quadrant story — a pure additive fit can't
tell Q1 apart from "a bit of Q2 plus a bit of Q4." Outcomes are z-scored so
coefficients are directly comparable across rows.

```{r pca-outcome-regression}
scores_out <- scores %>%
  left_join(outcome_wide %>% select(iso3, all_of(outcome_meta$column)),
            by = "iso3")

pca_lm <- map_dfr(seq_len(nrow(outcome_meta)), function(i) {
  col <- outcome_meta$column[i]
  d <- scores_out %>%
    filter(!is.na(.data[[col]])) %>%
    mutate(y = as.numeric(scale(.data[[col]])))
  fit <- lm(y ~ PC1 + PC2 + PC1:PC2, data = d)
  cf <- coef(fit)
  tibble(bucket = outcome_meta$bucket[i],
         label  = outcome_meta$label[i],
         n      = nrow(d),
         r2     = summary(fit)$r.squared,
         PC1    = cf["PC1"],
         PC2    = cf["PC2"],
         `PC1:PC2` = cf["PC1:PC2"])
})

pca_lm %>%
  mutate(across(c(r2, PC1, PC2, `PC1:PC2`), ~round(.x, 3))) %>%
  arrange(bucket, desc(r2)) %>%
  kable(caption = "OLS on z-scored outcomes. `PC1`, `PC2`, and `PC1:PC2` are standardized coefficients: how a one-standard-deviation move along each axis (or their product) shifts the outcome, in outcome-SD units. Compare to the best-depth XGBoost R^2 above to see how much signal lives outside the first two PCs.")
```

## Quadrant × outcome heatmap

Each cell is the mean z-score of an outcome among countries in that quadrant.
Blue = outcome runs high in this quadrant, red = runs low, grey = near the
sample average. Reading one row across the four quadrants is the
quadrant-labeling exercise: it says which combinations of institutional
arrangements are associated with better fiscal / health / education /
complexity outcomes.

```{r pca-quadrant-heatmap, fig.height=5.5}
quad_means <- scores_out %>%
  pivot_longer(all_of(outcome_meta$column),
               names_to = "outcome", values_to = "value") %>%
  filter(!is.na(value)) %>%
  left_join(outcome_meta %>% select(column, label, bucket),
            by = c("outcome" = "column")) %>%
  group_by(outcome) %>%
  mutate(z = as.numeric(scale(value))) %>%
  ungroup() %>%
  group_by(bucket, label, quadrant) %>%
  summarise(mean_z = mean(z), n = n(), .groups = "drop") %>%
  mutate(label = factor(label, levels = rev(outcome_meta$label)))

ggplot(quad_means, aes(quadrant, label, fill = mean_z)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%+.2f", mean_z)), size = 3) +
  scale_fill_gradient2(low = highlight, mid = "grey95",
                       high = gl$brand_blue, midpoint = 0,
                       name = "Mean z-score") +
  labs(x = NULL, y = NULL,
       subtitle = "Outcome mean (z) by PCA quadrant") +
  theme(legend.position = "bottom",
        panel.grid = element_blank())
```

## Quadrant count

```{r pca-quadrant-counts}
scores %>%
  count(quadrant) %>%
  kable(caption = "Countries per quadrant. Small-N quadrants make the row means above noisier.")
```

# Top bundle per outcome

For each outcome, the highest-weight bundle from its best-depth model — the
*set* of V-Dem features that the most-frequented tree paths repeatedly split on.
Weight is the total leaf cover routed through paths that use exactly this set,
so the top bundle is the "decision rule" the model uses most often, not
necessarily the bundle with the strongest direct correlation to the outcome.
Each feature is listed with its V-Dem codebook description for interpretability.

```{r top-bundle-per-outcome, results='asis'}
for (i in seq_len(nrow(outcome_meta))) {
  col <- outcome_meta$column[i]
  bd <- best_depth_df %>% filter(label == outcome_meta$label[i]) %>%
    pull(depth) %>% as.character()
  r <- results[[col]][[bd]]
  if (is.null(r) || nrow(r$bundles) == 0) next
  best <- r$bundles %>% slice_max(weight, n = 1, with_ties = FALSE)
  feats <- strsplit(best$features, "\\|")[[1]]

  cat("\n## ", r$label, " (", outcome_meta$bucket[i], ")\n\n", sep = "")
  cat("- **Best depth:** ", r$depth, "  \n", sep = "")
  cat("- **OOS R\\textsuperscript{2}:** ", round(r$r2_oos, 3), "  \n", sep = "")
  cat("- **Bundle size:** ", best$size, "  \n", sep = "")
  cat("- **Bundle weight:** ", round(best$weight, 1),
      " (top 1 of ", nrow(r$bundles), " distinct bundles)\n\n", sep = "")
  cat("**Features in this bundle:**\n\n")
  for (code in feats) {
    desc <- desc_lookup[code]
    if (is.na(desc)) desc <- "(no codebook description)"
    cat("- `", code, "` — ", desc, "\n", sep = "")
  }
  cat("\n")
}
```

# Notes and caveats

- **Cross-sectional.** Everything here is a single-year snapshot. No causal claims.
- **Transforms.** Infant and maternal mortality are modeled in logs. SHAP units for
  those outcomes are log-units of mortality.
- **Regularisation.** `min_child_weight = 5` keeps leaves from specialising on
  one or two countries with N in the 130-190 range. CV-tuned `nrounds` plus the
  depth sweep are the only moving pieces.
- **Bundle extraction.** A bundle is the *set* of distinct V-Dem features used on
  any split along the path from root to a given leaf. Weight sums leaf `Cover`
  (approximately the number of training samples reaching that leaf), aggregated
  across all leaves in all trees whose feature-set matches. Bundles are
  direction-free — they say "these features co-occur in decisions" rather than
  "high A + low B co-occur."
- **Feature overlap.** V-Dem indicators are highly collinear. When two similar
  features (e.g. two measures of executive respect for law) show up together in a
  bundle, it may mean the model is using one as a noisy proxy for the other
  rather than that they genuinely interact. Treat top bundles as hypotheses.
