---
title: "Deals and Development: Stakeholder Profiles from Institutional Data"
subtitle: "Mapping V-Dem and QoG indicators to the Pritchett-Sen-Werker typology"
author: "Institutional Complexity Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6.5
    fig_height: 4.5
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  cache = TRUE,
  cache.lazy = FALSE
)
```

# Framing

Pritchett, Sen & Werker (2018, *Deals and Development*, OUP) classify firms
on two axes — target market (export vs. domestic) × source of profit
(regulatory rents vs. market competition) — yielding four stakeholder
types: **magicians** (export, competitive), **workhorses** (domestic,
competitive), **rentiers** (export, regulatory rents), and
**powerbrokers** (domestic, regulatory rents). Each type demands a
different policy bundle along six axes (Table 1.1, p. 22):

|                  | Transp. rules | Common infra | Special. infra | Contract enf. | Subsidies | Exclusivity |
| ---------------- | :-----------: | :----------: | :------------: | :-----------: | :-------: | :---------: |
| **Magicians**    |       x       |       x      |        x       |       x       |     x     |             |
| **Workhorses**   |       x       |       x      |                |       x       |           |             |
| **Rentiers**     |               |              |        x       |       x       |     x     |      x      |
| **Powerbrokers** |               |              |                |               |           |      x      |

We treat realised institutional outcomes as **revealed bargaining power**:
a country whose institutions strongly supply the policy bundle a
stakeholder demands is one where that stakeholder has been winning. The
deliverable is a country-by-stakeholder fit score, interpreted as a
qualitative read on which coalition has shaped each country's
institutional landscape.

```{r packages}
library(tidyverse)
library(arrow)
library(countrycode)
library(ggrepel)
library(knitr)
library(tidytext)  # reorder_within, scale_y_reordered

source("~/dev/gl-design/skills/gl-ggplot/assets/theme_gl.R")
gl_setup()
```

# Data

## V-Dem and macro panel

```{r load-vdem, cache=TRUE}
vdem_raw <- read_parquet("data/shared-data/vdem/vdem.parquet")

latest_year <- vdem_raw %>%
  filter(year >= 2020) %>%
  group_by(year) %>%
  summarise(n_countries = n_distinct(country_name), .groups = "drop") %>%
  filter(n_countries >= 150) %>%
  pull(year) %>% max()

cat("Latest V-Dem year:", latest_year, "\n")
```

```{r indicator-map}
indicator_map <- tribble(
  ~axis,                    ~indicator,                ~source, ~sign,
  # Axis 1: transparent rules
  "transparent_rules",      "v2cltrnslw",              "vdem",   1,
  "transparent_rules",      "v2clrspct",               "vdem",   1,
  # Axis 2: common infrastructure
  "common_infra",           "v2dlencmps",              "vdem",   1,
  "common_infra",           "wdi_eg_elc_accs_zs",      "macro",  1,
  "common_infra",           "wdi_sh_h2o_basw_zs",      "macro",  1,
  # Axis 3: specialized infrastructure
  "specialized_infra",      "wdi_lp_lpi_infr_xq",      "macro",  1,
  "specialized_infra",      "wdi_is_shp_gcnw_xq",      "macro",  1,
  "specialized_infra",      "wdi_gb_xpd_rsdv_gd_zs",   "macro",  1,
  "specialized_infra",      "wdi_tx_val_tech_mf_zs",   "macro",  1,
  # Axis 4: contract enforcement
  "contract_enforcement",   "v2xcl_prpty",             "vdem",   1,
  "contract_enforcement",   "v2juncind",               "vdem",   1,
  "contract_enforcement",   "v2juhcind",               "vdem",   1,
  "contract_enforcement",   "v2juhccomp",              "vdem",   1,
  "contract_enforcement",   "v2jucorrdc",              "vdem",   1,
  "contract_enforcement",   "v2juaccnt",               "vdem",   1,
  # Axis 5: subsidies / clientelism (high score = MORE clientelism supplied).
  # v2psprlnks and v2elvotbuy are inputs to v2xnp_client, so we keep only the
  # composite plus the orthogonal v2dlencmps spending-mix indicator.
  "subsidies_clientelism",  "v2xnp_client",            "vdem",   1,
  "subsidies_clientelism",  "v2dlencmps",              "vdem",  -1,  # higher = public goods
  # Axis 6: exclusivity / barriers to entry (high = MORE exclusivity supplied).
  # Market-side only; political-exclusivity V-Dem indicators dropped.
  "exclusivity",            "qog_fi_reg",              "macro", -1,  # higher Fraser = less regulation
  "exclusivity",            "qog_bti_mo",              "macro", -1   # higher = better-organised market
)

axis_levels <- c("transparent_rules", "common_infra", "specialized_infra",
                 "contract_enforcement", "subsidies_clientelism", "exclusivity")
axis_labels <- c("Transparent\nrules", "Common\ninfra", "Specialized\ninfra",
                 "Contract\nenforcement", "Subsidies /\nclientelism", "Exclusivity")
names(axis_labels) <- axis_levels

kable(
  indicator_map %>% count(axis) %>% mutate(axis = factor(axis, levels = axis_levels)) %>%
    arrange(axis),
  col.names = c("Axis", "# indicators"),
  caption = "Indicators per policy-demand axis"
)
```

```{r pull-vdem}
vdem_cols <- indicator_map %>% filter(source == "vdem") %>% pull(indicator) %>% unique()

vdem_latest <- vdem_raw %>%
  filter(year == latest_year) %>%
  select(country_name, iso3 = country_text_id, all_of(vdem_cols))
```

```{r pull-macro, cache=TRUE}
macro_cols <- indicator_map %>% filter(source == "macro") %>% pull(indicator) %>% unique()

macro <- read_parquet(
  "data/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year", all_of(macro_cols))
)

# For each macro indicator, take the most recent non-NA value within the
# six years ending at latest_year.
window <- seq(latest_year - 5, latest_year)

pick_latest <- function(col) {
  macro %>%
    filter(year %in% window, !is.na(.data[[col]])) %>%
    group_by(countrycodeiso) %>%
    slice_max(year, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(iso3 = countrycodeiso, !!col := .data[[col]])
}

macro_latest <- map(macro_cols, pick_latest) %>%
  reduce(full_join, by = "iso3")
```

```{r combine}
panel <- vdem_latest %>%
  left_join(macro_latest, by = "iso3") %>%
  filter(!is.na(country_name))

cat("Panel:", nrow(panel), "countries ×",
    length(c(vdem_cols, macro_cols)), "indicators\n")
```

# Axis construction

For each of the six policy demands we pick a small basket of V-Dem
and QoG/macro indicators. The construction recipe is the same for every
axis:

1. **Percentile-rank globally.** Each indicator is converted to its
   percentile within the country panel, putting it on `[0, 1]` regardless
   of the original scale (V-Dem ordinal, QoG continuous, WDI in raw
   units, etc.). This avoids letting a single wide-ranged variable
   dominate the average.
2. **Sign-flip where needed.** Some indicators are coded so that higher
   values mean *less* of the demand being supplied (e.g. V-Dem's
   `v2psparban` runs from 0 = "parties banned" up to 4 = "no party ban",
   so high values mean *less* exclusivity). Those get flipped to
   `1 - pct` before averaging.
3. **Mean within axis.** The axis score is the unweighted mean of
   indicator scores. Missing values are dropped from the mean, so a
   country with partial coverage still gets scored on what's available.

The result is a country × 6-axis matrix on `[0, 1]` where higher = the
country supplies more of what that policy demand is asking for.

```{r axis-helpers}
gloss <- c(
  v2cltrnslw            = "Transparent laws with predictable enforcement",
  v2clrspct             = "Rigorous and impartial public administration",
  v2dlencmps            = "Particularistic vs. public-goods spending",
  wdi_eg_elc_accs_zs    = "Electricity access (% pop.)",
  wdi_sh_h2o_basw_zs    = "Basic drinking water access (% pop.)",
  wdi_lp_lpi_infr_xq    = "Logistics Performance Index — infrastructure",
  wdi_is_shp_gcnw_xq    = "Liner Shipping Connectivity Index",
  wdi_gb_xpd_rsdv_gd_zs = "R&D expenditure (% GDP)",
  wdi_tx_val_tech_mf_zs = "High-tech exports (% manufactured exports)",
  v2xcl_prpty           = "Property rights (V-Dem composite of men's and women's property rights)",
  v2juncind             = "Lower court independence",
  v2juhcind             = "High court independence",
  v2juhccomp            = "Government compliance with high-court rulings",
  v2jucorrdc            = "Judicial bribery (V-Dem coding: high = less bribery)",
  v2juaccnt             = "Judicial accountability for misconduct",
  v2xnp_client          = "Clientelism index (V-Dem composite)",
  qog_fi_reg            = "Fraser regulation index (high = less regulation)",
  qog_bti_mo            = "BTI organisation of market and competition"
)

axis_table <- function(ax) {
  indicator_map %>%
    filter(axis == ax) %>%
    transmute(
      Indicator = indicator,
      Source    = ifelse(source == "vdem", "V-Dem", "QoG/macro"),
      Direction = ifelse(sign == 1, "+", "- (flipped)"),
      Gloss     = gloss[indicator]
    ) %>%
    kable()
}
```

## Transparent rules

PSW define this as predictable, publicly known regulation — firms can
plan based on rules they can read, not deals they have to negotiate.
We use two complementary V-Dem indicators that capture the law-on-paper
and the law-in-practice:

- **`v2cltrnslw` — Transparent laws with predictable enforcement.**
  V-Dem question: *"Are the laws of the land clear, well publicized,
  coherent (consistent with each other), relatively stable from year
  to year, and enforced in a predictable manner?"* Ordinal 0–4,
  converted to interval by V-Dem's measurement model: 0 =
  "Transparency and predictability are almost non-existent. The laws
  of the land are created and/or enforced in completely arbitrary
  fashion." through 4 = "Transparency and predictability are very
  strong. The laws of the land are created and enforced in a
  non-arbitrary fashion." This is about the *quality and coherence of
  the legal code itself* — whether the rules are knowable and stable.

- **`v2clrspct` — Rigorous and impartial public administration.**
  V-Dem question: *"Are public officials rigorous and impartial in
  the performance of their duties?"* The codebook clarifies that
  "this question focuses on the extent to which public officials
  generally abide by the law and treat like cases alike, or
  conversely, the extent to which public administration is
  characterized by arbitrariness and biases (i.e., nepotism,
  cronyism, or discrimination)." Ordinal 0–4 (interval after the
  measurement model): 0 = "The law is not respected by public
  officials. Arbitrary or biased administration of the law is
  rampant." through 4 = "The law is generally fully respected by the
  public officials. Arbitrary or biased administration of the law is
  very limited." This captures the *bureaucratic* face of the same
  concept — street-level impartiality.

A country could have clear laws applied corruptly (high `v2cltrnslw`,
low `v2clrspct`) or murky laws applied conscientiously, so both are
needed. Both run positively; no sign-flipping.

```{r ax-tab-transparent}
axis_table("transparent_rules")
```

## Common infrastructure

The PSW demand is for the public goods every firm benefits from —
electricity, water, basic education, transport. We pair one V-Dem
fiscal-intent indicator with two WDI outcome variables that have
near-universal coverage:

- **`v2dlencmps` — Particularistic vs. public-goods spending.**
  V-Dem question: *"Considering the profile of social and
  infrastructural spending in the national budget, how
  'particularistic' or 'public goods' are most expenditures?"* The
  codebook clarifies that "particularistic spending is narrowly
  targeted on a specific corporation, sector, social group, region,
  party, or set of constituents… Public-goods spending is intended
  to benefit all communities within a society." Ordinal 0–4
  (interval after the measurement model), running from 0 = "Almost
  all of the social and infrastructure expenditures are
  particularistic" to 4 = "Almost all social and infrastructure
  expenditures are public-goods in character." Higher = more
  public-goods spending mix; this is the *fiscal intent* side of
  common-infra supply.

- **`wdi_eg_elc_accs_zs`** — share of population with access to
  electricity.

- **`wdi_sh_h2o_basw_zs`** — share of population with access to at
  least basic drinking water.

The two WDI variables capture realised supply of universal
infrastructure with broad country coverage. (Earlier drafts also
included secondary enrolment and PWT human capital, which we drop
here in favour of indicators more directly about physical-network
infrastructure rather than human capital.) All three run positively.

```{r ax-tab-common}
axis_table("common_infra")
```

## Specialized infrastructure

This is the demand for sector-specific public investment that
benefits a subset of firms — special economic zones, port facilities,
R&D infrastructure, regulatory bodies that handle complex industries.
V-Dem doesn't capture this directly, so we lean entirely on
outcome-side proxies. `wdi_lp_lpi_infr_xq` (Logistics Performance
Index — infrastructure) and `wdi_is_shp_gcnw_xq` (Liner Shipping
Connectivity Index) measure trade-enabling specialized
infrastructure; `wdi_gb_xpd_rsdv_gd_zs` (R&D as % GDP) and
`wdi_tx_val_tech_mf_zs` (high-tech export share) proxy R&D
infrastructure and its outputs. We deliberately exclude the Atlas
complexity indices (ECI, COI) here: they are downstream *outcomes*
of specialized infrastructure rather than a measure of the
institutional supply, and including them would risk circularity in
later sanity checks against complexity. All four run positively;
coverage is patchier than the other axes.

```{r ax-tab-specialized}
axis_table("specialized_infra")
```

## Contract enforcement

Courts that resolve commercial disputes predictably and protect
property. V-Dem has the strongest battery here. We checked whether
the property-rights composite already aggregates the judicial
indicators and could replace them: it does not. `v2xcl_prpty` is
built only from `v2clprptym` (men's property rights) and
`v2clprptyw` (women's property rights), with no judicial inputs.
Conversely the judicial battery feeds into `v2x_jucon` (judicial
constraints on the executive) and `v2x_rule`, neither of which we
use. So property rights and the judicial battery are non-redundant
and both are kept.

- **`v2xcl_prpty`** — V-Dem property-rights composite (men's and
  women's de facto rights to acquire, possess, inherit, and sell
  private property).
- **`v2juncind`** and **`v2juhcind`** — independence of lower courts
  and the high court from political pressure.
- **`v2juhccomp`** — government compliance with high-court rulings,
  the most demanding test of judicial enforcement (does the executive
  *actually* abide by rulings it dislikes).
- **`v2jucorrdc`** — judicial bribery; V-Dem's ordinal coding runs
  from "Always" (low) to "Never" (high), so high = less bribery and
  no flip is needed.
- **`v2juaccnt`** — accountability of judges who engage in
  misconduct.

All six run positively.

```{r ax-tab-contract}
axis_table("contract_enforcement")
```

## Subsidies / clientelism

PSW define "subsidies" broadly to include directed credit, tax breaks,
and any state transfer that benefits specific firms or coalitions
rather than the economy at large. The unifying logic is
particularistic distribution.

- **`v2xnp_client` — Clientelism Index.** A V-Dem Bayesian factor
  composite of how much politics is built on the targeted, contingent
  distribution of resources (goods, services, jobs, money) in
  exchange for political support. The codebook flags that this
  composite is built from inputs that include `v2psprlnks` (party
  linkages) and `v2elvotbuy` (vote buying) directly. For that reason
  we drop those two as separate axis components — keeping them
  alongside `v2xnp_client` would double-count their signal. Higher
  score = more clientelism.

- **`v2dlencmps`** (flipped) — the same particularistic-vs-public-goods
  budget-mix variable used in Axis 2. Flipped here so that *higher*
  reflects more particularistic spending. This complements
  `v2xnp_client` by capturing the *fiscal* face of clientelism rather
  than the political-relationships face, and is not an input to the
  composite.

The axis score is high when both political relationships and fiscal
spending flow through targeted, particularistic channels.

```{r ax-tab-subsidies}
axis_table("subsidies_clientelism")
```

## Exclusivity

Barriers to entry, licensing regimes, regulatory rents that protect
incumbents. We narrow this axis to the *market-side* face of
exclusivity, dropping the political-exclusivity V-Dem indicators
(`v2x_neopat`, `v2psparban`, `v2psoppaut`) used in earlier drafts.
That is a deliberate scope decision — the kept indicators are
clean, well-coded, and squarely about market entry, but they cannot
distinguish open-market/closed-politics regimes (e.g. Singapore-type)
from open-market/open-politics regimes. A reader interested in the
political face of exclusivity should consult the V-Dem indicators
separately. Both kept indicators are flipped so that higher = more
exclusivity:

- **`qog_fi_reg`** — Fraser Institute composite of business, credit,
  and labour-market regulation. The Fraser coding has *higher*
  scores for *less* restrictive regulation, so we flip: higher in
  our axis = more restrictive regulation, more exclusivity.
- **`qog_bti_mo`** — Bertelsmann Transformation Index "Organisation
  of the market and competition." Higher BTI = a more competitive,
  open-market environment, so we flip: higher in our axis = a more
  closed/captured market environment.

The axis score is high when the regulatory environment favours
incumbents over entrants.

```{r ax-tab-exclusivity}
axis_table("exclusivity")
```

# Structural variable — resource rents

The rentier and powerbroker stakeholder types both profit from
regulatory rents and demand near-identical policy bundles; what
distinguishes them is whether the rent-seeking sector is export-facing
(rentiers: oil, mining, plantations) or domestic (powerbrokers: telco,
banking, state contracts). The institutional indicators alone cannot
separate them. Following PSW's own use of sector size to make this
distinction, we pull **total natural resource rents as % of GDP**
(WDI `NY.GDP.TOTL.RT.ZS`, summing oil, gas, coal, mineral and forest
rents) and fold it directly into the fit calculation below as a
multiplicative weight on the rentier and powerbroker columns.

```{r resource-rents}
# Resource rents move slowly (it's a structural feature of the economy),
# so we grab the latest available value since 2010 rather than the tight
# window we use for institutional indicators. This recovers Venezuela
# (2014), Eritrea (2011), South Sudan (2015), Syria (2020) and others
# whose WDI series stopped being reported in recent years.
rent_latest <- read_parquet(
  "data/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year", "wdi_ny_gdp_totl_rt_zs")
) %>%
  filter(!is.na(wdi_ny_gdp_totl_rt_zs), year >= 2010) %>%
  group_by(countrycodeiso) %>% slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(iso3 = countrycodeiso,
            resource_rents_pct_gdp = wdi_ny_gdp_totl_rt_zs,
            resource_rents_year    = year)

cat("Countries with resource-rents data:", nrow(rent_latest), "\n")
cat("Year coverage:\n")
print(table(rent_latest$resource_rents_year))
```

# Axis scoring

```{r score-axes}
percentile_rank <- function(x) {
  r <- rank(x, na.last = "keep")
  r / sum(!is.na(x))
}

# Score every indicator
indicator_scores <- panel %>%
  pivot_longer(c(-country_name, -iso3),
               names_to = "indicator", values_to = "value") %>%
  inner_join(indicator_map, by = "indicator") %>%
  group_by(indicator) %>%
  mutate(pct = percentile_rank(value)) %>%
  ungroup() %>%
  mutate(score = ifelse(sign == 1, pct, 1 - pct))

# Average within axis (drop NA so countries with partial coverage still score)
axis_scores <- indicator_scores %>%
  group_by(country_name, iso3, axis) %>%
  summarise(axis_score = mean(score, na.rm = TRUE),
            n_indicators = sum(!is.na(score)),
            .groups = "drop")

# Coverage: keep countries with at least 1 non-NA indicator on every axis
ok_countries <- axis_scores %>%
  group_by(country_name) %>%
  summarise(min_n = min(n_indicators), .groups = "drop") %>%
  filter(min_n >= 1) %>% pull(country_name)

axis_scores <- axis_scores %>% filter(country_name %in% ok_countries)

axis_wide <- axis_scores %>%
  select(country_name, iso3, axis, axis_score) %>%
  pivot_wider(names_from = axis, values_from = axis_score) %>%
  select(country_name, iso3, all_of(axis_levels))

cat("Countries with full axis coverage:", nrow(axis_wide), "\n")
```

# Stakeholder fit scores

```{r psw-prefs}
psw_prefs <- tribble(
  ~stakeholder,
  ~transparent_rules, ~common_infra, ~specialized_infra,
  ~contract_enforcement, ~subsidies_clientelism, ~exclusivity,
  "Magicians",    1, 1, 1, 1, 1, 0,
  "Workhorses",   1, 1, 0, 1, 0, 0,
  "Rentiers",     0, 0, 1, 1, 1, 1,
  "Powerbrokers", 0, 0, 0, 0, 0, 1
)

stakeholder_levels <- c("Magicians", "Workhorses", "Rentiers", "Powerbrokers")

pref_mat <- psw_prefs %>% column_to_rownames("stakeholder") %>% as.matrix()
profile_mat <- axis_wide %>% column_to_rownames("country_name") %>%
  select(-iso3) %>% as.matrix()
```

For each country and each stakeholder, the **institutional fit** is the
average of the country's axis scores over the axes that stakeholder
demands. Equivalently: "out of 1, how well does this country supply
what stakeholder X is asking for." We then apply the structural
adjustment described above — multiplying the rentier column by the
country's percentile-ranked resource rents / GDP, and the powerbroker
column by its complement — so that countries with sizeable export-rent
sectors carry rentier mass and resource-poor countries with the same
exclusivity profile concentrate on powerbrokers. Magicians and
workhorses are not adjusted; the export-vs-domestic split for the
market-competition stakeholders is left to the institutional axes
(no clean structural variable for non-resource competitive exports).
Finally we renormalise the four adjusted scores to sum to 1 to read
them as *stakeholder shares*.

The share interpretation has a built-in caveat. The four preference
vectors overlap (e.g., contract enforcement is wanted by everyone except
powerbrokers), so the share isn't a clean decomposition — it's a
similarity-based approximation of which stakeholder profile a country's
institutions most resemble.

```{r fit-scores}
# 1. Institutional-only fit: average of axis scores over demanded axes
fit_mat_inst <- profile_mat %*% t(pref_mat) /
  matrix(rowSums(pref_mat),
         nrow = nrow(profile_mat),
         ncol = nrow(pref_mat),
         byrow = TRUE)

# 2. Structural adjustment for rentier/powerbroker.
# Percentile-rank resource rents/GDP across countries with axis coverage.
rents_keyed <- tibble(country_name = rownames(fit_mat_inst)) %>%
  left_join(axis_wide %>% select(country_name, iso3), by = "country_name") %>%
  left_join(rent_latest, by = "iso3") %>%
  mutate(
    rents_pct = percentile_rank(resource_rents_pct_gdp),
    # Countries missing rents data get neutral weight (0.5)
    rents_pct = ifelse(is.na(rents_pct), 0.5, rents_pct)
  )

fit_mat <- fit_mat_inst
fit_mat[, "Rentiers"]     <- fit_mat_inst[, "Rentiers"]     * rents_keyed$rents_pct
fit_mat[, "Powerbrokers"] <- fit_mat_inst[, "Powerbrokers"] * (1 - rents_keyed$rents_pct)

# 3. Renormalise rows to sum to 1
fit_shares <- fit_mat / rowSums(fit_mat)

fit_long <- as_tibble(fit_mat, rownames = "country_name") %>%
  pivot_longer(-country_name, names_to = "stakeholder", values_to = "fit") %>%
  left_join(
    as_tibble(fit_shares, rownames = "country_name") %>%
      pivot_longer(-country_name, names_to = "stakeholder", values_to = "share"),
    by = c("country_name", "stakeholder")
  ) %>%
  left_join(
    as_tibble(fit_mat_inst, rownames = "country_name") %>%
      pivot_longer(-country_name, names_to = "stakeholder", values_to = "fit_inst"),
    by = c("country_name", "stakeholder")
  ) %>%
  mutate(stakeholder = factor(stakeholder, levels = stakeholder_levels)) %>%
  left_join(axis_wide %>% select(country_name, iso3), by = "country_name") %>%
  left_join(rent_latest, by = "iso3")

# Country labels reused across plots
label_set <- c("USA", "DEU", "JPN", "KOR", "CHN", "VNM", "BGD", "KHM",
               "MYS", "IND", "MEX", "TUR", "BRA", "RUS", "SAU", "VEN",
               "AGO", "NGA", "ETH", "RWA", "LBR", "MAR")
```

## Top countries by stakeholder fit

```{r top-by-stakeholder, fig.width=6.5, fig.height=8.5}
top15 <- fit_long %>%
  group_by(stakeholder) %>%
  slice_max(fit, n = 15) %>%
  ungroup() %>%
  mutate(country_name = reorder_within(country_name, fit, stakeholder))

ggplot(top15, aes(x = fit, y = country_name, fill = stakeholder)) +
  geom_col() +
  facet_wrap(~ stakeholder, scales = "free_y", ncol = 2) +
  scale_y_reordered() +
  labs(x = "Fit score (mean of demanded-axis scores)",
       y = NULL,
       title = "Top 15 countries by stakeholder fit",
       subtitle = paste0("Higher fit = institutions supply more of what stakeholder demands. ",
                         "V-Dem ", latest_year, "."))
```

```{r top-tables}
top10_table <- fit_long %>%
  group_by(stakeholder) %>%
  slice_max(fit, n = 10) %>%
  mutate(rank = row_number(),
         entry = sprintf("%s (%.2f)", country_name, fit)) %>%
  ungroup() %>%
  select(rank, stakeholder, entry) %>%
  pivot_wider(names_from = stakeholder, values_from = entry) %>%
  select(rank, all_of(stakeholder_levels))

kable(top10_table, caption = "Top 10 country fit per stakeholder (fit score in parens)")
```

## Stakeholder profile of selected countries

A few well-known cases — including those Pritchett, Sen & Werker
themselves apply the framework to in their book (Bangladesh, Cambodia,
Liberia, Malaysia, Rwanda) — alongside reference economies and
resource-dependent states.

```{r selected-profiles, fig.width=6.5, fig.height=5}
selected_iso <- c("BGD", "KHM", "LBR", "MYS", "RWA",  # PSW case studies
                  "USA", "DEU", "JPN", "KOR",          # advanced economies
                  "CHN", "VNM", "ETH",                 # state-led
                  "SAU", "RUS", "VEN", "AGO", "AZE",   # resource-heavy
                  "IND", "BRA", "MEX", "TUR", "ARG",   # large EM
                  "PAK", "MAR", "NGA")                 # other comparators

selected_shares <- fit_long %>%
  filter(iso3 %in% selected_iso) %>%
  mutate(country_name = fct_reorder2(country_name,
                                     stakeholder == "Magicians", share))

ggplot(selected_shares, aes(x = share, y = country_name, fill = stakeholder)) +
  geom_col() +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_gl() +
  labs(x = "Stakeholder share (fit, renormalised)",
       y = NULL,
       fill = NULL,
       title = "Stakeholder shares — selected countries",
       subtitle = "Renormalised so the four stakeholder fits sum to 100% per country") +
  theme(legend.position = "bottom")
```

## Axis profile of archetypal countries

For one country at the top of each stakeholder's ranking, what does its
6-axis institutional profile look like?

```{r archetype-axes, fig.width=6.5, fig.height=4.5}
archetypes <- fit_long %>%
  group_by(stakeholder) %>%
  slice_max(fit, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(stakeholder, country_name)

archetype_axes <- axis_scores %>%
  inner_join(archetypes, by = "country_name") %>%
  mutate(axis = factor(axis, levels = axis_levels, labels = axis_labels[axis_levels]),
         stakeholder = factor(stakeholder, levels = stakeholder_levels)) %>%
  mutate(label = paste0(country_name, "\n(", stakeholder, ")"))

# Order facet labels by stakeholder
archetype_axes <- archetype_axes %>%
  mutate(label = factor(label,
                        levels = archetype_axes %>%
                          distinct(stakeholder, label) %>%
                          arrange(stakeholder) %>% pull(label)))

ggplot(archetype_axes, aes(x = axis, y = axis_score, group = label)) +
  geom_col(aes(fill = stakeholder)) +
  facet_wrap(~ label, ncol = 2) +
  scale_fill_gl() +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Axis score (percentile)",
       fill = NULL,
       title = "Six-axis institutional profile of each archetype",
       subtitle = "Top-ranked country on each stakeholder's fit score") +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 7))
```

# Diagnostic: why the structural adjustment matters

Before the rents adjustment, the institutional fit alone fails to
separate rentier-leaning from powerbroker-leaning countries — the
demanded policy bundles overlap heavily on exclusivity, and
powerbroker fit reduces mechanically to the exclusivity score. The
following 2×2 plots the *institutional-only* powerbroker fit (i.e.
`fit_inst`, before applying the rents weight) against resource rents
/ GDP. Countries in the top-right have high institutional powerbroker
fit *and* a large export-rentier sector — the rents adjustment
reallocates their share towards rentiers. Countries in the
bottom-right have the same institutional profile but no rent base —
they keep their share on powerbrokers.

```{r pb-vs-rents, fig.width=6.5, fig.height=5}
pb_rents <- fit_long %>%
  filter(stakeholder == "Powerbrokers") %>%
  filter(!is.na(resource_rents_pct_gdp))

pb_cut   <- median(pb_rents$fit_inst, na.rm = TRUE)
rent_cut <- 5

ggplot(pb_rents, aes(x = fit_inst, y = resource_rents_pct_gdp)) +
  geom_vline(xintercept = pb_cut, linetype = "dashed", alpha = 0.4) +
  geom_hline(yintercept = rent_cut, linetype = "dashed", alpha = 0.4) +
  geom_point(alpha = 0.5) +
  geom_text_repel(
    data = ~ filter(.x, iso3 %in% label_set |
                         resource_rents_pct_gdp > 15 |
                         (fit_inst > quantile(fit_inst, 0.85, na.rm = TRUE) &
                          resource_rents_pct_gdp > rent_cut)),
    aes(label = iso3), size = 2.8, max.overlaps = 30
  ) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Powerbroker fit (institutional only, pre-adjustment)",
       y = "Total natural resource rents (% GDP)",
       title = "Diagnostic: institutional fit alone collapses powerbroker and rentier",
       subtitle = paste0("Top-right cases shift toward rentier under the structural ",
                         "adjustment; bottom-right stay on powerbroker."))
```

```{r pb-rents-table}
pb_rents %>%
  mutate(
    quadrant = case_when(
      fit_inst >= pb_cut & resource_rents_pct_gdp >= rent_cut ~ "Reallocated to rentier",
      fit_inst >= pb_cut & resource_rents_pct_gdp <  rent_cut ~ "Stays on powerbroker",
      fit_inst <  pb_cut & resource_rents_pct_gdp >= rent_cut ~ "Resource-rich, weak powerbroker fit",
      TRUE                                                     ~ "Magician/workhorse territory"
    )
  ) %>%
  count(quadrant, name = "n_countries") %>%
  arrange(desc(n_countries)) %>%
  kable(caption = "Country count by quadrant (using institutional-only PB fit)")
```

## Spider plot: Burundi vs Cambodia

A radar plot summarising the four-stakeholder share profile. Burundi
and Cambodia look almost identical on institutional fit (both
high-powerbroker, modest-rentier), but the integrated calculation
splits them cleanly: Burundi's 14% resource-rents/GDP pulls its share
toward rentiers, Cambodia's 0.8% pulls it toward powerbrokers. Axes
are arranged to preserve the PSW 2×2: market-competition stakeholders
(magicians, workhorses) at top, regulatory-rents stakeholders
(powerbrokers, rentiers) at bottom; export-facing types (magicians,
rentiers) on the left, domestic types (workhorses, powerbrokers) on
the right.

```{r country-spider, fig.width=6.5, fig.height=5}
spider_order <- c("Magicians", "Workhorses", "Powerbrokers", "Rentiers")
focus_iso    <- c("BDI", "KHM")

# Angular positions (radians, measured CCW from positive x-axis), arranged
# so factor levels go clockwise starting at the top.
n_axes <- length(spider_order)
axes_df <- tibble(
  stakeholder = factor(spider_order, levels = spider_order),
  angle       = pi/2 - 2 * pi * (seq_len(n_axes) - 1) / n_axes
)

# Country vertices in cartesian coordinates
spider_data <- fit_long %>%
  filter(iso3 %in% focus_iso) %>%
  mutate(stakeholder = factor(stakeholder, levels = spider_order)) %>%
  inner_join(axes_df, by = "stakeholder") %>%
  arrange(country_name, stakeholder) %>%
  mutate(x = share * cos(angle),
         y = share * sin(angle))

# Close the polygon loop by repeating the first vertex per country
spider_closed <- spider_data %>%
  group_by(country_name) %>%
  group_modify(~ bind_rows(.x, slice(.x, 1))) %>%
  ungroup()

# Concentric grid rings + axis spokes
grid_rings <- expand_grid(r = c(0.25, 0.5, 0.75, 1.0),
                          theta = seq(0, 2 * pi, length.out = 120)) %>%
  mutate(x = r * cos(theta), y = r * sin(theta), grp = factor(r))

axis_spokes <- axes_df %>%
  mutate(x_end = cos(angle), y_end = sin(angle))

axis_text <- axes_df %>%
  mutate(x_lbl = 1.13 * cos(angle),
         y_lbl = 1.13 * sin(angle),
         hjust = case_when(cos(angle) >  0.3 ~ 0,
                           cos(angle) < -0.3 ~ 1,
                           TRUE              ~ 0.5))

ring_labels <- tibble(r = c(0.25, 0.5, 0.75)) %>%
  mutate(x = 0, y = r, label = scales::percent(r, accuracy = 1))

ggplot() +
  geom_path(data = grid_rings,
            aes(x, y, group = grp), color = "grey85") +
  geom_segment(data = axis_spokes,
               aes(x = 0, y = 0, xend = x_end, yend = y_end),
               color = "grey85") +
  geom_polygon(data = spider_closed,
               aes(x, y, group = country_name,
                   fill = country_name, color = country_name),
               alpha = 0.25, linewidth = 0.8) +
  geom_point(data = spider_data,
             aes(x, y, color = country_name), size = 2.4) +
  geom_text(data = axis_text,
            aes(x_lbl, y_lbl, label = stakeholder, hjust = hjust),
            size = 3.5) +
  geom_text(data = ring_labels,
            aes(x, y, label = label),
            size = 2.6, color = "grey55", vjust = -0.4) +
  coord_equal(xlim = c(-1.35, 1.35), ylim = c(-1.2, 1.2), clip = "off") +
  scale_color_gl() +
  scale_fill_gl() +
  labs(color = NULL, fill = NULL,
       title = "Stakeholder share — Burundi vs Cambodia",
       subtitle = "After the rents adjustment: Burundi tips to rentier, Cambodia to powerbroker") +
  theme_void(base_family = ggplot2::theme_get()$text$family) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey30"))
```

```{r country-spider-rents}
fit_long %>%
  filter(iso3 %in% focus_iso, stakeholder == "Powerbrokers") %>%
  select(country_name, resource_rents_pct_gdp) %>%
  kable(col.names = c("Country", "Resource rents (% GDP)"),
        digits = 1,
        caption = "Structural variable: resource rents as % of GDP")
```

# Sanity checks

## Magician fit vs. Economic Complexity

If the framework holds, magician-style institutions (transparent rules,
contract enforcement, public-goods spending) should be associated with
diversified, sophisticated export baskets. We expect a positive but
imperfect relationship — magician-friendly institutions are a
*precondition* for complexity, not a mechanical determinant.

```{r magician-vs-eci, fig.width=6.5, fig.height=4.5}
eci_latest <- read_parquet(
  "data/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year", "atl_sitc_eci")
) %>%
  filter(!is.na(atl_sitc_eci)) %>%
  group_by(countrycodeiso) %>% slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(iso3 = countrycodeiso, eci = atl_sitc_eci)

magician_eci <- fit_long %>%
  filter(stakeholder == "Magicians") %>%
  inner_join(eci_latest, by = "iso3")

ggplot(magician_eci, aes(x = fit, y = eci)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = highlight) +
  geom_text_repel(
    data = ~ filter(.x, iso3 %in% label_set),
    aes(label = iso3), size = 2.8, max.overlaps = 30
  ) +
  labs(x = "Magician fit score",
       y = "Economic Complexity Index (latest)",
       title = "Magician institutions vs. economic complexity",
       subtitle = sprintf("r = %.2f",
                          cor(magician_eci$fit, magician_eci$eci, use = "pairwise")))
```

## Powerbroker vs. magician fit

These two stakeholder types have nearly orthogonal demand vectors —
magicians want everything except exclusivity, powerbrokers want only
exclusivity (and benefit from the absence of everything else). Their
fits across countries should be strongly negatively correlated.

```{r mag-vs-pb, fig.width=6.5, fig.height=5}
mag_pb <- fit_long %>%
  filter(stakeholder %in% c("Magicians", "Powerbrokers")) %>%
  select(country_name, iso3, stakeholder, fit) %>%
  pivot_wider(names_from = stakeholder, values_from = fit)

ggplot(mag_pb, aes(x = Magicians, y = Powerbrokers)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = highlight) +
  geom_text_repel(
    data = ~ filter(.x, iso3 %in% label_set),
    aes(label = iso3), size = 2.8, max.overlaps = 30
  ) +
  labs(x = "Magician fit", y = "Powerbroker fit",
       title = "Powerbroker world is the inverse of magician world",
       subtitle = sprintf("r = %.2f",
                          cor(mag_pb$Magicians, mag_pb$Powerbrokers, use = "pairwise")))
```

# Outputs

```{r write-outputs}
fit_out <- fit_long %>%
  select(country_name, iso3, stakeholder,
         fit_inst, fit, share,
         resource_rents_pct_gdp, resource_rents_year) %>%
  arrange(country_name, stakeholder)

axis_out <- axis_wide %>%
  left_join(rent_latest, by = "iso3")

dir.create("outputs", showWarnings = FALSE)
write_csv(fit_out, "outputs/deals_stakeholder_fit.csv")
write_csv(axis_out, "outputs/deals_axis_scores.csv")

cat("Wrote: outputs/deals_stakeholder_fit.csv (",
    nrow(fit_out), "rows)\n")
cat("Wrote: outputs/deals_axis_scores.csv (",
    nrow(axis_out), "rows)\n")
```

# Caveats and next steps

- **The share interpretation is approximate.** Preference vectors
  overlap (most stakeholders want contract enforcement; only powerbrokers
  affirmatively want exclusivity), so the renormalised shares aren't a
  proper decomposition of bargaining power — they're a similarity index
  rescaled.
- **Powerbroker institutional fit reduces to "high exclusivity."** With
  only one demanded axis, the unadjusted powerbroker fit is
  mechanically the exclusivity score. The structural adjustment using
  resource rents / GDP fixes this in the headline shares (rentier vs
  powerbroker now separate), but the institutional-only `fit_inst`
  column in the output retains the mechanical equivalence and should
  be read with that in mind.
- **Resource-rents data is point-in-time per country.** WDI's
  `NY.GDP.TOTL.RT.ZS` series stops being reported for several
  rent-economies in recent years (Eritrea 2011, Venezuela 2014, South
  Sudan 2015, Syria 2020). We pick the latest non-NA observation since
  2010 per country and expose the year as `resource_rents_year` in the
  output. Resource-rent shares move slowly so this is a reasonable
  approximation; readers comparing across countries should still note
  that Venezuela's rent share is fixed at its pre-collapse 2014 level.
- **Magicians and workhorses are correlated by design.** They share four
  of five demanded axes, so their rankings will track each other; the
  separation lives almost entirely in specialized infrastructure and
  subsidies/clientelism.
- **No infrastructure stock data in V-Dem.** Common-infra and
  specialized-infra rely on WDI access/quality variables, which have
  more measurement noise and patchier coverage than V-Dem.
- **Exclusivity is market-side only.** Dropping the V-Dem political
  exclusivity indicators (party bans, opposition autonomy,
  neopatrimonial rule) means the axis cannot separate
  open-market/closed-politics regimes from open-market/open-politics
  ones. A country with vibrant business competition but no political
  contestability scores low on exclusivity here even though PSW would
  read it as a powerbroker-friendly environment.
- **PSW's per-country rents-space estimates** (e.g. Liberia ~10/25/61/3,
  Malawi ~22/37/36/small for rentiers/powerbrokers/workhorses/magicians)
  could serve as ground truth for a few cases. Worth a follow-up
  comparison.

The unsupervised counterpart (archetype analysis on the 6-axis matrix,
recovering 4 latent stakeholder types from the data alone) is a
companion exercise documented in `deals-and-development.md`.
