---
title: "The Country Space"
subtitle: "Construction, Variable Choice, and Multi-Year Replication"
author: "Institutional Complexity Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6.5
    fig_height: 4.5
    latex_engine: xelatex
    includes:
      in_header: ~/dev/gl-design/skills/gl-ggplot/assets/gl_pdf.tex
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  cache = TRUE,
  cache.lazy = FALSE
)
```

# Overview

This document is the clean writeup of how we construct a 2D **country space** from V-Dem
governance indicators. It pulls together the version of the pipeline we have settled on
after experimenting with several alternatives in `institutional_complexity.Rmd` (the
"kitchen") and applies it to six benchmark years from **1975 to 2025**.

The pipeline in one sentence: take every continuous V-Dem indicator with adequate coverage
in the target year, z-score across countries so each indicator contributes equally, and
embed the resulting country x feature matrix in 2D with UMAP using Euclidean distance.

We deliberately avoid hand-picking indicators. The motivation, the consequences, and a few
sensitivity checks live in Section 3.

```{r packages}
library(tidyverse)
library(arrow)
library(uwot)
library(ggrepel)
library(countrycode)
library(knitr)
library(patchwork)
```

```{r gl-setup, cache=FALSE}
# Uncached so showtext / sysfonts re-attach on every knit and gl_setup()
# can self-install its knitr chunk hook. See deals-development.Rmd for the
# rationale.
source("~/dev/gl-design/skills/gl-ggplot/assets/theme_gl.R")
gl_setup()
```

```{r load-vdem}
vdem_raw       <- read_parquet("data/shared-data/vdem/vdem.parquet")
vdem_codebook  <- read_parquet("data/shared-data/vdem/vdem_codebook.parquet")
```

# 1. Why a country space?

The Atlas of Economic Complexity uses a *product space* — a 2D map of products, with
countries positioned by what they produce — to study how productive capabilities cluster
and how countries diversify. The institutional analogue swaps "products" for V-Dem's
expert-coded measures of governance: rule of law, electoral integrity, judicial
independence, civil-society autonomy, and so on. A country's vector across these
measures is its **institutional fingerprint**, and the country space is the 2D
projection of those fingerprints.

The point is not the absolute coordinates (any 2D embedding of a high-dimensional space
is heuristic). The point is that countries with similar institutional architectures end
up close together, and the resulting layout exposes regional and regime-type structure
that is otherwise hidden in tables of correlation matrices.

# 2. The data: V-Dem, year by year

V-Dem provides expert-coded scores for ~500 governance variables per country per year,
with full back-coverage to 1789 for established polities. Most of the variables we use
are produced by a Bayesian IRT measurement model that pools expert codings into a
continuous latent score (typically in roughly $[-4, +4]$). This continuity is what makes
Euclidean distance a sensible thing to compute.

We load the parquet snapshot maintained at `data/shared-data/vdem/vdem.parquet` and the
companion codebook. Coverage by benchmark year:

```{r year-coverage}
benchmark_years <- c(1975, 1985, 1995, 2005, 2015, 2025)

vdem_raw %>%
  filter(year %in% benchmark_years) %>%
  group_by(year) %>%
  summarise(n_countries = n_distinct(country_name), .groups = "drop") %>%
  kable(caption = "Country (polity) coverage by benchmark year")
```

V-Dem's polity count grows over the late 20th century as decolonization and the Soviet
collapse fragment large polities into many smaller ones. We treat each year's panel as
its own population: indicator selection and country selection are both done within the
target year (see Section 3).

# 3. Variable choice

This is the most consequential design decision in the whole pipeline. Two reasonable
positions exist:

1. **Hand-pick** a curated set of, say, 20-40 governance indicators, chosen on
   conceptual grounds. This is what most institutional-economics papers do.
2. **Take all of them** and let the metric do the work. This is what we do.

We pick (2) on the grounds that any hand-curated subset embeds the analyst's prior about
what governance "is" — judiciary vs. media vs. elections — and a country space is
supposed to be a *prior-light* map. The cost is that V-Dem ships indicator families
that triangulate the same latent concept (multiple measures of judicial independence,
multiple measures of media freedom). With min-max + Euclidean, those families implicitly
upweight the dimensions they cover. We accept this for now and flag it at the end of
this section. The Mahalanobis-whitened variant in `institutional_complexity.Rmd` is the
robustness check.

## What "continuous indicators" means concretely

V-Dem's codebook tags every variable with a `vartype`:

- **C** — Component: an expert-coded measurement, usually delivered as the IRT
  latent score. These are the raw signals.
- **A** — Aggregate / **D** — Derived: indices built from components (e.g., the
  liberal-democracy index). Including these would double-count the components they are
  built from.

We keep only `vartype == "C"`. From those we further filter:

- **Drop suffix variants.** Each base indicator ships with `_codelow`, `_codehigh`,
  `_sd`, `_osp`, `_ord`, `_mean`, `_nr` companions (uncertainty bounds, ordinal
  recodings, raw means, response counts). Only the base IRT score is kept.
- **Drop multi-coder per-rater columns** (`_1`, `_2`, ...). These are individual
  rater's codings, not aggregated scores.
- **Drop two discrete-categorical components**: `v2regopploc` ("regime opposition
  location") and `v2regsuploc` ("regime support location") are integer-coded categorical
  variables, not IRT scores. Treating them as continuous numbers is meaningless and
  they would dominate Euclidean distance because their nominal-integer scale is much
  wider than the IRT $[-4, +4]$ range.
- **Coverage filter.** Within the target year, an indicator must be observed for at
  least 50% of countries. Rare or near-empty columns add noise without signal.
- **Drop a small set of near-duplicate indicators** (see below).

The filter is stated as a rule, not a list of approved tags. Applied to a different
year, it will produce a slightly different indicator set as V-Dem coverage evolves
(some indicators were added in later codebook waves; some have spotty coverage in
the 1970s).

## Pruning near-duplicate indicators

The "all C-type indicators" rule keeps several indicator pairs/triples that V-Dem ships
as conceptual duplicates — most obviously, every civil-liberty indicator coded
separately for men and women, plus a few cause/effect pairs (high vs. lower court
independence; general CSO repression vs. religious-organization repression). Empirically
these pairs sit at $|r| \geq 0.85$ across the 2025 cross-section and contribute almost
nothing beyond the base indicator. Concretely, hierarchical clustering on $1 - |r|$ at
a $0.15$ cut surfaces 24 redundancy clusters covering 61 of the 192 indicators
(see `outputs/redundancy_clusters_2025.csv` for the full table).

We do **not** prune all 24 — many of those clusters bundle conceptually distinct
mechanisms that just happen to co-vary in 2025 (executive-corruption vs.
public-sector-corruption; censorship effort vs. self-censorship; polarization vs.
political violence). Collapsing those would lose interpretable signal in panel or
counterfactual analyses.

We do prune a conservative list of **18 indicators** where the redundancy is structural
rather than coincidental — gender-paired siblings, court-level pairs, religious-org-
specific copies of general-CSO measures, and a handful of within-family duplicates. The
v-dem advisor confirmed each as low-information beyond its kept counterpart. The full
list is in the `redundant_drops` vector in the next chunk.

The motivation is *not* dimensionality reduction for its own sake (a 174-vs-192-feature
UMAP is essentially indistinguishable). The motivation is interpretability: when a
country sits unusually far from its neighbors, we want that to mean something
substantively distinct, not "scored low on women's property rights AND men's property
rights AND access to justice for women AND..." with each gender-paired indicator
counting again toward distance.

## The construction function

```{r build-country-space}
discrete_vars <- c("v2regopploc", "v2regsuploc")

# Near-duplicate indicators dropped per the redundancy analysis. Each tag here was
# identified at |r| >= 0.85 against another indicator that we keep, and confirmed as
# conceptually redundant by the v-dem advisor. Comments give the indicator we keep
# in its place.
redundant_drops <- c(
  # Gender-paired siblings (drop the f/w version, keep the m version)
  "v2cldmovew",   # vs v2cldmovem  — freedom of domestic movement
  "v2cldiscw",    # vs v2cldiscm   — freedom of discussion
  "v2clslavef",   # vs v2clslavem  — freedom from forced labor
  "v2clprptyw",   # vs v2clprptym  — property rights
  "v2clacjstw",   # vs v2clacjstm  — access to justice
  # Court-level pairs (keep general / high court)
  "v2juncind",    # vs v2juhcind   — high court independence subsumes lower court
  "v2juhccomp",   # vs v2jucomp    — general compliance with judiciary subsumes high court
  # Religious-org-specific copies of general CSO measures (keep general)
  "v2csrlgrep",   # vs v2csreprss  — general CSO repression
  "v2csrlgcon",   # vs v2cscnsult  — general CSO consultation
  # Cabinet appoint/dismiss pair (keep appoint)
  "v2exdfdmhs",   # vs v2exdfcbhs  — HOS appoints cabinet in practice
  # Internet vs social-media shutdown twin (keep internet shutdown)
  "v2smgovsm",    # vs v2smgovshut — government internet shutdown in practice
  # Within-family near-duplicates flagged safe by the advisor
  "v2lgqstexp",   # vs v2lginvstp  — legislature investigates is more general
  "v2lglegplo",   # vs v2lgcomslo  — committee system is the structural measure
  "v2merange",    # vs v2mebias    — bias is the more cited media-fairness measure
  "v2dlreason",   # vs v2dlcommon  — common-good framing is more theoretically loaded
  "v2psoppaut",   # vs v2cseeorgs  — CSO entry/exit subsumes opposition autonomy
  "v2cafexch",    # vs v2cafres    — research/teach is the foundational right
  "v2caconmob"    # vs v2cagenmob  — general mass mobilization subsumes concentration
)

# Region labels and lookup based on V-Dem's politico-geographic 6-region scheme.
# A country's region doesn't change over time, so we pin it from its most recent
# observation in V-Dem.
region_labels <- c(
  "1" = "E. Europe & Central Asia",
  "2" = "Latin America & Caribbean",
  "3" = "Middle East & N. Africa",
  "4" = "Sub-Saharan Africa",
  "5" = "W. Europe & N. America",
  "6" = "East Asia & Pacific"
)

region_lookup <- vdem_raw %>%
  group_by(country_name) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(country_name, e_regionpol_6C) %>%
  mutate(region = region_labels[as.character(e_regionpol_6C)]) %>%
  select(country_name, region) %>%
  deframe()

# Per-codebook category labels for indicators (used in Section 6 sensitivity).
prefix_categories <- c(
  "el" = "Elections",            "ps" = "Political Parties",
  "cl" = "Civil Liberties",      "ju" = "Judiciary",
  "lg" = "Legislature",          "me" = "Media",
  "ex" = "Executive",            "dl" = "Deliberation",
  "ca" = "Academic & Cultural",  "cs" = "Civil Society",
  "pe" = "Political Equality",   "st" = "State Capacity",
  "sv" = "Sovereignty",          "dd" = "Direct Democracy",
  "ed" = "Education",            "sm" = "Social Movements & Protest",
  "re" = "Regime",               "te" = "Technology & Media Environment"
)

# Cache the column-name screen — it's expensive and year-invariant.
suffixes        <- c("_codelow", "_codehigh", "_sd", "_osp",
                     "_ord", "_mean", "_nr", "_rec")
all_cols        <- names(vdem_raw)
is_suffix       <- vapply(all_cols, function(x)
                          any(vapply(suffixes, endsWith, logical(1), x = x)),
                          logical(1))
is_numbered     <- grepl("_\\d+$", all_cols)
base_cols       <- all_cols[!is_suffix & !is_numbered]
component_codes <- vdem_codebook %>%
                     filter(vartype == "C") %>% pull(tag) %>% unique()
candidate_codes <- intersect(component_codes, base_cols) %>%
                     setdiff(discrete_vars) %>%
                     setdiff(redundant_drops)
candidate_codes <- candidate_codes[
  vapply(candidate_codes, function(x) is.numeric(vdem_raw[[x]]), logical(1))
]

build_country_space <- function(target_year,
                                min_indicator_coverage = 0.5,
                                max_country_missing    = 0.5,
                                seed = 42,
                                n_neighbors = 15,
                                min_dist = 0.2,
                                n_epochs = 500) {
  year_data <- vdem_raw %>% filter(year == target_year)
  n_total   <- nrow(year_data)

  coverage <- vapply(candidate_codes,
                     function(x) sum(!is.na(year_data[[x]])),
                     integer(1))
  threshold <- min_indicator_coverage * n_total
  indicators <- candidate_codes[coverage >= threshold]

  raw <- year_data %>%
    select(country_name, all_of(indicators)) %>%
    mutate(p_missing = rowSums(is.na(across(all_of(indicators)))) /
                       length(indicators)) %>%
    filter(p_missing < max_country_missing) %>%
    select(-p_missing) %>%
    column_to_rownames("country_name") %>%
    as.matrix()

  # Column-mean imputation. Within-year, within-feature mean is a reasonable
  # neutral fill since the IRT score is approximately mean-centered globally.
  for (j in seq_len(ncol(raw))) {
    na_idx <- is.na(raw[, j])
    if (any(na_idx)) raw[na_idx, j] <- mean(raw[, j], na.rm = TRUE)
  }

  Mz <- scale(raw, center = TRUE, scale = TRUE)
  # Drop any feature with zero within-year variance (would produce NaN after scale).
  Mz <- Mz[, !is.nan(colSums(Mz)), drop = FALSE]

  set.seed(seed)
  emb <- umap(Mz, n_neighbors = n_neighbors,
              min_dist = min_dist, n_epochs = n_epochs)

  coords <- tibble(
    country = rownames(Mz),
    UMAP1   = emb[, 1],
    UMAP2   = emb[, 2],
    region  = unname(region_lookup[country])
  ) %>%
    mutate(region = replace_na(region, "Other"))

  list(
    year         = target_year,
    n_countries  = nrow(Mz),
    n_indicators = ncol(Mz),
    indicators   = colnames(Mz),
    Mz           = Mz,
    coords       = coords
  )
}
```

# 4. Building the matrix

We illustrate on the most recent year (2025) and then run all six.

```{r build-2025}
sp_2025 <- build_country_space(2025)

cat("Year 2025\n")
cat("  Countries:", sp_2025$n_countries, "\n")
cat("  Indicators:", sp_2025$n_indicators, "\n")
cat("  Mean per-feature SD after z-scoring (should be 1):",
    round(mean(apply(sp_2025$Mz, 2, sd)), 3), "\n")
```

The matrix is `r sp_2025$n_countries` countries x `r sp_2025$n_indicators` indicators.
Each row is a country's institutional fingerprint, each column is z-scored across
countries so a one-unit move in any column equals one cross-country standard deviation
in that indicator. Pairwise Euclidean distances on `Mz` are then directly interpretable
as "total cross-country-standardized institutional difference" — large values mean
two countries differ by many SDs across many indicators.

## Why z-score, not min-max?

The earlier exploration in `institutional_complexity.Rmd` used min-max normalization
to rescale every indicator to $[0, 1]$. That works for the EC-style proximity
calculations but distorts Euclidean geometry: a feature with a few extreme tail values
gets squashed into the same $[0, 1]$ range as a feature with broad variation, so the
former contributes much less to inter-country distance than its actual signal warrants.
Z-scoring puts every indicator on the same SD scale and is the right choice when the
downstream operation is Euclidean distance.

## Why mean-impute, not drop?

Coverage at the indicator level is already filtered (>= 50% of countries). At the
country level we drop rows with > 50% missing on the kept indicators. The remaining
NAs (typically a few percent) are imputed at the column mean. This biases the imputed
country toward the global average on the missing dimension, which is the most
defensible neutral fill — it neither helps nor hurts the country's distance to its true
neighbors *on that dimension*, just shrinks variance toward zero on the imputed slots.

# 5. Embedding: UMAP with Euclidean distance

UMAP defaults to Euclidean distance, and we keep that default. We run on the full
z-scored matrix (`Mz`), not on a precomputed distance matrix or PCA truncation. That
keeps the local-neighborhood structure UMAP is designed to preserve in terms of the
metric we actually care about. We use mild hyperparameters
(`n_neighbors = 15`, `min_dist = 0.2`) — small enough to surface clusters, large
enough that the layout is not stringy.

Two caveats worth keeping in mind:

- UMAP layouts are stochastic. We set `set.seed(42)` for reproducibility, but a
  different seed (or a different indicator set on a re-run) can rotate, flip, or
  modestly rearrange the layout. Cluster structure is stable; absolute coordinates
  are not.
- 2D coordinates across years are **not aligned**. Each year is embedded
  independently. Compare structure (which clusters exist, who is in them, how
  isolated they are), not coordinates.

```{r plot-2025, fig.width=7, fig.height=7}
ggplot(sp_2025$coords, aes(UMAP1, UMAP2, color = region)) +
  geom_point(size = 2.4, alpha = 0.75) +
  geom_text_repel(aes(label = country),
                  size = 2.3, max.overlaps = 25,
                  segment.alpha = 0.3,
                  show.legend = FALSE) +
  labs(x = "UMAP 1", y = "UMAP 2",
       title = "Country space, 2025",
       color = NULL) +
  theme(legend.position = "bottom")

ggsave("imgs/country_space_2025.png", width = 9, height = 9, dpi = 300)
```

# 6. Multi-year replication

We run the same pipeline for the six benchmark years and lay them out as small
multiples. The indicator set is re-derived each year from the coverage rule, so
older years use a slightly leaner set (some V-Dem indicators were added in later
codebook waves; others have thinner pre-1990 coverage).

```{r build-all-years, cache=TRUE}
spaces <- map(benchmark_years, build_country_space)
names(spaces) <- as.character(benchmark_years)

space_summary <- map_dfr(spaces, function(s) {
  tibble(year = s$year, n_countries = s$n_countries,
         n_indicators = s$n_indicators)
})
kable(space_summary,
      caption = "Pipeline output by year (countries kept; indicators kept).")
```

```{r save-coords}
# Persist coordinates for downstream use.
all_coords <- map_dfr(spaces, function(s)
  s$coords %>% mutate(year = s$year)) %>%
  select(year, country, region, UMAP1, UMAP2)

write_csv(all_coords, "outputs/country_space_coords_by_year.csv")

# Also save the per-year matrices for any downstream analysis.
saveRDS(spaces, "scratch/country_spaces_by_year.rds")
```

How indicator availability changes over the panel:

```{r indicator-availability, fig.width=6.5, fig.height=3}
indicator_panel <- map_dfr(spaces, function(s)
  tibble(year = s$year, indicator = s$indicators)) %>%
  mutate(
    prefix   = sub("^v[23](\\w{2}).*", "\\1", indicator),
    category = unname(prefix_categories[prefix]),
    category = replace_na(category, "Other")
  )

indicator_panel %>%
  count(year, category) %>%
  mutate(year = factor(year)) %>%
  ggplot(aes(x = year, y = n, fill = category)) +
  geom_col(position = "stack") +
  labs(x = NULL, y = "Indicators kept", fill = "Category",
       title = "Indicator set composition by year")

save_fig("full", "country_space_indicator_panel.pdf")
```

```{r indicator-set-stability}
# How stable is the *set* of indicators across years?
sets <- map(spaces, "indicators")
core <- Reduce(intersect, sets)
union_all <- Reduce(union, sets)

tibble(
  Metric = c("Indicators kept in every year (core)",
             "Indicators kept in at least one year (union)",
             "Core / union ratio"),
  Value  = c(length(core),
             length(union_all),
             sprintf("%.2f", length(core) / length(union_all)))
) %>%
  kable(caption = "Indicator-set overlap across the six years")
```

The core set — indicators that pass the coverage filter in every benchmark year — is
what we'd use if we wanted **strict** cross-year comparability of the input matrix. We
report it but do not enforce it on the layouts: re-running with the year-specific
sets gives each panel the most informative space available for that year, which is
what a domain reader probably wants. Each year's space gets its own page below.

```{r plot-fn}
plot_country_space <- function(sp, label_size = 2.6, point_size = 2.4) {
  ggplot(sp$coords, aes(UMAP1, UMAP2, color = region)) +
    geom_point(size = point_size, alpha = 0.85) +
    geom_text_repel(
      aes(label = country),
      size = label_size, max.overlaps = 40,
      segment.alpha = 0.3, show.legend = FALSE
    ) +
    labs(
      x = "UMAP 1", y = "UMAP 2", color = NULL,
      title = sprintf("Country space, %d", sp$year),
      subtitle = sprintf("%d countries x %d V-Dem indicators (z-scored, UMAP, Euclidean)",
                         sp$n_countries, sp$n_indicators)
    ) +
    theme(legend.position = "bottom")
}

# Persist per-year PNGs alongside the inline plots below.
for (yr in benchmark_years) {
  ggsave(
    sprintf("imgs/country_space_%d.png", yr),
    plot_country_space(spaces[[as.character(yr)]]),
    width = 9, height = 9, dpi = 300
  )
}
```

\newpage

## 1975

```{r plot-1975, fig.width=7, fig.height=8.2, out.width="100%"}
plot_country_space(spaces[["1975"]])
```

\newpage

## 1985

```{r plot-1985, fig.width=7, fig.height=8.2, out.width="100%"}
plot_country_space(spaces[["1985"]])
```

\newpage

## 1995

```{r plot-1995, fig.width=7, fig.height=8.2, out.width="100%"}
plot_country_space(spaces[["1995"]])
```

\newpage

## 2005

```{r plot-2005, fig.width=7, fig.height=8.2, out.width="100%"}
plot_country_space(spaces[["2005"]])
```

\newpage

## 2015

```{r plot-2015, fig.width=7, fig.height=8.2, out.width="100%"}
plot_country_space(spaces[["2015"]])
```

\newpage

## 2025

```{r plot-2025-panel, fig.width=7, fig.height=8.2, out.width="100%"}
plot_country_space(spaces[["2025"]])
```

\newpage

# 7. What the panel shows

A few observations are worth flagging without over-interpreting:

- **Regional clustering is the dominant first-order pattern in every year.** That is
  reassuring: countries that share political-cultural neighborhoods do tend to share
  institutional features. It is also a bit of a Rorschach — the layout reflects
  genuine institutional similarity *and* the V-Dem coding tradition's decision to
  benchmark some indicators against regional norms.

- **The Western Europe & North America cluster is tight in every year.** This is
  the "consolidated liberal democracies" anchor.

- **The post-Soviet space appears suddenly in 1995.** Pre-1991 the USSR is one polity;
  post-1991 it fragments into 15 successor states which then drift across the country
  space at different speeds (the Baltics moving toward the W. European cluster, Central
  Asia moving toward the closed-autocracy region, Russia carving out its own neighborhood).

- **The Middle East & North Africa region is unusually compact in earlier decades and
  starts to spread in 2015-2025.** This is the post-Arab-Spring institutional
  reshuffling visible in the data.

- **East Asia & Pacific is bimodal across the whole panel.** Japan/Korea/Taiwan/Australia
  pull toward the W. European cluster; China and its closer neighbors form a distinct
  pole. The middle (Indonesia, Philippines, India in adjacent regions) shifts year by
  year.

These observations are descriptive. The mapping from country-space neighborhoods to
specific institutional features happens in `institutional_complexity.Rmd` (the
cluster-feature `z`-deviation analysis at the end of Section 6 there) and is not reproduced
here.

# 8. Limitations and what to read next

- **2025 is not the maximum-coverage year.** V-Dem's indicator availability declines as
  we approach the present (242 indicators pass the 50% rule in 2020-2021, 215 in 2022-2023,
  192 in 2024-2025). 2024 has identical coverage to 2025, so swapping to 2024 buys nothing;
  swapping to 2021 would gain ~50 indicators at the cost of four years of currency. We keep
  2025 as the endpoint, but the alternative schedule (1971/81/91/01/11/21) is a one-line
  change to `benchmark_years` if richer coverage matters more than recency.

- **All-continuous-indicators** is a deliberate choice, not the only one. The
  Mahalanobis-whitened variant in `institutional_complexity.Rmd` checks how much the
  feature-family bundling distorts the Euclidean layout (answer: not much in 2025,
  but the comparison is worth re-running for older years before drawing strong
  cross-year conclusions about *structure*).

- **Mean imputation** biases imputed countries toward the global mean *only on the
  missing dimensions*. If a country has 30%-50% missing coverage, the imputation can
  pull it noticeably toward the centroid. For the most recent years this is rare; for
  1975 and 1985 it is more common, especially among newly independent states. The
  per-country missingness distribution is a useful diagnostic to add if a specific
  country's position seems off.

- **Coordinates are not comparable across years.** Treat each panel as its own map.
  A formal "trajectory" analysis — same country across years — would require either
  a joint embedding (concat the matrices, embed once) or a Procrustes-style
  alignment of per-year layouts. Both are reasonable next steps and are sketched in
  `case-studies/cambodia/cambodia-history.Rmd` for a single country, but not
  generalized.

- **Variable selection is the next thing to revisit.** The "all C-type indicators"
  rule is a starting point. A natural follow-up is to check whether dropping the
  top-loading indicator family (e.g., "all elections-related indicators") materially
  changes the layout — that would tell us how much of the country-space is being
  driven by a single conceptual axis vs. genuinely multidimensional institutional
  variation.
