---
title: "Deals and Development: Country Reads Across Three Spaces"
subtitle: "PSW's rents space, deals space, and political settlement, indicator by indicator"
author: "Institutional Complexity Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6.5
    fig_height: 4.5
    latex_engine: xelatex
    includes:
      in_header: ~/dev/gl-design/skills/gl-ggplot/assets/gl_pdf.tex
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  cache = TRUE,
  cache.lazy = FALSE
)
```

# Framing

Pritchett, Sen & Werker (2018, *Deals and Development*, OUP) classify firms
on two axes — target market (export vs. domestic) × source of profit
(regulatory rents vs. market competition) — yielding four stakeholder
types: **magicians** (export, competitive), **workhorses** (domestic,
competitive), **rentiers** (export, regulatory rents), and
**powerbrokers** (domestic, regulatory rents). Each type demands a
different policy bundle along six axes (Table 1.1, p. 22):

|                  | Transp. rules | Common infra | Special. infra | Contract enf. | Subsidies | Exclusivity |
| ---------------- | :-----------: | :----------: | :------------: | :-----------: | :-------: | :---------: |
| **Magicians**    |       x       |       x      |        x       |       x       |     x     |             |
| **Workhorses**   |       x       |       x      |                |       x       |           |             |
| **Rentiers**     |               |              |        x       |       x       |     x     |      x      |
| **Powerbrokers** |               |              |                |               |           |      x      |

PSW's framework has three analytical lenses, not one:

1. **The rents space** — the 4×6 stakeholder × policy-demand table
   above. The orthodox PSW reading uses sectoral GDP shares to locate
   a country's coalition; we instead read it institutionally — for
   each country, which of the demanded policies are actually being
   supplied? — without committing to a single dominant stakeholder.
2. **The deals space** — Pritchett's two axes of how firms transact
   with the state: *open vs closed* (can outsiders make deals or only
   insiders) and *ordered vs disordered* (can you rely on a deal once
   you have it). Mature liberal economies are open/ordered;
   developmental autocracies are closed/ordered; predatory states
   closed/disordered.
3. **The political settlement** — Khan (2010), adopted by PSW: the
   *vertical* power of the ruling coalition over excluded factions
   and the *horizontal* power of the leadership over its own allies.
   The 2×2 yields dominant-party, vulnerable-authoritarian-coalition,
   weak-dominant-party, and competitive-clientelist regimes.

This document builds each of the three reads from V-Dem and QoG
indicators, then renders a handful of country profiles showing the
joint picture. An earlier draft tried to compress the rents-space
read into a single stakeholder-fit score per country; that approach
is archived in `_archive_deals-development_stakeholder-fit.Rmd` and
was discarded as overreaching the data.

```{r packages}
library(tidyverse)
library(arrow)
library(countrycode)
library(ggrepel)
library(knitr)
library(tidytext)  # reorder_within, scale_y_reordered
```

```{r gl-setup, cache=FALSE}
# Both source() and gl_setup() must run uncached so the GL theme applies to
# any newly-rendered chunk on every knit. Sourcing theme_gl.R re-attaches
# showtext / sysfonts / ggthemes (knitr's cache does not replay library()
# side effects). gl_setup() also self-registers a knitr chunk hook that
# re-applies the theme before each chunk runs.
source("~/dev/gl-design/skills/gl-ggplot/assets/theme_gl.R")
gl_setup()
```

# Data

## V-Dem and macro panel

```{r load-vdem, cache=TRUE}
vdem_raw <- read_parquet("data/shared-data/vdem/vdem.parquet")

latest_year <- vdem_raw %>%
  filter(year >= 2020) %>%
  group_by(year) %>%
  summarise(n_countries = n_distinct(country_name), .groups = "drop") %>%
  filter(n_countries >= 150) %>%
  pull(year) %>% max()

cat("Latest V-Dem year:", latest_year, "\n")
```

```{r indicator-map}
indicator_map <- tribble(
  ~axis,                    ~indicator,                ~source, ~sign,
  # Axis 1: transparent rules
  "transparent_rules",      "v2cltrnslw",              "vdem",   1,
  "transparent_rules",      "v2clrspct",               "vdem",   1,
  # Axis 2: common infrastructure
  "common_infra",           "v2dlencmps",              "vdem",   1,
  "common_infra",           "wdi_eg_elc_accs_zs",      "macro",  1,
  "common_infra",           "wdi_sh_h2o_basw_zs",      "macro",  1,
  # Axis 3: specialized infrastructure
  "specialized_infra",      "wdi_lp_lpi_infr_xq",      "macro",  1,
  "specialized_infra",      "wdi_is_shp_gcnw_xq",      "macro",  1,
  "specialized_infra",      "wdi_gb_xpd_rsdv_gd_zs",   "macro",  1,
  "specialized_infra",      "wdi_tx_val_tech_mf_zs",   "macro",  1,
  # Axis 4: contract enforcement
  "contract_enforcement",   "v2xcl_prpty",             "vdem",   1,
  "contract_enforcement",   "v2juncind",               "vdem",   1,
  "contract_enforcement",   "v2juhcind",               "vdem",   1,
  "contract_enforcement",   "v2juhccomp",              "vdem",   1,
  "contract_enforcement",   "v2jucorrdc",              "vdem",   1,
  "contract_enforcement",   "v2juaccnt",               "vdem",   1,
  # Axis 5: subsidies / clientelism (high score = MORE clientelism supplied).
  # v2psprlnks and v2elvotbuy are inputs to v2xnp_client, so we keep only the
  # composite plus the orthogonal v2dlencmps spending-mix indicator.
  "subsidies_clientelism",  "v2xnp_client",            "vdem",   1,
  "subsidies_clientelism",  "v2dlencmps",              "vdem",  -1,  # higher = public goods
  # Axis 6: exclusivity / barriers to entry (high = MORE exclusivity supplied).
  # Market-side only; political-exclusivity V-Dem indicators dropped.
  "exclusivity",            "qog_fi_reg",              "macro", -1,  # higher Fraser = less regulation
  "exclusivity",            "qog_bti_mo",              "macro", -1   # higher = better-organised market
)

axis_levels <- c("transparent_rules", "common_infra", "specialized_infra",
                 "contract_enforcement", "subsidies_clientelism", "exclusivity")
axis_labels <- c("Transparent\nrules", "Common\ninfra", "Specialized\ninfra",
                 "Contract\nenforcement", "Subsidies /\nclientelism", "Exclusivity")
names(axis_labels) <- axis_levels

kable(
  indicator_map %>% count(axis) %>% mutate(axis = factor(axis, levels = axis_levels)) %>%
    arrange(axis),
  col.names = c("Axis", "# indicators"),
  caption = "Indicators per policy-demand axis"
)
```

```{r pull-vdem}
# Indicators used directly in the rents-space axes
vdem_cols <- indicator_map %>% filter(source == "vdem") %>% pull(indicator) %>% unique()

# Additional V-Dem indicators used by the deals-space and political-settlement
# sections. Pulled here so the panel is self-contained.
vdem_extra <- c(
  # Horizontal power (advisor: dispersion across political groups)
  "v2psoppaut",      # opposition parties autonomy
  "v2lgoppart",      # legislature opposition party oversight
  "v2x_jucon",       # judicial constraints on the executive
  # Vertical power (advisor: elite-mass distribution)
  "v2pepwrses",      # political power distributed by SES
  "v2xeg_eqaccess",  # equal access to power across groups
  "v2cseeorgs"       # CSO entry and exit (low gov control = mass voice)
)

vdem_latest <- vdem_raw %>%
  filter(year == latest_year) %>%
  select(country_name, iso3 = country_text_id,
         all_of(c(vdem_cols, vdem_extra)))
```

```{r pull-macro, cache=TRUE}
macro_cols <- indicator_map %>% filter(source == "macro") %>% pull(indicator) %>% unique()

macro <- read_parquet(
  "data/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year", all_of(macro_cols))
)

# For each macro indicator, take the most recent non-NA value within the
# six years ending at latest_year.
window <- seq(latest_year - 5, latest_year)

pick_latest <- function(col) {
  macro %>%
    filter(year %in% window, !is.na(.data[[col]])) %>%
    group_by(countrycodeiso) %>%
    slice_max(year, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(iso3 = countrycodeiso, !!col := .data[[col]])
}

macro_latest <- map(macro_cols, pick_latest) %>%
  reduce(full_join, by = "iso3")
```

```{r combine}
panel <- vdem_latest %>%
  left_join(macro_latest, by = "iso3") %>%
  filter(!is.na(country_name))

cat("Panel:", nrow(panel), "countries ×",
    length(c(vdem_cols, macro_cols)), "indicators\n")
```

# Axis construction

For each of the six policy demands we pick a small basket of V-Dem
and QoG/macro indicators. The construction recipe is the same for every
axis:

1. **Percentile-rank globally.** Each indicator is converted to its
   percentile within the country panel, putting it on `[0, 1]` regardless
   of the original scale (V-Dem ordinal, QoG continuous, WDI in raw
   units, etc.). This avoids letting a single wide-ranged variable
   dominate the average.
2. **Sign-flip where needed.** Some indicators are coded so that higher
   values mean *less* of the demand being supplied (e.g. V-Dem's
   `v2psparban` runs from 0 = "parties banned" up to 4 = "no party ban",
   so high values mean *less* exclusivity). Those get flipped to
   `1 - pct` before averaging.
3. **Mean within axis.** The axis score is the unweighted mean of
   indicator scores. Missing values are dropped from the mean, so a
   country with partial coverage still gets scored on what's available.

The result is a country × 6-axis matrix on `[0, 1]` where higher = the
country supplies more of what that policy demand is asking for.

```{r axis-helpers}
gloss <- c(
  v2cltrnslw            = "Transparent laws with predictable enforcement",
  v2clrspct             = "Rigorous and impartial public administration",
  v2dlencmps            = "Particularistic vs. public-goods spending",
  wdi_eg_elc_accs_zs    = "Electricity access (% pop.)",
  wdi_sh_h2o_basw_zs    = "Basic drinking water access (% pop.)",
  wdi_lp_lpi_infr_xq    = "Logistics Performance Index — infrastructure",
  wdi_is_shp_gcnw_xq    = "Liner Shipping Connectivity Index",
  wdi_gb_xpd_rsdv_gd_zs = "R&D expenditure (% GDP)",
  wdi_tx_val_tech_mf_zs = "High-tech exports (% manufactured exports)",
  v2xcl_prpty           = "Property rights (V-Dem composite of men's and women's property rights)",
  v2juncind             = "Lower court independence",
  v2juhcind             = "High court independence",
  v2juhccomp            = "Government compliance with high-court rulings",
  v2jucorrdc            = "Judicial bribery (V-Dem coding: high = less bribery)",
  v2juaccnt             = "Judicial accountability for misconduct",
  v2xnp_client          = "Clientelism index (V-Dem composite)",
  qog_fi_reg            = "Fraser regulation index (high = less regulation)",
  qog_bti_mo            = "BTI organisation of market and competition"
)

axis_table <- function(ax) {
  indicator_map %>%
    filter(axis == ax) %>%
    transmute(
      Indicator = indicator,
      Source    = ifelse(source == "vdem", "V-Dem", "QoG/macro"),
      Direction = ifelse(sign == 1, "+", "- (flipped)"),
      Gloss     = gloss[indicator]
    ) %>%
    kable()
}
```

## Transparent rules

PSW define this as predictable, publicly known regulation — firms can
plan based on rules they can read, not deals they have to negotiate.
We use two complementary V-Dem indicators that capture the law-on-paper
and the law-in-practice:

- **`v2cltrnslw` — Transparent laws with predictable enforcement.**
  V-Dem question: *"Are the laws of the land clear, well publicized,
  coherent (consistent with each other), relatively stable from year
  to year, and enforced in a predictable manner?"* Ordinal 0–4,
  converted to interval by V-Dem's measurement model: 0 =
  "Transparency and predictability are almost non-existent. The laws
  of the land are created and/or enforced in completely arbitrary
  fashion." through 4 = "Transparency and predictability are very
  strong. The laws of the land are created and enforced in a
  non-arbitrary fashion." This is about the *quality and coherence of
  the legal code itself* — whether the rules are knowable and stable.

- **`v2clrspct` — Rigorous and impartial public administration.**
  V-Dem question: *"Are public officials rigorous and impartial in
  the performance of their duties?"* The codebook clarifies that
  "this question focuses on the extent to which public officials
  generally abide by the law and treat like cases alike, or
  conversely, the extent to which public administration is
  characterized by arbitrariness and biases (i.e., nepotism,
  cronyism, or discrimination)." Ordinal 0–4 (interval after the
  measurement model): 0 = "The law is not respected by public
  officials. Arbitrary or biased administration of the law is
  rampant." through 4 = "The law is generally fully respected by the
  public officials. Arbitrary or biased administration of the law is
  very limited." This captures the *bureaucratic* face of the same
  concept — street-level impartiality.

A country could have clear laws applied corruptly (high `v2cltrnslw`,
low `v2clrspct`) or murky laws applied conscientiously, so both are
needed. Both run positively; no sign-flipping.

```{r ax-tab-transparent}
axis_table("transparent_rules")
```

## Common infrastructure

The PSW demand is for the public goods every firm benefits from —
electricity, water, basic education, transport. We pair one V-Dem
fiscal-intent indicator with two WDI outcome variables that have
near-universal coverage:

- **`v2dlencmps` — Particularistic vs. public-goods spending.**
  V-Dem question: *"Considering the profile of social and
  infrastructural spending in the national budget, how
  'particularistic' or 'public goods' are most expenditures?"* The
  codebook clarifies that "particularistic spending is narrowly
  targeted on a specific corporation, sector, social group, region,
  party, or set of constituents… Public-goods spending is intended
  to benefit all communities within a society." Ordinal 0–4
  (interval after the measurement model), running from 0 = "Almost
  all of the social and infrastructure expenditures are
  particularistic" to 4 = "Almost all social and infrastructure
  expenditures are public-goods in character." Higher = more
  public-goods spending mix; this is the *fiscal intent* side of
  common-infra supply.

- **`wdi_eg_elc_accs_zs`** — share of population with access to
  electricity.

- **`wdi_sh_h2o_basw_zs`** — share of population with access to at
  least basic drinking water.

The two WDI variables capture realised supply of universal
infrastructure with broad country coverage. (Earlier drafts also
included secondary enrolment and PWT human capital, which we drop
here in favour of indicators more directly about physical-network
infrastructure rather than human capital.) All three run positively.

```{r ax-tab-common}
axis_table("common_infra")
```

## Specialized infrastructure

This is the demand for sector-specific public investment that
benefits a subset of firms — special economic zones, port facilities,
R&D infrastructure, regulatory bodies that handle complex industries.
V-Dem doesn't capture this directly, so we lean entirely on
outcome-side proxies. `wdi_lp_lpi_infr_xq` (Logistics Performance
Index — infrastructure) and `wdi_is_shp_gcnw_xq` (Liner Shipping
Connectivity Index) measure trade-enabling specialized
infrastructure; `wdi_gb_xpd_rsdv_gd_zs` (R&D as % GDP) and
`wdi_tx_val_tech_mf_zs` (high-tech export share) proxy R&D
infrastructure and its outputs. We deliberately exclude the Atlas
complexity indices (ECI, COI) here: they are downstream *outcomes*
of specialized infrastructure rather than a measure of the
institutional supply, and including them would risk circularity in
later sanity checks against complexity. All four run positively;
coverage is patchier than the other axes.

```{r ax-tab-specialized}
axis_table("specialized_infra")
```

## Contract enforcement

Courts that resolve commercial disputes predictably and protect
property. V-Dem has the strongest battery here. We checked whether
the property-rights composite already aggregates the judicial
indicators and could replace them: it does not. `v2xcl_prpty` is
built only from `v2clprptym` (men's property rights) and
`v2clprptyw` (women's property rights), with no judicial inputs.
Conversely the judicial battery feeds into `v2x_jucon` (judicial
constraints on the executive) and `v2x_rule`, neither of which we
use. So property rights and the judicial battery are non-redundant
and both are kept.

- **`v2xcl_prpty`** — V-Dem property-rights composite (men's and
  women's de facto rights to acquire, possess, inherit, and sell
  private property).
- **`v2juncind`** and **`v2juhcind`** — independence of lower courts
  and the high court from political pressure.
- **`v2juhccomp`** — government compliance with high-court rulings,
  the most demanding test of judicial enforcement (does the executive
  *actually* abide by rulings it dislikes).
- **`v2jucorrdc`** — judicial bribery; V-Dem's ordinal coding runs
  from "Always" (low) to "Never" (high), so high = less bribery and
  no flip is needed.
- **`v2juaccnt`** — accountability of judges who engage in
  misconduct.

All six run positively.

```{r ax-tab-contract}
axis_table("contract_enforcement")
```

## Subsidies / clientelism

PSW define "subsidies" broadly to include directed credit, tax breaks,
and any state transfer that benefits specific firms or coalitions
rather than the economy at large. The unifying logic is
particularistic distribution.

- **`v2xnp_client` — Clientelism Index.** A V-Dem Bayesian factor
  composite of how much politics is built on the targeted, contingent
  distribution of resources (goods, services, jobs, money) in
  exchange for political support. The codebook flags that this
  composite is built from inputs that include `v2psprlnks` (party
  linkages) and `v2elvotbuy` (vote buying) directly. For that reason
  we drop those two as separate axis components — keeping them
  alongside `v2xnp_client` would double-count their signal. Higher
  score = more clientelism.

- **`v2dlencmps`** (flipped) — the same particularistic-vs-public-goods
  budget-mix variable used in Axis 2. Flipped here so that *higher*
  reflects more particularistic spending. This complements
  `v2xnp_client` by capturing the *fiscal* face of clientelism rather
  than the political-relationships face, and is not an input to the
  composite.

The axis score is high when both political relationships and fiscal
spending flow through targeted, particularistic channels.

```{r ax-tab-subsidies}
axis_table("subsidies_clientelism")
```

## Exclusivity

Barriers to entry, licensing regimes, regulatory rents that protect
incumbents. We narrow this axis to the *market-side* face of
exclusivity, dropping the political-exclusivity V-Dem indicators
(`v2x_neopat`, `v2psparban`, `v2psoppaut`) used in earlier drafts.
That is a deliberate scope decision — the kept indicators are
clean, well-coded, and squarely about market entry, but they cannot
distinguish open-market/closed-politics regimes (e.g. Singapore-type)
from open-market/open-politics regimes. A reader interested in the
political face of exclusivity should consult the V-Dem indicators
separately. Both kept indicators are flipped so that higher = more
exclusivity:

- **`qog_fi_reg`** — Fraser Institute composite of business, credit,
  and labour-market regulation. The Fraser coding has *higher*
  scores for *less* restrictive regulation, so we flip: higher in
  our axis = more restrictive regulation, more exclusivity.
- **`qog_bti_mo`** — Bertelsmann Transformation Index "Organisation
  of the market and competition." Higher BTI = a more competitive,
  open-market environment, so we flip: higher in our axis = a more
  closed/captured market environment.

The axis score is high when the regulatory environment favours
incumbents over entrants.

```{r ax-tab-exclusivity}
axis_table("exclusivity")
```

# Structural variable — resource rents

The rentier and powerbroker stakeholder types both profit from
regulatory rents and demand near-identical policy bundles; what
distinguishes them is whether the rent-seeking sector is export-facing
(rentiers: oil, mining, plantations) or domestic (powerbrokers: telco,
banking, state contracts). The institutional indicators alone cannot
separate them. Following PSW's own use of sector size to make this
distinction, we pull **total natural resource rents as % of GDP**
(WDI `NY.GDP.TOTL.RT.ZS`, summing oil, gas, coal, mineral and forest
rents) and fold it directly into the fit calculation below as a
multiplicative weight on the rentier and powerbroker columns.

```{r resource-rents}
# Resource rents move slowly (it's a structural feature of the economy),
# so we grab the latest available value since 2010 rather than the tight
# window we use for institutional indicators. This recovers Venezuela
# (2014), Eritrea (2011), South Sudan (2015), Syria (2020) and others
# whose WDI series stopped being reported in recent years.
rent_latest <- read_parquet(
  "data/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year", "wdi_ny_gdp_totl_rt_zs")
) %>%
  filter(!is.na(wdi_ny_gdp_totl_rt_zs), year >= 2010) %>%
  group_by(countrycodeiso) %>% slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(iso3 = countrycodeiso,
            resource_rents_pct_gdp = wdi_ny_gdp_totl_rt_zs,
            resource_rents_year    = year)

cat("Countries with resource-rents data:", nrow(rent_latest), "\n")
cat("Year coverage:\n")
print(table(rent_latest$resource_rents_year))
```

# Axis scoring

```{r score-axes}
percentile_rank <- function(x) {
  r <- rank(x, na.last = "keep")
  r / sum(!is.na(x))
}

# Score every indicator
indicator_scores <- panel %>%
  pivot_longer(c(-country_name, -iso3),
               names_to = "indicator", values_to = "value") %>%
  inner_join(indicator_map, by = "indicator") %>%
  group_by(indicator) %>%
  mutate(pct = percentile_rank(value)) %>%
  ungroup() %>%
  mutate(score = ifelse(sign == 1, pct, 1 - pct))

# Average within axis (drop NA so countries with partial coverage still score)
axis_scores <- indicator_scores %>%
  group_by(country_name, iso3, axis) %>%
  summarise(axis_score = mean(score, na.rm = TRUE),
            n_indicators = sum(!is.na(score)),
            .groups = "drop")

# Coverage: keep countries with at least 1 non-NA indicator on every axis
ok_countries <- axis_scores %>%
  group_by(country_name) %>%
  summarise(min_n = min(n_indicators), .groups = "drop") %>%
  filter(min_n >= 1) %>% pull(country_name)

axis_scores <- axis_scores %>% filter(country_name %in% ok_countries)

axis_wide <- axis_scores %>%
  select(country_name, iso3, axis, axis_score) %>%
  pivot_wider(names_from = axis, values_from = axis_score) %>%
  select(country_name, iso3, all_of(axis_levels))

cat("Countries with full axis coverage:", nrow(axis_wide), "\n")
```


```{r focus-set}
# Focus countries used across the three reads and country profiles
focus_iso <- c("USA", "DEU", "CHN", "VNM", "KHM", "BDI",
               "SAU", "BGD", "RWA", "IND", "RUS", "MEX", "PAK")
```

# Rents space — policy-demand satisfaction

The rents-space lens asks, for each policy in PSW's Table 1.1 that a
given stakeholder demands, whether the country is supplying it. We
render this as a per-country 4×6 grid: cells where the stakeholder
demands the policy are coloured by the country's percentile band on
that policy axis (red = bottom third, yellow = middle third, green
= top third); cells where there's no demand are left blank.

Reading down a column tells you which stakeholder coalitions would be
served by the country's strongest policies. Reading across a row
tells you whether a given stakeholder's whole bundle is satisfied —
"lots of green" in the magician row means the institutions look
magician-friendly. Crucially we don't aggregate across rows, so
nothing is hidden behind a single-number score.

```{r psw-prefs-rents}
psw_prefs <- tribble(
  ~stakeholder,
  ~transparent_rules, ~common_infra, ~specialized_infra,
  ~contract_enforcement, ~subsidies_clientelism, ~exclusivity,
  "Magicians",    1, 1, 1, 1, 1, 0,
  "Workhorses",   1, 1, 0, 1, 0, 0,
  "Rentiers",     0, 0, 1, 1, 1, 1,
  "Powerbrokers", 0, 0, 0, 0, 0, 1
)

stakeholder_levels <- c("Magicians", "Workhorses", "Rentiers", "Powerbrokers")

psw_prefs_long <- psw_prefs %>%
  pivot_longer(-stakeholder, names_to = "axis", values_to = "demanded") %>%
  mutate(stakeholder = factor(stakeholder, levels = stakeholder_levels),
         axis = factor(axis, levels = axis_levels))

# Note on the "subsidies / clientelism" axis: high score = MORE clientelism
# supplied. Magicians and Workhorses have demanded = 0 here (they don't want
# clientelism), so those cells will be blank regardless. Rentiers and
# Powerbrokers have demanded = 1, and the colour reads correctly: high
# clientelism = green (their demand is met).
```

```{r rents-grid-fn}
axis_short_labels <- c(
  transparent_rules     = "Transp. rules",
  common_infra          = "Common infra",
  specialized_infra     = "Special. infra",
  contract_enforcement  = "Contract enf.",
  subsidies_clientelism = "Subsidies",
  exclusivity           = "Exclusivity"
)

make_rents_grid <- function(iso_set) {
  axis_long <- axis_wide %>%
    filter(iso3 %in% iso_set) %>%
    pivot_longer(-c(country_name, iso3),
                 names_to = "axis", values_to = "axis_score") %>%
    mutate(axis = factor(axis, levels = axis_levels))

  iso_to_country <- axis_wide %>% filter(iso3 %in% iso_set) %>%
    select(country_name, iso3)

  grid <- expand_grid(
    iso3 = iso_set,
    stakeholder = factor(stakeholder_levels, levels = stakeholder_levels),
    axis = factor(axis_levels, levels = axis_levels)
  ) %>%
    inner_join(psw_prefs_long, by = c("stakeholder", "axis")) %>%
    left_join(axis_long %>% select(iso3, axis, axis_score),
              by = c("iso3", "axis")) %>%
    inner_join(iso_to_country, by = "iso3") %>%
    mutate(
      band = case_when(
        demanded == 0       ~ NA_character_,
        is.na(axis_score)   ~ "Missing",
        axis_score < 1/3    ~ "Bottom third",
        axis_score < 2/3    ~ "Middle third",
        TRUE                ~ "Top third"
      ),
      band = factor(band,
                    levels = c("Bottom third", "Middle third", "Top third", "Missing")),
      country_name = factor(country_name,
                            levels = iso_to_country$country_name[
                              match(iso_set, iso_to_country$iso3)
                            ])
    )

  grid
}

rents_palette <- c(
  "Bottom third" = "#d73027",
  "Middle third" = "#fee08b",
  "Top third"    = "#1a9850",
  "Missing"      = "grey60"
)
```

```{r rents-grid-focus, fig.width=6.5, fig.height=6}
focus_grid <- make_rents_grid(focus_iso)

ggplot(focus_grid,
       aes(x = axis, y = fct_rev(stakeholder), fill = band)) +
  geom_tile(color = "white", linewidth = 0.6, na.rm = FALSE) +
  facet_wrap(~ country_name, ncol = 3) +
  scale_fill_manual(values = rents_palette, na.value = NA, drop = FALSE) +
  scale_x_discrete(labels = axis_short_labels) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Policy-demand satisfaction matrix by country",
       subtitle = "Coloured = stakeholder demands this policy; blank = no demand") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 7, angle = 45, hjust = 1, vjust = 1),
        axis.text.y = element_text(size = 8),
        strip.text  = element_text(size = 9, face = "bold"))
```

# Deals space — partial read

PSW's deals space asks two questions:

- **Closed vs open**: are state-firm deals available based on the
  firm's *identity* (you need to be in the regime club, ethnically
  tied, an SOE, politically connected) or based on its *actions*
  (any firm meeting criteria can transact)?
- **Ordered vs disordered**: will a deal be honoured regardless of
  subsequent political shifts, or only as long as it remains in the
  short-term political interest of the elite?

We are honest that **V-Dem cannot measure these well**. The closed/
open axis has weak proxies; the ordered/disordered axis is essentially
unreachable. We report what we have on openness with explicit
caveats and *omit* the order axis entirely rather than ship a
formal-rule-of-law composite that would be misread as "deals stick
through political shifts."

## Openness — what we can (partially) say

We approximate the action-vs-identity distinction with two
indicators, both reverse-coded where needed so high = more
action-based / open:

- **`v2clrspct`** — *"Are public officials rigorous and impartial in
  the performance of their duties?"* Codebook: *"the extent to which
  public officials generally abide by the law and treat like cases
  alike, or conversely, the extent to which public administration is
  characterized by arbitrariness and biases (i.e., nepotism, cronyism,
  or discrimination)."* Direct measure of bureaucratic identity-vs-
  action access.
- **`v2dlencmps`** — particularistic vs public-goods spending mix
  (high = public-goods). The fiscal expression: when state allocations
  flow through identity-targeted channels (specific firms, regions,
  cronies), deals are closed; when they flow through universal rules,
  they are open.

What's *not* in the composite, and why:

- *Political-openness indicators* (opposition autonomy, party ban,
  electoral democracy) measure who can run for office, not who can
  do business with the state. They belong to the political-settlement
  read, not to deals.
- *Petty corruption* (`v2excrptps`) is conceptually ambiguous: high
  petty corruption means *anyone with money* can transact, which is
  closer to PSW's "open and disordered" pole than to "closed". We
  drop it rather than commit to a sign.

```{r openness-axes}
openness_indicator_map <- tribble(
  ~indicator,    ~source, ~sign,
  "v2clrspct",   "vdem",  1,    # impartial admin (high = action-based)
  "v2dlencmps",  "vdem",  1     # public-goods spending (high = open)
)

openness_panel <- panel %>%
  pivot_longer(c(-country_name, -iso3),
               names_to = "indicator", values_to = "value") %>%
  inner_join(openness_indicator_map, by = "indicator") %>%
  group_by(indicator) %>%
  mutate(pct = percentile_rank(value)) %>%
  ungroup() %>%
  mutate(score = ifelse(sign == 1, pct, 1 - pct)) %>%
  group_by(country_name, iso3) %>%
  summarise(openness = mean(score, na.rm = TRUE), .groups = "drop")
```

```{r openness-table}
openness_indicator_map %>%
  transmute(
    Indicator = indicator,
    Direction = ifelse(sign == 1, "+", "- (flipped)"),
    Gloss     = c(
      v2clrspct  = "Rigorous and impartial public administration",
      v2dlencmps = "Particularistic vs public-goods spending (high = public goods)"
    )[indicator]
  ) %>%
  kable(caption = "Indicators in the openness composite")
```

```{r openness-ranking, fig.width=6.5, fig.height=5}
openness_focus <- openness_panel %>%
  filter(iso3 %in% focus_iso) %>%
  mutate(country_label = sprintf("%s (%s)", country_name, iso3)) %>%
  arrange(openness) %>%
  mutate(country_label = factor(country_label, levels = country_label))

ggplot(openness_focus, aes(x = openness, y = country_label)) +
  geom_segment(aes(x = 0, xend = openness, yend = country_label),
               color = "grey80") +
  geom_point(size = 3, color = highlight) +
  scale_x_continuous(limits = c(0, 1),
                     labels = scales::percent_format(accuracy = 1)) +
  labs(x = "Openness composite (percentile, higher = more action-based access)",
       y = NULL,
       title = "Openness — focus countries",
       subtitle = "Mean of v2clrspct and v2dlencmps percentile ranks")
```

## Order — what we cannot say

PSW's *ordered* deal is one that survives political shifts: even if
the government changes, even if it becomes inconvenient, the deal
sticks. V-Dem has nothing that directly measures inter-temporal
durability of state-firm commitments. The closest proxies —
`v2juhccomp` (executive compliance with court rulings), `v2juhcind`
(high court independence), `v2exrescon` (executive respects
constitution) — capture *current* executive discipline, not the
"deals survive a change in government" idea.

We deliberately omit an order composite from this section. A reader
interested in the ordered/disordered dimension should consult
contract-cancellation data (e.g. ICSID arbitration filings) or PRS
ICRG's "contract viability" sub-index — both outside V-Dem.

# Political settlement — vertical × horizontal power

The political-settlement framework asks two distinct questions about
how political power is distributed:

- **Horizontal power** — concentration of power *among political
  groups*. High = ONE dominant political group has institutional
  control and can resist any challenge to it. Low = power is
  dispersed across several political groups jockeying for control.
- **Vertical power** — distribution of power between *elites and
  non-elites*. High = elites monopolise political power, mass voice
  is negligible. Low = power is more evenly distributed across
  socioeconomic strata.

The 2×2 produces four types: **dominant party** (high horizontal,
high vertical), **vulnerable authoritarian coalition** (high
horizontal, low vertical), **weak dominant party** (low horizontal,
high vertical), and **competitive clientelist** (low/low).

V-Dem operationalises both concepts cleanly (much more so than the
deals space). We use three indicators per axis, all reverse-coded so
high score = concentration:

```{r settlement-axes}
settlement_indicator_map <- tribble(
  ~settle_axis,    ~indicator,       ~source, ~sign,
  # Horizontal: concentration among political groups
  "horizontal",    "v2psoppaut",     "vdem", -1,  # high opp aut = dispersed
  "horizontal",    "v2lgoppart",     "vdem", -1,  # high oversight = dispersed
  "horizontal",    "v2x_jucon",      "vdem", -1,  # high jucon = constrained = dispersed
  # Vertical: elite-mass concentration
  "vertical",      "v2pepwrses",     "vdem", -1,  # high SES equality = mass voice
  "vertical",      "v2xeg_eqaccess", "vdem", -1,  # high equal access = mass voice
  "vertical",      "v2cseeorgs",     "vdem", -1   # high CSO autonomy = mass voice
)

settle_panel <- panel %>%
  pivot_longer(c(-country_name, -iso3),
               names_to = "indicator", values_to = "value") %>%
  inner_join(settlement_indicator_map, by = "indicator") %>%
  group_by(indicator) %>%
  mutate(pct = percentile_rank(value)) %>%
  ungroup() %>%
  mutate(score = ifelse(sign == 1, pct, 1 - pct)) %>%
  group_by(country_name, iso3, settle_axis) %>%
  summarise(score = mean(score, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = settle_axis, values_from = score)
```

```{r settlement-table}
settlement_indicator_map %>%
  transmute(
    Axis      = ifelse(settle_axis == "vertical",
                       "Vertical (elite-mass)",
                       "Horizontal (group concentration)"),
    Indicator = indicator,
    Direction = ifelse(sign == 1, "+", "- (flipped)"),
    Gloss     = c(
      v2psoppaut     = "Opposition parties autonomy (high = dispersed)",
      v2lgoppart     = "Legislature opposition party oversight (high = dispersed)",
      v2x_jucon      = "Judicial constraints on the executive (high = constrained)",
      v2pepwrses     = "Power distributed by socioeconomic position (high = equal)",
      v2xeg_eqaccess = "Equal access to power across groups (high = equal)",
      v2cseeorgs     = "CSO entry and exit (high = autonomous CSOs)"
    )[indicator]
  ) %>%
  kable(caption = "Indicators by political-settlement axis. All flipped so high = concentration of power.")
```

```{r settlement-2x2, fig.width=6.5, fig.height=5}
settle_focus <- settle_panel %>% mutate(focus = iso3 %in% focus_iso)

ggplot(settle_panel, aes(x = horizontal, y = vertical)) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.3) +
  annotate("text", x = 0.05, y = 0.97, hjust = 0, vjust = 1,
           label = "Weak\ndominant\nparty", color = "grey40", size = 3, lineheight = 0.9) +
  annotate("text", x = 0.95, y = 0.97, hjust = 1, vjust = 1,
           label = "Dominant\nparty", color = "grey40", size = 3, lineheight = 0.9) +
  annotate("text", x = 0.05, y = 0.03, hjust = 0, vjust = 0,
           label = "Competitive\nclientelist", color = "grey40", size = 3, lineheight = 0.9) +
  annotate("text", x = 0.95, y = 0.03, hjust = 1, vjust = 0,
           label = "Vulnerable\nauthoritarian\ncoalition", color = "grey40",
           size = 3, lineheight = 0.9) +
  geom_point(alpha = 0.35) +
  geom_point(data = filter(settle_focus, focus),
             color = highlight, size = 2.4) +
  geom_text_repel(data = filter(settle_focus, focus),
                  aes(label = iso3),
                  size = 3, color = highlight, max.overlaps = 30) +
  scale_x_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
  labs(x = "Horizontal: concentration among political groups (high = one dominant)",
       y = "Vertical: elite-mass concentration (high = elite monopoly)",
       title = "Political settlement — horizontal × vertical",
       subtitle = "Horizontal: dispersion of authority across political groups. Vertical: elite vs mass voice.")
```

## Settlement — horizontal × managed-vs-contested (PC2 overlay)

The hand-picked vertical axis above measures elite-mass concentration,
which is one defensible reading of "political settlement". A
data-driven alternative comes from PC2 of the institutional-
complexity PCA (computed in `institutional_complexity.Rmd` over the
full 185-feature V-Dem matrix). PC2's top loadings — anti-system CSO
movements, political violence, mass mobilization, antagonistic
political camps, elite-organised offline action — read as a
**managed vs contested** axis: low PC2 = political conflict is
absorbed by institutions (Germany, Singapore, China, Saudi Arabia
all cluster here); high PC2 = political conflict is mobilized in the
streets (Lebanon, Pakistan, Brazil, Bangladesh, Mexico).

This axis maps onto Khan's typology more cleanly than the elite-mass
axis: dominant-party regimes are concentrated *and* managed;
vulnerable authoritarian coalitions are concentrated *but* contested
(Pakistan); competitive clientelism is dispersed *and* contested
(Lebanon); consolidated democracies are dispersed *and* managed
(Germany). The cost is that the elite-mass dimension drops out — we
keep `v2pepwrses` etc. in the country-profile table for context.

```{r pc2-load}
pca_scores <- read_csv("outputs/institutional_pca_scores.csv",
                       show_col_types = FALSE) %>%
  select(country_name, PC2, PC2_pct)

settle_with_pc2 <- settle_panel %>%
  left_join(pca_scores, by = "country_name") %>%
  mutate(contested = PC2_pct / 100)  # rescale to [0,1]

cat("Countries with PC2:", sum(!is.na(settle_with_pc2$PC2)),
    " / ", nrow(settle_with_pc2), "\n")
```

```{r settlement-2x2-pc2, fig.width=6.5, fig.height=5}
settle_pc2_focus <- settle_with_pc2 %>%
  mutate(focus = iso3 %in% focus_iso) %>%
  filter(!is.na(contested))

ggplot(settle_pc2_focus, aes(x = horizontal, y = contested)) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.3) +
  annotate("text", x = 0.05, y = 0.97, hjust = 0, vjust = 1,
           label = "Competitive\nclientelist",
           color = "grey40", size = 3, lineheight = 0.9) +
  annotate("text", x = 0.95, y = 0.97, hjust = 1, vjust = 1,
           label = "Vulnerable\nauthoritarian\ncoalition",
           color = "grey40", size = 3, lineheight = 0.9) +
  annotate("text", x = 0.05, y = 0.03, hjust = 0, vjust = 0,
           label = "Consolidated\ndemocracy",
           color = "grey40", size = 3, lineheight = 0.9) +
  annotate("text", x = 0.95, y = 0.03, hjust = 1, vjust = 0,
           label = "Dominant\nparty",
           color = "grey40", size = 3, lineheight = 0.9) +
  geom_point(alpha = 0.35) +
  geom_point(data = filter(settle_pc2_focus, focus),
             color = highlight, size = 2.4) +
  geom_text_repel(data = filter(settle_pc2_focus, focus),
                  aes(label = iso3),
                  size = 3, color = highlight, max.overlaps = 30) +
  scale_x_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
  labs(x = "Horizontal: concentration among political groups (high = one dominant)",
       y = "Contested State (high = mobilized / chaotic politics)",
       title = "Political settlement — horizontal × contested (PC2)",
       subtitle = "PC2: high = anti-system mobilization, polarization, political violence")
```

# Country profiles

The three reads brought together for each focus country: where it
sits on the deals 2×2, where it sits on the political-settlement 2×2,
its rents-space row reads (which stakeholder bundles are best
satisfied), and its resource-rents share for context.

```{r country-profiles-data}
country_profile <- bind_rows(
  openness_panel %>%
    transmute(country_name, iso3, metric = "openness", score = openness) %>%
    mutate(space = "Deals"),
  settle_panel %>%
    pivot_longer(c(vertical, horizontal),
                 names_to = "metric", values_to = "score") %>%
    mutate(space = "Settlement"),
  settle_with_pc2 %>%
    transmute(country_name, iso3, metric = "contested",
              score = contested, space = "Settlement")
) %>%
  mutate(score = round(score, 2))

# Stakeholder satisfaction: average percentile across each row of the rents
# grid (i.e. mean axis_score over demanded cells per stakeholder).
stake_satisfaction <- axis_wide %>%
  pivot_longer(-c(country_name, iso3),
               names_to = "axis", values_to = "axis_score") %>%
  inner_join(psw_prefs_long, by = "axis") %>%
  filter(demanded == 1) %>%
  group_by(country_name, iso3, stakeholder) %>%
  summarise(satisfaction = mean(axis_score, na.rm = TRUE), .groups = "drop")
```

```{r country-profiles-table}
deals_focus_long <- country_profile %>%
  filter(iso3 %in% focus_iso) %>%
  unite("col", space, metric, sep = " — ") %>%
  pivot_wider(names_from = col, values_from = score)

stake_focus_wide <- stake_satisfaction %>%
  filter(iso3 %in% focus_iso) %>%
  mutate(satisfaction = round(satisfaction, 2)) %>%
  pivot_wider(names_from = stakeholder, values_from = satisfaction)

profile_table <- deals_focus_long %>%
  left_join(stake_focus_wide, by = c("country_name", "iso3")) %>%
  mutate(country_name = factor(country_name,
                               levels = axis_wide$country_name[
                                 match(focus_iso, axis_wide$iso3)
                               ])) %>%
  arrange(country_name) %>%
  select(country_name, iso3,
         `Deals — openness`,
         `Settlement — horizontal`, `Settlement — vertical`,
         `Settlement — contested`,
         Magicians, Workhorses, Rentiers, Powerbrokers)

kable(profile_table,
      digits = 2,
      caption = "Country profile table — three reads side by side. Stakeholder columns are mean of percentile scores across that stakeholder's demanded axes (higher = more of the demanded bundle is supplied).")
```

# Outputs

```{r write-outputs}
dir.create("outputs", showWarnings = FALSE)

write_csv(axis_wide, "outputs/deals_axis_scores.csv")
write_csv(openness_panel, "outputs/deals_openness_scores.csv")
write_csv(settle_panel, "outputs/settlement_scores.csv")
write_csv(stake_satisfaction, "outputs/rents_stakeholder_satisfaction.csv")

cat("Wrote: outputs/deals_axis_scores.csv (",  nrow(axis_wide),  " rows)\n", sep = "")
cat("Wrote: outputs/deals_openness_scores.csv (", nrow(openness_panel), " rows)\n", sep = "")
cat("Wrote: outputs/settlement_scores.csv (",  nrow(settle_panel), " rows)\n", sep = "")
cat("Wrote: outputs/rents_stakeholder_satisfaction.csv (",
    nrow(stake_satisfaction), " rows)\n", sep = "")
```

# Caveats and next steps

- **Three reads, not one composite.** We deliberately don't compress
  these into a single country score. PSW's framework treats them as
  complementary lenses — a country can look magician-friendly on rents
  but competitive-clientelist on settlement (Bangladesh) — and the
  policy implications differ.
- **Political-settlement indicators are proxies.** Khan's vertical /
  horizontal power isn't directly measured by V-Dem; we approximate
  with electoral openness and personalism composites. Country reads
  near the diagonals (vulnerable-authoritarian vs dominant-party,
  competitive-clientelist vs weak-dominant-party) should be treated
  as suggestive rather than authoritative.
- **The rents-space grid is a *supply* read, not a *power* read.**
  A country whose institutions look magician-friendly is one where
  magicians *would prosper*, not necessarily one where magicians
  hold political power. The book's reading of "who actually wins"
  requires sectoral data that isn't in this panel.
- **No infrastructure-stock data in V-Dem.** Common-infra and
  specialized-infra rely on WDI access/quality measures, which have
  more measurement noise than the V-Dem axes.
- **Resource-rents context.** Rents data is point-in-time per country
  (latest non-NA since 2010). Eritrea (2011) and Venezuela (2014)
  series stop being reported in recent years; reads on those use
  pre-collapse levels.
