---
title: "Deals and Development: Stakeholder Profiles from Institutional Data"
subtitle: "Mapping V-Dem and QoG indicators to the Pritchett-Sen-Werker typology"
author: "Institutional Complexity Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6.5
    fig_height: 4.5
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  cache = TRUE,
  cache.lazy = FALSE
)
```

# Framing

Pritchett, Sen & Werker (2018, *Deals and Development*, OUP) classify firms
on two axes — target market (export vs. domestic) × source of profit
(regulatory rents vs. market competition) — yielding four stakeholder
types: **magicians** (export, competitive), **workhorses** (domestic,
competitive), **rentiers** (export, regulatory rents), and
**powerbrokers** (domestic, regulatory rents). Each type demands a
different policy bundle along six axes (Table 1.1, p. 22):

|                  | Transp. rules | Common infra | Special. infra | Contract enf. | Subsidies | Exclusivity |
| ---------------- | :-----------: | :----------: | :------------: | :-----------: | :-------: | :---------: |
| **Magicians**    |       x       |       x      |        x       |       x       |     x     |             |
| **Workhorses**   |       x       |       x      |                |       x       |           |             |
| **Rentiers**     |               |              |        x       |       x       |     x     |      x      |
| **Powerbrokers** |               |              |                |               |           |      x      |

We treat realised institutional outcomes as **revealed bargaining power**:
a country whose institutions strongly supply the policy bundle a
stakeholder demands is one where that stakeholder has been winning. The
deliverable is a country-by-stakeholder fit score, interpreted as a
qualitative read on which coalition has shaped each country's
institutional landscape.

```{r packages}
library(tidyverse)
library(arrow)
library(countrycode)
library(ggrepel)
library(knitr)
library(tidytext)  # reorder_within, scale_y_reordered

source("~/dev/gl-design/skills/gl-ggplot/assets/theme_gl.R")
gl_setup()
```

# Data

## V-Dem and macro panel

```{r load-vdem, cache=TRUE}
vdem_raw <- read_parquet("~/dev/shared-data/vdem/vdem.parquet")

latest_year <- vdem_raw %>%
  filter(year >= 2020) %>%
  group_by(year) %>%
  summarise(n_countries = n_distinct(country_name), .groups = "drop") %>%
  filter(n_countries >= 150) %>%
  pull(year) %>% max()

cat("Latest V-Dem year:", latest_year, "\n")
```

```{r indicator-map}
indicator_map <- tribble(
  ~axis,                    ~indicator,                ~source, ~sign,
  # Axis 1: transparent rules
  "transparent_rules",      "v2cltrnslw",              "vdem",   1,
  "transparent_rules",      "v2x_rule",                "vdem",   1,
  "transparent_rules",      "v2clrspct",               "vdem",   1,
  "transparent_rules",      "wgi_rle",                 "macro",  1,
  "transparent_rules",      "wgi_rqe",                 "macro",  1,
  # Axis 2: common infrastructure
  "common_infra",           "v2dlencmps",              "vdem",   1,
  "common_infra",           "wdi_eg_elc_accs_zs",      "macro",  1,
  "common_infra",           "wdi_sh_h2o_basw_zs",      "macro",  1,
  "common_infra",           "wdi_se_sec_enrr",         "macro",  1,
  "common_infra",           "pwt_hc",                  "macro",  1,
  # Axis 3: specialized infrastructure
  "specialized_infra",      "wdi_lp_lpi_infr_xq",      "macro",  1,
  "specialized_infra",      "wdi_is_shp_gcnw_xq",      "macro",  1,
  "specialized_infra",      "wdi_gb_xpd_rsdv_gd_zs",   "macro",  1,
  "specialized_infra",      "wdi_tx_val_tech_mf_zs",   "macro",  1,
  "specialized_infra",      "atl_sitc_eci",            "macro",  1,
  "specialized_infra",      "atl_sitc_coi",            "macro",  1,
  # Axis 4: contract enforcement
  "contract_enforcement",   "v2xcl_prpty",             "vdem",   1,
  "contract_enforcement",   "v2juncind",               "vdem",   1,
  "contract_enforcement",   "v2juhcind",               "vdem",   1,
  "contract_enforcement",   "v2juhccomp",              "vdem",   1,
  "contract_enforcement",   "v2jucorrdc",              "vdem",   1,
  "contract_enforcement",   "v2juaccnt",               "vdem",   1,
  # Axis 5: subsidies / clientelism (high score = MORE clientelism supplied)
  "subsidies_clientelism",  "v2xnp_client",            "vdem",   1,
  "subsidies_clientelism",  "v2dlencmps",              "vdem",  -1,  # higher = public goods
  "subsidies_clientelism",  "v2psprlnks",              "vdem",  -1,  # higher = programmatic
  "subsidies_clientelism",  "v2elvotbuy",              "vdem",  -1,  # higher = less vote buying
  # Axis 6: exclusivity / barriers to entry (high = MORE exclusivity supplied)
  "exclusivity",            "v2x_neopat",              "vdem",   1,
  "exclusivity",            "v2psparban",              "vdem",  -1,  # higher = no party ban
  "exclusivity",            "v2psoppaut",              "vdem",  -1,  # higher = more opp autonomy
  "exclusivity",            "qog_fi_reg",              "macro", -1,  # higher Fraser = less regulation
  "exclusivity",            "qog_bti_mo",              "macro", -1   # higher = better-organised market
)

axis_levels <- c("transparent_rules", "common_infra", "specialized_infra",
                 "contract_enforcement", "subsidies_clientelism", "exclusivity")
axis_labels <- c("Transparent\nrules", "Common\ninfra", "Specialized\ninfra",
                 "Contract\nenforcement", "Subsidies /\nclientelism", "Exclusivity")
names(axis_labels) <- axis_levels

kable(
  indicator_map %>% count(axis) %>% mutate(axis = factor(axis, levels = axis_levels)) %>%
    arrange(axis),
  col.names = c("Axis", "# indicators"),
  caption = "Indicators per policy-demand axis"
)
```

```{r pull-vdem}
vdem_cols <- indicator_map %>% filter(source == "vdem") %>% pull(indicator) %>% unique()

vdem_latest <- vdem_raw %>%
  filter(year == latest_year) %>%
  select(country_name, iso3 = country_text_id, all_of(vdem_cols))
```

```{r pull-macro, cache=TRUE}
macro_cols <- indicator_map %>% filter(source == "macro") %>% pull(indicator) %>% unique()

macro <- read_parquet(
  "~/dev/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year", all_of(macro_cols))
)

# For each macro indicator, take the most recent non-NA value within the
# six years ending at latest_year.
window <- seq(latest_year - 5, latest_year)

pick_latest <- function(col) {
  macro %>%
    filter(year %in% window, !is.na(.data[[col]])) %>%
    group_by(countrycodeiso) %>%
    slice_max(year, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(iso3 = countrycodeiso, !!col := .data[[col]])
}

macro_latest <- map(macro_cols, pick_latest) %>%
  reduce(full_join, by = "iso3")
```

```{r combine}
panel <- vdem_latest %>%
  left_join(macro_latest, by = "iso3") %>%
  filter(!is.na(country_name))

cat("Panel:", nrow(panel), "countries ×",
    length(c(vdem_cols, macro_cols)), "indicators\n")
```

# Axis construction

For each of the six policy demands we pick a small basket of V-Dem
and QoG/macro indicators. The construction recipe is the same for every
axis:

1. **Percentile-rank globally.** Each indicator is converted to its
   percentile within the country panel, putting it on `[0, 1]` regardless
   of the original scale (V-Dem ordinal, QoG continuous, WDI in raw
   units, etc.). This avoids letting a single wide-ranged variable
   dominate the average.
2. **Sign-flip where needed.** Some indicators are coded so that higher
   values mean *less* of the demand being supplied (e.g. V-Dem's
   `v2psparban` runs from 0 = "parties banned" up to 4 = "no party ban",
   so high values mean *less* exclusivity). Those get flipped to
   `1 - pct` before averaging.
3. **Mean within axis.** The axis score is the unweighted mean of
   indicator scores. Missing values are dropped from the mean, so a
   country with partial coverage still gets scored on what's available.

The result is a country × 6-axis matrix on `[0, 1]` where higher = the
country supplies more of what that policy demand is asking for.

```{r axis-helpers}
gloss <- c(
  v2cltrnslw            = "Transparent laws with predictable enforcement",
  v2x_rule              = "Rule of law (V-Dem composite)",
  v2clrspct             = "Rigorous and impartial public administration",
  wgi_rle               = "WGI Rule of Law",
  wgi_rqe               = "WGI Regulatory Quality",
  v2dlencmps            = "Particularistic vs. public-goods spending",
  wdi_eg_elc_accs_zs    = "Electricity access (% pop.)",
  wdi_sh_h2o_basw_zs    = "Basic drinking water access (% pop.)",
  wdi_se_sec_enrr       = "Secondary school enrolment (gross %)",
  pwt_hc                = "Penn World Tables human-capital index",
  wdi_lp_lpi_infr_xq    = "Logistics Performance Index — infrastructure",
  wdi_is_shp_gcnw_xq    = "Liner Shipping Connectivity Index",
  wdi_gb_xpd_rsdv_gd_zs = "R&D expenditure (% GDP)",
  wdi_tx_val_tech_mf_zs = "High-tech exports (% manufactured exports)",
  atl_sitc_eci          = "Economic Complexity Index",
  atl_sitc_coi          = "Complexity Outlook Index",
  v2xcl_prpty           = "Property rights (V-Dem composite)",
  v2juncind             = "Lower court independence",
  v2juhcind             = "High court independence",
  v2juhccomp            = "Government compliance with high-court rulings",
  v2jucorrdc            = "Judicial bribery (V-Dem coding: high = less bribery)",
  v2juaccnt             = "Judicial accountability for misconduct",
  v2xnp_client          = "Clientelism index",
  v2psprlnks            = "Party linkages (clientelistic to programmatic)",
  v2elvotbuy            = "Vote buying (V-Dem coding: high = less)",
  v2x_neopat            = "Neopatrimonial Rule index",
  v2psparban            = "Party ban (V-Dem coding: high = no ban)",
  v2psoppaut            = "Opposition party autonomy",
  qog_fi_reg            = "Fraser regulation index (high = less regulation)",
  qog_bti_mo            = "BTI organisation of market and competition"
)

axis_table <- function(ax) {
  indicator_map %>%
    filter(axis == ax) %>%
    transmute(
      Indicator = indicator,
      Source    = ifelse(source == "vdem", "V-Dem", "QoG/macro"),
      Direction = ifelse(sign == 1, "+", "- (flipped)"),
      Gloss     = gloss[indicator]
    ) %>%
    kable()
}
```

## Transparent rules

PSW define this as predictable, publicly known regulation — firms can
plan based on rules they can read, not deals they have to negotiate.
V-Dem's `v2cltrnslw` is a near-verbatim match. We add `v2x_rule` (the
V-Dem rule-of-law composite) for breadth, and `v2clrspct` to capture
the *bureaucratic* face of transparency: do public officials apply
rules even-handedly. From QoG we cross-reference with the World Bank
governance indicators `wgi_rle` (Rule of Law) and `wgi_rqe` (Regulatory
Quality), which are noisier but the most widely cited benchmarks. All
five run the same direction (higher = more transparency) so no
sign-flipping is needed.

```{r ax-tab-transparent}
axis_table("transparent_rules")
```

## Common infrastructure

The PSW demand is for the public goods every firm benefits from —
electricity, water, basic education, transport. V-Dem doesn't measure
infrastructure stocks, but `v2dlencmps` (particularistic vs.
public-goods budget mix) is a fiscal proxy for whether the state
*spends* on broadly shared infrastructure rather than targeted
patronage. The remaining four indicators are outcome-side WDI/PWT
measures: electricity access, basic water access, secondary
enrolment, and the Penn World Tables human-capital index. Together
they capture both spending intent and realised supply. All five
run positively.

```{r ax-tab-common}
axis_table("common_infra")
```

## Specialized infrastructure

This is the demand for sector-specific public investment that
benefits a subset of firms — special economic zones, port facilities,
R&D infrastructure, regulatory bodies that handle complex industries.
V-Dem doesn't capture this directly, so we lean entirely on
outcome-side proxies. `wdi_lp_lpi_infr_xq` (Logistics Performance
Index — infrastructure) and `wdi_is_shp_gcnw_xq` (port connectivity)
measure trade-enabling specialized infrastructure;
`wdi_gb_xpd_rsdv_gd_zs` (R&D as % GDP) and `wdi_tx_val_tech_mf_zs`
(high-tech export share) proxy R&D infrastructure and its outputs;
`atl_sitc_eci` and `atl_sitc_coi` measure economic complexity itself
— diversified, sophisticated economies are the ones for which
specialized institutional support has actually been built. All
positively oriented.

```{r ax-tab-specialized}
axis_table("specialized_infra")
```

## Contract enforcement

Courts that resolve commercial disputes predictably and protect
property. V-Dem has the strongest battery here. `v2xcl_prpty` is the
property-rights composite. `v2juncind` (lower court) and `v2juhcind`
(high court) capture freedom from political pressure on judges.
`v2juhccomp` is more demanding still — it asks whether the government
*actually complies* with rulings it dislikes, the real test of
judicial enforcement. `v2jucorrdc` measures judicial bribery; V-Dem's
ordinal coding runs from "Always" (low) to "Never" (high), so high =
less bribery and no flip is needed. `v2juaccnt` captures whether
judges face accountability when caught in misconduct. All six run
positively.

```{r ax-tab-contract}
axis_table("contract_enforcement")
```

## Subsidies / clientelism

PSW define "subsidies" broadly to include directed credit, tax breaks,
and any state transfer that benefits specific firms or coalitions
rather than the economy at large. The unifying logic is
particularistic distribution. V-Dem's `v2xnp_client` (clientelism
index) is the cleanest match — a Bayesian composite of how much
politics is built on selective material incentives. We supplement
with three indicators that need flipping to face the right way:
`v2dlencmps` is the same particularistic-vs-public-goods variable
used for common infrastructure (high there means public goods, the
*opposite* of clientelistic spending, so we flip); `v2psprlnks` runs
from clientelistic (low) to programmatic (high) party linkages, also
flipped; `v2elvotbuy` is V-Dem's vote-buying indicator with the usual
"low = systematic, high = none" coding, flipped. The axis score is
high when more of the policy mix flows through targeted, particularistic
channels.

```{r ax-tab-subsidies}
axis_table("subsidies_clientelism")
```

## Exclusivity

Barriers to entry, licensing regimes, regulatory rents that protect
incumbents. V-Dem's `v2x_neopat` (Neopatrimonial Rule) is the cleanest
political-side measure — rule by personal authority *substitutes* for
impersonal market access, which is what exclusivity is in PSW's frame.
`v2psparban` (party ban) and `v2psoppaut` (opposition autonomy) capture
political exclusivity in a strict sense; both need flipping (high values
mean *less* banning / *more* opposition autonomy, i.e. less exclusivity).
From QoG we add two market-competition measures, both flipped:
`qog_fi_reg` is Fraser's business/credit/labour regulation score where
*higher* means *less* restrictive regulation; `qog_bti_mo` is BTI's
organisation-of-market-and-competition where higher means a *more*
competitive market. The axis score is high when the institutional
environment favours incumbents over entrants.

```{r ax-tab-exclusivity}
axis_table("exclusivity")
```

# Axis scoring

```{r score-axes}
percentile_rank <- function(x) {
  r <- rank(x, na.last = "keep")
  r / sum(!is.na(x))
}

# Score every indicator
indicator_scores <- panel %>%
  pivot_longer(c(-country_name, -iso3),
               names_to = "indicator", values_to = "value") %>%
  inner_join(indicator_map, by = "indicator") %>%
  group_by(indicator) %>%
  mutate(pct = percentile_rank(value)) %>%
  ungroup() %>%
  mutate(score = ifelse(sign == 1, pct, 1 - pct))

# Average within axis (drop NA so countries with partial coverage still score)
axis_scores <- indicator_scores %>%
  group_by(country_name, iso3, axis) %>%
  summarise(axis_score = mean(score, na.rm = TRUE),
            n_indicators = sum(!is.na(score)),
            .groups = "drop")

# Coverage: keep countries with at least 1 non-NA indicator on every axis
ok_countries <- axis_scores %>%
  group_by(country_name) %>%
  summarise(min_n = min(n_indicators), .groups = "drop") %>%
  filter(min_n >= 1) %>% pull(country_name)

axis_scores <- axis_scores %>% filter(country_name %in% ok_countries)

axis_wide <- axis_scores %>%
  select(country_name, iso3, axis, axis_score) %>%
  pivot_wider(names_from = axis, values_from = axis_score) %>%
  select(country_name, iso3, all_of(axis_levels))

cat("Countries with full axis coverage:", nrow(axis_wide), "\n")
```

# Stakeholder fit scores

```{r psw-prefs}
psw_prefs <- tribble(
  ~stakeholder,
  ~transparent_rules, ~common_infra, ~specialized_infra,
  ~contract_enforcement, ~subsidies_clientelism, ~exclusivity,
  "Magicians",    1, 1, 1, 1, 1, 0,
  "Workhorses",   1, 1, 0, 1, 0, 0,
  "Rentiers",     0, 0, 1, 1, 1, 1,
  "Powerbrokers", 0, 0, 0, 0, 0, 1
)

stakeholder_levels <- c("Magicians", "Workhorses", "Rentiers", "Powerbrokers")

pref_mat <- psw_prefs %>% column_to_rownames("stakeholder") %>% as.matrix()
profile_mat <- axis_wide %>% column_to_rownames("country_name") %>%
  select(-iso3) %>% as.matrix()
```

For each country and each stakeholder, the **fit score** is the average
of the country's axis scores over the axes that stakeholder demands.
Equivalently: "out of 1, how well does this country supply what
stakeholder X is asking for." We then renormalise the four scores to
sum to 1 to read them as a *stakeholder share*.

The share interpretation has a built-in caveat. The four preference
vectors overlap (e.g., contract enforcement is wanted by everyone except
powerbrokers), so the share isn't a clean decomposition — it's a
similarity-based approximation of which stakeholder profile a country's
institutions most resemble.

```{r fit-scores}
# Average of country's axis scores over the axes demanded by each stakeholder
fit_mat <- profile_mat %*% t(pref_mat) / matrix(rowSums(pref_mat),
                                                nrow = nrow(profile_mat),
                                                ncol = nrow(pref_mat),
                                                byrow = TRUE)

# Renormalise rows to sum to 1
fit_shares <- fit_mat / rowSums(fit_mat)

fit_long <- as_tibble(fit_mat, rownames = "country_name") %>%
  pivot_longer(-country_name, names_to = "stakeholder", values_to = "fit") %>%
  left_join(
    as_tibble(fit_shares, rownames = "country_name") %>%
      pivot_longer(-country_name, names_to = "stakeholder", values_to = "share"),
    by = c("country_name", "stakeholder")
  ) %>%
  mutate(stakeholder = factor(stakeholder, levels = stakeholder_levels)) %>%
  left_join(axis_wide %>% select(country_name, iso3), by = "country_name")
```

## Top countries by stakeholder fit

```{r top-by-stakeholder, fig.width=6.5, fig.height=8.5}
top15 <- fit_long %>%
  group_by(stakeholder) %>%
  slice_max(fit, n = 15) %>%
  ungroup() %>%
  mutate(country_name = reorder_within(country_name, fit, stakeholder))

ggplot(top15, aes(x = fit, y = country_name, fill = stakeholder)) +
  geom_col() +
  facet_wrap(~ stakeholder, scales = "free_y", ncol = 2) +
  scale_y_reordered() +
  labs(x = "Fit score (mean of demanded-axis scores)",
       y = NULL,
       title = "Top 15 countries by stakeholder fit",
       subtitle = paste0("Higher fit = institutions supply more of what stakeholder demands. ",
                         "V-Dem ", latest_year, "."))
```

```{r top-tables}
top10_table <- fit_long %>%
  group_by(stakeholder) %>%
  slice_max(fit, n = 10) %>%
  mutate(rank = row_number(),
         entry = sprintf("%s (%.2f)", country_name, fit)) %>%
  ungroup() %>%
  select(rank, stakeholder, entry) %>%
  pivot_wider(names_from = stakeholder, values_from = entry) %>%
  select(rank, all_of(stakeholder_levels))

kable(top10_table, caption = "Top 10 country fit per stakeholder (fit score in parens)")
```

## Stakeholder profile of selected countries

A few well-known cases — including those Pritchett, Sen & Werker
themselves apply the framework to in their book (Bangladesh, Cambodia,
Liberia, Malaysia, Rwanda) — alongside reference economies and
resource-dependent states.

```{r selected-profiles, fig.width=6.5, fig.height=5}
selected_iso <- c("BGD", "KHM", "LBR", "MYS", "RWA",  # PSW case studies
                  "USA", "DEU", "JPN", "KOR",          # advanced economies
                  "CHN", "VNM", "ETH",                 # state-led
                  "SAU", "RUS", "VEN", "AGO", "AZE",   # resource-heavy
                  "IND", "BRA", "MEX", "TUR", "ARG",   # large EM
                  "PAK", "MAR", "NGA")                 # other comparators

selected_shares <- fit_long %>%
  filter(iso3 %in% selected_iso) %>%
  mutate(country_name = fct_reorder2(country_name,
                                     stakeholder == "Magicians", share))

ggplot(selected_shares, aes(x = share, y = country_name, fill = stakeholder)) +
  geom_col() +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_gl() +
  labs(x = "Stakeholder share (fit, renormalised)",
       y = NULL,
       fill = NULL,
       title = "Stakeholder shares — selected countries",
       subtitle = "Renormalised so the four stakeholder fits sum to 100% per country") +
  theme(legend.position = "bottom")
```

## Axis profile of archetypal countries

For one country at the top of each stakeholder's ranking, what does its
6-axis institutional profile look like?

```{r archetype-axes, fig.width=6.5, fig.height=4.5}
archetypes <- fit_long %>%
  group_by(stakeholder) %>%
  slice_max(fit, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(stakeholder, country_name)

archetype_axes <- axis_scores %>%
  inner_join(archetypes, by = "country_name") %>%
  mutate(axis = factor(axis, levels = axis_levels, labels = axis_labels[axis_levels]),
         stakeholder = factor(stakeholder, levels = stakeholder_levels)) %>%
  mutate(label = paste0(country_name, "\n(", stakeholder, ")"))

# Order facet labels by stakeholder
archetype_axes <- archetype_axes %>%
  mutate(label = factor(label,
                        levels = archetype_axes %>%
                          distinct(stakeholder, label) %>%
                          arrange(stakeholder) %>% pull(label)))

ggplot(archetype_axes, aes(x = axis, y = axis_score, group = label)) +
  geom_col(aes(fill = stakeholder)) +
  facet_wrap(~ label, ncol = 2) +
  scale_fill_gl() +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Axis score (percentile)",
       fill = NULL,
       title = "Six-axis institutional profile of each archetype",
       subtitle = "Top-ranked country on each stakeholder's fit score") +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 7))
```

# Sanity checks

## Magician fit vs. Economic Complexity

If the framework holds, magician-style institutions (transparent rules,
contract enforcement, public-goods spending) should be associated with
diversified, sophisticated export baskets. We expect a positive but
imperfect relationship — magician-friendly institutions are a
*precondition* for complexity, not a mechanical determinant.

```{r magician-vs-eci, fig.width=6.5, fig.height=4.5}
eci_latest <- read_parquet(
  "~/dev/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year", "atl_sitc_eci")
) %>%
  filter(!is.na(atl_sitc_eci)) %>%
  group_by(countrycodeiso) %>% slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(iso3 = countrycodeiso, eci = atl_sitc_eci)

magician_eci <- fit_long %>%
  filter(stakeholder == "Magicians") %>%
  inner_join(eci_latest, by = "iso3")

label_set <- c("USA", "DEU", "JPN", "KOR", "CHN", "VNM", "BGD", "KHM",
               "MYS", "IND", "MEX", "TUR", "BRA", "RUS", "SAU", "VEN",
               "AGO", "NGA", "ETH", "RWA", "LBR", "MAR")

ggplot(magician_eci, aes(x = fit, y = eci)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = highlight) +
  geom_text_repel(
    data = ~ filter(.x, iso3 %in% label_set),
    aes(label = iso3), size = 2.8, max.overlaps = 30
  ) +
  labs(x = "Magician fit score",
       y = "Economic Complexity Index (latest)",
       title = "Magician institutions vs. economic complexity",
       subtitle = sprintf("r = %.2f",
                          cor(magician_eci$fit, magician_eci$eci, use = "pairwise")))
```

## Powerbroker vs. magician fit

These two stakeholder types have nearly orthogonal demand vectors —
magicians want everything except exclusivity, powerbrokers want only
exclusivity (and benefit from the absence of everything else). Their
fits across countries should be strongly negatively correlated.

```{r mag-vs-pb, fig.width=6.5, fig.height=5}
mag_pb <- fit_long %>%
  filter(stakeholder %in% c("Magicians", "Powerbrokers")) %>%
  select(country_name, iso3, stakeholder, fit) %>%
  pivot_wider(names_from = stakeholder, values_from = fit)

ggplot(mag_pb, aes(x = Magicians, y = Powerbrokers)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = highlight) +
  geom_text_repel(
    data = ~ filter(.x, iso3 %in% label_set),
    aes(label = iso3), size = 2.8, max.overlaps = 30
  ) +
  labs(x = "Magician fit", y = "Powerbroker fit",
       title = "Powerbroker world is the inverse of magician world",
       subtitle = sprintf("r = %.2f",
                          cor(mag_pb$Magicians, mag_pb$Powerbrokers, use = "pairwise")))
```

# Outputs

```{r write-outputs}
fit_out <- fit_long %>%
  select(country_name, iso3, stakeholder, fit, share) %>%
  arrange(country_name, stakeholder)

axis_out <- axis_wide

dir.create("outputs", showWarnings = FALSE)
write_csv(fit_out, "outputs/deals_stakeholder_fit.csv")
write_csv(axis_out, "outputs/deals_axis_scores.csv")

cat("Wrote: outputs/deals_stakeholder_fit.csv (",
    nrow(fit_out), "rows)\n")
cat("Wrote: outputs/deals_axis_scores.csv (",
    nrow(axis_out), "rows)\n")
```

# Caveats and next steps

- **The share interpretation is approximate.** Preference vectors
  overlap (most stakeholders want contract enforcement; only powerbrokers
  affirmatively want exclusivity), so the renormalised shares aren't a
  proper decomposition of bargaining power — they're a similarity index
  rescaled.
- **Powerbrokers reduce to "high exclusivity."** With only one demanded
  axis, the powerbroker fit is mechanically the exclusivity score. That
  is by construction of PSW's matrix, not a flaw in scoring, but worth
  flagging when reading rankings.
- **Magicians and workhorses are correlated by design.** They share four
  of five demanded axes, so their rankings will track each other; the
  separation lives almost entirely in specialized infrastructure and
  subsidies/clientelism.
- **No infrastructure stock data in V-Dem.** Common-infra and
  specialized-infra rely on WDI access/quality variables, which have
  more measurement noise and patchier coverage than V-Dem.
- **PSW's per-country rents-space estimates** (e.g. Liberia ~10/25/61/3,
  Malawi ~22/37/36/small for rentiers/powerbrokers/workhorses/magicians)
  could serve as ground truth for a few cases. Worth a follow-up
  comparison.

The unsupervised counterpart (archetype analysis on the 6-axis matrix,
recovering 4 latent stakeholder types from the data alone) is a
companion exercise documented in `deals-and-development.md`.
