---
title: "Institutional Complexity Analysis"
subtitle: "Applying Economic Complexity Methods to V-Dem Governance Data"
author: "Institutional Complexity Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6.5
    fig_height: 4.5
    latex_engine: xelatex
    includes:
      in_header: ~/dev/gl-design/skills/gl-ggplot/assets/gl_pdf.tex
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  cache = TRUE,
  cache.lazy = FALSE
)
```

# Introduction

Economic complexity theory, developed by Hidalgo and Hausmann (2009), provides a powerful
framework for understanding why some countries are wealthier than others. The core insight is
elegant: countries that can produce many products -- especially products that few other countries
can produce -- tend to be more economically sophisticated.

Here we apply this same logic to **institutions and governance**. Using data from the
Varieties of Democracy (V-Dem) project, we ask: which countries have built complex institutional
portfolios? Which institutional features are "rare" -- present only in countries with deep
institutional capacity? And how do institutions cluster together in an "institution space"?

The analogy works as follows:

| Economic Complexity | Institutional Complexity |
|---|---|
| Countries export products | Countries have institutional features |
| Revealed Comparative Advantage | Continuous V-Dem scores (0-1 normalized) |
| Product Complexity Index (PCI) | Institutional Feature Complexity (IFC) |
| Economic Complexity Index (ECI) | Institutional Complexity Index (ICI) |
| Product Space | Institution Space |

# Setup and Data

## Load packages

```{r packages}
library(tidyverse)
library(arrow)
library(uwot)
library(ggrepel)
library(countrycode)
library(WDI)
library(pheatmap)
library(igraph)
library(knitr)
library(DT)
library(patchwork)
```

```{r gl-setup, cache=FALSE}
# Uncached so showtext / sysfonts re-attach on every knit and gl_setup()
# can self-install its knitr chunk hook. See deals-development.Rmd for the
# rationale.
source("~/dev/gl-design/skills/gl-ggplot/assets/theme_gl.R")
gl_setup()
```

## Load V-Dem data

We load the full V-Dem dataset and codebook from the shared data directory
(`data/shared-data/vdem/`).

```{r load-vdem, cache=TRUE}
vdem_raw <- read_parquet("data/shared-data/vdem/vdem.parquet")
vdem_codebook <- read_parquet("data/shared-data/vdem/vdem_codebook.parquet")

# Check the most recent years and coverage
year_coverage <- vdem_raw %>%
  filter(year >= 2020) %>%
  group_by(year) %>%
  summarise(
    n_countries = n_distinct(country_name),
    .groups = "drop"
  )

kable(year_coverage, caption = "Country coverage by year (2020+)")
```

```{r select-year}
# Use the most recent year with substantial coverage
latest_year <- year_coverage %>%
  filter(n_countries >= 150) %>%
  pull(year) %>%
  max()

cat("Selected year:", latest_year, "\n")
```

## Select institutional indicators

Rather than hand-picking a subset, we use **all non-index component indicators** from V-Dem.
The V-Dem codebook classifies each variable by type:

- **C** (Component): Expert-coded assessments -- these are the actual institutional measures
- **A** (Aggregate): Aggregated indices built from components
- **D** (Derived): Derived/composite indices

We use only type-C indicators, which represent the individual expert-coded assessments of
institutional features. This avoids double-counting (indices are linear combinations of
components) and gives us the most granular view of institutional capabilities.

We further filter to indicators with good coverage in the selected year (>= 150 countries)
and exclude text/name variables.

```{r select-indicators}
# Category labels based on V-Dem prefix naming convention
prefix_categories <- c(
  "el" = "Elections",
  "ps" = "Political Parties",
  "cl" = "Civil Liberties",
  "ju" = "Judiciary",
  "lg" = "Legislature",
  "me" = "Media",
  "ex" = "Executive",
  "dl" = "Deliberation",
  "ca" = "Academic & Cultural",
  "cs" = "Civil Society",
  "pe" = "Political Equality",
  "st" = "State Capacity",
  "sv" = "Sovereignty",
  "dd" = "Direct Democracy",
  "ed" = "Education",
  "sm" = "Social Movements & Protest",
  "re" = "Regime",
  "te" = "Technology & Media Environment"
)

# Select all C-type (component) indicators from the codebook
component_codes <- vdem_codebook %>%
  filter(vartype == "C") %>%
  pull(tag)

# Keep only v2* base indicators (no suffix variants, no text columns)
suffixes <- c("_codelow", "_codehigh", "_sd", "_osp", "_ord", "_mean", "_nr")
all_cols <- names(vdem_raw)
is_suffix <- sapply(all_cols, function(x) any(sapply(suffixes, function(s) endsWith(x, s))))
is_numbered <- grepl("_\\d+$", all_cols)
base_cols <- all_cols[!is_suffix & !is_numbered]

# Intersect with codebook C-type
candidate_codes <- intersect(component_codes, base_cols)

# Filter to selected year and check coverage
latest_data <- vdem_raw %>% filter(year == latest_year)
coverage <- sapply(candidate_codes, function(x) sum(!is.na(latest_data[[x]])))

# Keep indicators with >= 150 countries and that are numeric
numeric_check <- sapply(candidate_codes, function(x) is.numeric(vdem_raw[[x]]))
indicator_codes <- candidate_codes[coverage >= 150 & numeric_check]

# Build indicator_info from codebook
indicator_info <- vdem_codebook %>%
  filter(tag %in% indicator_codes) %>%
  transmute(
    code = tag,
    description = name,
    prefix = sub("^v[23](\\w{2}).*", "\\1", tag),
    category = prefix_categories[prefix]
  ) %>%
  mutate(category = replace_na(category, "Other")) %>%
  select(-prefix) %>%
  distinct(code, .keep_all = TRUE)

cat("Total component indicators selected:", nrow(indicator_info), "\n")
cat("\nCategories:\n")
indicator_info %>% count(category, sort = TRUE) %>% print(n = 20)
```

```{r filter-data}
# Check which of our selected indicators actually exist in the dataset
available_codes <- intersect(indicator_info$code, names(vdem_raw))
missing_codes <- setdiff(indicator_info$code, names(vdem_raw))

if (length(missing_codes) > 0) {
  cat("Indicators not found in dataset (will be dropped):\n")
  cat(paste(missing_codes, collapse = ", "), "\n\n")
}

indicator_info <- indicator_info %>% filter(code %in% available_codes)
cat("Indicators available:", nrow(indicator_info), "\n")

# Filter to selected year and extract relevant columns
vdem_filtered <- vdem_raw %>%
  filter(year == latest_year) %>%
  select(
    country_name, country_text_id, COWcode,
    e_regionpol, e_regionpol_6C,
    all_of(indicator_info$code)
  ) %>%
  # Drop countries with too much missing data (> 50% missing)
  mutate(n_missing = rowSums(is.na(across(all_of(indicator_info$code))))) %>%
  filter(n_missing < ncol(across(all_of(indicator_info$code))) * 0.5) %>%
  select(-n_missing)

cat("Countries retained:", nrow(vdem_filtered), "\n")

# Region lookups (used throughout)
region_lookup <- vdem_filtered %>%
  select(country_name, e_regionpol_6C) %>%
  deframe()

region_labels <- c(
  "1" = "Eastern Europe & Central Asia",
  "2" = "Latin America & Caribbean",
  "3" = "Middle East & North Africa",
  "4" = "Sub-Saharan Africa",
  "5" = "Western Europe & North America",
  "6" = "East Asia & Pacific"
)
```

## Remove discrete variables

V-Dem's C-type indicators are produced by a Bayesian IRT measurement model that converts
ordinal expert codings into continuous latent scores (typically ranging from roughly -4 to +4).
However, two indicators are discrete integers rather than continuous IRT output:

- **v2regopploc** — "Regime opposition location" (integer-coded)
- **v2regsuploc** — "Regime support location" (integer-coded)

Since our continuous analysis requires all indicators to sit on comparable continuous scales,
we drop these two.

```{r remove-discrete}
discrete_vars <- c("v2regopploc", "v2regsuploc")

# Document what we're removing
indicator_info %>%
  filter(code %in% discrete_vars) %>%
  select(Code = code, Category = category, Description = description) %>%
  kable(caption = "Discrete variables removed from analysis")

indicator_info <- indicator_info %>%
  filter(!code %in% discrete_vars)

cat("Indicators after removing discrete variables:", nrow(indicator_info), "\n")
```

## Indicator catalog

Here is the full set of institutional indicators we're working with:

```{r indicator-table}
kable(
  indicator_info %>% select(Code = code, Category = category, Description = description),
  caption = "Institutional indicators used in this analysis"
)
```

# Continuous Mcp Matrix

Rather than binarizing indicators (which discards information about *how much* of a feature
a country has), we construct a **continuous** country × feature matrix with values compressed
to [0, 1]. This preserves the full variation in institutional quality across countries.

## Min-max normalization

Each indicator is rescaled to [0, 1] using min-max normalization across countries:

$$x_{cp}^{*} = \frac{x_{cp} - \min_c(x_{cp})}{\max_c(x_{cp}) - \min_c(x_{cp})}$$

This means 0 = worst observed score across all countries, 1 = best observed score. The
continuous matrix can be used directly in EC-style calculations — higher values indicate
stronger "presence" of a feature, analogous to higher export intensity in trade data.

```{r continuous-mcp}
# Build the continuous 0-1 matrix
cont_df <- vdem_filtered %>% select(country_name, all_of(indicator_info$code))

for (ind in indicator_info$code) {
  values <- cont_df[[ind]]
  rng <- range(values, na.rm = TRUE)
  if (rng[2] > rng[1]) {
    cont_df[[ind]] <- (values - rng[1]) / (rng[2] - rng[1])
  } else {
    cont_df[[ind]] <- 0
  }
}

# Replace remaining NAs with 0
cont_df <- cont_df %>%
  mutate(across(all_of(indicator_info$code), ~replace_na(.x, 0)))

# Create the continuous Mcp matrix
Mcp <- cont_df %>%
  column_to_rownames("country_name") %>%
  as.matrix()

cat("Continuous Mcp matrix dimensions:", nrow(Mcp), "countries x", ncol(Mcp), "features\n")
cat("Value range:", round(min(Mcp), 3), "-", round(max(Mcp), 3), "\n")
cat("Mean value:", round(mean(Mcp), 3), "\n")
cat("Median value:", round(median(Mcp), 3), "\n")
```

```{r mcp-summary}
# Country-level summary: mean institutional score across features
country_means <- rowMeans(Mcp)
cat("\nMean institutional score by country:\n")
cat("  Mean:", round(mean(country_means), 3), "\n")
cat("  SD:", round(sd(country_means), 3), "\n")
cat("  Range:", round(min(country_means), 3), "-", round(max(country_means), 3), "\n")

# Feature-level summary: mean score across countries
feature_means <- colMeans(Mcp)
cat("\nMean country score by feature:\n")
cat("  Mean:", round(mean(feature_means), 3), "\n")
cat("  SD:", round(sd(feature_means), 3), "\n")
cat("  Range:", round(min(feature_means), 3), "-", round(max(feature_means), 3), "\n")
```

# Complexity Calculation

## Diversity and Ubiquity (0th order)

With a continuous Mcp matrix, **diversity** becomes the sum of institutional scores across
all features (total institutional intensity), and **ubiquity** becomes the sum across all
countries for a given feature. These are the continuous generalizations of the binary count
measures.

```{r diversity-ubiquity}
# Diversity: sum of continuous feature scores per country
kc0 <- rowSums(Mcp)
# Ubiquity: sum of continuous country scores per feature
kp0 <- colSums(Mcp)

diversity_df <- tibble(
  country = names(kc0),
  diversity = as.numeric(kc0)
) %>%
  arrange(desc(diversity))

# Top and bottom by diversity
cat("Top 10 countries by diversity (# institutional features):\n")
head(diversity_df, 10) %>% kable()

cat("\nBottom 10 countries by diversity:\n")
tail(diversity_df, 10) %>% kable()
```

```{r ubiquity-plot, fig.height=6}
ubiquity_df <- tibble(
  indicator = names(kp0),
  ubiquity = as.numeric(kp0)
) %>%
  left_join(indicator_info, by = c("indicator" = "code")) %>%
  arrange(ubiquity)

ggplot(ubiquity_df, aes(x = ubiquity, fill = category)) +
  geom_histogram(bins = 30, color = "white") +
  labs(x = "Ubiquity (sum of country scores)", y = "Count of features", fill = "Category")

save_fig("full", "ubiquity_distribution.pdf")
```

## Method of Reflections

The Method of Reflections iteratively refines diversity and ubiquity. With continuous Mcp
values, each iteration updates a country's score as the Mcp-weighted average of feature
scores, and vice versa. This is the continuous generalization of the binary method.

```{r method-of-reflections}
# Method of Reflections (continuous generalization)
n_iterations <- 20

# Initialize
kc <- matrix(0, nrow = nrow(Mcp), ncol = n_iterations + 1)
kp <- matrix(0, nrow = ncol(Mcp), ncol = n_iterations + 1)
rownames(kc) <- rownames(Mcp)
rownames(kp) <- colnames(Mcp)

kc[, 1] <- kc0
kp[, 1] <- kp0

for (n in 1:n_iterations) {
  # Update country complexity: weighted average of feature complexity
  for (c_idx in 1:nrow(Mcp)) {
    if (kc0[c_idx] > 0) {
      kc[c_idx, n + 1] <- (1 / kc0[c_idx]) * sum(Mcp[c_idx, ] * kp[, n])
    }
  }

  # Update feature complexity: weighted average of country complexity
  for (p_idx in 1:ncol(Mcp)) {
    if (kp0[p_idx] > 0) {
      kp[p_idx, n + 1] <- (1 / kp0[p_idx]) * sum(Mcp[, p_idx] * kc[, n])
    }
  }
}

# ECI from Method of Reflections
kc_even <- kc[, n_iterations + 1]
eci_mor <- (kc_even - mean(kc_even)) / sd(kc_even)

kp_even <- kp[, n_iterations + 1]
pci_mor <- (kp_even - mean(kp_even)) / sd(kp_even)

cat("Method of Reflections converged after", n_iterations, "iterations.\n")
```

## Eigenvalue Method

The eigenvalue method is more elegant and mathematically equivalent to the Method of Reflections
at convergence. We construct the matrix $\tilde{M}_{cc'}$ and take its second eigenvector.

$$\tilde{M}_{cc'} = \sum_p \frac{M_{cp} M_{c'p}}{k_{c,0} k_{p,0}}$$

```{r eigenvalue-method}
# Build Mcc_tilde
# Mcc_tilde = (1/kc0) * Mcp %*% diag(1/kp0) %*% t(Mcp)

# Handle zeros
kc0_safe <- ifelse(kc0 == 0, 1, kc0)
kp0_safe <- ifelse(kp0 == 0, 1, kp0)

Mcc_tilde <- diag(1 / kc0_safe) %*% Mcp %*% diag(1 / kp0_safe) %*% t(Mcp)
rownames(Mcc_tilde) <- rownames(Mcp)
colnames(Mcc_tilde) <- rownames(Mcp)

# Eigen decomposition
eig <- eigen(Mcc_tilde)

# The first eigenvector is trivial (all same sign), use second
eci_eigen_raw <- Re(eig$vectors[, 2])
names(eci_eigen_raw) <- rownames(Mcp)

# Standardize
eci_eigen <- (eci_eigen_raw - mean(eci_eigen_raw)) / sd(eci_eigen_raw)

# Ensure sign convention: positive ECI = more complex
# Check correlation with diversity -- complex countries should have higher diversity
if (cor(eci_eigen, kc0) < 0) {
  eci_eigen <- -eci_eigen
}

# Similarly for PCI via Mpp_tilde
Mpp_tilde <- diag(1 / kp0_safe) %*% t(Mcp) %*% diag(1 / kc0_safe) %*% Mcp
rownames(Mpp_tilde) <- colnames(Mcp)
colnames(Mpp_tilde) <- colnames(Mcp)

eig_p <- eigen(Mpp_tilde)
pci_eigen_raw <- Re(eig_p$vectors[, 2])
names(pci_eigen_raw) <- colnames(Mcp)
pci_eigen <- (pci_eigen_raw - mean(pci_eigen_raw)) / sd(pci_eigen_raw)

if (cor(pci_eigen, kp0, use = "complete.obs") > 0) {
  # More ubiquitous features should be LESS complex
  pci_eigen <- -pci_eigen
}

cat("Correlation between MoR and eigenvalue ECI:",
    round(cor(eci_mor, eci_eigen), 3), "\n")
```

## Results: Institutional Complexity Index

We use the eigenvalue method as our primary measure (it's more robust), with the Method of
Reflections as a check.

```{r eci-results}
eci_df <- tibble(
  country = names(eci_eigen),
  ICI = eci_eigen,
  ICI_MoR = eci_mor,
  diversity = kc0
) %>%
  arrange(desc(ICI))

cat("Top 20 countries by Institutional Complexity Index:\n")
eci_df %>%
  head(20) %>%
  mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  kable()

cat("\nBottom 20 countries by Institutional Complexity Index:\n")
eci_df %>%
  tail(20) %>%
  mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  kable()
```

```{r eci-distribution}
ggplot(eci_df, aes(x = ICI)) +
  geom_histogram(bins = 30, fill = gl$brand_blue, color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = highlight) +
  labs(x = "ICI (standardized)", y = "Count")

save_fig("full", "eci_distribution.pdf")
```

## Results: Institutional Feature Complexity

Which institutional features are most "complex" -- i.e., found predominantly in
institutionally sophisticated countries?

```{r pci-results}
pci_df <- tibble(
  indicator = names(pci_eigen),
  IFC = pci_eigen,
  ubiquity = kp0
) %>%
  left_join(indicator_info, by = c("indicator" = "code")) %>%
  arrange(desc(IFC))

cat("Top 20 most complex institutional features:\n")
pci_df %>%
  head(20) %>%
  select(indicator, description, category, IFC, ubiquity) %>%
  mutate(IFC = round(IFC, 3)) %>%
  kable()

cat("\nBottom 20 (most ubiquitous/least complex) features:\n")
pci_df %>%
  tail(20) %>%
  select(indicator, description, category, IFC, ubiquity) %>%
  mutate(IFC = round(IFC, 3)) %>%
  kable()
```

```{r pci-by-category}
ggplot(pci_df, aes(x = reorder(category, IFC, median), y = IFC, fill = category)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Institutional Feature Complexity (IFC)")

save_fig("full", "pci_by_category.pdf")
```

# Proximity and Relatedness

## Method 1: Co-location proximity (EC-style)

We calculate proximity between institutional features using a continuous generalization of
the conditional probability approach from Hidalgo et al. (2007). With continuous Mcp values,
co-occurrence becomes the dot product of feature columns, and conditioning becomes division
by the feature's total weight:

$$\phi_{pp'} = \min\left(\frac{\sum_c M_{cp} \cdot M_{cp'}}{\sum_c M_{cp'}}, \frac{\sum_c M_{cp} \cdot M_{cp'}}{\sum_c M_{cp}}\right)$$

This is the natural continuous extension: when Mcp is binary, this reduces to the standard
min conditional probability.

```{r colocation-feature-proximity, cache=TRUE}
n_features <- ncol(Mcp)
phi_pp <- matrix(0, n_features, n_features)
rownames(phi_pp) <- colnames(Mcp)
colnames(phi_pp) <- colnames(Mcp)

for (i in 1:n_features) {
  for (j in i:n_features) {
    if (i == j) {
      phi_pp[i, j] <- 1
    } else {
      co_occur <- sum(Mcp[, i] * Mcp[, j])
      p_given_j <- if (kp0[j] > 0) co_occur / kp0[j] else 0
      p_given_i <- if (kp0[i] > 0) co_occur / kp0[i] else 0
      phi_pp[i, j] <- min(p_given_i, p_given_j)
      phi_pp[j, i] <- phi_pp[i, j]
    }
  }
}

cat("Co-location feature proximity matrix:", nrow(phi_pp), "x", ncol(phi_pp), "\n")
cat("Mean proximity:", round(mean(phi_pp[upper.tri(phi_pp)]), 3), "\n")
```

```{r colocation-country-proximity, cache=TRUE}
n_countries <- nrow(Mcp)
phi_cc <- matrix(0, n_countries, n_countries)
rownames(phi_cc) <- rownames(Mcp)
colnames(phi_cc) <- rownames(Mcp)

for (i in 1:n_countries) {
  for (j in i:n_countries) {
    if (i == j) {
      phi_cc[i, j] <- 1
    } else {
      co_occur <- sum(Mcp[i, ] * Mcp[j, ])
      p_given_j <- if (kc0[j] > 0) co_occur / kc0[j] else 0
      p_given_i <- if (kc0[i] > 0) co_occur / kc0[i] else 0
      phi_cc[i, j] <- min(p_given_i, p_given_j)
      phi_cc[j, i] <- phi_cc[i, j]
    }
  }
}

cat("Co-location country proximity matrix:", nrow(phi_cc), "x", ncol(phi_cc), "\n")
cat("Mean proximity:", round(mean(phi_cc[upper.tri(phi_cc)]), 3), "\n")
```

## Method 2: Spearman correlation proximity

An alternative approach uses **rank correlation** between indicator pairs across countries
(for feature proximity) and between country pairs across indicators (for country proximity).
Unlike co-location, which captures the degree to which high values co-occur, Spearman
correlation captures the full monotonic relationship — whether countries that score high on
one feature also tend to score high on another, *and* whether countries that score low on one
also score low on the other.

```{r spearman-feature-proximity, cache=TRUE}
# Feature proximity: Spearman correlation between indicator columns
spearman_pp <- cor(Mcp, method = "spearman")

# Shift to [0, 1] for comparability with co-location proximity
spearman_pp_01 <- (spearman_pp + 1) / 2

cat("Spearman feature proximity matrix:", nrow(spearman_pp), "x", ncol(spearman_pp), "\n")
cat("Mean raw Spearman correlation:", round(mean(spearman_pp[upper.tri(spearman_pp)]), 3), "\n")
cat("Mean rescaled [0,1]:", round(mean(spearman_pp_01[upper.tri(spearman_pp_01)]), 3), "\n")
```

```{r spearman-country-proximity, cache=TRUE}
# Country proximity: Spearman correlation between country rows
spearman_cc <- cor(t(Mcp), method = "spearman")

# Shift to [0, 1]
spearman_cc_01 <- (spearman_cc + 1) / 2

cat("Spearman country proximity matrix:", nrow(spearman_cc), "x", ncol(spearman_cc), "\n")
cat("Mean raw Spearman correlation:", round(mean(spearman_cc[upper.tri(spearman_cc)]), 3), "\n")
cat("Mean rescaled [0,1]:", round(mean(spearman_cc_01[upper.tri(spearman_cc_01)]), 3), "\n")
```

## PCA on the continuous matrix

Rather than extracting a single complexity index from the eigenvalue method (which uses only
the second eigenvector), we can use PCA on the full continuous matrix to capture multiple
dimensions of institutional variation. The first few principal components may reveal distinct
axes of institutional development — not just "more vs less" but qualitatively different
patterns.

```{r pca-analysis}
# PCA on the country x feature matrix
pca_result <- prcomp(Mcp, center = TRUE, scale. = TRUE)

# Variance explained
var_explained <- summary(pca_result)$importance[2, 1:10]
cum_var <- summary(pca_result)$importance[3, 1:10]

tibble(
  PC = paste0("PC", 1:10),
  `Variance explained` = round(var_explained * 100, 1),
  `Cumulative %` = round(cum_var * 100, 1)
) %>% kable(caption = "PCA variance explained (first 10 components)")
```

```{r pca-country-scores}
# Country scores on first few PCs
pca_country <- tibble(
  country = rownames(Mcp),
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  PC3 = pca_result$x[, 3]
) %>%
  mutate(
    region = region_labels[as.character(region_lookup[country])],
    region = replace_na(region, "Other")
  )

cat("Correlation between ECI (eigenvalue) and PC1:",
    round(cor(eci_eigen, pca_result$x[, 1]), 3), "\n")

# Persist country PC scores for downstream analyses (e.g. deals-development.Rmd
# uses PC2 as a "managed vs contested" settlement axis).
pca_export <- as.data.frame(pca_result$x[, 1:5])
pca_export$country_name <- rownames(pca_export)
pca_export <- pca_export[, c("country_name", paste0("PC", 1:5))]
for (k in 1:5) {
  pca_export[[paste0("PC", k, "_pct")]] <-
    round(ecdf(pca_export[[paste0("PC", k)]])(pca_export[[paste0("PC", k)]]) * 100, 1)
}
dir.create("outputs", showWarnings = FALSE)
write.csv(pca_export, "outputs/institutional_pca_scores.csv", row.names = FALSE)
```

## Institution Space (Network)

The institution space is a network where nodes are institutional features and edges connect
features that tend to co-occur across countries.

```{r institution-space-network}
# Threshold for edges: use a percentile-based cutoff
threshold_p <- quantile(phi_pp[upper.tri(phi_pp)], 0.75)

# Build adjacency matrix
adj_pp <- phi_pp
adj_pp[adj_pp < threshold_p] <- 0
diag(adj_pp) <- 0

# Create igraph object
g_inst <- graph_from_adjacency_matrix(adj_pp, mode = "undirected", weighted = TRUE)

# Add attributes
V(g_inst)$category <- indicator_info$category[match(V(g_inst)$name, indicator_info$code)]
V(g_inst)$complexity <- pci_eigen[V(g_inst)$name]
V(g_inst)$ubiquity <- kp0[V(g_inst)$name]

cat("Institution space: ", vcount(g_inst), "nodes,", ecount(g_inst), "edges\n")
cat("Mean degree:", round(mean(degree(g_inst)), 1), "\n")
```

## Country Space (Network)

```{r country-space-network}
threshold_c <- quantile(phi_cc[upper.tri(phi_cc)], 0.80)

adj_cc <- phi_cc
adj_cc[adj_cc < threshold_c] <- 0
diag(adj_cc) <- 0

g_country <- graph_from_adjacency_matrix(adj_cc, mode = "undirected", weighted = TRUE)

V(g_country)$region <- region_labels[as.character(
  region_lookup[V(g_country)$name]
)]
V(g_country)$ICI <- eci_eigen[V(g_country)$name]

cat("Country space:", vcount(g_country), "nodes,", ecount(g_country), "edges\n")
```

# UMAP Visualizations

## Institution Space (UMAP)

We use UMAP to embed the institution proximity matrix into 2D for visualization. UMAP
preserves local structure well, making it ideal for revealing clusters of related institutions.

```{r umap-institutions, cache=TRUE}
# Convert proximity to distance, use cmdscale (MDS) as fallback since uwot
# doesn't support precomputed distance matrices well on all versions
inst_dist <- as.dist(1 - phi_pp)

# Use MDS first, then UMAP on the MDS coordinates for robust embedding.
# Larger n_neighbors and min_dist produce a more balanced, blob-shaped layout
# than the stringy horseshoe that small values give on this near-1D proximity.
set.seed(2024)
mds_inst <- cmdscale(inst_dist, k = min(10, ncol(Mcp) - 1))
umap_inst <- umap(mds_inst, n_neighbors = 60, min_dist = 0.8, n_epochs = 500)

umap_inst_df <- tibble(
  indicator = colnames(Mcp),
  UMAP1 = umap_inst[, 1],
  UMAP2 = umap_inst[, 2],
  ubiquity = kp0,
  IFC = pci_eigen
) %>%
  left_join(indicator_info, by = c("indicator" = "code"))
```

```{r umap-inst-plot, fig.width=6.5, fig.height=6.5}
# Extend GL palette for 15+ categories
n_cats <- length(unique(umap_inst_df$category))
cat_colors <- colorRampPalette(gl_palettes$categorical)(n_cats)

# Human-readable labels: truncate then wrap
umap_inst_df <- umap_inst_df %>%
  mutate(label = str_trunc(description, 35) %>% str_wrap(width = 20))

ggplot(umap_inst_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = category), size = 2.5, alpha = 0.75) +
  geom_text_repel(
    aes(label = label),
    size = 1.8, max.overlaps = 20, segment.alpha = 0.3, lineheight = 0.85
  ) +
  scale_color_manual(values = cat_colors, name = "Category") +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme(legend.position = "bottom")

ggsave("imgs/umap_institution_space_colocation.png", width = 9, height = 9, dpi = 300)
```

## Country Space (UMAP)

```{r umap-countries, cache=TRUE}
set.seed(42)
country_dist <- as.dist(1 - phi_cc)
mds_country <- cmdscale(country_dist, k = min(10, nrow(Mcp) - 1))
umap_country <- umap(mds_country, n_neighbors = 15, min_dist = 0.2, n_epochs = 500)

umap_country_df <- tibble(
  country = rownames(Mcp),
  UMAP1 = umap_country[, 1],
  UMAP2 = umap_country[, 2],
  ICI = eci_eigen,
  diversity = kc0
)

# Add region
umap_country_df <- umap_country_df %>%
  mutate(
    region = region_labels[as.character(region_lookup[country])],
    region = replace_na(region, "Other")
  )
```

```{r umap-country-plot, fig.width=6.5, fig.height=6.5}
ggplot(umap_country_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = region), size = 2.5, alpha = 0.75) +
  geom_text_repel(
    aes(label = country),
    size = 2.5, max.overlaps = 25, segment.alpha = 0.3
  ) +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme(legend.position = "bottom")

ggsave("imgs/umap_country_space_colocation.png", width = 9, height = 9, dpi = 300)
```

## Spearman Institution Space (UMAP)

```{r umap-institutions-spearman, cache=TRUE}
inst_dist_sp <- as.dist(1 - spearman_pp_01)

set.seed(42)
mds_inst_sp <- cmdscale(inst_dist_sp, k = min(10, ncol(Mcp) - 1))
umap_inst_sp <- umap(mds_inst_sp, n_neighbors = 15, min_dist = 0.1, n_epochs = 500)

umap_inst_sp_df <- tibble(
  indicator = colnames(Mcp),
  UMAP1 = umap_inst_sp[, 1],
  UMAP2 = umap_inst_sp[, 2],
  ubiquity = kp0,
  IFC = pci_eigen
) %>%
  left_join(indicator_info, by = c("indicator" = "code")) %>%
  mutate(label = str_trunc(description, 35) %>% str_wrap(width = 20))
```

## Clustering the Spearman Institution Space

The Spearman embedding is more compact and symmetric than the co-location one, so it is a
better substrate for unsupervised cluster discovery. We run k-means directly on the 2D UMAP
coordinates and use the elbow (within-cluster sum of squares) and silhouette width to pick
the number of clusters.

```{r kmeans-inst-spearman-diagnostics, cache=TRUE, fig.width=6.5, fig.height=3}
library(cluster)

coords_sp <- umap_inst_sp[, 1:2]

ks <- 2:12
cluster_diag <- map_dfr(ks, function(k) {
  set.seed(1)
  km <- kmeans(coords_sp, centers = k, nstart = 50, iter.max = 200)
  sil <- silhouette(km$cluster, dist(coords_sp))
  tibble(k = k, wss = km$tot.withinss, sil = mean(sil[, 3]))
})

p_wss <- ggplot(cluster_diag, aes(k, wss)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = ks) +
  labs(x = "k", y = "Within-cluster SS")
p_sil <- ggplot(cluster_diag, aes(k, sil)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = ks) +
  labs(x = "k", y = "Mean silhouette width")

p_wss + p_sil

save_fig("full", "kmeans_diagnostics.pdf")
```

The silhouette width peaks at small k (best at k=3), while the WSS shows a soft elbow around
k=3--4. We pick **k = 6** — it trades a modest drop in silhouette for a richer partition that
separates conceptually distinct clusters (e.g., judicial rule-of-law features from
deliberative/consultative features) that get lumped together at lower k.

```{r kmeans-inst-spearman-fit}
set.seed(1)
N_CLUSTERS <- 6
km_inst <- kmeans(coords_sp, centers = N_CLUSTERS, nstart = 50, iter.max = 200)

# Map kmeans cluster IDs -> interpretive labels via landmark indicators.
# (k-means IDs are arbitrary; using landmark codes makes label assignment
# stable even if the ID ordering shifts.)
landmark_labels <- tribble(
  ~code,                  ~label,
  "v2dlreason",           "Deliberation & Civic Voice",
  "v2excrptps",           "Clean & Impartial State",
  "v2juhcind",            "Rule of Law & Checks on Power",
  "v2elmulpar",           "Inclusive Electoral Democracy",
  "v2regoppgroupssize",   "Executive Control & Mobilization",
  "v2mecenefi",           "Repression & Censorship"
)

cluster_id_to_label <- tibble(
  indicator = colnames(Mcp),
  cluster = km_inst$cluster
) %>%
  inner_join(landmark_labels, by = c("indicator" = "code")) %>%
  distinct(cluster, label) %>%
  arrange(cluster)

stopifnot(nrow(cluster_id_to_label) == N_CLUSTERS)

cluster_labels <- setNames(cluster_id_to_label$label,
                           as.character(cluster_id_to_label$cluster))

umap_inst_sp_df <- umap_inst_sp_df %>%
  mutate(
    cluster = km_inst$cluster,
    cluster_label = factor(cluster_labels[as.character(cluster)],
                           levels = landmark_labels$label)
  )

# Summary table: one row per cluster, showing size and top categories
cluster_summary <- umap_inst_sp_df %>%
  group_by(cluster, cluster_label) %>%
  summarise(
    n = n(),
    top_categories = paste(
      head(names(sort(table(category), decreasing = TRUE)), 3),
      collapse = ", "
    ),
    .groups = "drop"
  ) %>%
  arrange(cluster)

kable(cluster_summary,
      col.names = c("Cluster", "Label", "n", "Top V-Dem categories"),
      caption = "K-means clusters (k = 6) on the Spearman institution UMAP")
```

### Cluster membership

For each cluster we list every institutional feature it contains:

```{r kmeans-inst-spearman-members, results='asis'}
for (cl in sort(unique(umap_inst_sp_df$cluster))) {
  members <- umap_inst_sp_df %>% filter(cluster == cl)
  cat("\n\n**Cluster", cl, "—", cluster_labels[as.character(cl)],
      "** (n = ", nrow(members), ")\n\n", sep = "")
  members %>%
    arrange(category, description) %>%
    select(Code = indicator, Category = category, Description = description) %>%
    kable() %>%
    print()
}
```

### Cluster-colored Institution Space

Replacing the V-Dem category coloring with the discovered clusters. A bold label marks
each cluster's centroid.

```{r umap-inst-spearman-cluster-plot, fig.width=6.5, fig.height=6.5}
# Cluster centroids for label placement
cluster_centroids <- umap_inst_sp_df %>%
  group_by(cluster_label) %>%
  summarise(
    UMAP1 = median(UMAP1),
    UMAP2 = median(UMAP2),
    .groups = "drop"
  )

cluster_colors <- colorRampPalette(gl_palettes$categorical)(N_CLUSTERS)
names(cluster_colors) <- levels(umap_inst_sp_df$cluster_label)

ggplot(umap_inst_sp_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = cluster_label), size = 2.5, alpha = 0.75,
             show.legend = FALSE) +
  geom_text_repel(
    aes(label = label, color = cluster_label),
    size = 1.8, max.overlaps = 20, segment.alpha = 0.3, lineheight = 0.85,
    show.legend = FALSE
  ) +
  geom_label(
    data = cluster_centroids,
    aes(label = cluster_label, fill = cluster_label),
    size = 3.2, fontface = "bold", color = "white",
    label.padding = unit(0.35, "lines"), label.r = unit(0.25, "lines"),
    show.legend = FALSE
  ) +
  scale_color_manual(values = cluster_colors) +
  scale_fill_manual(values = cluster_colors) +
  labs(x = "UMAP 1", y = "UMAP 2")

ggsave("imgs/umap_institution_space_spearman_clusters.png",
       width = 9, height = 9, dpi = 300)
```

## Spearman Country Space (UMAP)

```{r umap-countries-spearman, cache=TRUE}
set.seed(42)
country_dist_sp <- as.dist(1 - spearman_cc_01)
mds_country_sp <- cmdscale(country_dist_sp, k = min(10, nrow(Mcp) - 1))
umap_country_sp <- umap(mds_country_sp, n_neighbors = 15, min_dist = 0.2, n_epochs = 500)

umap_country_sp_df <- tibble(
  country = rownames(Mcp),
  UMAP1 = umap_country_sp[, 1],
  UMAP2 = umap_country_sp[, 2],
  ICI = eci_eigen,
  diversity = kc0
) %>%
  mutate(
    region = region_labels[as.character(region_lookup[country])],
    region = replace_na(region, "Other")
  )
```

```{r umap-country-spearman-plot, fig.width=6.5, fig.height=6.5}
ggplot(umap_country_sp_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = region), size = 2.5, alpha = 0.75) +
  geom_text_repel(
    aes(label = country),
    size = 2.5, max.overlaps = 25, segment.alpha = 0.3
  ) +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme(legend.position = "bottom")

ggsave("imgs/umap_country_space_spearman.png", width = 9, height = 9, dpi = 300)
```

## Clustering the Spearman Country Space

We repeat the same k-means exercise on the country UMAP. Because k-means clusters are just
groups of points in embedding space, we can then look *back* at the raw Mcp matrix and ask:
for each cluster of countries, which institutional features do they share — and which are
they collectively deficient on?

```{r kmeans-country-spearman-diagnostics, cache=TRUE, fig.width=6.5, fig.height=3}
coords_country_sp <- umap_country_sp[, 1:2]

country_cluster_diag <- map_dfr(2:12, function(k) {
  set.seed(1)
  km <- kmeans(coords_country_sp, centers = k, nstart = 50, iter.max = 200)
  sil <- silhouette(km$cluster, dist(coords_country_sp))
  tibble(k = k, wss = km$tot.withinss, sil = mean(sil[, 3]))
})

p_wss_c <- ggplot(country_cluster_diag, aes(k, wss)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 2:12) +
  labs(x = "k", y = "Within-cluster SS")
p_sil_c <- ggplot(country_cluster_diag, aes(k, sil)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 2:12) +
  labs(x = "k", y = "Mean silhouette width")

p_wss_c + p_sil_c

save_fig("full", "kmeans_country_diagnostics.pdf")
```

The silhouette is highest at k=2 (a coarse democracy/autocracy split), with a secondary
plateau around k=6--7. We pick **k = 6** — it surfaces a richer taxonomy (closed autocracies,
conflict states, consolidating democracies, etc.) while still being interpretable.

```{r kmeans-country-spearman-fit}
set.seed(1)
N_CLUSTERS_C <- 6
km_country <- kmeans(coords_country_sp, centers = N_CLUSTERS_C,
                     nstart = 50, iter.max = 200)

# Landmark-country labelling: each landmark uniquely names a cluster.
country_landmark_labels <- tribble(
  ~country,      ~label,
  "Sweden",      "Advanced Liberal Democracies",
  "Poland",      "Consolidating Democracies",
  "Brazil",      "Contentious Electoral Democracies",
  "China",       "Closed Autocracies",
  "Haiti",       "Conflict & Breakdown States",
  "Mauritius",   "Small Open Democracies"
)

country_cluster_map <- tibble(
  country = rownames(Mcp),
  cluster = km_country$cluster
) %>%
  inner_join(country_landmark_labels, by = "country") %>%
  distinct(cluster, label) %>%
  arrange(cluster)

stopifnot(nrow(country_cluster_map) == N_CLUSTERS_C)

country_cluster_labels <- setNames(country_cluster_map$label,
                                   as.character(country_cluster_map$cluster))

umap_country_sp_df <- umap_country_sp_df %>%
  mutate(
    cluster = km_country$cluster,
    cluster_label = factor(
      country_cluster_labels[as.character(cluster)],
      levels = country_landmark_labels$label
    )
  )

# Summary: cluster size + representative members
country_cluster_summary <- umap_country_sp_df %>%
  group_by(cluster, cluster_label) %>%
  summarise(
    n = n(),
    representative = paste(head(country, 6), collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(cluster)

kable(country_cluster_summary,
      col.names = c("Cluster", "Label", "n", "First 6 members"),
      caption = "K-means clusters (k = 6) on the Spearman country UMAP")
```

### What each country cluster has (and lacks) in common

For every cluster we compute each institutional feature's mean across its members and compare
to the global mean (z-score vs. the global SD). The top positive z's are the cluster's shared
**distinctive strengths**; the top negatives are shared **distinctive weaknesses**.

```{r kmeans-country-cluster-features, results='asis'}
country_global_means <- colMeans(Mcp)
country_global_sds <- apply(Mcp, 2, sd)

indicator_desc <- setNames(indicator_info$description, indicator_info$code)

for (cl in sort(unique(umap_country_sp_df$cluster))) {
  cl_label <- country_cluster_labels[as.character(cl)]
  cl_members <- umap_country_sp_df %>% filter(cluster == cl) %>% pull(country)

  cat("\n\n**Cluster", cl, "—", cl_label,
      "** (n = ", length(cl_members), ")\n\n", sep = "")

  cl_means <- colMeans(Mcp[cl_members, , drop = FALSE])
  z <- (cl_means - country_global_means) / country_global_sds

  top_high <- sort(z, decreasing = TRUE) %>% head(8)
  top_low  <- sort(z, decreasing = FALSE) %>% head(8)

  cat("_Members:_ ", paste(cl_members, collapse = ", "), "\n\n", sep = "")

  high_tbl <- tibble(
    Feature = indicator_desc[names(top_high)],
    Code = names(top_high),
    `z vs global` = round(unname(top_high), 2)
  )
  low_tbl <- tibble(
    Feature = indicator_desc[names(top_low)],
    Code = names(top_low),
    `z vs global` = round(unname(top_low), 2)
  )

  cat("\n_Shared strengths:_\n\n")
  print(kable(high_tbl))
  cat("\n_Shared weaknesses:_\n\n")
  print(kable(low_tbl))
}
```

### Cluster-colored Country Space

```{r umap-country-spearman-cluster-plot, fig.width=6.5, fig.height=6.5}
country_centroids <- umap_country_sp_df %>%
  group_by(cluster_label) %>%
  summarise(UMAP1 = median(UMAP1), UMAP2 = median(UMAP2), .groups = "drop")

country_cluster_colors <- colorRampPalette(gl_palettes$categorical)(N_CLUSTERS_C)
names(country_cluster_colors) <- levels(umap_country_sp_df$cluster_label)

ggplot(umap_country_sp_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = cluster_label), size = 2.5, alpha = 0.75,
             show.legend = FALSE) +
  geom_text_repel(
    aes(label = country, color = cluster_label),
    size = 2.5, max.overlaps = 25, segment.alpha = 0.3,
    show.legend = FALSE
  ) +
  geom_label(
    data = country_centroids,
    aes(label = cluster_label, fill = cluster_label),
    size = 3.2, fontface = "bold", color = "white",
    label.padding = unit(0.35, "lines"), label.r = unit(0.25, "lines"),
    show.legend = FALSE
  ) +
  scale_color_manual(values = country_cluster_colors) +
  scale_fill_manual(values = country_cluster_colors) +
  labs(x = "UMAP 1", y = "UMAP 2")

ggsave("imgs/umap_country_space_spearman_clusters.png",
       width = 9, height = 9, dpi = 300)
```

### Economic Complexity Overlay

Do the institutional clusters line up with economic complexity? We recolor the same
country UMAP by each country's latest Atlas SITC **ECI**, keeping the cluster labels
in place. Clusters that line up with a single color band share an ECI regime;
clusters that span the gradient contain institutional archetypes spread across the
complexity distribution.

```{r umap-country-eci-plot, fig.width=6.5, fig.height=6.5}
iso3_lookup <- vdem_filtered %>%
  select(country_name, country_text_id) %>%
  deframe()

eci_latest <- arrow::read_parquet(
  "data/shared-data/growth-lab/glmacro_master_alldata.parquet",
  col_select = c("countrycodeiso", "year", "atl_sitc_eci")
) %>%
  filter(!is.na(atl_sitc_eci)) %>%
  group_by(countrycodeiso) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(iso3 = countrycodeiso, eci = atl_sitc_eci, eci_year = year)

umap_eci_df <- umap_country_sp_df %>%
  mutate(iso3 = unname(iso3_lookup[country])) %>%
  left_join(eci_latest, by = "iso3")

cat("Countries without matching ECI:",
    sum(is.na(umap_eci_df$eci)), "of", nrow(umap_eci_df), "\n")

ggplot(umap_eci_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = eci), size = 2.8, alpha = 0.9) +
  geom_text_repel(
    aes(label = country),
    size = 2.5, max.overlaps = 25, segment.alpha = 0.3
  ) +
  geom_label(
    data = country_centroids,
    aes(label = cluster_label, fill = cluster_label),
    size = 3.2, fontface = "bold", color = "white",
    label.padding = unit(0.35, "lines"), label.r = unit(0.25, "lines"),
    show.legend = FALSE
  ) +
  scale_color_gradient2(
    low = highlight, mid = "grey85", high = gl$brand_blue,
    midpoint = 0, name = "ECI", na.value = "grey75"
  ) +
  scale_fill_manual(values = country_cluster_colors) +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme(legend.position = "bottom")

ggsave("imgs/umap_country_space_spearman_eci.png",
       width = 9, height = 9, dpi = 300)
```

## Z-scored IRT Institution Space (UMAP)

We mirror the country-space logic for institutions. Each institution is a vector in
country-space (one entry per country); two institutions are close if they vary similarly
across countries. To make every country contribute equally to inter-institution distance,
we z-score across countries (i.e., scale the columns of the transposed matrix). The
imputation step from the country pipeline produces `Mcp_raw` and is reused here.

```{r institution-zscore-matrix, cache=TRUE}
# Raw IRT country × feature matrix (no min-max normalization)
Mcp_raw <- vdem_filtered %>%
  select(country_name, all_of(indicator_info$code)) %>%
  column_to_rownames("country_name") %>%
  as.matrix()

# Impute NAs with column mean (IRT scores are roughly mean-centered globally,
# so a within-sample column mean is a reasonable neutral imputation)
for (j in seq_len(ncol(Mcp_raw))) {
  na_idx <- is.na(Mcp_raw[, j])
  if (any(na_idx)) {
    Mcp_raw[na_idx, j] <- mean(Mcp_raw[, j], na.rm = TRUE)
  }
}

# Institution × country matrix; z-score each country column so each country
# contributes equally to distances between institutions
Mpc_z <- scale(t(Mcp_raw), center = TRUE, scale = TRUE)

cat("Institution z-scored matrix:", nrow(Mpc_z), "features x", ncol(Mpc_z), "countries\n")
cat("Per-country SD after scaling (should be 1):",
    round(mean(apply(Mpc_z, 2, sd)), 3), "\n")
```

```{r umap-institution-zscore, cache=TRUE}
set.seed(42)
umap_inst_z <- umap(Mpc_z, n_neighbors = 15, min_dist = 0.1, n_epochs = 500)

umap_inst_z_df <- tibble(
  indicator = rownames(Mpc_z),
  UMAP1 = umap_inst_z[, 1],
  UMAP2 = umap_inst_z[, 2]
) %>%
  left_join(indicator_info, by = c("indicator" = "code"))
```

```{r umap-institution-zscore-plot, fig.width=6.5, fig.height=6.5}
n_cats_z <- length(unique(umap_inst_z_df$category))
cat_colors_z <- colorRampPalette(gl_palettes$categorical)(n_cats_z)

umap_inst_z_df <- umap_inst_z_df %>%
  mutate(label = str_trunc(description, 35) %>% str_wrap(width = 20))

ggplot(umap_inst_z_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = category), size = 2.5, alpha = 0.75) +
  geom_text_repel(
    aes(label = label),
    size = 1.8, max.overlaps = 20, segment.alpha = 0.3, lineheight = 0.85
  ) +
  scale_color_manual(values = cat_colors_z, name = "Category") +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme(legend.position = "bottom")

ggsave("imgs/umap_institution_space_zscored.png", width = 9, height = 9, dpi = 300)
```

## Z-scored IRT Country Space (UMAP)

The Spearman pipeline operates on min-max normalized scores and rank correlations, which
discards the "1 unit ~= 1 SD" property of V-Dem's IRT measurement model. As an alternative,
we build a country embedding directly on the raw IRT scores: impute missing values with
the feature mean, z-score each indicator across the country sample (so every feature
contributes equally to Euclidean distance), and feed the coordinates straight to UMAP.
No MDS step is needed — the input is already Euclidean.

```{r country-zscore-matrix, cache=TRUE}
# Z-score each feature across countries: 1 unit = 1 sample SD on that indicator
Mcp_z <- scale(Mcp_raw, center = TRUE, scale = TRUE)

cat("Z-scored matrix:", nrow(Mcp_z), "countries x", ncol(Mcp_z), "features\n")
cat("Per-feature SD after scaling (should be 1):",
    round(mean(apply(Mcp_z, 2, sd)), 3), "\n")
```

```{r umap-country-zscore, cache=TRUE}
set.seed(42)
umap_country_z <- umap(Mcp_z, n_neighbors = 15, min_dist = 0.2, n_epochs = 500)

umap_country_z_df <- tibble(
  country = rownames(Mcp_z),
  UMAP1 = umap_country_z[, 1],
  UMAP2 = umap_country_z[, 2]
) %>%
  mutate(
    region = region_labels[as.character(region_lookup[country])],
    region = replace_na(region, "Other")
  )
```

```{r umap-country-zscore-plot, fig.width=6.5, fig.height=6.5}
ggplot(umap_country_z_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = region), size = 2.5, alpha = 0.75) +
  geom_text_repel(
    aes(label = country),
    size = 2.5, max.overlaps = 25, segment.alpha = 0.3
  ) +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme(legend.position = "bottom")

ggsave("imgs/umap_country_space_zscored.png", width = 9, height = 9, dpi = 300)
```

## Clustering the Z-scored Country Space

K-means runs on the **full z-scored matrix** (`Mcp_z`), not on the 2D UMAP coordinates.
The UMAP is for visualization; clusters should reflect the actual high-dimensional feature
space. We use the elbow + silhouette diagnostic across k = 2..12 to pick the cluster count.

```{r kmeans-country-zscore-diagnostics, cache=TRUE, fig.width=6.5, fig.height=3}
country_z_diag <- map_dfr(2:12, function(k) {
  set.seed(1)
  km <- kmeans(Mcp_z, centers = k, nstart = 50, iter.max = 200)
  sil <- silhouette(km$cluster, dist(Mcp_z))
  tibble(k = k, wss = km$tot.withinss, sil = mean(sil[, 3]))
})

p_wss_z <- ggplot(country_z_diag, aes(k, wss)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 2:12) +
  labs(x = "k", y = "Within-cluster SS")
p_sil_z <- ggplot(country_z_diag, aes(k, sil)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 2:12) +
  labs(x = "k", y = "Mean silhouette width")

p_wss_z + p_sil_z
save_fig("full", "kmeans_country_zscore_diagnostics.pdf")
```

```{r kmeans-country-zscore-fit, cache=TRUE}
set.seed(1)
N_CLUSTERS_Z <- 6
km_country_z <- kmeans(Mcp_z, centers = N_CLUSTERS_Z, nstart = 50, iter.max = 200)

umap_country_z_df <- umap_country_z_df %>%
  mutate(cluster_z = km_country_z$cluster)

# Per-cluster z-deviation: since Mcp_z is already centered/scaled across countries,
# colMeans on cluster members IS the z-score deviation from the global mean.
cluster_summaries_z <- map(sort(unique(km_country_z$cluster)), function(cl) {
  members <- umap_country_z_df %>% filter(cluster_z == cl) %>% pull(country)
  z <- colMeans(Mcp_z[members, , drop = FALSE])
  high_idx <- order(z, decreasing = TRUE)[1:12]
  low_idx  <- order(z, decreasing = FALSE)[1:12]
  list(
    cluster = cl,
    n = length(members),
    members = members,
    top_strengths = tibble(
      code = names(z)[high_idx],
      description = indicator_desc[names(z)[high_idx]],
      z = unname(z[high_idx])
    ),
    top_weaknesses = tibble(
      code = names(z)[low_idx],
      description = indicator_desc[names(z)[low_idx]],
      z = unname(z[low_idx])
    )
  )
})
names(cluster_summaries_z) <- as.character(sort(unique(km_country_z$cluster)))

# Persist for downstream labeling
saveRDS(cluster_summaries_z, "scratch/cluster_summaries_z.rds")
```

```{r kmeans-country-zscore-print, results='asis'}
for (cl in sort(unique(km_country_z$cluster))) {
  s <- cluster_summaries_z[[as.character(cl)]]
  cat("\n\n**Cluster ", cl, "** (n = ", s$n, ")\n\n", sep = "")
  cat("_Members:_ ", paste(s$members, collapse = ", "), "\n\n", sep = "")
  cat("\n_Top distinctive strengths (z vs global):_\n\n")
  print(kable(s$top_strengths, digits = 2))
  cat("\n_Top distinctive weaknesses (z vs global):_\n\n")
  print(kable(s$top_weaknesses, digits = 2))
}
```

### Cluster labels

Labels were generated by inspecting each cluster's distinctive strengths/weaknesses and
member list (the tables above). They emphasize **institutional type**, not regime quality
alone — e.g. clusters 4 and 6 are both closed regimes, but cluster 4 is distinguished by
high state capacity (especially digital surveillance/control infrastructure) while cluster
6 reads as personalist/totalitarian with extreme repression and no civil society.

```{r kmeans-country-zscore-labels}
country_z_cluster_labels <- c(
  "1" = "Conflict-Prone Repressive Regimes",
  "2" = "Consolidated Liberal Democracies",
  "3" = "Contentious Hybrid Regimes",
  "4" = "Closed Capable Autocracies",
  "5" = "Stable Middle Democracies",
  "6" = "Personalist / Totalitarian Regimes"
)

umap_country_z_df <- umap_country_z_df %>%
  mutate(
    cluster_label_z = factor(
      country_z_cluster_labels[as.character(cluster_z)],
      levels = unname(country_z_cluster_labels)
    )
  )

# Verify all clusters labelled
stopifnot(!any(is.na(umap_country_z_df$cluster_label_z)))
```

### Cluster-colored Country Space

Same UMAP coordinates as before, recolored by cluster, with the cluster label drawn at
each cluster's centroid (median UMAP position).

```{r umap-country-zscore-cluster-plot, fig.width=6.5, fig.height=6.5}
country_z_centroids <- umap_country_z_df %>%
  group_by(cluster_label_z) %>%
  summarise(UMAP1 = median(UMAP1), UMAP2 = median(UMAP2), .groups = "drop")

country_z_cluster_colors <- colorRampPalette(gl_palettes$categorical)(N_CLUSTERS_Z)
names(country_z_cluster_colors) <- levels(umap_country_z_df$cluster_label_z)

ggplot(umap_country_z_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = cluster_label_z), size = 2.5, alpha = 0.75,
             show.legend = FALSE) +
  geom_text_repel(
    aes(label = country, color = cluster_label_z),
    size = 2.5, max.overlaps = 25, segment.alpha = 0.3,
    show.legend = FALSE
  ) +
  geom_label(
    data = country_z_centroids,
    aes(label = cluster_label_z, fill = cluster_label_z),
    size = 3.2, fontface = "bold", color = "white",
    label.padding = unit(0.35, "lines"), label.r = unit(0.25, "lines"),
    show.legend = FALSE
  ) +
  scale_color_manual(values = country_z_cluster_colors) +
  scale_fill_manual(values = country_z_cluster_colors) +
  labs(x = "UMAP 1", y = "UMAP 2")

ggsave("imgs/umap_country_space_zscored_clusters.png",
       width = 9, height = 9, dpi = 300)
```

## Clustering on the 2D UMAP coordinates

For comparison, we re-run k-means on the **2D UMAP coordinates** (`UMAP1`, `UMAP2`) rather
than the full `Mcp_z` matrix. UMAP preserves local structure but distorts global gaps, so
the clusters discovered here are the visible point-blobs in the chart. We keep k = 6 and
the same downstream labeling pipeline.

```{r kmeans-country-zscore-2d-diagnostics, cache=TRUE, fig.width=6.5, fig.height=3}
coords_z_2d <- as.matrix(umap_country_z_df[, c("UMAP1", "UMAP2")])

country_z_2d_diag <- map_dfr(2:12, function(k) {
  set.seed(1)
  km <- kmeans(coords_z_2d, centers = k, nstart = 50, iter.max = 200)
  sil <- silhouette(km$cluster, dist(coords_z_2d))
  tibble(k = k, wss = km$tot.withinss, sil = mean(sil[, 3]))
})

p_wss_z_2d <- ggplot(country_z_2d_diag, aes(k, wss)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 2:12) +
  labs(x = "k", y = "Within-cluster SS")
p_sil_z_2d <- ggplot(country_z_2d_diag, aes(k, sil)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 2:12) +
  labs(x = "k", y = "Mean silhouette width")

p_wss_z_2d + p_sil_z_2d
save_fig("full", "kmeans_country_zscore_2d_diagnostics.pdf")
```

```{r kmeans-country-zscore-2d-fit, cache=TRUE}
set.seed(1)
N_CLUSTERS_Z_2D <- 6
km_country_z_2d <- kmeans(coords_z_2d, centers = N_CLUSTERS_Z_2D,
                          nstart = 50, iter.max = 200)

umap_country_z_df <- umap_country_z_df %>%
  mutate(cluster_z_2d = km_country_z_2d$cluster)

# Cluster summaries: even though clustering ran on UMAP coords, we still characterize
# clusters by their distinctive Mcp_z features so labels remain interpretable.
cluster_summaries_z_2d <- map(sort(unique(km_country_z_2d$cluster)), function(cl) {
  members <- umap_country_z_df %>% filter(cluster_z_2d == cl) %>% pull(country)
  z <- colMeans(Mcp_z[members, , drop = FALSE])
  high_idx <- order(z, decreasing = TRUE)[1:12]
  low_idx  <- order(z, decreasing = FALSE)[1:12]
  list(
    cluster = cl,
    n = length(members),
    members = members,
    top_strengths = tibble(
      code = names(z)[high_idx],
      description = indicator_desc[names(z)[high_idx]],
      z = unname(z[high_idx])
    ),
    top_weaknesses = tibble(
      code = names(z)[low_idx],
      description = indicator_desc[names(z)[low_idx]],
      z = unname(z[low_idx])
    )
  )
})
names(cluster_summaries_z_2d) <- as.character(sort(unique(km_country_z_2d$cluster)))

saveRDS(cluster_summaries_z_2d, "scratch/cluster_summaries_z_2d.rds")
```

```{r kmeans-country-zscore-2d-print, results='asis'}
for (cl in sort(unique(km_country_z_2d$cluster))) {
  s <- cluster_summaries_z_2d[[as.character(cl)]]
  cat("\n\n**Cluster ", cl, "** (n = ", s$n, ")\n\n", sep = "")
  cat("_Members:_ ", paste(s$members, collapse = ", "), "\n\n", sep = "")
  cat("\n_Top distinctive strengths (z vs global):_\n\n")
  print(kable(s$top_strengths, digits = 2))
  cat("\n_Top distinctive weaknesses (z vs global):_\n\n")
  print(kable(s$top_weaknesses, digits = 2))
}
```

### Cluster labels (2D UMAP clustering)

The 2D-UMAP partition shifts the taxonomy in two visible ways relative to the full-matrix
version: it **merges** capable autocracies (China, Russia, Saudi) with personalist/total
regimes (North Korea, Belarus, Eritrea) into one closed-autocracies cluster — they sit
adjacent in the embedding, so 2D k-means can't separate them — and it **splits** the
broad "stable middle democracies" group along regional/structural lines, surfacing a
distinct polarized post-communist + Latin American cluster.

```{r kmeans-country-zscore-2d-labels}
country_z_2d_cluster_labels <- c(
  "1" = "Developing / Mid-tier Democracies",
  "2" = "Fragile / Conflict-Prone States",
  "3" = "Small Stable Democracies",
  "4" = "Polarized Post-Communist & Latin Democracies",
  "5" = "Closed Autocracies",
  "6" = "Consolidated Liberal Democracies"
)

umap_country_z_df <- umap_country_z_df %>%
  mutate(
    cluster_label_z_2d = factor(
      country_z_2d_cluster_labels[as.character(cluster_z_2d)],
      levels = unname(country_z_2d_cluster_labels)
    )
  )

stopifnot(!any(is.na(umap_country_z_df$cluster_label_z_2d)))
```

### Cluster-colored Country Space (2D UMAP clustering)

```{r umap-country-zscore-2d-cluster-plot, fig.width=6.5, fig.height=6.5}
country_z_2d_centroids <- umap_country_z_df %>%
  group_by(cluster_label_z_2d) %>%
  summarise(UMAP1 = median(UMAP1), UMAP2 = median(UMAP2), .groups = "drop")

country_z_2d_cluster_colors <- colorRampPalette(gl_palettes$categorical)(N_CLUSTERS_Z_2D)
names(country_z_2d_cluster_colors) <- levels(umap_country_z_df$cluster_label_z_2d)

ggplot(umap_country_z_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = cluster_label_z_2d), size = 2.5, alpha = 0.75,
             show.legend = FALSE) +
  geom_text_repel(
    aes(label = country, color = cluster_label_z_2d),
    size = 2.5, max.overlaps = 25, segment.alpha = 0.3,
    show.legend = FALSE
  ) +
  geom_label(
    data = country_z_2d_centroids,
    aes(label = cluster_label_z_2d, fill = cluster_label_z_2d),
    size = 3.2, fontface = "bold", color = "white",
    label.padding = unit(0.35, "lines"), label.r = unit(0.25, "lines"),
    show.legend = FALSE
  ) +
  scale_color_manual(values = country_z_2d_cluster_colors) +
  scale_fill_manual(values = country_z_2d_cluster_colors) +
  labs(x = "UMAP 1", y = "UMAP 2")

ggsave("imgs/umap_country_space_zscored_clusters_2d.png",
       width = 9, height = 9, dpi = 300)
```

## Mahalanobis IRT Country Space (UMAP)

The z-scored space treats each indicator as one independent unit of variation, but V-Dem
ships families of correlated indicators (multiple measures of judicial independence,
electoral fairness, etc.) that triangulate latent concepts. Mahalanobis distance whitens
with the feature covariance — equivalently, projects onto principal components and rescales
each to unit variance. We **truncate to the top-k PCs** rather than going full rank: with
~180 countries the full covariance is ill-conditioned, and full-rank Mahalanobis lifts
noisy tail components up to the same weight as PC1, which we don't want.

```{r country-mahalanobis-matrix, cache=TRUE}
# PCA on the z-scored country × feature matrix.
# (scale.=TRUE on Mcp_raw is identical to PCA on Mcp_z.)
pca_country_z <- prcomp(Mcp_raw, center = TRUE, scale. = TRUE)

var_cum_z <- summary(pca_country_z)$importance[3, ]
# Truncate at 80% variance, capped at 10 PCs to stay in the high-signal regime.
K_PCS <- min(which(var_cum_z >= 0.80)[1], 10)

cat("Cumulative variance, PC1..15:\n")
print(round(var_cum_z[1:min(15, length(var_cum_z))], 3))
cat("\nKeeping K =", K_PCS, "PCs (",
    round(100 * var_cum_z[K_PCS], 1), "% of variance)\n")

# Whitened coordinates: PC scores divided by sqrt(eigenvalue) so each retained
# direction has unit variance. Euclidean distance in this space equals
# Mahalanobis distance restricted to the top-K_PCS subspace.
sds <- pca_country_z$sdev[1:K_PCS]
Mcp_mahal <- sweep(pca_country_z$x[, 1:K_PCS], 2, sds, "/")

cat("Mahalanobis coords:", nrow(Mcp_mahal), "countries x",
    ncol(Mcp_mahal), "whitened PCs\n")
```

```{r umap-country-mahalanobis, cache=TRUE}
set.seed(42)
umap_country_m <- umap(Mcp_mahal, n_neighbors = 15, min_dist = 0.2, n_epochs = 500)

umap_country_m_df <- tibble(
  country = rownames(Mcp_mahal),
  UMAP1 = umap_country_m[, 1],
  UMAP2 = umap_country_m[, 2]
) %>%
  mutate(
    region = region_labels[as.character(region_lookup[country])],
    region = replace_na(region, "Other")
  )
```

```{r umap-country-mahalanobis-plot, fig.width=6.5, fig.height=6.5}
ggplot(umap_country_m_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = region), size = 2.5, alpha = 0.75) +
  geom_text_repel(
    aes(label = country),
    size = 2.5, max.overlaps = 25, segment.alpha = 0.3
  ) +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme(legend.position = "bottom")

ggsave("imgs/umap_country_space_mahalanobis.png", width = 9, height = 9, dpi = 300)
```

### Mahalanobis vs. z-scored: how much does whitening change neighborhoods?

Mantel-style Spearman correlation between pairwise distances under the two metrics, both
in the high-dimensional input space and in the resulting 2D UMAPs. High agreement means
the indicator bundling isn't actually distorting the z-scored space; low agreement means
whitening is doing real work.

```{r mahalanobis-vs-zscore-compare}
stopifnot(identical(rownames(Mcp_z), rownames(Mcp_mahal)))

d_z     <- as.matrix(dist(Mcp_z))
d_mahal <- as.matrix(dist(Mcp_mahal))
upper   <- upper.tri(d_z)
cat("High-dim distance correlation (z-scored vs Mahalanobis, Spearman):",
    round(cor(d_z[upper], d_mahal[upper], method = "spearman"), 3), "\n")

zc <- umap_country_z_df %>% arrange(country)
mc <- umap_country_m_df %>% arrange(country)
stopifnot(identical(zc$country, mc$country))

d_z_umap     <- as.matrix(dist(zc[, c("UMAP1", "UMAP2")]))
d_mahal_umap <- as.matrix(dist(mc[, c("UMAP1", "UMAP2")]))
upper_u <- upper.tri(d_z_umap)
cat("2D UMAP distance correlation (z-scored vs Mahalanobis, Spearman):",
    round(cor(d_z_umap[upper_u], d_mahal_umap[upper_u], method = "spearman"), 3), "\n")
```

## Alternative Clustering: Original Features vs. PCA vs. UMAP

The clusters above were fit on the 2D UMAP coordinates — i.e., k-means is partitioning the
points you actually see in the chart. UMAP preserves local neighborhood structure but distorts
global gaps, and the embedding inherits stochasticity from its own hyperparameters. To check
how much that choice is driving the results, we re-fit k-means with the same K under two
alternatives:

- **Original features**: k-means directly on the rows of `Mcp` (countries) or its transpose
  (institutions). Distances are Euclidean in the full institutional-feature space.
- **PCA scores**: k-means on the leading principal components covering 80% of variance.
  This denoises the high-dim space without forcing it into 2D.

We compare the three partitions using the **Adjusted Rand Index** (ARI): 1 = identical
clusters (up to relabelling), 0 = chance agreement, negative = worse than chance.

```{r alt-clustering-fit}
adjustedRandIndex <- function(a, b) {
  tab <- table(a, b)
  n  <- sum(tab)
  ni <- rowSums(tab); nj <- colSums(tab)
  choose2 <- function(x) x * (x - 1) / 2
  sum_ij <- sum(choose2(tab))
  sum_i  <- sum(choose2(ni)); sum_j <- sum(choose2(nj))
  total  <- choose2(n)
  expected <- sum_i * sum_j / total
  max_idx  <- (sum_i + sum_j) / 2
  if (max_idx == expected) return(0)
  (sum_ij - expected) / (max_idx - expected)
}

# Country space: PCA scores already exist (pca_result on Mcp)
var_cum_c <- summary(pca_result)$importance[3, ]
n_pc_country <- which(var_cum_c >= 0.80)[1]
country_pcs <- pca_result$x[, 1:n_pc_country]

# Institution space: PCA on the transposed matrix (indicators as rows)
pca_inst <- prcomp(t(Mcp), center = TRUE, scale. = TRUE)
var_cum_i <- summary(pca_inst)$importance[3, ]
n_pc_inst <- which(var_cum_i >= 0.80)[1]
inst_pcs <- pca_inst$x[, 1:n_pc_inst]

set.seed(1)
km_country_orig <- kmeans(Mcp, centers = N_CLUSTERS_C,
                          nstart = 50, iter.max = 200)
set.seed(1)
km_country_pca  <- kmeans(country_pcs, centers = N_CLUSTERS_C,
                          nstart = 50, iter.max = 200)
set.seed(1)
km_inst_orig    <- kmeans(t(Mcp), centers = N_CLUSTERS,
                          nstart = 50, iter.max = 200)
set.seed(1)
km_inst_pca     <- kmeans(inst_pcs, centers = N_CLUSTERS,
                          nstart = 50, iter.max = 200)

ari_country <- tibble(
  Comparison = c("UMAP vs Original features",
                 "UMAP vs PCA scores",
                 "Original vs PCA"),
  ARI = c(
    adjustedRandIndex(km_country$cluster, km_country_orig$cluster),
    adjustedRandIndex(km_country$cluster, km_country_pca$cluster),
    adjustedRandIndex(km_country_orig$cluster, km_country_pca$cluster)
  )
)

ari_inst <- tibble(
  Comparison = c("UMAP vs Original features",
                 "UMAP vs PCA scores",
                 "Original vs PCA"),
  ARI = c(
    adjustedRandIndex(km_inst$cluster, km_inst_orig$cluster),
    adjustedRandIndex(km_inst$cluster, km_inst_pca$cluster),
    adjustedRandIndex(km_inst_orig$cluster, km_inst_pca$cluster)
  )
)

kable(ari_country, digits = 3,
      caption = sprintf("Country clusters: agreement across methods (PCA uses %d components, 80%% variance)",
                        n_pc_country))
kable(ari_inst, digits = 3,
      caption = sprintf("Institution clusters: agreement across methods (PCA uses %d components, 80%% variance)",
                        n_pc_inst))
```

To diagnose how *crisp* each partition actually is, we compute the mean silhouette width
**in the same feature space the clustering was done on** (so the UMAP-based clusters are
scored against UMAP coordinates, the raw-feature clusters against `Mcp`, etc.). Higher is
better; near zero means cluster boundaries are weak.

```{r alt-clustering-silhouette}
sil_mean <- function(km, X) mean(silhouette(km$cluster, dist(X))[, 3])

silhouette_tbl <- tribble(
  ~Space,        ~Method,             ~`Mean silhouette`,
  "Country",     "UMAP coordinates",  sil_mean(km_country, coords_country_sp),
  "Country",     "Original features", sil_mean(km_country_orig, Mcp),
  "Country",     "PCA scores",        sil_mean(km_country_pca, country_pcs),
  "Institution", "UMAP coordinates",  sil_mean(km_inst, coords_sp),
  "Institution", "Original features", sil_mean(km_inst_orig, t(Mcp)),
  "Institution", "PCA scores",        sil_mean(km_inst_pca, inst_pcs)
)
kable(silhouette_tbl, digits = 3,
      caption = "Mean silhouette width per clustering, scored in its own feature space")
```

Two findings come out of this:

1. **The UMAP-based clusters score artificially high (~0.40) because UMAP has already
   pre-separated points in 2D**, so any k-means partition over those coordinates falls into
   pre-existing visual gaps. The raw-feature and PCA clusterings, scored honestly in their
   high-dim native spaces, sit around 0.10–0.16 for both countries and institutions.

2. **The institution data does not have a strong K=6 structure in native space.** Sweeping
   k from 2 to 12 with k-means on `t(Mcp)`, the silhouette peaks at k=2 (0.20–0.25) and
   drops as k grows. Hierarchical clustering on Spearman-correlation distance shows the same
   pattern. So the institutional space is genuinely closer to a 2-axis (democratic vs.
   authoritarian feature bundles) structure with weaker substructure beneath it — UMAP +
   k-means makes the K=6 partition look cleaner than the underlying data supports.

So the UMAP-based clusters remain the most useful *for visualization and storytelling*
(they map cleanly onto the chart), but the alt clusterings are the more honest description
of what the high-dim space actually contains. The country side is more robust across all
three methods (regime typologies do recur); the institution side is dominated by one
strong democratic-vs-authoritarian axis with the K=6 carving being a stylistic choice on
top of that.

```{r alt-clustering-plots, fig.width=6.5, fig.height=6.5}
# Cluster labels were generated by inspecting each cluster's members
# (ranked by Euclidean distance to the centroid in the clustering's own
# feature space) and naming the dominant institutional / regime character.
# See scratch/dump_alt_clusters.R for the full member-distance dump.
country_orig_labels <- tribble(
  ~cluster, ~label,
  1, "Emerging Democracies",
  2, "Closed Autocracies",
  3, "Conflict-Affected Fragile States",
  4, "Hybrid Electoral Autocracies",
  5, "Liberal Democracies",
  6, "Repressive Authoritarian Regimes"
)
country_pca_labels <- tribble(
  ~cluster, ~label,
  1, "Hybrid Electoral Autocracies",
  2, "Liberal Democracies",
  3, "Low-Capacity Electoral Democracies",
  4, "Collapsed States & Civil Wars",
  5, "Closed Authoritarian Regimes",
  6, "Backsliding Middle Democracies"
)
inst_orig_labels <- tribble(
  ~cluster, ~label,
  1, "Contentious Politics & Mobilization",
  2, "Inclusive Programmatic Parties",
  3, "Executive Discretionary Powers",
  4, "Rule of Law & Anti-Corruption",
  5, "Civil Liberties & Media Freedom",
  6, "Personal Freedoms & Sovereignty"
)
inst_pca_labels <- tribble(
  ~cluster, ~label,
  1, "Personal Freedoms & State Sovereignty",
  2, "Social Inclusion & Party Organization",
  3, "Executive Discretion & Digital Repression",
  4, "Mass Mobilization & Polarization",
  5, "Bureaucratic Integrity & Accountability",
  6, "Civil Liberties & Media Pluralism"
)

plot_alt <- function(df, key_col, cluster_vec, label_tbl, title) {
  wrapped <- str_wrap(label_tbl$label, width = 18)
  id_to_label <- setNames(wrapped, as.character(label_tbl$cluster))
  df$alt_id <- cluster_vec[df[[key_col]]]
  df$alt <- factor(id_to_label[as.character(df$alt_id)], levels = wrapped)

  centroids <- df %>%
    group_by(alt) %>%
    summarise(UMAP1 = median(UMAP1), UMAP2 = median(UMAP2), .groups = "drop")

  n_clust <- length(levels(df$alt))
  pal <- colorRampPalette(gl_palettes$categorical)(n_clust)

  ggplot(df, aes(x = UMAP1, y = UMAP2)) +
    geom_point(aes(color = alt), size = 2, alpha = 0.7, show.legend = FALSE) +
    geom_label_repel(
      data = centroids,
      aes(label = alt, fill = alt),
      size = 2.6, fontface = "bold", color = "white",
      label.padding = unit(0.25, "lines"), label.r = unit(0.2, "lines"),
      box.padding = unit(0.4, "lines"), point.padding = unit(0.2, "lines"),
      min.segment.length = 0, segment.color = "grey40",
      seed = 1, max.overlaps = Inf,
      show.legend = FALSE
    ) +
    scale_color_manual(values = pal) +
    scale_fill_manual(values = pal) +
    labs(x = "UMAP 1", y = "UMAP 2", title = title)
}

p_country_orig <- plot_alt(umap_country_sp_df, "country",
                           setNames(km_country_orig$cluster, rownames(Mcp)),
                           country_orig_labels,
                           "Country clusters from Original features")
p_country_pca  <- plot_alt(umap_country_sp_df, "country",
                           setNames(km_country_pca$cluster, rownames(Mcp)),
                           country_pca_labels,
                           "Country clusters from PCA scores")
p_inst_orig <- plot_alt(umap_inst_sp_df, "indicator",
                        setNames(km_inst_orig$cluster, colnames(Mcp)),
                        inst_orig_labels,
                        "Institution clusters from Original features")
p_inst_pca  <- plot_alt(umap_inst_sp_df, "indicator",
                        setNames(km_inst_pca$cluster, colnames(Mcp)),
                        inst_pca_labels,
                        "Institution clusters from PCA scores")

(p_country_orig + p_country_pca) / (p_inst_orig + p_inst_pca)

ggsave("imgs/alt_clustering_comparison.png",
       width = 9, height = 9, dpi = 200)
```

Each panel re-uses the Spearman UMAP layout (so positions are unchanged) but recolors points
by the alternative cluster assignments. Where contiguous regions of the embedding share a
single color, the alt-method recovers the same group as UMAP-k-means; where colors fragment
across the embedding, the alt-method is partitioning along axes that UMAP collapsed (or
vice versa).

# Country Profiles

## Institutional Portfolios in Spearman Space

How do individual countries occupy the institution space? Each panel shows every
institutional feature, with point size and color intensity proportional to the country's
continuous score (0 = absent/weak, 1 = strongest observed). The Spearman embedding is
used here because it respects both shared strengths *and* shared weaknesses — two countries
that both score unusually low on a feature are brought closer together.

```{r country-profiles-institution-space, fig.width=6.5, fig.height=9}
focus_countries <- c("Cambodia", "Pakistan", "Venezuela",
                     "Morocco", "Bolivia", "Argentina")

# Summary table
eci_df %>%
  filter(country %in% focus_countries) %>%
  select(country, ICI, diversity) %>%
  arrange(desc(ICI)) %>%
  kable(col.names = c("Country", "ICI", "Total intensity"),
        digits = 2)

# Build a long data frame using Spearman institution UMAP
profile_df <- map_dfr(focus_countries, function(cname) {
  umap_inst_sp_df %>%
    mutate(
      country = cname,
      score = Mcp[cname, indicator]
    )
})

ggplot(profile_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(color = gl$text_muted, size = 0.5, alpha = 0.3) +
  geom_point(aes(size = score, color = score)) +
  scale_size_continuous(range = c(0.5, 2.5), guide = "none") +
  scale_color_gradient2(low = highlight, mid = "grey85", high = gl$brand_blue,
                        midpoint = 0.5, limits = c(0, 1), name = "Score") +
  facet_wrap(~country, ncol = 2) +
  labs(x = "UMAP 1", y = "UMAP 2")

save_fig("full_tall", "country_profiles_institution_space.pdf")
```

## Nearest Institutional Peers

For each focus country, we identify its closest neighbors based on Spearman correlation
of institutional profiles. Two countries are "institutional peers" when their scores move
together across all features — including shared *weaknesses* (both scoring unusually low)
which are just as informative as shared strengths.

```{r nearest-peers, results='asis'}
n_peers <- 5

for (cname in focus_countries) {
  # Get Spearman similarities to all other countries
  sims <- spearman_cc[cname, ]
  sims <- sort(sims[names(sims) != cname], decreasing = TRUE)
  peers <- names(sims)[1:n_peers]
  peer_cors <- sims[1:n_peers]

  cat("\n\n###", cname, "\n\n")

  # Peer table with both proximity measures
  coloc_sims <- phi_cc[cname, peers]

  tibble(
    Peer = peers,
    `Spearman r` = round(peer_cors, 3),
    `Co-location proximity` = round(coloc_sims, 3)
  ) %>% kable() %>% print()

  # What makes them peers? Find features where the country and its top peer
  # are BOTH distinctively far from the global mean in the same direction
  top_peer <- peers[1]
  country_scores <- Mcp[cname, ]
  peer_scores <- Mcp[top_peer, ]
  global_means <- colMeans(Mcp)
  global_sds <- apply(Mcp, 2, sd)

  # Z-scores relative to global distribution
  country_z <- (country_scores - global_means) / global_sds
  peer_z <- (peer_scores - global_means) / global_sds

  # Shared distinctive features: both far from mean in same direction
  shared_distinctive <- tibble(
    indicator = names(country_scores),
    country_score = country_scores,
    peer_score = peer_scores,
    country_z = country_z,
    peer_z = peer_z,
    same_direction = sign(country_z) == sign(peer_z),
    combined_extremity = abs(country_z) + abs(peer_z)
  ) %>%
    filter(same_direction, abs(country_z) > 0.5, abs(peer_z) > 0.5) %>%
    left_join(indicator_info, by = c("indicator" = "code")) %>%
    arrange(desc(combined_extremity))

  # Show top shared distinctive features split by direction
  shared_high <- shared_distinctive %>% filter(country_z > 0) %>% head(5)
  shared_low <- shared_distinctive %>% filter(country_z < 0) %>% head(5)

  if (nrow(shared_high) > 0) {
    cat("\n\n**Shared strengths with", top_peer, "(both score well above global mean):**\n\n")
    shared_high %>%
      select(description, category,
             !!sym(cname) := country_score,
             !!sym(top_peer) := peer_score) %>%
      mutate(across(where(is.numeric), ~round(.x, 2))) %>%
      kable() %>% print()
  }

  if (nrow(shared_low) > 0) {
    cat("\n\n**Shared weaknesses with", top_peer, "(both score well below global mean):**\n\n")
    shared_low %>%
      select(description, category,
             !!sym(cname) := country_score,
             !!sym(top_peer) := peer_score) %>%
      mutate(across(where(is.numeric), ~round(.x, 2))) %>%
      kable() %>% print()
  }
}
```

## What Makes Each Country Distinct?

For each focus country, we show its most extreme features — the indicators where it scores
furthest from the global mean. These are the features that most define the country's
institutional character, whether strengths or weaknesses.

```{r country-distinct-features, fig.width=6.5, fig.height=9, results='asis'}
global_means <- colMeans(Mcp)
global_sds <- apply(Mcp, 2, sd)

for (cname in focus_countries) {
  scores <- Mcp[cname, ]
  z_scores <- (scores - global_means) / global_sds

  extremes <- tibble(
    indicator = names(scores),
    score = scores,
    z = z_scores
  ) %>%
    left_join(indicator_info, by = c("indicator" = "code")) %>%
    arrange(z)

  # Top 10 highest and lowest
  top10 <- bind_rows(
    extremes %>% head(10) %>% mutate(end = "Lowest"),
    extremes %>% tail(10) %>% mutate(end = "Highest")
  ) %>%
    mutate(
      label = str_trunc(description, 45),
      label = fct_reorder(label, z)
    )

  p <- ggplot(top10, aes(x = z, y = label, fill = end)) +
    geom_col(show.legend = FALSE) +
    geom_vline(xintercept = 0, color = gl$text_dark, linewidth = 0.3) +
    scale_fill_manual(values = c("Lowest" = highlight, "Highest" = gl$brand_blue)) +
    labs(x = "Standard deviations from global mean", y = NULL)

  cat("\n\n### ", cname, "\n\n", sep = "")
  print(p)
}
```

## Cluster Placement & Anchoring Features

Each focus country sits in one of the six country clusters identified earlier. This section
unpacks *why*. For every country we compute:

1. The **Spearman correlation** of its full institutional profile with each cluster's mean
   vector — how closely its overall pattern matches each institutional archetype.
2. The features **anchoring** it to its cluster (where the country's score is close to its
   own cluster mean *and* far from the runner-up cluster mean), weighted by how much the
   two cluster means actually differ on that feature.
3. The features **pulling** it toward the runner-up cluster — outlier dimensions where the
   country looks more like the alternative archetype.

```{r country-cluster-anchors, results='asis'}
# Pre-compute per-cluster mean vectors (labelled by cluster name)
country_cluster_members <- split(rownames(Mcp), km_country$cluster)
cluster_mean_matrix <- sapply(country_cluster_members, function(cs)
  colMeans(Mcp[cs, , drop = FALSE]))
colnames(cluster_mean_matrix) <- country_cluster_labels[colnames(cluster_mean_matrix)]

indicator_desc_lookup <- setNames(indicator_info$description, indicator_info$code)

for (cname in focus_countries) {
  cl <- km_country$cluster[rownames(Mcp) == cname]
  cl_label <- country_cluster_labels[as.character(cl)]

  cv <- Mcp[cname, ]
  sims <- apply(cluster_mean_matrix, 2,
                function(v) cor(cv, v, method = "spearman"))

  sims_df <- tibble(Cluster = names(sims),
                    `Spearman r` = round(sims, 3)) %>%
    arrange(desc(`Spearman r`))

  alt_label <- sims_df$Cluster[2]
  own_mu <- cluster_mean_matrix[, cl_label]
  alt_mu <- cluster_mean_matrix[, alt_label]

  # anchor score = (closer to own than alt) × (how different own/alt are on this feature)
  pull  <- abs(cv - alt_mu) - abs(cv - own_mu)
  gap   <- abs(own_mu - alt_mu)
  score <- pull * gap

  feat_tbl <- tibble(
    Feature = str_trunc(indicator_desc_lookup[names(cv)], 45),
    country_val = round(unname(cv), 2),
    own_val = round(unname(own_mu), 2),
    alt_val = round(unname(alt_mu), 2),
    score = round(score, 4)
  ) %>% arrange(desc(score))

  cat("\n\n### ", cname, " — placed in *", cl_label, "*\n\n", sep = "")

  # Correlation table
  print(kable(sims_df,
              caption = paste("Profile correlation of", cname,
                              "with each cluster mean")))

  # Top anchoring features
  cat("\n*Features anchoring ", cname, " to ", cl_label,
      " (vs. runner-up ", alt_label, "):*\n\n", sep = "")
  print(kable(feat_tbl %>% head(8) %>% select(-score),
              col.names = c("Feature", cname, cl_label, alt_label)))

  # Top features pulling toward the alternative
  cat("\n*Features where ", cname, " instead resembles ", alt_label, ":*\n\n",
      sep = "")
  print(kable(feat_tbl %>% tail(5) %>% arrange(score) %>% select(-score),
              col.names = c("Feature", cname, cl_label, alt_label)))
}
```

# Comparing Proximity Methods

How similar are the co-location and Spearman proximity matrices? If they agree strongly,
the institutional space structure is robust to the choice of proximity measure. If they
diverge, they're capturing different aspects of institutional relationships.

## Element-wise comparison

```{r proximity-comparison}
# Feature proximity comparison
coloc_upper <- phi_pp[upper.tri(phi_pp)]
spear_upper <- spearman_pp_01[upper.tri(spearman_pp_01)]

cor_feature_prox <- cor(coloc_upper, spear_upper, method = "spearman")

cat("Feature proximity matrices:\n")
cat("  Spearman rank correlation:", round(cor_feature_prox, 3), "\n")

# Country proximity comparison
coloc_cc_upper <- phi_cc[upper.tri(phi_cc)]
spear_cc_upper <- spearman_cc_01[upper.tri(spearman_cc_01)]

cor_country_prox <- cor(coloc_cc_upper, spear_cc_upper, method = "spearman")

cat("\nCountry proximity matrices:\n")
cat("  Spearman rank correlation:", round(cor_country_prox, 3), "\n")
```

```{r proximity-scatter, fig.height=4.5}
# Scatter: co-location vs Spearman proximity (features)
prox_compare_df <- tibble(
  colocation = coloc_upper,
  spearman = spear_upper
)

p1 <- ggplot(prox_compare_df, aes(x = colocation, y = spearman)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", color = highlight, se = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = gl$text_muted) +
  labs(x = "Co-location proximity", y = "Spearman proximity (rescaled)")

# Scatter: co-location vs Spearman proximity (countries)
prox_compare_cc_df <- tibble(
  colocation = coloc_cc_upper,
  spearman = spear_cc_upper
)

p2 <- ggplot(prox_compare_cc_df, aes(x = colocation, y = spearman)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", color = highlight, se = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = gl$text_muted) +
  labs(x = "Co-location proximity", y = "Spearman proximity (rescaled)")

p1 + p2 + plot_annotation(tag_levels = "A")

save_fig("full", "proximity_comparison_scatter.pdf")
```

## Comparing embeddings

Do the two proximity methods produce similar institution spaces? We compare by computing
the Spearman correlation between pairwise distances in each UMAP embedding.

```{r embedding-comparison}
# Pairwise Euclidean distances in each institution UMAP
dist_coloc <- as.matrix(dist(umap_inst[, 1:2]))
dist_spear <- as.matrix(dist(umap_inst_sp[, 1:2]))

# Correlation between pairwise distances (Mantel-like test)
d_coloc_upper <- dist_coloc[upper.tri(dist_coloc)]
d_spear_upper <- dist_spear[upper.tri(dist_spear)]

cat("Institution UMAP distance correlation (Spearman):",
    round(cor(d_coloc_upper, d_spear_upper, method = "spearman"), 3), "\n")

# Same for country UMAP
dist_coloc_c <- as.matrix(dist(umap_country[, 1:2]))
dist_spear_c <- as.matrix(dist(umap_country_sp[, 1:2]))

d_coloc_c_upper <- dist_coloc_c[upper.tri(dist_coloc_c)]
d_spear_c_upper <- dist_spear_c[upper.tri(dist_spear_c)]

cat("Country UMAP distance correlation (Spearman):",
    round(cor(d_coloc_c_upper, d_spear_c_upper, method = "spearman"), 3), "\n")
```

## ECI vs PCA comparison

The eigenvalue-based ECI extracts a single dimension of complexity from the co-location
structure. PCA extracts dimensions of variance from the raw data. How do they relate?

```{r eci-vs-pca}
compare_df <- tibble(
  country = rownames(Mcp),
  ICI = eci_eigen,
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  diversity = kc0
) %>%
  mutate(
    region = region_labels[as.character(region_lookup[country])],
    region = replace_na(region, "Other")
  )

ggplot(compare_df, aes(x = ICI, y = PC1)) +
  geom_point(aes(color = region), size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", color = gl$text_muted, linetype = "dashed", se = TRUE, alpha = 0.15) +
  geom_text_repel(aes(label = country), size = 2.5, max.overlaps = 15) +
  labs(x = "ICI (eigenvalue method)", y = "PC1 score")

save_fig("full", "eci_vs_pca.pdf")

cat("Correlation ICI vs PC1:", round(cor(compare_df$ICI, compare_df$PC1), 3), "\n")
cat("Correlation ICI vs PC2:", round(cor(compare_df$ICI, compare_df$PC2), 3), "\n")
```

# Additional Visualizations

## Heatmap of the Mcp Matrix

```{r heatmap, fig.width=6.5, fig.height=6.5}
# Order columns by ubiquity (most ubiquitous = left) and rows by diversity (most diverse = top)
col_order <- names(sort(kp0, decreasing = TRUE))
row_order <- names(sort(kc0, decreasing = TRUE))

Mcp_ordered <- Mcp[row_order, col_order]

pheatmap(
  Mcp_ordered,
  color = colorRampPalette(c("grey95", gl$brand_blue))(100),
  show_rownames = FALSE,
  show_colnames = FALSE,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize = 8
)
```

## ICI vs GDP per Capita

```{r gdp-data, cache=TRUE}
# Get GDP per capita from World Development Indicators
gdp_data <- WDI(
  indicator = "NY.GDP.PCAP.PP.KD",
  start = latest_year - 1,
  end = latest_year,
  extra = TRUE
) %>%
  filter(!is.na(NY.GDP.PCAP.PP.KD)) %>%
  group_by(country) %>%
  slice_max(year, n = 1) %>%
  ungroup() %>%
  select(country, iso3c, gdp_pc = NY.GDP.PCAP.PP.KD)

# Match to our country names using countrycode
eci_gdp <- eci_df %>%
  mutate(
    iso3c = countrycode(country, "country.name", "iso3c",
                        warn = FALSE)
  ) %>%
  left_join(gdp_data, by = "iso3c") %>%
  filter(!is.na(gdp_pc))

cat("Countries matched with GDP data:", nrow(eci_gdp), "\n")
```

```{r eci-vs-gdp}
# Add regions
eci_gdp <- eci_gdp %>%
  mutate(
    region = region_labels[as.character(
      region_lookup[country.x]
    )],
    region = replace_na(region, "Other")
  )

cor_gdp <- cor(eci_gdp$ICI, log(eci_gdp$gdp_pc), use = "complete.obs")

ggplot(eci_gdp, aes(x = ICI, y = gdp_pc)) +
  geom_point(aes(color = region), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = gl$text_muted, linetype = "dashed", se = TRUE, alpha = 0.15) +
  geom_text_repel(
    aes(label = country.x),
    size = 2.5, max.overlaps = 15
  ) +
  scale_y_log10(labels = scales::dollar_format()) +
  labs(
    x = "Institutional Complexity Index (ICI)",
    y = "GDP per capita (PPP, constant $, log scale)"
  )

save_fig("full", "ici_vs_gdp.pdf")
```

## ICI vs Liberal Democracy Index

```{r eci-vs-libdem}
# Get liberal democracy index
libdem <- vdem_raw %>%
  filter(year == latest_year) %>%
  select(country_name, v2x_libdem) %>%
  filter(!is.na(v2x_libdem))

eci_libdem <- eci_df %>%
  left_join(libdem, by = c("country" = "country_name")) %>%
  filter(!is.na(v2x_libdem)) %>%
  mutate(
    region = region_labels[as.character(region_lookup[country])],
    region = replace_na(region, "Other")
  )

cor_libdem <- cor(eci_libdem$ICI, log(eci_libdem$v2x_libdem), use = "complete.obs")

ggplot(eci_libdem, aes(x = ICI, y = v2x_libdem)) +
  geom_point(aes(color = region), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = gl$text_muted, linetype = "dashed", se = TRUE, alpha = 0.15) +
  geom_text_repel(
    aes(label = country),
    size = 2.5, max.overlaps = 15
  ) +
  scale_y_log10() +
  labs(
    x = "Institutional Complexity Index (ICI)",
    y = "V-Dem Liberal Democracy Index (log scale)"
  )

save_fig("full", "ici_vs_libdem.pdf")
```

## ICI by World Region

```{r eci-by-region}
eci_regions <- eci_df %>%
  mutate(
    region = region_labels[as.character(region_lookup[country])],
    region = replace_na(region, "Other")
  ) %>%
  filter(region != "Other")

ggplot(eci_regions, aes(x = reorder(region, ICI, median), y = ICI, fill = region)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE, outlier.alpha = 0.3) +
  geom_jitter(aes(color = region), width = 0.15, alpha = 0.5, size = 1.5, show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Institutional Complexity Index (ICI)")

save_fig("full", "ici_by_region.pdf")
```

# Discussion

## What does institutional complexity tell us?

The analogy between economic and institutional complexity reveals several patterns:

**High-complexity countries** tend to be established democracies with deep, interlocking
institutional frameworks. These countries don't just have individual strong institutions --
they have *rare combinations* of institutions that reinforce each other. Just as economically
complex countries produce products that require many complementary capabilities, institutionally
complex countries maintain governance features that require deep state capacity, civic culture,
and historical development.

**The institution space** reveals how governance features cluster. Features that are close
together in the UMAP embedding tend to co-occur -- they represent institutional "packages"
that countries adopt together. The structure of this space suggests that some institutional
reforms may be easier to adopt once neighboring capabilities are in place.

**Feature complexity** is particularly informative. The most "complex" institutional features --
those found primarily in institutionally sophisticated countries -- tend to involve subtle
aspects of governance: rigorous and impartial public administration, judicial accountability,
freedom from self-censorship in media, and deliberative quality of political discourse.
These are features that require not just formal rules but deeply embedded norms and practices.

## Methodological notes

This analysis uses a **continuous Mcp matrix** (min-max normalized to [0, 1]) rather than
the traditional binary matrix. This preserves the full variation in V-Dem's IRT-derived
scores and avoids arbitrary binarization thresholds. Two discrete integer variables
(v2regopploc, v2regsuploc) were excluded since they do not sit on comparable continuous
scales.

We compute proximity using two methods:

1. **Co-location proximity** (EC-style): the continuous generalization of the Hidalgo et al.
   minimum conditional probability. Captures whether high values tend to co-occur.

2. **Spearman rank correlation**: captures the full monotonic relationship between indicators
   (or countries), including agreement at the bottom of the distribution. Rescaled to [0, 1]
   for comparability.

The comparison between these methods tests whether the institution space structure is
driven primarily by which countries score *high* on features (co-location) or by the full
rank ordering (Spearman). PCA provides a multi-dimensional view complementary to the single
ECI axis.

## Caveats

1. **V-Dem data reflects expert assessments**, not objective measures. The normalization
   choices (min-max vs z-score vs rank) affect results.

2. **The analogy has limits.** Countries don't "choose" institutions the way firms choose
   products to export. Institutional development is path-dependent, often violent, and
   constrained by geography, history, and international power relations.

3. **Correlation with GDP and democracy is expected** given that many V-Dem indicators
   directly measure democratic quality. The interesting question is whether the *complexity*
   measure adds information beyond simple additive indices.

4. **The method is sensitive to indicator selection.** A different set of indicators would
   produce different complexity rankings. Our selection, while broad, reflects choices about
   what counts as an "institution."

5. **Cross-sectional analysis only.** A natural extension would track institutional complexity
   over time to study how countries build (or lose) institutional sophistication.

## Future directions

- Panel analysis: how does institutional complexity evolve over time?
- Predictive power: does ICI predict future economic growth or democratic stability
  beyond what simpler indices capture?
- Institutional "product space" analysis: which institutional upgrades are most
  feasible for countries at different positions in the institution space?
- Comparison with actual Economic Complexity Index: are economically and institutionally
  complex countries the same?
- Alternative normalizations: rank-based or z-score normalization instead of min-max

---

*Analysis generated on `r Sys.Date()`. Data from V-Dem (loaded from `data/shared-data/vdem/`)
and World Development Indicators.*
