---
title: "Pooled-Panel PCA"
subtitle: "Institutional axes that hold up across 1970-2025"
author: "Institutional Complexity Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6.5
    fig_height: 4.5
    latex_engine: xelatex
    includes:
      in_header: ~/dev/gl-design/skills/gl-ggplot/assets/gl_pdf.tex
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  cache = TRUE,
  cache.lazy = FALSE
)
```

```{r packages, cache=FALSE}
library(tidyverse)
library(arrow)
library(ggrepel)
library(knitr)
library(patchwork)
```

```{r gl-setup, cache=FALSE}
source("~/dev/gl-design/skills/gl-ggplot/assets/theme_gl.R")
gl_setup()
```

# Why pool

The country-space writeup builds a fresh PCA on **only** the 2025 V-Dem
cross-section. The cambodia-analysis trajectory chart then projects 1970-2024
data through that 2025 PCA and reads off PC2 over time. The driver decomposition
on Cambodia 2006-2025 surfaced the cost of that choice: 9 of the 12 top PC2
movers belong to V-Dem's Digital Society Survey (`v2sm*`), a module that
didn't exist before 2000 and accounts for ~39% of the absolute contribution
mass to PC2 (vs. its 18% share of the feature set). Fitting on 2025 means the
PCA's axes inherit 2025's variance structure — a structure dominated by a
post-2000 module that for two-thirds of the trajectory we project through is
entirely imputed.

A pooled-panel PCA addresses this directly. We stack every (country, year)
observation from 1970-2025 into one tall matrix, z-score across the pool, and
fit one PCA. The loadings then reflect variance over the **whole window we
care about**, not the present. Indicators with sparse historical coverage
contribute less to the loading geometry because their pool-wide variance is
small (they're approximately constant at the imputed mean for half the panel).

What we lose: the exact scaling of axis units in any single year is no longer
a "within-year SD" — it's a "within-pool SD". That's fine for the questions
we want to ask of these axes ("how big a move is Cambodia 2006 → 2025
relative to the kinds of moves countries make in this dataset?").

# Data

```{r load-vdem, cache=TRUE}
vdem_raw      <- read_parquet("data/shared-data/vdem/vdem.parquet")
vdem_codebook <- read_parquet("data/shared-data/vdem/vdem_codebook.parquet")
```

```{r panel-window}
panel_start <- 1970
panel_end   <- vdem_raw %>%
  filter(year >= 2020) %>%
  group_by(year) %>%
  summarise(n_countries = n_distinct(country_name), .groups = "drop") %>%
  filter(n_countries >= 150) %>%
  pull(year) %>% max()

cat(sprintf("Panel window: %d-%d\n", panel_start, panel_end))
```

## Indicator selection

We mirror the country-space writeup's filter:

- vartype `"C"` only (component-type IRT scores; aggregates would double-count)
- drop the two discrete-categorical components V-Dem ships (`v2regopploc`,
  `v2regsuploc`)
- drop the 18 conceptually-redundant indicators flagged in
  `outputs/redundancy_clusters_2025.csv`
- drop the suffix variants (`_codelow`, `_codehigh`, `_sd`, `_osp`, `_ord`,
  `_mean`, `_nr`, `_rec`) and per-rater columns (`_1`, `_2`, ...)

The new piece: a **panel-wide coverage filter**. An indicator must be
observed in at least 50% of (country, year) cells across the 1970-`r panel_end`
panel. This is the rule that decides whether the Digital Society Survey
indicators (`v2sm*`, coded only 2000-`r panel_end`) survive — they're observed
in roughly `r round(100 * (panel_end - 1999) / (panel_end - panel_start + 1))`%
of years, which is right at the threshold.

```{r indicator-filter, cache=TRUE}
discrete_vars <- c("v2regopploc", "v2regsuploc")

redundant_drops <- c(
  "v2cldmovew", "v2cldiscw", "v2clslavef", "v2clprptyw", "v2clacjstw",
  "v2juncind", "v2juhccomp", "v2csrlgrep", "v2csrlgcon", "v2exdfdmhs",
  "v2smgovsm", "v2lgqstexp", "v2lglegplo", "v2merange", "v2dlreason",
  "v2psoppaut", "v2cafexch", "v2caconmob"
)

suffixes      <- c("_codelow", "_codehigh", "_sd", "_osp",
                   "_ord", "_mean", "_nr", "_rec")
all_cols      <- names(vdem_raw)
is_suffix     <- vapply(all_cols, function(x)
                        any(vapply(suffixes, endsWith, logical(1), x = x)),
                        logical(1))
is_numbered   <- grepl("_\\d+$", all_cols)
base_cols     <- all_cols[!is_suffix & !is_numbered]

component_codes <- vdem_codebook %>%
  filter(vartype == "C") %>% pull(tag) %>% unique()

candidate_codes <- intersect(component_codes, base_cols) %>%
  setdiff(discrete_vars) %>%
  setdiff(redundant_drops)

candidate_codes <- candidate_codes[
  vapply(candidate_codes, function(x) is.numeric(vdem_raw[[x]]), logical(1))
]

panel_window <- vdem_raw %>%
  filter(year >= panel_start, year <= panel_end)

n_cells <- nrow(panel_window)

coverage <- vapply(candidate_codes,
                   function(x) sum(!is.na(panel_window[[x]])),
                   integer(1))

coverage_threshold <- 0.5
indicators <- candidate_codes[coverage >= coverage_threshold * n_cells]

cat(sprintf("Candidate indicators: %d\n", length(candidate_codes)))
cat(sprintf("Pass >=%.0f%% pool coverage: %d\n",
            100 * coverage_threshold, length(indicators)))
```

```{r indicator-coverage-summary}
prefix_categories <- c(
  "el" = "Elections",            "ps" = "Political Parties",
  "cl" = "Civil Liberties",      "ju" = "Judiciary",
  "lg" = "Legislature",          "me" = "Media",
  "ex" = "Executive",            "dl" = "Deliberation",
  "ca" = "Mobilization & Academic", "cs" = "Civil Society",
  "pe" = "Political Equality",   "st" = "State Capacity",
  "sv" = "Sovereignty",          "dd" = "Direct Democracy",
  "ed" = "Education",            "sm" = "Digital Society / Social Media",
  "re" = "Regime",               "te" = "Technology & Media Environment"
)

cov_df <- tibble(
  code     = candidate_codes,
  pool_cov = coverage / n_cells,
  prefix   = sub("^v[23](\\w{2}).*", "\\1", candidate_codes),
  kept     = pool_cov >= coverage_threshold
) %>%
  mutate(category = unname(prefix_categories[prefix]),
         category = replace_na(category, "Other"))

cov_df %>%
  count(category, kept) %>%
  pivot_wider(names_from = kept, values_from = n,
              values_fill = 0) %>%
  rename(`Dropped (<50% pool cov)` = `FALSE`,
         `Kept` = `TRUE`) %>%
  mutate(Total = `Dropped (<50% pool cov)` + Kept) %>%
  arrange(desc(Total)) %>%
  kable(caption = "How the panel-coverage filter cuts each V-Dem category.")
```

```{r v2sm-coverage-note}
sm_codes <- candidate_codes[grepl("^v2sm", candidate_codes)]
sm_kept  <- intersect(sm_codes, indicators)
cat(sprintf("v2sm* (Digital Society) candidates: %d\n", length(sm_codes)))
cat(sprintf("v2sm* surviving the 50%% filter: %d\n",  length(sm_kept)))
```

The Digital Society block is largely culled; whatever survives is whatever
V-Dem coded retroactively for some pre-2000 years.

## Country-year row construction

```{r build-panel, cache=TRUE}
panel_mat_df <- panel_window %>%
  select(country_name, year, all_of(indicators)) %>%
  mutate(p_missing = rowSums(is.na(across(all_of(indicators)))) /
                     length(indicators)) %>%
  filter(p_missing < 0.5) %>%
  select(-p_missing)

cat(sprintf("Country-year rows kept: %d (of %d in window)\n",
            nrow(panel_mat_df), n_cells))

panel_mat_df %>%
  count(year) %>%
  ggplot(aes(year, n)) +
  geom_col(fill = gl$brand_blue, width = 0.8) +
  scale_x_continuous(breaks = seq(1970, panel_end, 10)) +
  labs(x = NULL, y = "Country-years kept",
       title = "Panel composition by year",
       subtitle = "Rows that pass the >=50% feature-coverage filter.") +
  theme(legend.position = "none")
```

## Imputation and z-scoring

Mean-impute remaining NAs at the **pool** level (not within-year), then
z-score each indicator across the pool. The pool-mean fill is the most
defensible neutral value for a country-year that's missing a feature: it
treats the imputed cell as "an average country-year on this dimension."

```{r normalize, cache=TRUE}
M_raw <- panel_mat_df %>%
  select(all_of(indicators)) %>%
  as.matrix()

# Pool-mean imputation
for (j in seq_len(ncol(M_raw))) {
  na_idx <- is.na(M_raw[, j])
  if (any(na_idx)) M_raw[na_idx, j] <- mean(M_raw[, j], na.rm = TRUE)
}

# z-score across pool
M_z <- scale(M_raw, center = TRUE, scale = TRUE)
M_z <- M_z[, !is.nan(colSums(M_z)), drop = FALSE]

cat(sprintf("Final matrix: %d country-years x %d features\n",
            nrow(M_z), ncol(M_z)))
cat(sprintf("Mean per-feature SD after z-scoring (sanity, should be 1): %.3f\n",
            mean(apply(M_z, 2, sd))))
```

# Fit the PCA

```{r fit-pca, cache=TRUE}
pca_pool <- prcomp(M_z, center = FALSE, scale. = FALSE)
# scale.=FALSE because we already z-scored; centering is the column means of z
# which are zero by construction.

var_exp <- pca_pool$sdev^2 / sum(pca_pool$sdev^2)
cum_var <- cumsum(var_exp)
```

## How much variance does each PC explain?

```{r scree, fig.height=3.6}
scree_df <- tibble(PC = seq_along(var_exp),
                   var_exp = var_exp,
                   cum_var = cum_var)

p_bar <- ggplot(scree_df %>% slice(1:20),
                aes(factor(PC), var_exp)) +
  geom_col(fill = gl$brand_blue, width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = "PC", y = "Variance explained",
       title = "Scree (first 20 PCs)")

p_cum <- ggplot(scree_df %>% slice(1:20),
                aes(PC, cum_var)) +
  geom_line(linewidth = 0.7, color = gl$brand_blue) +
  geom_point(size = 1.2, color = gl$brand_blue) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  scale_x_continuous(breaks = c(1, 5, 10, 15, 20)) +
  labs(x = "PC", y = "Cumulative variance",
       title = "Cumulative variance")

p_bar + p_cum
```

```{r variance-table}
scree_df %>%
  slice(1:10) %>%
  transmute(PC,
            `Eigenvalue (var)`     = round(pca_pool$sdev[1:10]^2, 2),
            `Variance explained`   = scales::percent(var_exp, accuracy = 0.1),
            `Cumulative variance`  = scales::percent(cum_var, accuracy = 0.1)) %>%
  kable(caption = "First 10 principal components.")
```

The first PC absorbs the dominant democracy/autocracy axis and is large; the
next several PCs each capture meaningful but smaller chunks. We carry PC1-PC4
forward.

# What does each PC capture?

For each of the first four PCs we list the indicators with the largest
positive and negative loadings, grouped by V-Dem category. A loading is
the indicator's coefficient in that PC's direction in the standardized
feature space — ie how much a one-SD move in that indicator shifts a
country-year along this PC.

```{r loadings-helper}
codebook_lookup <- vdem_codebook %>%
  transmute(code = tag, name) %>%
  distinct(code, .keep_all = TRUE)

loadings_df <- as_tibble(pca_pool$rotation[, 1:4],
                          rownames = "code") %>%
  mutate(prefix   = sub("^v[23](\\w{2}).*", "\\1", code),
         category = unname(prefix_categories[prefix]),
         category = replace_na(category, "Other")) %>%
  left_join(codebook_lookup, by = "code")

top_loadings <- function(pc, k = 8, side = c("pos", "neg")) {
  side <- match.arg(side)
  d <- loadings_df %>%
    transmute(code, name, category, loading = .data[[pc]])
  if (side == "pos") {
    d %>% arrange(desc(loading)) %>% slice_head(n = k)
  } else {
    d %>% arrange(loading) %>% slice_head(n = k)
  }
}
```

```{r print-loadings, results='asis'}
for (pc in c("PC1", "PC2", "PC3", "PC4")) {
  cat(sprintf("\n## %s — top positive loadings\n\n", pc))
  print(kable(top_loadings(pc, 8, "pos") %>%
                transmute(Code = code, Indicator = name, Category = category,
                          Loading = round(loading, 3))))
  cat(sprintf("\n## %s — top negative loadings\n\n", pc))
  print(kable(top_loadings(pc, 8, "neg") %>%
                transmute(Code = code, Indicator = name, Category = category,
                          Loading = round(loading, 3))))
}
```

## PC2 in detail: top and bottom 15 loadings with codebook definitions

PC2 is the second-most-important institutional axis after democracy and the
one whose interpretation requires care — several top contributors have
inverted scales, so the loading sign and the meaning of "high indicator
score" don't line up the way the indicator names suggest. We lay out the
top and bottom 15 PC2 loadings with V-Dem's full question text and response
options so the reading developed in the takeaways section below is
independently verifiable from primary materials.

```{r pc2-detail-helper}
format_indicator_block <- function(code, loading, codebook) {
  row <- codebook[codebook$tag == code, ]
  if (nrow(row) == 0) return(invisible(NULL))
  row <- row[1, ]
  cat(sprintf("\n**`%s` --- %s** (loading %+.3f)\n\n",
              code, row$name, loading))
  if (!is.na(row$question) && nchar(row$question) > 0)
    cat(sprintf("*Question.* %s\n\n", row$question))
  if (!is.na(row$responses) && nchar(row$responses) > 0)
    cat(sprintf("*Responses.* %s\n\n", row$responses))
}
```

### PC2 --- top 15 positive loadings

These pull a country-year toward the positive pole of PC2.

```{r pc2-pos15-detail, results='asis'}
pc2_pos15 <- top_loadings("PC2", 15, "pos")
for (i in seq_len(nrow(pc2_pos15))) {
  format_indicator_block(pc2_pos15$code[i], pc2_pos15$loading[i], vdem_codebook)
}
```

### PC2 --- top 15 negative loadings

These pull a country-year toward the negative pole of PC2. Several
indicators here are coded so a *high* indicator score means *less* of the
named feature --- for example `v2psparban` (Party ban) runs 0 = "all
parties banned" to 4 = "no parties banned", so a high indicator score (no
bans) contributes negatively to PC2, meaning *more* party bans pull a
country toward the *positive* pole of PC2. Each block below includes the
full response scale so the directionality is unambiguous.

```{r pc2-neg15-detail, results='asis'}
pc2_neg15 <- top_loadings("PC2", 15, "neg")
for (i in seq_len(nrow(pc2_neg15))) {
  format_indicator_block(pc2_neg15$code[i], pc2_neg15$loading[i], vdem_codebook)
}
```

## Loading mass by V-Dem category, per PC

The same redundancy worry as before: a PC that draws disproportionately from
one V-Dem family is partly capturing module-internal covariance rather than
generic institutional structure. The table below shows, for each of the first
four PCs, the share of squared loading mass falling in each category.

```{r loading-mass-by-category}
pc_long <- loadings_df %>%
  pivot_longer(starts_with("PC"), names_to = "pc", values_to = "loading")

pc_mass <- pc_long %>%
  group_by(pc, category) %>%
  summarise(sq_mass = sum(loading^2),
            .groups = "drop") %>%
  group_by(pc) %>%
  mutate(pct = sq_mass / sum(sq_mass)) %>%
  ungroup()

pc_mass %>%
  pivot_wider(id_cols = category, names_from = pc, values_from = pct,
              values_fill = 0) %>%
  mutate(across(starts_with("PC"),
                ~scales::percent(.x, accuracy = 0.1))) %>%
  arrange(desc(PC1)) %>%
  kable(caption = "Share of squared loading mass by V-Dem category, by PC.")
```

# Sample country trajectories

A pooled PCA is only useful if its axes track the substantive history we
think we know. We project a small panel of well-known countries into PC1
and PC2 over the full window.

```{r country-pc-scores, cache=TRUE}
panel_with_scores <- panel_mat_df %>%
  select(country_name, year) %>%
  bind_cols(as_tibble(pca_pool$x[, 1:5]))

focus <- c("Germany", "Sweden", "United States of America", "Singapore",
           "China", "Russia", "Cambodia", "Vietnam", "Iran",
           "South Korea", "Poland", "Brazil", "South Africa")
focus_scores <- panel_with_scores %>%
  filter(country_name %in% focus) %>%
  mutate(country_name = factor(country_name, levels = focus))
```

```{r pc1-trajectories, fig.height=4.6}
ggplot(focus_scores, aes(year, PC1, color = country_name)) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 0.6) +
  scale_x_continuous(breaks = seq(1970, panel_end, 10)) +
  labs(x = NULL, y = "PC1 score (pooled)",
       title = "PC1 trajectories, 1970-present",
       subtitle = "Sample panel. Sign and meaning of PC1: see top-loadings table above.",
       color = NULL) +
  theme(legend.position = "right",
        legend.text = element_text(size = 7))
```

```{r pc2-trajectories, fig.height=4.6}
ggplot(focus_scores, aes(year, PC2, color = country_name)) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 0.6) +
  scale_x_continuous(breaks = seq(1970, panel_end, 10)) +
  labs(x = NULL, y = "PC2 score (pooled)",
       title = "PC2 trajectories, 1970-present",
       subtitle = "Sample panel. Sign and meaning of PC2: see top-loadings table above.",
       color = NULL) +
  theme(legend.position = "right",
        legend.text = element_text(size = 7))
```

## Two-axis snapshot at year endpoints

```{r endpoint-scatter, fig.width=7, fig.height=8}
endpoint_years <- c(1970, 1995, panel_end)

ep_df <- panel_with_scores %>%
  filter(year %in% endpoint_years) %>%
  mutate(year_lab = factor(year, levels = endpoint_years))

focus_label <- c("Germany", "Sweden", "United States of America", "Singapore",
                 "China", "Russia", "Cambodia", "Vietnam",
                 "South Korea", "Brazil", "Pakistan", "Saudi Arabia",
                 "Soviet Union")

ggplot(ep_df, aes(PC1, PC2)) +
  geom_hline(yintercept = 0, color = "grey80", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "grey80", linewidth = 0.3) +
  geom_point(color = "grey55", alpha = 0.4, size = 1.2) +
  geom_point(data = ~ filter(.x, country_name %in% focus_label),
             color = highlight, size = 2.2) +
  geom_text_repel(data = ~ filter(.x, country_name %in% focus_label),
                  aes(label = country_name),
                  size = 2.5, color = highlight,
                  segment.alpha = 0.4, max.overlaps = 30) +
  facet_wrap(~ year_lab, ncol = 1) +
  labs(x = "PC1", y = "PC2",
       title = "Pooled-PCA snapshots at three years",
       subtitle = "Same axes (pooled rotation) at three endpoint years. Compare layouts to read panel-PCA stability.") +
  theme(strip.text = element_text(face = "bold"))
```

# Key takeaways: PC1 and PC2 as analytical tools

The pooled PCA produces dozens of components but only the first two carry
enough variance and conceptual coherence to deploy on case studies. Both
have a clear interpretation that survives careful reading of the loadings.
Together they give a two-dimensional institutional space we can use on
individual countries to separate regime-type movement from state-building
movement. This section codifies what each axis means and how to read
movement on it; downstream case-study writeups should refer back here
rather than re-deriving the interpretation each time.

## PC1: openness of the regime (the democracy axis)

**Positive pole.** Liberal democracy: freedom of expression and discussion,
free press (low government censorship, low harassment of journalists, low
media corruption), free academic and cultural expression, peaceful
assembly, autonomous electoral management, no civil society repression,
freedom from torture, transparent laws with predictable enforcement,
access to justice, rigorous public administration, broad consultation in
deliberation.

**Negative pole.** Personalist/ideological autocracy with a state-
controlled information environment: state-owned print and broadcast media
with political influence, control of entertainment content, political
teacher firing/hiring with no teacher autonomy or independent unions,
person-of-the-leader and ideological executive legitimation, HOS veto and
dissolution power in practice, state-administered mass organizations,
mobilization for autocracy.

**How to read movement.** A move on PC1 is regime change in the obvious
sense --- democratization (positive direction) or autocratization
(negative direction). PC1 is the axis any V-Dem PCA recovers; the pooled
fit just confirms it's robust across the 1970--`r panel_end` window and
not an artifact of post-2000 measurement.

## PC2: stateness --- how much of social life the state encompasses

This is the axis whose interpretation took work to nail down (see the
top-15 detail above). After fixing directionality on the reverse-coded
indicators, PC2 contrasts two opposing **modes of governance** that cut
across regime type.

**Positive pole --- "the state encompasses society."** Bundles two
clusters that the data hold together:

1. *Broad, equal public-goods delivery.* Equal access to education,
   health, public services across income / urban-rural / social group /
   gender; state jobs accessible across income; universalistic rather
   than means-tested welfare. State authority over territory; uniform
   subnational administration of elections and civil liberties. Clean
   everyday administration (no legislative corruption, no vote-buying,
   no electoral violence).
2. *State control of the non-state sphere.* State-owned broadcast media;
   state ownership of the economy; biased state-aligned media; media
   self-censorship; government control over CSO entry and exit; low
   university autonomy; party bans and barriers to party formation; no
   judicial review constraining the executive.

These look like opposites at first --- equality of delivery sounds
liberal, control of the non-state sphere sounds illiberal --- but the data
bundle them because they share a common precondition: **infrastructural
power** in Michael Mann's sense. A state with reach to deliver basic
services uniformly across territory is the same state with reach to
license every broadcaster, register every CSO, supervise every
university, and gate every party-formation rule. Encompassment cuts
both ways.

**Negative pole --- "society encompasses the state."** Private economy,
private media, autonomous universities, autonomous CSOs, critical media,
critical academics, judicial review constraining the executive, easy
party formation and large opposition --- alongside political violence by
non-state actors, anti-system CSO movements, and mass mobilization for
democracy. The unifying thread is *non-state actors having visible
weight*, whether that weight takes benign-constitutional form (a
constitutional court overriding the executive) or violent-contentious
form (insurgent movements attacking it). PC2 lumps them because both
register as the state's encompassment of society being incomplete.

**How to read movement.**

- Positive PC2 movement = state-building / state-encompassment deepening
  (administrative reach into delivery + control of the non-state sphere).
- Negative PC2 movement = state-encompassment shrinking, either through
  deliberate liberal-pluralist constraint (judicial review, free CSOs)
  or through state weakening (non-state violence, contested authority).

PC2 is **orthogonal to PC1 by construction**. Welfare democracies and
developmental autocracies sit together at high PC2 (they share
encompassment despite opposing each other on PC1). Liberal-market
democracies and weak/contested polities sit together at low PC2 (they
share non-state weight on the state, despite opposing each other on PC1).
This is the empirical signature of the long-standing political-theory
claim (Tilly, Mann, Migdal) that state capacity is a separate dimension
from regime type.

## How to deploy PC1 and PC2 in case studies

When projecting a country into this space, report and read both axes
separately. The diagnostic patterns to look for:

- **Regime change with stable stateness.** PC1 moves, PC2 flat. Many
  third-wave democratization cases --- institutions liberalize without
  the state's reach into economy, media, civil society shifting much.
- **State-building without democratization.** PC2 up, PC1 flat. The
  developmental-state pattern --- encompassment deepens (administrative
  delivery + control of non-state sphere) without regime opening.
  Singapore, post-Doi Moi Vietnam, late-Hun-Sen Cambodia are candidates.
- **State collapse with regime liberalization.** PC2 down, PC1 up. The
  post-Soviet 1990s pattern --- media, parties, civil society, and
  economy de-state-ize while the regime formally opens. Russia
  1989--1996 is the textbook case.
- **State reconsolidation under autocracy.** PC2 up, PC1 down. The
  recovery from the previous pattern --- regime closes, state reabsorbs
  media, parties, CSOs. Russia 2000s.

A case-study writeup should locate the country in (PC1, PC2) space at
endpoints, plot the trajectory, and decompose any large move into its
PC1 and PC2 components. The four patterns above are the ones to look for;
their absence is just as informative as their presence.

# Persist outputs

```{r persist}
dir.create("outputs/pooled_pca", showWarnings = FALSE)

saveRDS(list(
  rotation    = pca_pool$rotation,
  sdev        = pca_pool$sdev,
  feature_means = colMeans(M_raw),
  feature_sds   = apply(M_raw, 2, sd),
  indicators    = colnames(M_z),
  panel_start   = panel_start,
  panel_end     = panel_end,
  n_country_years = nrow(M_z)
), "outputs/pooled_pca/pca_pool.rds")

panel_with_scores %>%
  select(country_name, year, starts_with("PC")) %>%
  write_csv("outputs/pooled_pca/country_year_pc_scores.csv")

# Loading table for downstream use
loadings_df %>%
  select(code, name, category, PC1, PC2, PC3, PC4) %>%
  write_csv("outputs/pooled_pca/pca_loadings_pc1_pc4.csv")

cat("Wrote: outputs/pooled_pca/pca_pool.rds\n")
cat("Wrote: outputs/pooled_pca/country_year_pc_scores.csv (", nrow(panel_with_scores), " rows)\n", sep="")
cat("Wrote: outputs/pooled_pca/pca_loadings_pc1_pc4.csv\n")
```

# Notes and what to read into this

- **Sign convention.** PC sign is arbitrary in `prcomp`. We don't flip
  anything in this file; if a downstream caller wants "high PC1 =
  liberal democracy", inspect the top loadings of PC1 above and apply
  a sign flip when reading the saved scores. The reproducible move is
  to store both the rotation and the sign convention you settled on.

- **Pooled vs. snapshot PCAs are answering different questions.**
  Pooled = "what are the dominant institutional axes across our
  development-relevant window?" Snapshot = "what's the shape of
  institutional space *now*?" The 2025 snapshot embeds modules
  (Digital Society) that are absent from most of the pool. The
  trajectory chart in `case-studies/cambodia/cambodia-analysis.Rmd`
  should switch to the pooled rotation if its claim is "Cambodia's
  institutions over time."

- **Coverage filter.** The 50% pool-coverage threshold is the lever
  that decides whether `v2sm*` enters the PCA. With 50% almost all
  v2sm indicators get dropped (they're observed in ~46% of cells).
  Tightening to 70% is conservative; loosening to 30% would re-admit
  them. We start at 50% and report the count above; the file is set
  up so the threshold is a single-line change.

- **Imputation.** Pool-mean imputation makes the v2sm pre-2000 cells
  approximately constant at the pool mean — neutral by construction
  but it's still synthetic data. If we wanted to be stricter we'd
  drop any indicator whose post-2000 mean differs materially from its
  full-pool mean (which would catch indicators where pre-2000 cells
  have been imputed to a value far from where the indicator actually
  sits when measured). Worth doing as a sensitivity if PC2's
  interpretation hinges on a specific v2sm indicator that the filter
  let through.
